---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dsprrr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/dsprrr)](https://CRAN.R-project.org/package=dsprrr)
[![R-CMD-check](https://github.com/JamesHWade/dsprrr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JamesHWade/dsprrr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/JamesHWade/dsprrr/graph/badge.svg)](https://app.codecov.io/gh/JamesHWade/dsprrr)
<!-- badges: end -->

`dsprrr` is an R implementation of [DSPy](https://github.com/stanfordnlp/dspy), providing a framework for building **principled, test-driven, and optimizable** applications using Large Language Models (LLMs). It moves beyond simple prompt engineering to a structured programming model where LLM workflows are treated as programs that can be systematically improved.

## Philosophy

`dsprrr` brings the elegance of R's pipe operator (`|>`) to LLM programming. By designing composable, pipe-friendly APIs and integrating with the tidyverse ethos, `dsprrr` makes developing robust AI systems feel like a natural extension of R-based data analysis workflows. Build LLM pipelines the R way—with clear, readable, and chainable operations.

## Installation

You can install the development version of dsprrr from GitHub:

```r
# install.packages("pak")
pak::pak("JamesHWade/dsprrr")
```

## Quick Start

```{r quickstart, eval=FALSE}
library(dsprrr)
library(ellmer)

# Configure your LLM
llm <- chat_openai(model = "gpt-4o-mini")

# Clean, consistent API with pipes!
result <- signature("text -> sentiment: enum('positive', 'negative', 'neutral')") |>
  module(type = "predict", template = "Analyze: {text}") |>
  run(text = "This package makes LLM programming so intuitive!", .llm = llm)

print(result)  # "positive"
```

## Core Concepts

### 1. Pipe-Friendly Design

dsprrr embraces R's pipe operator (`|>`) for building LLM workflows:

```{r pipe-examples, eval=FALSE}
# Build modules with pipes - one consistent way!
classifier <- signature("text -> sentiment") |>
  module(type = "predict", template = "Text: {text}\nSentiment:")

# Execute with pipes
result <- classifier |>
  run(text = "I love how this integrates with R!", .llm = llm)

# Or do it all in one pipeline
result <- signature("text -> summary") |>
  module(type = "predict", template = "Summarize: {text}") |>
  run(text = "Long text...", .llm = llm)

# Future: Chain operations for complex workflows
# analysis <- text |>
#   extract_key_points() |>
#   summarize(max_words = 100) |>
#   translate(to = "Spanish")
```

### 2. Clear and Consistent API

One way to do each thing - no confusion:

```{r api-overview, eval=FALSE}
# Step 1: Define what you want (signature)
sig <- signature("text -> summary: string[50, 200]")

# Step 2: Create how to do it (module)
mod <- sig |> module(type = "predict", template = "Summarize: {text}")

# Step 3: Execute it (run)
result <- mod |> run(text = "Your text here...", .llm = llm)
```

### 3. Flexible Signature Creation

Define operations using string notation or explicit specifications:

```{r signatures, eval=FALSE}
# String notation (recommended for most cases)
qa_sig <- signature("context, question -> answer")
math_sig <- signature("problem -> solution: float")
extract_sig <- signature("text -> entities: array(string)")

# NEW: Multiple outputs support!
analysis_sig <- signature("text -> sentiment: string, confidence: float, keywords: list[string]")

# NEW: Advanced type annotations
optional_sig <- signature("doc -> title: string, subtitle: Optional[string]")
dict_sig <- signature("data -> metadata: dict[string, list[string]]")

# Explicit notation for complex structures
complex_sig <- signature(
  inputs = list(
    input("data", description = "Raw data to analyze")
  ),
  output_type = type_object(
    summary = type_string(),
    score = type_number(),
    tags = type_array(type_string())
  ),
  instructions = "Analyze the data comprehensively"
)

# NEW: Field-level descriptions for better LLM guidance
detailed_sig <- signature(
  inputs = list(
    input("text",
          type_string(description = "Document to analyze for sentiment"))
  ),
  output_type = type_object(
    sentiment = type_enum(
      c('positive', 'negative', 'neutral'),
      description = "Overall sentiment of the document"
    )
  )
)

# Or simpler for basic types - pass description directly to input()
simple_sig <- signature(
  inputs = list(
    input("text", "string", description = "Text to analyze"),
    input("max_length", "integer", description = "Maximum output length")
  )
)
```

## Complete Example: Building a Question-Answer System

Here's how to build a QA system using the clean, consistent dsprrr API:

```{r qa-example, eval=FALSE}
library(dsprrr)
library(ellmer)

# Configure LLM
llm <- chat_openai(model = "gpt-4o-mini")

# Simple pipeline with string notation
answer <- signature("context, question -> answer") |>
  module(type = "predict",
         template = "Context: {context}\n\nQuestion: {question}") |>
  run(
    context = "The tidyverse is a collection of R packages for data science.",
    question = "What is the tidyverse?",
    .llm = llm
  )

# With structured output and demonstrations
qa_module <- signature(
  inputs = list(
    input("context", description = "Background information"),
    input("question", description = "Question to answer")
  ),
  output_type = type_object(
    answer = type_string(),
    confidence = type_number()
  ),
  instructions = "Answer based on context. Include confidence score."
) |>
  module(
    type = "predict",
    demos = list(
      list(
        inputs = list(
          context = "R is a programming language for statistical computing.",
          question = "What is R used for?"
        ),
        output = list(
          answer = "Statistical computing",
          confidence = 0.95
        )
      )
    )
  )

# Execute with pipes
result <- qa_module |>
  run(
    context = "dsprrr brings DSPy to R with clean, consistent APIs.",
    question = "What does dsprrr do?",
    .llm = llm
  )
```

## Automatic Prompt Optimization (NEW!)

Optimize your LLM programs automatically with dsprrr's compilation framework:

```{r compilation-example, eval=FALSE}
# Create a sentiment classifier
classifier <- signature("text -> sentiment: enum('positive', 'negative', 'neutral')") |>
  module(type = "predict", template = "Text: {text}\nSentiment:")

# Prepare training data
trainset <- dsp_trainset(
  text = c("I love this product!", "Terrible experience", "It's okay I guess"),
  sentiment = c("positive", "negative", "neutral")
)

# Method 1: Simple few-shot optimization with LabeledFewShot
optimized <- compile_module(
  program = classifier,
  teleprompter = LabeledFewShot(k = 2),  # Use 2 examples
  trainset = trainset
)

# The optimized module now includes demonstrations!
result <- optimized |> run(text = "Amazing service!", .llm = llm)

# Method 2: Grid search over instruction variants
variants <- data.frame(
  id = c("terse", "detailed", "analytical"),
  instructions_suffix = c(
    ". Be concise.",
    ". Provide detailed reasoning.",
    ". Analyze sentiment analytically."
  )
)

optimized_grid <- compile_module(
  program = classifier,
  teleprompter = GridSearchTeleprompter(
    variants = variants,
    metric = metric_exact_match(field = "sentiment"),
    k = 2
  ),
  trainset = trainset
)

# Check optimization results
print(optimized_grid@config$best_variant)  # Which variant performed best
print(optimized_grid@config$all_scores)     # Scores for all variants
```

### Metrics for Evaluation

dsprrr provides flexible metrics for evaluating and optimizing your programs:

```{r metrics-example, eval=FALSE}
# Exact match for classification
metric <- metric_exact_match(field = "sentiment")

# F1 score for text generation
metric <- metric_f1(field = "answer")

# Custom metrics
length_metric <- metric_custom(function(pred, exp) {
  nchar(pred$summary) <= 100  # Ensure summaries are concise
}, name = "length_check")

# Threshold-based metrics
f1_threshold <- metric_threshold(
  metric_f1(field = "answer"),
  threshold = 0.8,
  comparison = ">="
)

# Evaluate your module
results <- evaluate_dsp(
  module = optimized,
  dataset = test_data,
  metric = metric_exact_match(field = "sentiment"),
  llm = llm
)
print(results$mean_score)
```

## Key Features (NEW in v0.1.0)

### Batch Processing & Dataset Support
Process multiple inputs efficiently with automatic parallelization:

```{r batch-processing, eval=FALSE}
# Process multiple inputs at once
results <- classifier |>
  run(text = c("Great!", "Terrible", "Okay"), .llm = llm)

# Process entire datasets with run_dataset()
data <- tibble(
  text = c("I love this!", "Not good", "It's fine")
)
results_df <- classifier |> run_dataset(data, .llm = llm)
```

### Structured Returns
Get detailed metadata with every call:

```{r structured-returns, eval=FALSE}
result <- classifier |>
  run(text = "Amazing!", .llm = llm, .return_format = "structured")

# Access components
result$output      # The actual result
result$chat        # The LLM chat object
result$metadata    # Timing, tokens, etc.
```

### vitals Integration
Seamless interoperability with the vitals evaluation framework:

```{r vitals-integration, eval=FALSE}
library(vitals)

# Use dsprrr modules as vitals solvers
task <- Task$new(
  dataset = test_data,
  solver = as_function(classifier, .llm = llm),
  scorer = model_graded_qa()
)
task$eval()
```

## Roadmap

`dsprrr` is under active development. Current implementation status:

### ✅ Milestone 1: Core Foundation (COMPLETED - Dec 2024)
- **S7 Architecture**: Robust `Signature` and `Predict` classes with validators
- **Clean API**: Consistent `signature()`, `module()`, `run()` functions
- **DSPy Notation**: String signatures like `"text -> sentiment: enum('pos', 'neg')"`
- **Type System**: Flexible input types with ellmer integration
- **Pipe-Friendly**: Full support for R's `|>` operator
- **Testing**: 340+ passing tests with comprehensive coverage
- **Documentation**: Getting-started vignette with practical examples

### ✅ Milestone 2: Enhanced Integration (COMPLETED - Jan 2025)
- [x] **Batch Processing**: Process multiple inputs with automatic parallelization
- [x] **Dataset Support**: Native `run_dataset()` for tibbles/data frames
- [x] **Structured Returns**: Detailed metadata with every LLM call
- [x] **vitals Integration**: Seamless interoperability with evaluation framework
- [x] **Function Interface**: Convert modules to callable functions
- [x] `Teleprompter` base class for optimization strategies
- [x] `compile()` generic for automatic prompt optimization
- [x] Bootstrap few-shot learning with `LabeledFewShot`
- [x] Grid search optimization with `GridSearchTeleprompter`
- [x] Comprehensive metric helpers
- [x] Module state management (`reset_copy`, `deepcopy`, `is_compiled`)
- [x] Evaluation framework with `evaluate_dsp()`

### 📋 Milestone 3: Ecosystem & Ergonomics (Planned)
- [ ] `evaluate()` generic for systematic testing
- [ ] Advanced error handling and debugging tools
- [ ] Provider-specific helpers (OpenAI, Anthropic, etc.)
- [ ] pkgdown documentation website

### 🔮 Future Vision
- **Advanced Modules**: ChainOfThought, ReAct agents, multi-step reasoning
- **Optimization**: MIPRO, BetterTogether, and other DSPy optimizers
- **Deployment**: Integration with pins (caching) and vetiver (API deployment)
- **Extensions**: Custom modules and teleprompters via S7 inheritance

## Getting Started

Check out our comprehensive [Getting Started vignette](vignettes/getting-started.Rmd) for detailed examples and best practices.

```r
# View the vignette
vignette("getting-started", package = "dsprrr")
```

## Development

### Running Tests and Vignettes

The package uses [vcr](https://docs.ropensci.org/vcr/) to cache API responses, allowing tests and vignettes to run without API keys:

```r
# Run tests (uses cached responses)
devtools::test()

# Build vignettes (uses cached responses)
devtools::build_vignettes()
```

### Recording New Cassettes

Maintainers with API keys can update cached responses:

```bash
# Set your API keys
export OPENAI_API_KEY="your-key"
export ANTHROPIC_API_KEY="your-key"

# Record new cassettes
Rscript inst/scripts/record-cassettes.R
```

## Contributing

We welcome contributions! Please see our [contribution guidelines](CONTRIBUTING.md) for details.

## License

MIT © James Wade

## Acknowledgments

This package is inspired by [DSPy](https://github.com/stanfordnlp/dspy) from Stanford NLP. Special thanks to the R Consortium Object-Oriented Programming Working Group for the S7 system that powers this implementation.
