---
title: "Advanced Optimizer Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Optimizer Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

dsprrr provides a comprehensive suite of DSPy-inspired optimizers (teleprompters) for automatically improving your LLM programs. This guide covers the advanced optimizers beyond basic few-shot learning.

For basic optimization concepts, see `vignette("compilation-optimization")`.

## Quick Reference: Choosing an Optimizer

| If you want... | Use this optimizer | Complexity |
|----------------|-------------------|------------|
| Add labeled examples as demos | `LabeledFewShot` | Low |
| Bootstrap demos from LLM outputs | `BootstrapFewShot` | Medium |
| Bootstrap + search multiple configs | `BootstrapFewShotWithRandomSearch` | Medium |
| Dynamic per-query demo selection | `KNNFewShot` | Medium |
| Optimize instructions (not demos) | `COPRO` | Medium |
| Joint instruction + demo optimization | `MIPROv2` | High |
| Focus on hard examples | `SIMBA` | Medium |
| Multi-objective optimization | `GEPA` | High |
| Combine multiple strategies | `Ensemble` | Low |

### Decision Tree

```
Start here: Do you have labeled training data?
│
├─ NO → Use zero-shot or manually write demos
│
└─ YES → How much data?
    │
    ├─ < 30 examples → LabeledFewShot (simple few-shot)
    │
    └─ 30+ examples → What do you want to optimize?
        │
        ├─ Just add demos → BootstrapFewShot
        │
        ├─ Demos + search configs → BootstrapFewShotWithRandomSearch
        │
        ├─ Different demos per query → KNNFewShot
        │
        ├─ Improve instructions → COPRO
        │
        ├─ Both instructions + demos → MIPROv2
        │
        ├─ Handle hard cases better → SIMBA
        │
        ├─ Balance quality vs cost → GEPA
        │
        └─ Combine multiple optimized modules → Ensemble
```

## Dataset Sizing Guidance

| Dataset Size | Recommended Optimizers | Notes |
|--------------|----------------------|-------|
| 10-30 examples | `LabeledFewShot` | Minimal optimization |
| 30-100 examples | `BootstrapFewShot`, `COPRO` | Good starting point |
| 100-300 examples | `BootstrapFewShotWithRandomSearch`, `MIPROv2` | Meaningful search |
| 300+ examples | All optimizers | Full optimization potential |

**Key principles:**

- **Train/validation split**: Use 70-80% for training, 20-30% for validation
- **Diverse examples**: Ensure coverage of edge cases and all output categories
- **Quality over quantity**: 50 high-quality examples beat 500 noisy ones

```{r dataset-split}
library(dsprrr)

# Split your dataset properly
full_data <- tibble::tibble(

  question = c(...),  # Your examples
  answer = c(...)
)

set.seed(42)
n <- nrow(full_data)
train_idx <- sample(n, size = floor(0.7 * n))

trainset <- full_data[train_idx, ]
valset <- full_data[-train_idx, ]

# Or use the built-in helper
splits <- split_dataset(full_data, train_pct = 0.7, seed = 42)
trainset <- splits$train
valset <- splits$val
```

## Setup

```{r setup}
library(dsprrr)
library(ellmer)

# Configure your LLM
llm <- chat_openai(model = "gpt-4o-mini")

# Example training data for demonstrations
trainset <- dsp_trainset(
  question = c(
    "What is the capital of France?",
    "Who wrote Romeo and Juliet?",
    "What is the chemical symbol for gold?",
    "When did World War II end?",
    "What is the largest planet in our solar system?"
  ),
  answer = c(
    "Paris",
    "William Shakespeare",
    "Au",
    "1945",
    "Jupiter"
  )
)

# Base module to optimize
qa_module <- module(
  signature("question -> answer"),
  type = "predict"
)
```

## BootstrapFewShot

Bootstraps demonstrations by running the module on training examples and keeping successful outputs as demos.

**Best for:** When you want the LLM to generate its own demonstration format.

```{r bootstrap-fewshot}
tp <- BootstrapFewShot(
  metric = metric_exact_match(field = "answer"),
  max_bootstrapped_demos = 4L,
  max_labeled_demos = 2L,
  max_rounds = 3L,
  max_errors = 5L,
  seed = 42L
)

compiled <- compile(tp, qa_module, trainset, .llm = llm)

# Check what demos were bootstrapped
print(compiled$demos)

# Run the optimized module
result <- run(compiled, question = "What is the speed of light?", .llm = llm)
```

**Parameters:**

- `max_bootstrapped_demos`: Maximum LLM-generated demos to include
- `max_labeled_demos`: Maximum labeled examples from trainset
- `max_rounds`: Bootstrapping iterations
- `metric`: Evaluation metric (defaults to exact match)

## BootstrapFewShotWithRandomSearch

Combines bootstrapping with random search over configurations. Produces multiple candidate programs and selects the best.

**Best for:** When you want to explore different demo combinations and find the optimal configuration.

```{r bootstrap-rs}
tp <- BootstrapFewShotWithRandomSearch(
  metric = metric_exact_match(field = "answer"),
  max_bootstrapped_demos = 4L,
  max_labeled_demos = 2L,
  num_candidate_programs = 8L,
  num_threads = 4L,
  seed = 42L
)

compiled <- compile(tp, qa_module, trainset, valset = valset, .llm = llm)

# Access the candidate programs and their scores
candidates <- compiled$config$optimizer$candidate_programs
scores <- compiled$config$optimizer$candidate_scores

# The compiled module is the best performer
print(compiled$config$optimizer$best_score)
```

**Parameters:**

- `num_candidate_programs`: Number of configurations to try
- `num_threads`: Parallel evaluation threads
- All `BootstrapFewShot` parameters are inherited

**Tip:** The candidate programs from this optimizer work great with `Ensemble`:

```{r bootstrap-to-ensemble}
# Take top 3 candidates and ensemble them
top_candidates <- candidates[1:3]
top_scores <- scores[1:3]

ens <- ensemble(
  top_candidates,
  reduce_fn = reduce_weighted_vote(),
  weights = top_scores
)
```

## KNNFewShot

Selects demonstrations dynamically based on similarity to the input query. Uses embeddings to find the most relevant examples.
**Best for:** Tasks where example relevance varies significantly by query.

```{r knn-fewshot}
tp <- KNNFewShot(
  k = 3L,
  vectorizer = function(texts) {
    # Use any embedding function
    ragnar::embed_openai(texts)
  },
  metric = metric_exact_match(field = "answer"),
  seed = 42L
)

compiled <- compile(tp, qa_module, trainset, .llm = llm)

# Each query now gets personalized demos
result <- run(compiled, question = "What is DNA made of?", .llm = llm)
```

**Parameters:**

- `k`: Number of nearest neighbors to use as demos
- `vectorizer`: Function that converts text to embeddings
- Supports caching embeddings for efficiency

## COPRO (Coordinate Prompt Optimization)

Optimizes instructions through coordinate ascent. Generates and tests instruction variants to find the best wording.

**Best for:** When your task benefits from better instructions rather than more demos.

```{r copro}
# Optionally use a different model for instruction generation
prompt_llm <- chat_openai(model = "gpt-4o")

tp <- COPRO(
  metric = metric_exact_match(field = "answer"),
  prompt_model = prompt_llm,  # Model to generate instruction candidates
  breadth = 5L,               # Candidates per iteration
  depth = 3L,                 # Number of iterations
  init_temperature = 1.4,
  seed = 42L
)

compiled <- compile(tp, qa_module, trainset, valset = valset, .llm = llm)

# Check the optimized instructions
print(compiled$signature@instructions)

# View optimization history
history <- compiled$config$optimizer$history
print(history)
```

**Parameters:**

- `breadth`: Number of instruction candidates per iteration
- `depth`: Number of coordinate ascent iterations
- `prompt_model`: LLM for generating instructions (can differ from task LLM)
- `init_temperature`: Temperature for instruction generation

**How it works:**

1. Starts with current instructions as baseline
2. Generates `breadth` instruction variants
3. Evaluates each on validation set
4. Keeps the best, uses failed examples to improve
5. Repeats for `depth` iterations

## MIPROv2

Multi-prompt Instruction Proposal Optimizer. Jointly optimizes instructions and demonstrations using Bayesian optimization.

**Best for:** Maximum optimization when you have sufficient data and compute budget.

```{r miprov2}
tp <- MIPROv2(
  metric = metric_exact_match(field = "answer"),
  num_candidates = 10L,
  init_temperature = 1.0,
  prompt_model = chat_openai(model = "gpt-4o"),
  num_trials = 20L,          # Bayesian optimization trials
  minibatch_size = 25L,
  minibatch_full_eval_steps = 10L,
  seed = 42L
)

compiled <- compile(tp, qa_module, trainset, valset = valset, .llm = llm)

# MIPROv2 optimizes both instructions and demos
print(compiled$signature@instructions)
print(compiled$demos)
```

**Parameters:**

- `num_candidates`: Instruction candidates to generate
- `num_trials`: Bayesian optimization budget
- `minibatch_size`: Examples per evaluation batch
- Supports `log_dir` for detailed trial logging

**Presets:**

```{r mipro-presets}
# Light preset (faster, less thorough)
tp_light <- MIPROv2(
  metric = metric_exact_match(),
  num_candidates = 5L,
  num_trials = 10L
)

# Heavy preset (slower, more thorough)
tp_heavy <- MIPROv2(
  metric = metric_exact_match(),
  num_candidates = 20L,
  num_trials = 50L,
  minibatch_size = 50L
)
```

## SIMBA (Self-Improving Model-Based Augmentation)

Focuses optimization on hard examples that the model struggles with.

**Best for:** When your model performs well on average but fails on edge cases.

```{r simba}
tp <- SIMBA(
  metric = metric_exact_match(field = "answer"),
  num_candidates = 5L,
  max_bootstrapped_demos = 4L,
  max_labeled_demos = 2L,
  max_iterations = 3L,
  seed = 42L
)

compiled <- compile(tp, qa_module, trainset, valset = valset, .llm = llm)

# SIMBA iteratively improves on hard examples
print(compiled$config$optimizer$hard_examples)
```

**How it works:**

1. Identifies examples where the model fails
2. Generates targeted demos for those cases
3. Re-evaluates and repeats
4. Builds a demo set that covers edge cases

## GEPA (Guided Evolutionary Prompt Algorithm)

Multi-objective optimization balancing quality vs. cost (or other objectives).

**Best for:** Production systems where you need to balance accuracy against token usage or latency.

```{r gepa}
tp <- GEPA(
  metric = metric_exact_match(field = "answer"),
  population_size = 10L,
  num_generations = 5L,
  mutation_rate = 0.1,
  crossover_rate = 0.7,
  objectives = c("quality", "cost"),
  seed = 42L
)

compiled <- compile(tp, qa_module, trainset, valset = valset, .llm = llm)

# GEPA returns Pareto-optimal solutions
pareto_front <- compiled$config$optimizer$pareto_front
print(pareto_front)
```

**Parameters:**

- `population_size`: Number of candidates per generation
- `num_generations`: Evolution iterations
- `objectives`: What to optimize (quality, cost, latency)
- `mutation_rate`, `crossover_rate`: Genetic algorithm parameters

## Ensemble

Combines multiple compiled modules using voting or aggregation strategies.

**Best for:** Maximum robustness by combining diverse optimization strategies.

```{r ensemble}
# First, compile modules with different strategies
mod_bootstrap <- compile(BootstrapFewShot(), qa_module, trainset, .llm = llm)
mod_copro <- compile(COPRO(), qa_module, trainset, valset = valset, .llm = llm)
mod_knn <- compile(KNNFewShot(k = 3L), qa_module, trainset, .llm = llm)

# Combine with ensemble
ens <- ensemble(
  list(mod_bootstrap, mod_copro, mod_knn),
  reduce_fn = reduce_majority()
)

# Or use validation scores as weights
ens_weighted <- ensemble(
  list(mod_bootstrap, mod_copro, mod_knn),
  reduce_fn = reduce_weighted_vote(),
  weights = c(0.85, 0.90, 0.82)  # Validation scores
)

# Run the ensemble
result <- run(ens, question = "What is photosynthesis?", .llm = llm)
```

### Reduce Functions

```{r reduce-functions}
# Majority voting (default)
reduce_majority()

# Weighted voting using module weights
reduce_weighted_vote()

# Just take the first successful output
reduce_first()

# Score outputs with a metric
reduce_best_by_metric(
  metric = metric_f1(field = "answer")
)
```

### Ensemble via Teleprompter

```{r ensemble-teleprompter}
# Ensemble teleprompter wraps existing compiled modules
tp <- Ensemble(
  reduce_fn = reduce_weighted_vote(),
  weights = c(0.9, 0.85, 0.8)
)

ens <- compile(tp, programs = list(mod1, mod2, mod3))
```

## Tracking and Logging

All optimizers support logging for debugging and reproducibility.

### Trial Logging

```{r logging}
# Enable logging to a directory
tp <- BootstrapFewShotWithRandomSearch(
  metric = metric_exact_match(),
  log_dir = "optimization_logs/experiment_001",
  seed = 42L
)

compiled <- compile(tp, qa_module, trainset, .llm = llm)

# Log directory contains:
# - trials.jsonl: All trial results
# - best_program.rds: Serialized best module
# - config.json: Optimizer configuration
```

### Analyzing Trials

```{r analyze-trials}
# Read trial logs
trials <- read_trials_jsonl("optimization_logs/experiment_001/trials.jsonl")

# Examine trial results
print(trials)

# Plot optimization progress
library(ggplot2)
ggplot(trials, aes(x = trial_id, y = score)) +
  geom_line() +
  geom_point() +
  labs(title = "Optimization Progress", x = "Trial", y = "Score")
```

### Accessing Optimizer State

```{r optimizer-state}
# After compilation, access optimizer metadata
compiled$config$optimizer$name
compiled$config$optimizer$params
compiled$config$optimizer$best_score
compiled$config$optimizer$trials

# For BootstrapFewShotWithRandomSearch
compiled$config$optimizer$candidate_programs
compiled$config$optimizer$candidate_scores

# For COPRO
compiled$config$optimizer$history

# For GEPA
compiled$config$optimizer$pareto_front
```

## Reproducibility

All optimizers support deterministic seeds:

```{r reproducibility}
# Set seed for reproducibility
tp <- BootstrapFewShot(
  metric = metric_exact_match(),
  seed = 42L
)

# Same seed = same results
compiled1 <- compile(tp, qa_module, trainset, .llm = llm)
compiled2 <- compile(tp, qa_module, trainset, .llm = llm)

# Demos will be identical
identical(compiled1$demos, compiled2$demos)  # TRUE
```

## Error Handling

Optimizers gracefully handle LLM errors:

```{r error-handling}
tp <- BootstrapFewShot(
  metric = metric_exact_match(),
  max_errors = 10L  # Allow up to 10 errors before failing
)

# Compilation continues despite some failed examples
compiled <- compile(tp, qa_module, trainset, .llm = llm)

# Check how many errors occurred
compiled$config$optimizer$n_errors
```

## Performance Tips

### 1. Start Simple, Then Advance

```{r progressive-optimization}
# Step 1: Try LabeledFewShot first
simple <- compile(LabeledFewShot(k = 3L), qa_module, trainset, .llm = llm)
simple_score <- evaluate(simple, valset, metric_exact_match(), .llm = llm)$mean_score

# Step 2: If not good enough, try bootstrapping
if (simple_score < 0.8) {
  bootstrap <- compile(BootstrapFewShot(), qa_module, trainset, .llm = llm)
  bootstrap_score <- evaluate(bootstrap, valset, metric_exact_match(), .llm = llm)$mean_score
}

# Step 3: If still not good enough, try COPRO
if (bootstrap_score < 0.85) {
  copro <- compile(COPRO(), qa_module, trainset, valset = valset, .llm = llm)
}
```

### 2. Use Parallel Evaluation

```{r parallel}
# BootstrapFewShotWithRandomSearch supports parallel evaluation
tp <- BootstrapFewShotWithRandomSearch(
  num_candidate_programs = 16L,
  num_threads = 8L  # Evaluate 8 candidates in parallel
)
```

### 3. Cache Embeddings for KNNFewShot

```{r caching}
# Pre-compute and cache embeddings
trainset_embeddings <- ragnar::embed_openai(trainset$question)

# Pass cached embeddings to avoid recomputation
tp <- KNNFewShot(
  k = 3L,
  vectorizer = ragnar::embed_openai,
  cached_embeddings = trainset_embeddings
)
```

### 4. Monitor Costs

```{r cost-monitoring}
# Track costs during optimization
session_cost()  # Total cost so far

# After compilation
compiled$config$optimizer$total_cost
```

## Summary

| Optimizer | Optimizes | Data Needs | Compute Cost |
|-----------|-----------|------------|--------------|
| `LabeledFewShot` | Demos | Low (10+) | Very Low |
| `BootstrapFewShot` | Demos | Medium (30+) | Low |
| `BootstrapFewShotWithRandomSearch` | Demos + Config | Medium (50+) | Medium |
| `KNNFewShot` | Dynamic Demos | Medium (50+) | Low |
| `COPRO` | Instructions | Medium (30+) | Medium |
| `MIPROv2` | Instructions + Demos | High (100+) | High |
| `SIMBA` | Hard Example Demos | Medium (50+) | Medium |
| `GEPA` | Multi-objective | High (100+) | High |
| `Ensemble` | Combines modules | N/A | Varies |

## Next Steps

- Start with `LabeledFewShot` or `BootstrapFewShot` for quick wins
- Graduate to `COPRO` or `MIPROv2` for production optimization
- Use `Ensemble` to combine your best models
- See `vignette("compilation-optimization")` for basic optimization concepts
- See `vignette("vitals-integration")` for evaluation with vitals package
