---
title: "Advanced ellmer Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced ellmer Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = dsprrr:::eval_vignette()
)
```

```{r setup}
library(dsprrr)
library(ellmer)
```

## Introduction

dsprrr is built on top of ellmer, Posit's R package for LLM interactions. This vignette explores advanced integration patterns that leverage ellmer's full capabilities.

## Parallel Processing

dsprrr supports two concurrent backends for batch operations. Batch execution
is sequential unless you supply `concurrency_control()`.

### Method 1: mirai

Use mirai for multi-process parallelism:

```{r parallel-mirai, eval = FALSE}
mod <- module(signature("text -> sentiment"))

# Process multiple items in parallel using mirai
results <- run(
  mod,
  text = c("I love this!", "This is terrible", "It's okay"),
  .concurrency = concurrency_control(
    backend = "mirai",
    max_active = 3L
  )
)
```

### Method 2: ellmer Native

For more efficient parallelism, use ellmer's native `parallel_chat_structured()`:

```{r parallel-ellmer, eval = FALSE}
results <- run(
  mod,
  text = c("I love this!", "This is terrible", "It's okay"),
  .concurrency = concurrency_control(
    backend = "ellmer",
    max_active = 3L
  )
)
```

The ellmer method is more efficient because:
- Single process (no R subprocess overhead)
- Native async HTTP requests
- Better error handling
- Automatic rate limit handling

## Converting Modules to ellmer Tools

dsprrr modules can be converted to ellmer tools for use in agentic workflows:

```{r as-ellmer-tool, eval = FALSE}
# Create a sentiment analysis module
sentiment_mod <- module(
  signature("text -> sentiment: enum('positive', 'negative', 'neutral')"))

# Convert to an ellmer tool
sentiment_tool <- as_ellmer_tool(
  sentiment_mod,
  name = "analyze_sentiment",
  description = "Analyze the sentiment of text"
)

# The tool can now be registered with any Chat
chat <- chat_openai()
chat$register_tool(sentiment_tool)

# The LLM can now use the sentiment tool
chat$chat("Analyze the sentiment of: 'I love this product!'")
```

## Leveraging ellmer's Cost Tracking

ellmer provides robust token and cost tracking. dsprrr integrates with this via accessor functions:

```{r cost-tracking, eval = FALSE}
# After running some predictions
mod <- module(signature("question -> answer"))
result <- run(
  mod,
  question = "What is 2+2?",
  .llm = chat_openai(),
  .return_format = "structured"
)

# Get cost and token info from results using public accessors
get_cost(result)    # Cost in dollars
get_tokens(result)  # Token counts

# For session-wide aggregates
session_cost()
```

### Token Usage from Batch Results

```{r token-usage, eval = FALSE}
# Run batch predictions
mod <- module(signature("q -> a"))
results <- run(mod, q = c("Hello", "World"), .return_format = "batch")

# Get aggregated cost and tokens from batch
get_cost(results)    # Total cost for batch
get_tokens(results)  # Total tokens for batch
```

## Advanced Chat Patterns

### Using Chat Objects Across Multiple Calls

```{r chat-reuse, eval = FALSE}
# Create a Chat and reuse it
chat <- chat_openai()
mod <- module(signature("q -> a"))

# Reuse the same runtime across calls
result1 <- run(mod, q = "What is R?", .llm = chat)

result2 <- run(mod, q = "What about Python?", .llm = chat)
```

### Default Chat Management

```{r default-chat, eval = FALSE}
# Set a default Chat for module execution
set_default_chat(chat_openai(model = "gpt-4o"))

# run() uses the default when neither the call nor module supplies a Chat
result <- run(module(signature("q -> a")), q = "What is 2+2?")

# Check current configuration
dsprrr_sitrep()

# Clear when done
clear_default_chat()
```

## Multimodal Support

dsprrr inherits ellmer's multimodal capabilities:
```{r multimodal, eval = FALSE}
mod <- module(
  signature("image, question -> answer"))

# Pass an image via ellmer Content objects
result <- run(
  mod,
  image = ellmer::ContentImageRemote("https://example.com/image.jpg"),
  question = "What is in this image?"
)
```

## Streaming Responses

For long-form generation, use streaming:

```{r streaming, eval = FALSE}
mod <- module(
  signature("topic -> essay"))

# Stream with callback - pass named arguments directly
mod$stream(
  topic = "The future of AI",
  callback = function(chunk) cat(chunk)
)

# Or use async streaming with promises
library(promises)
stream_async(mod, topic = "The future of AI") %...>%
  print()  # Prints the final result when complete
```

## Error Handling and Retries

ellmer handles retries automatically. Configure via options:

```{r retries, eval = FALSE}
# Set max retries (default is 3)
options(ellmer_max_tries = 5)

# Set timeout (default is 60 seconds)
options(ellmer_timeout = 120)

# dsprrr wraps errors with helpful context
tryCatch(
  run(module(signature("q -> a")), q = "test"),
  error = function(e) {
    # Error includes model info, suggestions, etc.
    print(e)
  }
)
```

## Summary

Key integration points with ellmer:

1. **Parallel processing**: Choose between mirai (multi-process) or ellmer native (async HTTP)
2. **Tool integration**: Convert modules to ellmer tools with `as_ellmer_tool()`
3. **Cost tracking**: Use `get_cost()`, `get_tokens()`, and `session_cost()` for usage tracking
4. **Chat management**: Use `set_default_chat()` for session-wide defaults
5. **Multimodal**: Pass ellmer Content objects for images, PDFs, etc.
6. **Streaming**: Use `stream()` or `stream_async()` for long-form generation
7. **Error handling**: Benefit from ellmer's automatic retries and dsprrr's context-rich errors
