---
title: "Chaining Modules and Pipelines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chaining Modules and Pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
# Examples are illustrative and require an LLM client, so they are not evaluated.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview
Chaining lets you build multi-step LLM workflows by passing outputs from one
module to the next. dsprrr provides two approaches:

- `%>>%` for quick, readable pipelines with automatic field matching
- `pipeline()` + `step()` for explicit control over mappings and selections

Both approaches produce a pipeline module you can run with `run()` or
`run_dataset()` just like any other module.

## Create Building-Block Modules
Start with small modules that do one thing well:

```{r}
library(dsprrr)
library(ellmer)

mod_extract <- signature("document -> facts") |>
  module(type = "predict", template = "Extract key facts: {document}")

mod_answer <- signature("facts, question -> answer") |>
  module(type = "predict", template = "Use facts: {facts}\nQ: {question}")

mod_format <- signature("answer -> response") |>
  module(type = "predict", template = "Format: {answer}")
```

## Simple Chaining with `%>>%`
When field names align, outputs connect automatically:

```{r}
qa_pipeline <- mod_extract %>>% mod_answer %>>% mod_format

llm <- chat_openai()
result <- run(
  qa_pipeline,
  document = "...",
  question = "What happened?",
  .llm = llm
)
```

## Map Inputs When Names Differ
If a downstream module expects a different input name, map fields explicitly:

```{r}
mod_retrieve <- signature("query -> documents") |>
  module(type = "predict", template = "Retrieve docs for: {query}")

mod_summarize <- signature("context -> summary") |>
  module(type = "predict", template = "Summarize: {context}")

rag_pipeline <- pipeline(
  mod_retrieve,
  step(mod_summarize, map = c(documents = "context"))
)

result <- run(rag_pipeline, query = "dsprrr pipelines", .llm = llm)
```

## Inject Static Inputs
Pass constants through `step()` when they do not come from upstream outputs:

```{r}
mod_answer <- signature("facts, question, tone -> answer") |>
  module(type = "predict")

qa_pipeline <- pipeline(
  mod_extract,
  step(mod_answer, tone = "concise"),
  mod_format
)
```

## Select Outputs to Pass Forward
Keep only specific fields from a structured output:

```{r}
mod_reason <- signature("question -> answer, reasoning") |>
  module(type = "chain_of_thought")

mod_present <- signature("answer -> response") |>
  module(type = "predict")

pipeline_filtered <- pipeline(
  step(mod_reason, select = "answer"),
  mod_present
)
```

## Explicit Pipelines with `pipeline()` and `step()`
For more control (or when you prefer not to use `%>>%`), build a pipeline
explicitly:

```{r}
explicit_pipeline <- pipeline(
  mod_retrieve,
  step(mod_summarize, map = c(documents = "context")),
  mod_format
)

result <- run(explicit_pipeline, query = "...", .llm = llm)
```

## Batch Execution with `run_dataset()`
Pipelines work with data frames the same way as single modules.
`run_dataset()` adds a `result` column to your input data:

```{r}
library(tibble)

questions <- tibble(
  document = c("Doc A", "Doc B"),
  question = c("What changed?", "What is the summary?")
)

results <- run_dataset(
  qa_pipeline,
  questions,
  .llm = llm
)

# Inspect results
results$result
```

## Tips
- Prefer small, composable modules. They are easier to debug and optimize.
- Use `trace_summary()` and `export_traces()` to inspect multi-step behavior.
- When field names differ between steps, use the `map` argument to `step()`.
