---
title: "Chain modules into pipelines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chain modules into pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
# Chunks that call a model are not evaluated; chunks that only build modules are.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

A pipeline runs modules in order and passes each step's outputs to the next
step. This page shows how to build one, how data moves between steps, and how
to run, inspect and compile the result. A pipeline is itself a module, so
`run()`, `run_dataset()`, `evaluate()` and `compile()` work on it.

## How data moves between steps

Step 1 receives the inputs you pass to `run()`. Every later step receives only
the previous step's outputs, plus any static inputs you attach to that step.
Pipeline inputs are not forwarded past step 1.

The example on this page turns release notes into a headline for customers:

| Step | Module | Receives | Produces |
|---|---|---|---|
| 1 | `list_changes` | `notes`, the pipeline input | `changes` |
| 2 | `write_summary` | `changes` from step 1, and `audience`, a static input | `summary` |
| 3 | `write_headline` | `text`, which is step 2's `summary` renamed | `headline` |

## Build the steps

Each step is an ordinary module that does one thing:

```{r steps}
#| eval: true
#| message: false
library(dsprrr)

list_changes <- module(signature(
  "notes -> changes: list[str]",
  instructions = "List each user-visible change as one short sentence."
))
write_summary <- module(signature(
  "changes, audience -> summary",
  instructions = "Summarize the changes in two sentences for the given audience."
))
write_headline <- module(signature(
  "text -> headline",
  instructions = "Write a headline of at most eight words for the text."
))
```

## Connect them with pipeline() and step()

`pipeline()` takes modules in order. Wrap a module in `step()` when it needs
wiring:

```{r pipeline}
#| eval: true
release <- pipeline(
  list_changes,
  step(write_summary, audience = "customers"),
  step(write_headline, map = c(summary = "text"))
)
release
```

`step()` takes three kinds of wiring:

| Argument | Effect | Example |
|---|---|---|
| Named values | Static inputs, the same on every run. They win over an upstream field with the same name. | `audience = "customers"` |
| `map` | Renames fields from the previous step. Names are that step's outputs, values are this step's inputs. | `map = c(summary = "text")` |
| `select` | Keeps only the listed fields of this step's outputs for the next step. | `select = "summary"` |

Run the pipeline like any module. The result holds the last step's outputs,
here a list with one element, `headline`:

```{r run}
llm <- ellmer::chat_openai(model = "gpt-6-luna")
notes <- paste(
  "v2.3: Rewrote CSV export, now twice as fast.",
  "Added a dark mode.",
  "Fixed a bug that logged users out after five minutes."
)
run(release, notes = notes, .llm = llm)
```

## Chain with %>>%

When each module's inputs are exactly the previous module's outputs, `%>>%`
builds the same kind of pipeline with less typing:

```{r pipe-operator}
brief <- list_changes %>>%
  module(signature("changes -> summary")) %>>%
  module(signature("summary -> headline"))
run(brief, notes = notes, .llm = llm)
```

`%>>%` accepts only modules, so use `pipeline()` when a step needs static
inputs, `map` or `select`.

## Drop fields with select

Every output of a step moves on to the next one, and a Predict module
without a template writes every input it receives into its prompt.
`chain_of_thought()` adds a `reasoning` output, so without `select` the
headline step below would also read the summary's reasoning:

```{r select}
careful <- pipeline(
  list_changes,
  step(chain_of_thought("changes -> summary"), select = "summary"),
  module(signature("summary -> headline"))
)
```

## When a later step needs an original input

Suppose step 1 extracts facts from a document and step 2 answers a question
about those facts. As a pipeline, `run()` stops with "Missing inputs for
pipeline step 2", because `question` went only to step 1. If the value is the
same on every run, attach it as a static input with `step()`. Otherwise,
write the flow as a function with `module_fn()`, which passes the active chat
to a `.llm` argument:

```{r module-fn}
facts_from <- module(signature("document -> facts: list[str]"))
answer_from <- module(signature("facts: list[str], question -> answer"))

qa <- module_fn(
  "document, question -> answer",
  function(document, question, .llm) {
    facts <- run(facts_from, document = document, .llm = .llm)$facts
    run(answer_from, facts = facts, question = question, .llm = .llm)
  }
)
run(qa, document = notes, question = "Is there a dark mode now?", .llm = llm)
```

The trade-off is optimization: `compile()` cannot see inside a function, so
it refuses `module_fn()` programs. Compile `facts_from` and `answer_from` on
their own instead.

## Run many inputs

`run_dataset()` runs the pipeline once per row and returns the input columns
plus a `result` list-column:

```{r run-dataset}
releases <- tibble::tibble(
  notes = c(notes, "v2.4: Added single sign-on. Search is faster.")
)
results <- run_dataset(release, releases, .llm = llm)
results$result
```

## See what each step did

The result keeps only the last step's outputs. Each run also records every
step's inputs and outputs in the pipeline's traces:

```{r step-outputs}
run(release, notes = notes, .llm = llm)
latest <- release$state$traces[[length(release$state$traces)]]
latest$step_inputs
latest$step_outputs
```

A structured result reports usage for the whole run (`n_steps`,
`provider_calls`, `total_tokens`, `cost`) and keeps each step's own metadata,
including the prompt it sent, in `step_metadata`:

```{r usage}
result <- run(release, notes = notes, .llm = llm, .return_format = "structured")
result$metadata$total_tokens
cat(result$metadata$step_metadata[[2]]$prompt)
```

## Compile the whole pipeline

`BootstrapFewShot()` compiles a pipeline jointly: it runs the whole pipeline
on each training row and scores the final output with your metric. When a run
passes, every step's inputs and outputs from that run become a demonstration
for that step, so only the final output needs a label:

```{r compile}
past_releases <- tibble::tibble(
  notes = c(
    "v2.1: Added two-factor login. Fixed slow search on large projects.",
    "v2.2: New PDF export. Removed the legacy API."
  ),
  headline = c(
    "Two-factor login and faster search",
    "PDF export arrives, legacy API retires"
  )
)

compiled <- compile(
  release,
  BootstrapFewShot(
    metric = metric_f1(field = "headline"),
    metric_threshold = 0.5,
    max_bootstrapped_demos = 2L
  ),
  past_releases,
  .llm = llm
)
```

`compile()` returns a new pipeline and leaves `release` unchanged. A useful
training set has more than two rows; [Compile and
optimize](compilation-optimization.html) covers data, metrics and the other
teleprompters.
