---
title: "Quick Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
# Skip evaluation during R CMD check or CI (cassettes may not match current code)
# This prevents build failures when VCR cassettes are stale
in_check <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
            identical(Sys.getenv("NOT_CRAN"), "false")

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

# Only set up VCR if we're actually evaluating
if (should_eval) {
  # Ensure no cassettes are left over from previous vignettes
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::setup_knitr()
}
```

A quick reference for common dsprrr operations.

## Setup & Configuration

```{r setup}
library(dsprrr)
library(ellmer)
```

### Configure Default LLM

```{r configure}
#| eval: false
# Auto-detect from environment variables (OPENAI_API_KEY, ANTHROPIC_API_KEY, etc.)
dsp_configure()

# Explicitly set provider and model
dsp_configure(provider = "openai", model = "gpt-4o-mini")
dsp_configure(provider = "anthropic", model = "claude-3-5-sonnet-latest")

# With additional parameters
dsp_configure(provider = "openai", model = "gpt-4o", temperature = 0.7)
```

### Check Configuration Status

```{r sitrep}
#| eval: false
dsprrr_sitrep()
#> dsprrr Configuration
#> ────────────────────────────────────────
#> Default Chat: ✓ Active
#>   Provider: OpenAI
#>   Model: gpt-4o-mini
```

---

## Signature Notation

Signatures define what goes in and what comes out.

### Basic Format

```
inputs -> outputs
```

### Input Types

| Notation | Meaning |
|----------|---------|
| `question` | Single string input |
| `context, question` | Multiple inputs |
| `items: list[string]` | Typed list input |

### Output Types

| Notation | R/ellmer Type | Use Case |
|----------|---------------|----------|
| `answer` | `type_string()` | Free-form text |
| `answer: string` | `type_string()` | Explicit string |
| `score: int` | `type_integer()` | Whole numbers |
| `score: float` | `type_number()` | Decimal numbers |
| `score: number[0, 100]` | `type_number()` | Bounded numbers |
| `valid: bool` | `type_boolean()` | True/false |

### Constrained Types

| Notation | Description |
|----------|-------------|
| `sentiment: enum('pos', 'neg', 'neutral')` | Fixed choices |
| `label: Literal['a', 'b', 'c']` | Python-style enum |
| `summary: string[50, 200]` | String with length hints |

### Collection Types

| Notation | Description |
|----------|-------------|
| `tags: list[string]` | List of strings |
| `items: array(string)` | Array notation |
| `words: string[]` | Bracket notation |
| `data: dict[string, int]` | Dictionary/map |

### Optional Types

| Notation | Description |
|----------|-------------|
| `note: Optional[string]` | May be null |

### Multiple Outputs

```{r multi-output}
# Multiple output fields
sig <- signature("question -> answer, confidence: float")

# With different types
sig <- signature("text -> sentiment: enum('pos', 'neg'), score: float")
```

### With Instructions

```{r instructions}
sig <- signature(
  "context, question -> answer",
  instructions = "Answer based only on the provided context. Be concise."
)
```

---

## Creating Modules

### Quick: `dsp()` for One-Off Calls

```{r dsp-quick}
#| cassette: true
# With explicit Chat
chat <- chat_openai(model = "gpt-4o-mini")
chat |> dsp("question -> answer", question = "What is 2+2?")
```

The auto-detected form (when you have a default Chat configured):

```{r dsp-auto}
#| eval: false
# With auto-detected Chat (uses configured default)
dsp("question -> answer", question = "What is 2+2?")
```

### Reusable: `as_module()` for Repeated Use

```{r as-module}
#| cassette: true
# Create from Chat
classifier <- chat_openai(model = "gpt-4o-mini") |>
  as_module("text -> sentiment: enum('positive', 'negative', 'neutral')")

# Use repeatedly
classifier$predict(text = "Love it!")
classifier$predict(text = "Hate it!")
```
### Full Control: `signature()` + `module()`

```{r full-module}
# For optimization and complex configurations
sig <- signature("context, question -> answer")
mod <- module(sig, type = "predict")

# With custom template
mod <- module(
  sig,
  type = "predict",
  template = "Context:\n{context}\n\nQuestion: {question}"
)
```

---

## Module Types Decision Tree

```
What do you need?
│
├─ Simple text in/out → type = "predict" (PredictModule)
│   └─ Q&A, classification, summarization, extraction
│
└─ Tool use / multi-step reasoning → type = "react" (ReactModule)
    └─ Agents, search, calculations, API calls
```

| Type | Class | Use Case |
|------|-------|----------|
| `"predict"` | `PredictModule` | Standard text generation |
| `"react"` | `ReactModule` | Tool-calling agents |

---

## Running Modules

### Single Execution

```{r run-single}
#| eval: false
# Using run()
result <- run(mod, question = "What is R?", .llm = chat_openai())

# Using predict method
result <- mod$predict(question = "What is R?")
```

### Batch Processing

```{r run-batch}
#| eval: false
# Vector inputs
results <- mod$predict(text = c("Great!", "Awful!", "Meh"))

# Data frame
new_data <- data.frame(text = c("A", "B", "C"))
results <- predict(mod, new_data = new_data)

# run_dataset for full control
results <- run_dataset(mod, dataset, .llm = llm)
```

### Show Prompt

```{r show-prompt}
#| eval: false
# See the prompt being sent
result <- run(mod, question = "Test", .llm = llm, .show_prompt = TRUE)
```

---

## Metrics & Evaluation

### Built-in Metrics

| Metric | Use Case |
|--------|----------|
| `metric_exact_match()` | Exact string equality |
| `metric_f1()` | Token overlap similarity |
| `metric_contains()` | Output contains expected substring |
| `metric_field_match(fields)` | Match specific output fields |
| `metric_threshold(metric, 0.8)` | Apply threshold to any metric |
| `metric_custom(fn)` | Custom scoring function |

### Metric Selection Guide

```
What are you measuring?
│
├─ Exact correctness → metric_exact_match()
│   └─ Facts, names, simple answers
│
├─ Approximate match → metric_f1() or metric_threshold(metric_f1(), 0.8)
│   └─ Paraphrased answers, spelling tolerance
│
├─ Contains key info → metric_contains()
│   └─ Important terms must appear
│
├─ Specific fields → metric_field_match(c("answer", "score"))
│   └─ Multi-field outputs
│
└─ Custom logic → metric_custom(my_scorer_fn)
    └─ Domain-specific evaluation
```

### Running Evaluation

```{r evaluate}
#| eval: false
# Basic evaluation
result <- evaluate(mod, testset, metric = metric_exact_match(), .llm = llm)
result$mean_score
#> 0.85

# With dsp()
result <- evaluate_dsp(
  "question -> answer",
  testset,
  metric = metric_exact_match()
)
```

---

## Optimization

### Grid Search

```{r grid-search}
#| eval: false
# Search over parameters
mod$optimize_grid(
  devset = train_data,
  metric = metric_exact_match(),
  .llm = llm,
  parameters = list(
    prompt_style = c("concise", "detailed"),
    temperature = c(0.0, 0.3, 0.7)
  )
)

# Check results
module_trials(mod)
module_metrics(mod)
```

### Teleprompters

```{r teleprompters}
#| eval: false
# Few-shot learning
tp <- LabeledFewShot(k = 4L, metric = metric_exact_match())
compiled <- compile(tp, mod, trainset, .llm = llm)

# Grid search teleprompter
tp <- GridSearchTeleprompter(
  instructions = c("Be concise", "Be thorough"),
  metric = metric_exact_match()
)
compiled <- compile(tp, mod, trainset, .llm = llm)
```

---

## Debugging & Inspection

### Inspect Last Call

```{r last-prompt}
#| eval: false
# Get the last prompt sent
get_last_prompt()
#> ── Last Prompt ──
#> Model: gpt-4o-mini via OpenAI
#> Input tokens: 45
#>
#> Question: What is 2+2?
```

### Prompt History

```{r history}
#| eval: false
# View recent prompts
inspect_history(n = 5)

# Clear history
clear_prompt_history()
```

### Module Inspection

```{r mod-inspect}
#| eval: false
# Detailed module state
mod$inspect()

# Last trace from dsp()
get_last_trace()
```

### Traces

```{r traces}
#| eval: false
# Get all traces from a module
traces <- mod$get_traces()

# Export to tibble
export_traces(mod)

# Summary statistics
summarize_traces(mod)

# Clear traces
clear_traces(mod)
```

---

## Production

### Save/Restore Configuration

```{r persistence}
#| eval: false
# Save optimized config to pins board
pin_module_config(mod, board, "my-classifier-v1")

# Restore later
mod <- restore_module_config(board, "my-classifier-v1", .llm = llm)
```

### Vitals Integration

```{r vitals}
#| eval: false
# Convert module to vitals solver
solver <- as_vitals_solver(mod)

# Use with vitals::eval_task()
library(vitals)
result <- eval_task(my_task, solver)

# Convert vitals scorer to dsprrr metric
my_metric <- as_dsprrr_metric(vitals_scorer)
```

### Project Template

```{r template}
#| eval: false
# Create production-ready project structure
use_dsprrr_template("my_project")

# Validate workflow configuration
validate_workflow("workflow.yml")
```

---
## Common Patterns

### RAG (Retrieval-Augmented Generation)

```{r rag}
#| eval: false
sig <- signature(
  "context, question -> answer",
  instructions = "Answer based only on the provided context."
)
mod <- module(sig, type = "predict")

# Use with retrieved context
result <- run(mod,
  context = retrieved_docs,
  question = user_query,
  .llm = llm
)
```

### Classification

```{r classification}
#| eval: false
classifier <- chat_openai() |>
  as_module("text -> label: enum('spam', 'not_spam')")

labels <- classifier$predict(text = emails)
```

### Extraction

```{r extraction}
#| eval: false
extractor <- chat_openai() |>
  as_module("text -> entities: list[string], summary: string")

result <- extractor$predict(text = document)
result$entities
result$summary
```

### Chain of Thought (via instructions)

```{r cot}
sig <- signature(
  "problem -> answer",
  instructions = "Think step by step. Show your reasoning before the final answer."
)
```
