---
title: "Compile and optimize"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compile and optimize}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  # The cassettes were recorded with the response cache off. Keep it off so
  # every recorded request is replayed, in order.
  dsprrr::configure_cache(enable_memory = FALSE, enable_disk = FALSE)

  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

`compile()` runs an optimizer, called a teleprompter, over a module and a set
of labeled examples, and returns a copy of the module with new demos or
instructions. This guide goes through one complete run: define the task, the
labeled data and the metric, compile with two teleprompters, score the result
on rows the optimizer never saw, check what changed, and save the compiled
module.

## Setup

```{r setup}
#| eval: true
#| message: false
library(dsprrr)
library(ellmer)

llm <- chat_openai(model = "gpt-6-luna")
search_llm <- chat_openai(
  model = "gpt-6-luna",
  params = params(reasoning_effort = "low")
)
```

`llm` runs and scores the module. `search_llm` thinks less, which makes it
cheaper, and is used only while searching. The output on this page was
recorded with `gpt-5-mini` (runs) and `gpt-4o-mini` (search) and is replayed
when the site is built, so `gpt-6-luna` may give different answers and scores.

## Task, data and metric

The running example triages customer emails. The module returns three fields:
a category, a priority, and whether the email needs a reply.

```{r task}
#| eval: true
email_classifier <- signature(
  inputs = list(
    input("email", description = "Customer email text")
  ),
  output_type = type_object(
    category = type_enum(
      values = c("complaint", "inquiry", "feedback", "spam")
    ),
    priority = type_enum(values = c("urgent", "normal", "low")),
    needs_response = type_boolean()
  ),
  instructions = "Classify customer emails by category and priority."
) |>
  module()
```

Optimizers learn from a training set. A separate test set, which no optimizer
sees, tells you how the result does on new emails:

```{r data}
#| eval: true
email_trainset <- tibble::tibble(
  email = c(
    "Your product broke after 2 days! I want a refund immediately!",
    "Hi, I'm wondering if you ship to Canada?",
    "Just wanted to say your customer service is excellent.",
    "CLICK HERE FOR FREE PRIZES!!! Limited time offer!!!"
  ),
  category = c("complaint", "inquiry", "feedback", "spam"),
  priority = c("urgent", "normal", "low", "low"),
  needs_response = c(TRUE, TRUE, FALSE, FALSE)
)

test_emails <- tibble::tibble(
  email = c(
    "The package was damaged during shipping.",
    "Do you offer student discounts?",
    "Your team went above and beyond. Thank you!"
  ),
  category = c("complaint", "inquiry", "feedback"),
  priority = c("normal", "low", "low"),
  needs_response = c(TRUE, TRUE, FALSE)
)
```

Seven rows are enough to show the calls, not to trust the scores;
[Metrics and evaluation](concepts-why-metrics-matter.html) covers how much
data a comparison needs. If your labeled rows are in one table,
`split_dataset(data, prop = 0.8, seed = 1)` returns a `train` and a `val`
part.

dsprrr calls a metric with the module's output and the whole data row, so
`field` names both the output field and the column to compare it with:

```{r metric}
#| eval: true
category_metric <- metric_exact_match(field = "category")

category_metric(
  list(category = "spam", priority = "low", needs_response = FALSE),
  email_trainset[4, ]
)
```

This metric checks the category only. The evaluation below shows what that
leaves out.

Before compiling anything, score the module as it is. That baseline is the
number a compiled module has to beat:

```{r baseline}
#| eval: false
baseline <- email_classifier |>
  evaluate(test_emails, metric = category_metric, .llm = llm)
baseline$mean_score
```

## Compile with LabeledFewShot

`LabeledFewShot` samples `k` training rows and attaches them to the module as
demos. It makes no model calls, and it does not score its choices: the
metric's `field` only tells it which column holds each demo's output.

```{r labeled}
#| eval: true
optimized_simple <- email_classifier |>
  compile(LabeledFewShot(k = 3L, metric = category_metric), email_trainset)

str(optimized_simple$demos[[1]])
```

`compile()` returns a new module and leaves its input alone. The optimization
record lists the training rows that became demos:

```{r labeled-status}
#| eval: true
optimized_simple$is_compiled()
email_classifier$is_compiled()
optimization_result(optimized_simple)$lineage$selected_rows
```

Each demo carries one output column, here `category`. The model sees three
examples of how an email maps to a category and none of how priority is set.

## Search instructions with GridSearchTeleprompter

`GridSearchTeleprompter` scores a fixed set of variants and keeps the best.
Each row of `variants` is one candidate: `instructions_suffix` is appended to
the module's instructions, an `instructions` column replaces them, and a
`template` column replaces the prompt template. It also attaches `k` demos
drawn from the training rows.

```{r compilation-process}
#| cassette: true
email_variants <- data.frame(
  id = c("professional", "friendly", "analytical"),
  instructions_suffix = c(
    ". Use professional business judgment.",
    ". Consider customer satisfaction and relationship.",
    ". Analyze linguistic patterns and intent markers."
  )
)

grid_search <- GridSearchTeleprompter(
  variants = email_variants,
  metric = category_metric,
  k = 2L
)

# The validation rows and demos are drawn with R's random number generator
set.seed(37)
optimized_grid <- email_classifier |>
  compile(grid_search, email_trainset, .llm = search_llm)

optimized_grid |>
  run(
    email = "I've been waiting for 2 hours on hold! This is unacceptable!",
    .llm = llm
  )
```

The search ran on `search_llm`, and the result is used with `llm`. A variant
wins because it suited the model that ran the search, so confirm the result
with the model you will deploy, on held-out rows.

## Score on held-out data

`evaluate()` runs every test row and scores it:

```{r evaluation}
#| cassette: true
results <- optimized_grid |>
  evaluate(test_emails, metric = category_metric, .llm = llm)
results$mean_score
results$scores

strict_metric <- metric_field_match(
  c("category", "priority", "needs_response")
)
strict_results <- optimized_grid |>
  evaluate(test_emails, metric = strict_metric, .llm = llm)
strict_results$mean_score
```

Every category is right, but only one email in three is right on all three
fields. The predictions show where the difference comes from:

```{r misses}
tibble::tibble(
  email = test_emails$email,
  expected = test_emails$priority,
  predicted = vapply(results$predictions, `[[`, character(1), "priority")
)
```

Both misses are priorities. Neither the metric nor the demos covered priority,
so the search had no way to improve it: an optimizer improves only what its
metric measures. Compare `results$mean_score` with the baseline from before,
on the same rows, and keep the compiled module only if it wins there.

## Inspect what changed

`optimization_result()` reports the same fields for every optimizer:
`best_score`, `best_params`, `trials`, `lineage`, `budget`, `stop_reason`, and
optimizer-specific details under `extensions`.

```{r inspect}
res <- optimization_result(optimized_grid)
res
res$best_params$id
res$trials[c("trial_id", "score", "n_evaluated")]
```

Each variant was scored on a single email, so all three tied and the first one
won. Without a `valset`, `GridSearchTeleprompter` holds out a random fifth of
`trainset` (rounded up, at most `eval_sample_size` rows) to score the
variants, which is one row here. With real data, pass a validation set large enough to separate
the candidates:

```{r grid-valset}
#| eval: false
splits <- split_dataset(labeled_emails, prop = 0.8, seed = 1)

optimized_grid <- email_classifier |>
  compile(grid_search, splits$train, valset = splits$val, .llm = search_llm)
```

The winning instructions are in `optimized_grid$signature@instructions`, and
the demos in `optimized_grid$demos`.

## Save and reload

`save_program()` writes a compiled module to disk, and `load_program()` reads
it back. The file holds the signature, instructions, demos and optimization
record. Chats and credentials are not saved, so pass `.llm` when you run the
restored module. Here the `LabeledFewShot` module from earlier makes the round
trip:

```{r save}
#| eval: true
path <- tempfile(fileext = ".rds")
save_program(optimized_simple, path)

restored <- load_program(path)
restored
```

[Tutorial 6](tutorial-deploy-to-production.html) covers saving and reusing
modules in more depth.

## Other optimizers

Every teleprompter plugs into the same call. `BootstrapFewShot`, for example,
runs the module on training rows and keeps its own outputs that pass the
metric as demos. Those demos carry all three output fields, where the labeled
ones above carried only the category:

```{r bootstrap}
#| eval: false
bootstrapped <- email_classifier |>
  compile(
    BootstrapFewShot(metric = category_metric, max_labeled_demos = 0L),
    email_trainset,
    .llm = llm
  )
```

`compile()` always returns a copy. To tune settings such as reasoning effort on the
module itself, use `optimize_grid()`, which updates the module in place
([Tutorial 5](tutorial-optimize-your-module.html) walks through it).

[Choose an optimizer](advanced-optimization.html) compares the eleven
teleprompters and shows how to call each one.
