---
title: "Automatic Prompt Optimization with dsprrr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automatic Prompt Optimization with dsprrr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(dsprrr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = eval_vignette()
)

vcr::setup_knitr()
```

## Introduction

Welcome to **automatic prompt optimization** – dsprrr's most powerful feature. Instead of manually tweaking prompts and examples, you can let dsprrr systematically optimize your LLM programs using your training data and evaluation metrics.

As the DSPy philosophy teaches: *"it's unproductive to launch optimization runs using a poorly designed program or a bad metric."* That's why we recommend starting with the [Evaluation and Metrics](evaluation-metrics.html) vignette to learn how to evaluate your programs properly.

This vignette assumes you know how to define metrics and evaluate modules. We'll focus on:

- **`optimize_grid()`**: Direct grid search without teleprompters
- **Teleprompters**: Optimization strategies for improving programs
- **Compilation**: The `compile_module()` workflow
- **Tidymodels Integration**: Parameter sets, resampling, and cross-validation
- **Advanced Patterns**: Iterative refinement and best practices

## Setup

```{r setup}
library(dsprrr)
library(ellmer)

# Configure your LLM (we'll use Claude as an example)
llm <- chat_claude(
  api_key = Sys.getenv("ANTHROPIC_API_KEY"),
  model = "claude-3-5-sonnet-20241022"
)
```

## Quick Start: `optimize_grid()`

The `optimize_grid()` function lets you tune a module directly without using a teleprompter. It evaluates a grid of configuration values and records the results in `module$state$trials`.

```{r optimize-grid-quick, eval = eval_vignette()}
#| cassette: compilation-optimize-grid

# Create a sentiment classifier
sig_quick <- signature("review -> sentiment: string",
  instructions = "Respond with the sentiment label only."
)

sentiment_module <- module(
  signature = sig_quick,
  type = "predict",
  template = "Review: {review}\nSentiment:"
)

# Prepare development data
dev_reviews <- tibble::tibble(
  review = "I absolutely loved this!",
  target = "positive"
)

# Define a simple metric
exact_sentiment <- function(prediction, expected_row) {
  as.numeric(tolower(prediction) == tolower(expected_row$target))
}

# Optimize directly
optimize_grid(
  sentiment_module,
  devset = dev_reviews,
  metric = exact_sentiment,
  parameters = list(prompt_style = c("baseline", "energetic")),
  control = list(progress = FALSE)
)

# Check results
sentiment_module$state$trials[, c("trial_id", "score", "n_evaluated")]
sentiment_module$state$best_score
```

### Registering Parameter Metadata

Use `module_register_parameters()` to describe tunable fields once and reuse the metadata:

```{r parameter-registration, eval = FALSE}
library(dials)

module_register_parameters(sentiment_module, list(
  temperature = list(type = "double", range = c(0, 1)),
  prompt_style = list(type = "categorical", values = c("baseline", "energetic"))
))

# Generate a parameter set
param_set <- module_parameter_set(sentiment_module)
```

### Tidymodels Integration: Resampling

Combine `optimize_grid()` with rsample for cross-validation:

```{r yardstick-resampling, eval = FALSE}
library(rsample)
library(yardstick)

# Create folds
folds <- vfold_cv(dev_reviews, v = 3)

# Wrap yardstick metric
metric <- as_dsprrr_metric(
  metric_set(accuracy),
  truth = "target",
  estimate = "prediction",
  transform = function(pred) factor(pred, levels = levels(dev_reviews$target)),
  type = "yardstick"
)

# Optimize with resampling
optimize_grid(
  sentiment_module,
  devset = dev_reviews,
  resamples = folds,
  metric = metric,
  parameters = list(temperature = c(0, 0.5, 1)),
  control = list(progress = FALSE)
)

# Inspect results
module_trials_summary(sentiment_module)
module_metric_summary(sentiment_module)
```

## Teleprompters: Optimization Strategies

Teleprompters are reusable optimization strategies inspired by PyTorch modules. dsprrr provides two core teleprompters.

### LabeledFewShot

The simplest strategy: add labeled examples from your training data as demonstrations.

```{r teleprompter-labeled}
#| cassette: compilation-labeled-fewshot

# Create a sentiment classifier
classifier <- signature(
  "text -> sentiment: enum('positive', 'negative', 'neutral')"
) |>
  module(type = "predict", template = "Analyze sentiment: {text}\n\nSentiment:")

# Prepare training data
trainset <- dsp_trainset(
  text = c(
    "This product is amazing! Best purchase ever!",
    "Terrible experience. Would not recommend.",
    "It's okay, nothing special but works."
  ),
  sentiment = c(
    "positive",
    "negative",
    "neutral"
  )
)

# Create LabeledFewShot teleprompter
teleprompter <- LabeledFewShot(
  k = 2L,           # Use 2 examples as demonstrations
  sample = TRUE,    # Randomly sample from trainset
  seed = 42L        # For reproducibility
)

# Compile the module
optimized_classifier <- compile_module(
  program = classifier,
  teleprompter = teleprompter,
  trainset = trainset
)

# The optimized module now includes demonstrations
print(optimized_classifier$demos)  # Shows the 2 selected examples

# Use the optimized classifier
result <- optimized_classifier |>
  run(text = "This is fantastic!", .llm = llm)

print(result)  # "positive"
```

**When to use LabeledFewShot:**
- You have high-quality labeled examples
- You want a quick improvement without extensive tuning
- Your task benefits from in-context learning

### GridSearchTeleprompter

Test multiple instruction or template variants to find the best one:

```{r teleprompter-gridsearch}
#| cassette: compilation-gridsearch

# Create base QA module
qa_module <- signature("context, question -> answer") |>
  module(type = "predict")

# Define instruction variants to test
variants <- data.frame(
  id = c("concise", "detailed", "analytical", "step_by_step"),
  instructions_suffix = c(
    ". Answer concisely.",
    ". Provide a detailed answer with context.",
    ". Analyze the context carefully before answering.",
    ". Think step-by-step: 1) Understand context 2) Identify relevant info 3) Answer."
  ),
  stringsAsFactors = FALSE
)

# Prepare QA training data
qa_trainset <- dsp_trainset(
  context = c(
    "The R language was created by Ross Ihaka and Robert Gentleman in 1992.",
    "Python was created by Guido van Rossum and first released in 1991.",
    "JavaScript was created by Brendan Eich in just 10 days in 1995."
  ),
  question = c(
    "Who created R?",
    "When was Python released?",
    "How long did it take to create JavaScript?"
  ),
  answer = c(
    "Ross Ihaka and Robert Gentleman",
    "1991",
    "10 days"
  )
)

# Create GridSearchTeleprompter with evaluation metric
grid_teleprompter <- GridSearchTeleprompter(
  variants = variants,
  metric = metric_f1(field = "answer"),  # Evaluate using F1 score
  k = 2L,                                # Include 2 few-shot examples
  eval_sample_size = 10L,                # Smaller sample for faster execution
  verbose = TRUE                         # Show progress
)

# Compile with grid search
optimized_qa <- compile_module(
  program = qa_module,
  teleprompter = grid_teleprompter,
  trainset = qa_trainset
)

# Check which variant performed best
print(optimized_qa$config$best_variant)  # e.g., "analytical"
print(optimized_qa$config$best_score)    # e.g., 0.85
print(optimized_qa$config$all_scores)    # Scores for all variants

# Use the optimized module
answer <- optimized_qa |>
  run(
    context = "Ruby was created by Yukihiro Matsumoto in 1995.",
    question = "Who created Ruby?",
    .llm = llm
  )
```

**When to use GridSearchTeleprompter:**
- You have specific hypotheses about what works (instruction styles, templates)
- You want to compare multiple approaches systematically
- You're willing to invest compute time for better results

## The Compilation Process

The `compile_module()` function is the main interface for optimizing programs using teleprompters:

```{r compilation-process}
#| cassette: compilation-email-classifier

# Step 1: Create your base module
email_classifier <- signature(
  inputs = list(
    input("email", description = "Customer email text")
  ),
  output_type = type_object(
    category = type_enum(values = c("complaint", "inquiry", "feedback", "spam")),
    priority = type_enum(values = c("urgent", "normal", "low")),
    needs_response = type_boolean()
  ),
  instructions = "Classify customer emails by category and priority."
) |>
  module(type = "predict")

# Step 2: Prepare comprehensive training data
email_trainset <- dsp_trainset(
  email = c(
    "Your product broke after 2 days! I want a refund immediately!",
    "Hi, I'm wondering if you ship to Canada?",
    "Just wanted to say your customer service is excellent.",
    "CLICK HERE FOR FREE PRIZES!!! Limited time offer!!!"
  ),
  category = c("complaint", "inquiry", "feedback", "spam"),
  priority = c("urgent", "normal", "low", "low"),
  needs_response = c(TRUE, TRUE, FALSE, FALSE)
)

# Step 3: Choose optimization strategy
# Option A: Simple few-shot learning
simple_optimizer <- LabeledFewShot(k = 3L)

optimized_simple <- compile_module(
  program = email_classifier,
  teleprompter = simple_optimizer,
  trainset = email_trainset
)

# Option B: Grid search with custom variants
email_variants <- data.frame(
  id = c("professional", "friendly", "analytical"),
  instructions_suffix = c(
    ". Use professional business judgment.",
    ". Consider customer satisfaction and relationship.",
    ". Analyze linguistic patterns and intent markers."
  ),
  stringsAsFactors = FALSE
)

grid_optimizer <- GridSearchTeleprompter(
  variants = email_variants,
  metric = metric_field_match(c("category", "priority")),
  k = 2L,
  verbose = TRUE
)

optimized_grid <- compile_module(
  program = email_classifier,
  teleprompter = grid_optimizer,
  trainset = email_trainset
)

# Step 4: Use the optimized module
new_email <- "I've been waiting for 2 hours on hold! This is unacceptable!"
result <- optimized_grid |>
  run(email = new_email, .llm = llm)

print(result$category)        # "complaint"
print(result$priority)        # "urgent"
print(result$needs_response)  # TRUE
```

## Complete Workflow Example

Let's build and optimize a complete document analysis system:

```{r complete-example}
#| cassette: compilation-doc-analyzer

# 1. Define the task with structured output
doc_analyzer <- signature(
  inputs = list(
    input("document", description = "Document text to analyze")
  ),
  output_type = type_object(
    summary = type_string(description = "Concise summary of main points"),
    sentiment = type_enum(
      values = c("positive", "negative", "neutral", "mixed"),
      description = "Overall document sentiment"
    ),
    recommendation = type_enum(
      values = c("approve", "review", "reject"),
      description = "Recommended action"
    )
  ),
  instructions = "Analyze documents for key information and provide recommendations."
) |>
  module(type = "predict")

# 2. Create comprehensive training data
doc_trainset <- dsp_trainset(
  document = c(
    "Our Q3 revenue increased by 25% YoY, driven by strong product sales.
     Customer satisfaction scores reached an all-time high of 92%.
     We're expanding into three new markets next quarter.",

    "Project delays have caused budget overruns of 40%. Team morale is low.
     Three key engineers resigned last month. Urgent intervention needed.",

    "The new policy proposal includes both benefits and drawbacks.
     Cost savings of $2M are projected, but implementation challenges exist.
     Stakeholder feedback has been mixed, requiring further consultation."
  ),
  sentiment = c("positive", "negative", "mixed"),
  recommendation = c("approve", "reject", "review")
)

# 3. Define a composite metric (see evaluation-metrics vignette for details)
composite_metric <- function(pred, exp) {
  scores <- c(
    sentiment = metric_exact_match(field = "sentiment")(pred, exp),
    recommendation = metric_exact_match(field = "recommendation")(pred, exp)
  )
  mean(scores)  # Simple average
}

wrapped_metric <- metric_custom(composite_metric, name = "composite_score")

# 4. Test multiple optimization strategies
# Strategy A: Few-shot with different k values
results_by_k <- list()

for (k in c(1L, 2L, 3L)) {
  teleprompter <- LabeledFewShot(k = k, seed = 42L)

  optimized <- compile_module(
    program = doc_analyzer,
    teleprompter = teleprompter,
    trainset = doc_trainset
  )

  # Evaluate on validation set (in practice, use separate validation set)
  eval_result <- evaluate_dsp(
    module = optimized,
    dataset = doc_trainset,
    metric = wrapped_metric,
    .llm = llm,
    verbose = FALSE
  )

  results_by_k[[paste0("k_", k)]] <- eval_result$mean_score
}

print(results_by_k)  # See which k works best

# Strategy B: Grid search over different instruction styles
instruction_variants <- data.frame(
  id = c("executive", "technical"),
  instructions_suffix = c(
    ". Focus on business impact and executive decision-making.",
    ". Emphasize technical accuracy and detailed analysis."
  ),
  stringsAsFactors = FALSE
)

grid_optimizer <- GridSearchTeleprompter(
  variants = instruction_variants,
  metric = wrapped_metric,
  k = 2L,
  eval_sample_size = 10L,
  verbose = TRUE
)

best_analyzer <- compile_module(
  program = doc_analyzer,
  teleprompter = grid_optimizer,
  trainset = doc_trainset
)

# 5. Use the optimized analyzer
new_document <- "
The merger proposal shows promising synergies with projected cost savings of $5M annually.
However, cultural integration challenges are significant. Employee surveys show 60% concern
about job security. Regulatory approval is likely but will take 6-8 months.
"

analysis <- best_analyzer |>
  run(document = new_document, .llm = llm)

print(analysis$sentiment)       # Likely "mixed"
print(analysis$recommendation)  # Likely "review"
```

## Advanced Optimization Patterns

### Pattern 1: Cross-Validation for Robust Optimization

Don't optimize on your entire training set—use cross-validation:

```{r cross-validation, eval=FALSE}
# Split data for proper evaluation
split_data <- function(data, train_pct = 0.7) {
  n <- nrow(data)
  train_idx <- sample(n, size = floor(n * train_pct))

  list(
    train = data[train_idx, ],
    val = data[-train_idx, ]
  )
}

# Split your data
splits <- split_data(doc_trainset)

# Optimize on training set
optimizer <- GridSearchTeleprompter(
  variants = instruction_variants,
  metric = wrapped_metric,
  k = 2,
  verbose = FALSE
)

optimized_module <- compile_module(
  program = doc_analyzer,
  teleprompter = optimizer,
  trainset = splits$train
)

# Evaluate on validation set
val_results <- evaluate_dsp(
  module = optimized_module,
  dataset = splits$val,
  metric = wrapped_metric,
  .llm = llm
)

print(paste("Validation score:", val_results$mean_score))
```

### Pattern 2: Iterative Refinement

Optimize in stages for complex tasks:

```{r iterative-refinement, eval=FALSE}
# Stage 1: Optimize for accuracy with few-shot
stage1_module <- compile_module(
  program = doc_analyzer,
  teleprompter = LabeledFewShot(k = 3),
  trainset = doc_trainset
)

# Stage 2: Further optimize with grid search
stage2_variants <- data.frame(
  id = c("refined_1", "refined_2"),
  instructions_suffix = c(
    ". Prioritize accuracy over completeness.",
    ". Ensure all key points are captured."
  ),
  stringsAsFactors = FALSE
)

stage2_module <- compile_module(
  program = stage1_module,  # Start from stage 1 results
  teleprompter = GridSearchTeleprompter(
    variants = stage2_variants,
    metric = wrapped_metric,
    k = 1  # Add one more example
  ),
  trainset = doc_trainset
)
```

### Pattern 3: Using Tidymodels for Sophisticated Grids

Leverage dials for principled parameter specification:

```{r tidymodels-grid, eval=FALSE}
library(dials)
library(rsample)

# Register parameters
module_register_parameters(doc_analyzer, list(
  temperature = list(type = "double", range = c(0, 1)),
  max_tokens = list(type = "integer", range = c(100, 500)),
  instruction_style = list(
    type = "categorical",
    values = c("concise", "detailed", "analytical")
  )
))

# Generate a parameter set
param_set <- module_parameter_set(doc_analyzer)

# Create resamples for cross-validation
folds <- vfold_cv(doc_trainset, v = 3)

# Optimize with grid search and resampling
optimize_grid(
  doc_analyzer,
  devset = doc_trainset,
  resamples = folds,
  metric = wrapped_metric,
  parameters = param_set,
  control = list(
    grid_levels = 3,    # 3 levels per parameter
    progress = TRUE
  ),
  .llm = llm
)

# Inspect results
module_trials_summary(doc_analyzer)
module_metric_summary(doc_analyzer)
```

## Debugging and Troubleshooting

### Understanding Compilation Results

```{r debugging-compilation, eval=FALSE}
# Inspect what the compiler did
optimized <- compile_module(
  program = classifier,
  teleprompter = LabeledFewShot(k = 3),
  trainset = trainset
)

# Check compilation status
is_compiled(optimized)  # TRUE

# View selected demonstrations
print(optimized$demos)

# Check configuration
print(optimized$config$teleprompter)    # "LabeledFewShot"
print(optimized$config$compilation_k)   # 3

# For GridSearch, see which variant won
if (optimized$config$teleprompter == "GridSearchTeleprompter") {
  print(optimized$config$best_variant)
  print(optimized$config$all_scores)
}
```

### Module State Management

```{r state-management, eval=FALSE}
# Create fresh copy without compilation
fresh_module <- reset_copy(optimized)
is_compiled(fresh_module)  # FALSE
length(fresh_module$demos)  # 0

# Create deep copy preserving everything
backup_module <- deepcopy(optimized)
is_compiled(backup_module)  # TRUE
identical(backup_module$demos, optimized$demos)  # TRUE

# Modify copy without affecting original
backup_module$config$custom <- "modified"
is.null(optimized$config$custom)  # TRUE
```

### Common Issues and Solutions

```{r troubleshooting, eval=FALSE}
# Issue 1: Grid search takes too long
# Solution: Reduce eval_sample_size or number of variants
fast_grid <- GridSearchTeleprompter(
  variants = variants[1:2, ],      # Test fewer variants
  metric = metric,
  eval_sample_size = 10L,          # Smaller evaluation set
  verbose = FALSE
)

# Issue 2: Demonstrations not improving performance
# Solution: Ensure training data quality and diversity
check_diversity <- function(trainset, field) {
  values <- trainset[[field]]
  unique_ratio <- length(unique(values)) / length(values)

  if (unique_ratio < 0.8) {
    warning("Low diversity in ", field, ": ", round(unique_ratio * 100), "% unique")
  }

  table(values)
}

check_diversity(trainset, "sentiment")

# Issue 3: Optimization results are unstable
# Solution: Set temperature = 0 for deterministic outputs
# and use larger evaluation sets
```

## Performance Tips

### 1. Efficient Grid Search

Start with a coarse grid, then refine:

```{r performance-grid, eval=FALSE}
# Phase 1: Test major variants
coarse_variants <- data.frame(
  id = c("brief", "detailed"),
  instructions_suffix = c(". Be brief.", ". Be detailed."),
  stringsAsFactors = FALSE
)

coarse_result <- compile_module(
  program = module,
  teleprompter = GridSearchTeleprompter(
    variants = coarse_variants,
    metric = metric,
    eval_sample_size = 20L
  ),
  trainset = trainset
)

best_style <- coarse_result$config$best_variant

# Phase 2: Refine the winning approach with more variants
# ... (define fine_variants based on best_style)

fine_result <- compile_module(
  program = module,
  teleprompter = GridSearchTeleprompter(
    variants = fine_variants,
    metric = metric,
    eval_sample_size = 50L  # Larger sample for final decision
  ),
  trainset = trainset
)
```

### 2. Optimize Training Data

```{r performance-data, eval=FALSE}
# Ensure diversity in training examples
check_diversity <- function(trainset, field) {
  values <- trainset[[field]]
  unique_ratio <- length(unique(values)) / length(values)

  if (unique_ratio < 0.8) {
    warning("Low diversity in ", field, ": ", round(unique_ratio * 100), "% unique")
  }

  table(values)
}

check_diversity(trainset, "sentiment")

# Balance classes if needed
balance_trainset <- function(data, label_col, max_per_class = NULL) {
  labels <- unique(data[[label_col]])

  balanced <- lapply(labels, function(l) {
    subset_data <- data[data[[label_col]] == l, ]

    if (!is.null(max_per_class) && nrow(subset_data) > max_per_class) {
      subset_data[sample(nrow(subset_data), max_per_class), ]
    } else {
      subset_data
    }
  })

  do.call(rbind, balanced)
}
```

### 3. Cache LLM Responses

When running multiple optimization rounds, cache responses to save time and cost:

```{r performance-cache, eval=FALSE}
# Use vcr for caching during development
vcr::use_cassette("my_optimization", {
  optimized <- compile_module(
    program = module,
    teleprompter = grid_teleprompter,
    trainset = trainset
  )
})

# The second run will use cached responses
vcr::use_cassette("my_optimization", {
  optimized_again <- compile_module(
    program = module,
    teleprompter = grid_teleprompter,
    trainset = trainset
  )
})
```

## Summary

dsprrr's compilation framework transforms tedious prompt engineering into systematic optimization:

**Key Concepts:**

1. **`optimize_grid()`** – Direct grid search for quick tuning
2. **Teleprompters** – Reusable optimization strategies
   - `LabeledFewShot`: Add demonstrations from training data
   - `GridSearchTeleprompter`: Test multiple variants systematically
3. **Compilation** – `compile_module()` applies teleprompters to programs
4. **Tidymodels Integration** – Parameter sets, resampling, cross-validation

**Best Practices:**

- Start simple with `LabeledFewShot` for quick improvements
- Use `GridSearchTeleprompter` when you have specific hypotheses to test
- Always evaluate on held-out data, not training data
- Leverage tidymodels (`dials`, `rsample`, `yardstick`) for sophisticated workflows
- Use cross-validation for robust optimization
- Consider iterative refinement for complex tasks

**The Optimization Workflow:**

```
1. Define metrics (see evaluation-metrics vignette)
2. Prepare training data
3. Choose optimization strategy (LabeledFewShot or GridSearch)
4. Compile module
5. Evaluate on validation/test set
6. Iterate as needed
```

The compilation framework makes your LLM programs:

- **More accurate** through systematic optimization
- **More maintainable** by separating logic from prompt tuning
- **More portable** across different LLMs and use cases
- **More efficient** by finding the best prompts automatically

## Next Steps

- Review [Evaluation and Metrics](evaluation-metrics.html) if you haven't already
- Experiment with different teleprompters for your use case
- Try tidymodels integration for sophisticated optimization
- Check out [Integration with vitals](vitals-integration.html) for advanced evaluation workflows
- Build a library of optimized modules for common tasks

Remember: Let dsprrr handle the prompt engineering while you focus on building great applications!
