---
title: "Understanding Signatures and Modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding Signatures and Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article explains the design of dsprrr's two core abstractions: **Signatures** and **Modules**. Understanding why they work the way they do helps you use them effectively and extend them when needed.

## The Two-Object Architecture

dsprrr uses two fundamentally different object systems:

| Component | Object System | Mutability | Purpose |
|-----------|--------------|------------|---------|
| Signature | S7 | Immutable | Define *what* the module does |
| Module | R6 | Mutable | Manage *how* it executes |

This isn't arbitrary. Each choice solves specific problems that arise when building LLM applications.

## Signatures: Immutable Contracts

A Signature is an S7 class with three properties:

```{r}
Signature <- S7::new_class(
  "Signature",
  properties = list(
    inputs = S7::new_property(S7::class_list),
    output_type = S7::new_property(S7::class_any),
    instructions = S7::new_property(S7::class_character)
  )
)
```

### Why S7?

S7 provides **validated, immutable** objects. Once you create a signature, it doesn't change:

```{r}
sig <- signature("question -> answer")

# This would error - S7 objects are immutable
sig@inputs <- list()  # Error!
```

This matters for optimization. When you tune a module's prompts, the *interface* stays constant while the *implementation* varies. If signatures could mutate mid-optimization, you couldn't trust your evaluation results.

### Why Immutability Matters

Consider what happens during optimization:

```{r}
optimize_grid(  mod,
  data = train_data,
  metric = metric_exact_match(),
  parameters = list(temperature = c(0, 0.3, 0.7))
)
```

The optimizer creates copies of the module, varies parameters, and evaluates each configuration. If the signature were mutable, a bug in one trial could corrupt all subsequent trials. Immutability guarantees isolation.

This is the same reason functional programming prefers immutable data: you can reason about state because state doesn't silently change.

### The Type System

Signatures use ellmer's type system for outputs:

```{r}
# Simple string output
signature("text -> summary: string")

# Classification with fixed options
signature("text -> sentiment: enum('positive', 'negative', 'neutral')")

# Structured output
signature(
  inputs = list(input("article")),
  output_type = ellmer::type_object(
    headline = ellmer::type_string(),
    summary = ellmer::type_string(),
    keywords = ellmer::type_array(items = ellmer::type_string())
  )
)
```

Why ellmer types instead of R's type system? Because LLMs need schemas they can follow. When dsprrr sends a request to the LLM, it includes a JSON Schema derived from the ellmer type. The LLM constrains its output to match that schema.

This is **structured output**: the LLM returns valid JSON matching your specification, not free-form text you have to parse.

### DSPy-Style String Notation

The string parser provides a concise syntax:

```{r}
# These are equivalent
signature("context, question -> answer: string")

signature(
  inputs = list(
    input("context"),
    input("question")
  ),
  output_type = ellmer::type_string()
)
```

The string notation is:

- **Concise**: One line instead of many
- **Readable**: `context, question -> answer` is self-documenting
- **DSPy-compatible**: Familiar to Python DSPy users

Use string notation for simple cases. Use explicit notation when you need:
- Input descriptions
- Complex output types
- Custom validators

## Modules: Stateful Execution

Modules are R6 classes that wrap signatures with execution logic:

```{r}
Module <- R6::R6Class(
  "Module",
  public = list(
    signature = NULL,  # Immutable S7 reference
    config = NULL,     # Mutable configuration
    state = NULL,      # Mutable runtime state (traces, trials)
    chat = NULL,       # Optional ellmer Chat

    forward = function(batch, .llm = NULL) { ... },
    optimize_grid = function(data, metric, ...) { ... },
    reset = function(hard = FALSE) { ... }
  )
)
```

### Why R6?

R6 provides **reference semantics**: the object's state changes in place. This is essential for:

1. **Accumulating traces**: Each call adds to the trace history
2. **Storing optimization results**: The best configuration persists
3. **Managing LLM connections**: Chat objects maintain conversation state

With S3/S4/S7, you'd need to return modified copies from every function:

```{r}
# Without reference semantics (hypothetical)
result <- run(mod, question = "What is 2+2?", .llm = llm)
mod <- result$updated_module  # Have to remember to capture this!
```

With R6, the module updates itself:

```{r}
# With reference semantics (actual API)
result <- run(mod, question = "What is 2+2?", .llm = llm)
# mod is already updated with new traces
```

### The Signature-Module Relationship

A module **references** a signature but doesn't own it:

```{r}
sig <- signature("question -> answer")
mod1 <- module(sig)
mod2 <- module(sig)

# Both modules share the same signature
identical(mod1$signature, mod2$signature)  # TRUE
```

This reflects the conceptual relationship: the signature is the *interface* (what inputs and outputs look like), while the module is the *implementation* (how to generate outputs from inputs).

You can have multiple modules with the same signature but different configurations:

```{r}
sig <- signature("text -> sentiment: enum('positive', 'negative', 'neutral')")

# Same interface, different implementations
fast_classifier <- module(sig)
fast_classifier$config$temperature <- 0

careful_classifier <- module(sig)
careful_classifier$config$temperature <- 0.7
```

### State Management

Modules maintain rich state:

```{r}
mod$state <- list(
  traces = list(),           # Execution history
  cache = list(),            # Optional result caching
  compiled = FALSE,          # Optimization status
  optimization_history = list(),
  trials = tibble::tibble(), # Grid search results
  best_score = NULL,
  best_params = NULL,
  best_trial = NULL
)
```

**Traces** record every LLM call: inputs, outputs, latency, tokens, cost. This enables:
- Debugging: What prompt did the module actually send?
- Monitoring: How much are LLM calls costing?
- Analysis: Which inputs cause failures?

```{r}
# After running the module
mod$get_traces()
#> # A tibble: 10 x 8
#>    timestamp           latency_ms input_tokens output_tokens ...
```

**Compilation status** tracks whether the module has been optimized. A "compiled" module has found optimal parameters through grid search or teleprompter optimization:

```{r}
mod$is_compiled()
#> [1] FALSE

optimize_grid(mod, data = devset, metric = metric_exact_match())

mod$is_compiled()
#> [1] TRUE
```

### The `forward()` Method

Every module implements `forward()`:

```{r}
forward = function(batch, .llm = NULL, trace = TRUE, ...) {
  # 1. Build prompt from inputs and configuration
  prompt <- private$build_prompt(batch)

  # 2. Call LLM with structured output
  result <- llm$chat_structured(prompt, self$signature@output_type)

  # 3. Record trace if requested
  if (trace) {
    self$state$traces <- append(self$state$traces, list(trace_info))
  }

  # 4. Return tibble with output, chat, metadata
  tibble::tibble(
    output = list(result),
    chat = list(chat_object),
    metadata = list(metadata)
  )
}
```

The `forward()` method is the core execution path. The primary execution functions, `run()` and `run_dataset()`, eventually call `forward()`. `stats::predict()` is retained only as a data-frame interoperability method.

## Program Constructors

The constructor names make execution semantics explicit:

```{r}
module(sig)                              # Standard prediction
react(sig)                               # ReAct agent with tools
best_of_n(module(sig), N = 3)            # Reward-based selection
refine(module(sig), N = 3)               # Iterative refinement
multi_chain_comparison(sig)              # Multi-path reasoning
```

Each type is an R6 subclass that extends the base Module:

### PredictModule

The workhorse for most tasks. Generates a prompt from a template, calls the LLM, returns structured output.

```{r}
PredictModule <- R6::R6Class(
  "PredictModule",
  inherit = Module,
  public = list(
    template = NULL,  # glue template for prompt construction
    demos = NULL,     # Few-shot examples

    forward = function(batch, .llm = NULL, trace = TRUE, ...) {
      # Build prompt using template and demos
      # Call LLM with signature's output_type
      # Return results
    }
  )
)
```

### Wrapper Modules

Some modules wrap other modules to add capabilities:

```{r}
# BestOfN: Run base module N times, pick best via reward function
best_mod <- BestOfNModule$new(
  base_module = mod,
  n = 3,
  reward_fn = function(input, output) {
    # Return numeric score
  }
)

# RefineModule: Run base module, then refine with feedback
refine_mod <- RefineModule$new(
  base_module = mod,
  max_iterations = 3,
  refinement_prompt = "Improve this response: {previous_output}"
)
```

Wrappers compose because they maintain the same interface: input → output. You can wrap a wrapper:

```{r}
# BestOfN over Refine: 3 parallel refinement chains, pick best
robust_mod <- BestOfNModule$new(
  base_module = RefineModule$new(base_module = simple_mod),
  n = 3
)
```

## Copying and Independence

Modules need independent copies for optimization. The `copy()` method handles this:

```{r}
original <- module(sig)
original$config$temperature <- 0.5
original$state$traces <- list(trace1, trace2)

# Create independent copy
copied <- original$copy(deep = TRUE)

# Copies have same config but independent state
copied$config$temperature  # 0.5
copied$state$traces        # Empty list - state is reset

# Modifications don't affect original
copied$config$temperature <- 0.9
original$config$temperature  # Still 0.5
```

The `deep = TRUE` argument ensures nested structures are also copied. This is critical during grid search, where each candidate configuration needs complete isolation.

## Design Rationale Summary

| Design Choice | Problem It Solves |
|--------------|-------------------|
| S7 for Signatures | Immutability enables safe optimization |
| R6 for Modules | Reference semantics for stateful execution |
| ellmer types | Structured output via JSON Schema |
| Signature-Module separation | Interface vs implementation |
| Trace accumulation | Debugging, monitoring, analysis |
| `copy(deep=TRUE)` | Isolated candidates during optimization |

## Practical Implications

### When to Create New Signatures

Create a new signature when:
- The **inputs** change (different data your module needs)
- The **output structure** changes (different shape of results)
- The **task** changes (classification vs generation vs extraction)

Don't create a new signature when:
- You want different **prompt wording** (use configuration)
- You want different **temperature** (use configuration)
- You want different **demos** (use `mod$demos`)

### When to Create New Modules

Create a new module when:
- You need **independent state** (parallel optimization)
- You want to **preserve a configuration** (before vs after optimization)
- Different **execution contexts** (dev vs prod)

Use the same module when:

- **Sequential calls** in a workflow
- **Accumulating traces** for analysis
- The same logical "agent" in your application

### Extending the System

To create a custom module type:

```{r}
MyModule <- R6::R6Class(
  "MyModule",
  inherit = Module,
  public = list(
    forward = function(batch, .llm = NULL, trace = TRUE, ...) {
      # Your custom execution logic
      # Must return tibble with output, chat, metadata columns
    }
  )
)
```

The key contract: `forward()` takes a batch of inputs and returns a tibble. Honor this contract and your module works with `run()`, `run_dataset()`, `evaluate()`, and `optimize_grid()`.

## Connection to Optimization

The two-object architecture enables optimization:

1. **Signature defines the evaluation target**: What inputs and outputs look like
2. **Module provides the search space**: Configuration parameters to vary
3. **Immutability ensures isolation**: Each trial is independent
4. **State enables tracking**: Traces record what happened

When you call `optimize_grid()`:

```{r}
optimize_grid(  mod,
  data = train_data,
  metric = metric_exact_match(),
  parameters = list(temperature = c(0, 0.3, 0.7))
)
```

The optimizer:

1. Creates `copy(deep = TRUE)` for each candidate
2. Varies config parameters per the grid
3. Runs each copy's `forward()` on the devset
4. Computes metrics from (signature-defined) outputs
5. Updates the original module with best config
6. Records trial evidence in `optimization_result()$trials`

The architecture makes this workflow clean and reliable.

## Further Reading

- **[The DSPy Philosophy](concepts-dspy-philosophy.html)** - Why treat prompts as programs
- **[How Optimization Works](concepts-optimization-theory.html)** - Theory behind teleprompters
- **[Tutorial: Optimize Your Module](tutorial-optimize-your-module.html)** - Hands-on optimization
- **[Advanced Module Types](advanced-modules.html)** - BestOfN, Refine, ReAct
