---
title: "How dsprrr works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How dsprrr works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

dsprrr treats each LLM call as a small program with a declared interface, an
implementation you can inspect, and a score you can measure. This page explains
that model and the two objects that carry it: signatures, which declare what
goes in and comes out, and modules, which hold everything that can change. By
the end you will know what `run()`, `evaluate()` and `compile()` do to a module,
and when a plain prompt is the better tool.

```{r}
#| label: setup
#| eval: true
#| message: false
library(dsprrr)
```

## Programs you can measure

A hand-written prompt mixes three things in one string: the task's interface
(a review goes in, a label comes out), the wording that gets a model to do it,
and perhaps a few examples. When you change the wording or switch models, you
re-read a handful of outputs and decide whether they look better, with no
record of how often the old version was right.

dsprrr follows [DSPy](https://dspy.ai) (Declarative Self-improving Python;
see the [paper](https://arxiv.org/abs/2310.03714)) and keeps those parts
apart. A signature declares the interface. A module holds the wording, the
examples and the settings. A metric scores outputs against examples you trust,
and an optimizer searches over the module's wording and examples for the
version with the best score.

If you model data in R, the split is familiar. A formula such as `y ~ x`
states the model, `lm()` fits it to data, and you judge the fit on rows it
has not seen. A signature states the task, `compile()` fits the prompt to
labeled examples, and `evaluate()` scores the result on held-out rows.

## Before and after

Here is a sentiment classifier written as a plain ellmer prompt:

```{r}
chat <- ellmer::chat_openai(model = "gpt-4.1-mini")
chat$chat(
  "Is this review positive, negative or neutral? The battery died after a week."
)
```

The reply is a sentence you have to parse, the wording is frozen inside the
string, and you have no measure of how often it is right. The same task in
dsprrr:

```{r}
classify <- module(signature(
  "review -> sentiment: enum('positive', 'negative', 'neutral')",
  instructions = "Classify the sentiment of a product review."
))

run(classify, review = "The battery died after a week.", .llm = chat)
```

`run()` returns a named list such as `list(sentiment = "negative")`. The
signature's output type is sent to the provider as a JSON schema, so the answer
comes back as one of the three labels rather than a sentence to parse. With a
few labeled reviews you can measure the module and let an optimizer improve it:

```{r}
reviews <- tibble::tibble(
  review = c(
    "The battery died after a week.",
    "Does exactly what it says on the box.",
    "Arrived on time. Nothing special.",
    "Best purchase I have made this year.",
    "The strap broke on day two.",
    "Fine for the price.",
    "Customer service never replied.",
    "Works well, and setup took two minutes."
  ),
  sentiment = c(
    "negative", "positive", "neutral", "positive",
    "negative", "neutral", "negative", "positive"
  )
)
splits <- split_dataset(reviews, prop = 0.5, seed = 1)
metric <- metric_exact_match(field = "sentiment")

before <- evaluate(classify, splits$val, metric, .llm = chat)

tuned <- classify |>
  compile(LabeledFewShot(k = 3L), splits$train, .llm = chat)
after <- evaluate(tuned, splits$val, metric, .llm = chat)

c(before = before$mean_score, after = after$mean_score)
```

Eight rows are enough to show the calls, not to trust the numbers;
[Metrics and evaluation](concepts-why-metrics-matter.html) covers how much
data a comparison needs.

## Signatures declare the interface

A signature lists the inputs, the output fields with their types, and the
instructions. Printing one shows what the module will ask for:

```{r}
#| eval: true
sig <- signature(
  "review -> sentiment: enum('positive', 'negative', 'neutral')",
  instructions = "Classify the sentiment of a product review."
)
sig
```

The string notation accepts `str`, `int`, `float`, `bool`,
`enum('a', 'b')` (or `Literal['a', 'b']`), `list[...]` and `Optional[...]`,
and several outputs separated by commas:
`"review -> sentiment: enum('positive', 'negative'), confidence: float"`.
For input descriptions or nested objects, pass `inputs = list(input(...))`
and an ellmer type such as `ellmer::type_object()`; see
[Tutorial 3: Extract structured data](tutorial-structured-outputs.html).

Signatures are S7 objects, and S7 objects behave like values: modifying one
produces a new signature and leaves every other copy unchanged. That holds for
the helpers `with_instructions()`, `append_instructions()` and
`with_reasoning()`, and for direct assignment with `@<-`:

```{r}
#| eval: true
terse <- append_instructions(sig, "Reply with the label only.")
terse@instructions
sig@instructions

edited <- sig
edited@instructions <- "Label the review."
sig@instructions
```

This is what lets an optimizer try many instruction variants safely. Each
candidate gets its own signature; the one held by your baseline module never
changes underneath it.

## Modules hold everything that can change

A module is an R6 object that wraps a signature together with the parts that
vary between versions of the program:

| Field | Holds | Changed by |
|---|---|---|
| `mod$signature` | The interface | Optimizers that tune instructions swap in a new signature |
| `mod$demos` | Few-shot examples | `compile()` with `LabeledFewShot()` or `BootstrapFewShot()` |
| `mod$config` | Settings such as `temperature` | You, or `optimize_grid()` |
| `mod$chat` | An optional ellmer chat for this module | You, via `module(sig, chat = ...)` |
| `mod$state` | Traces and optimization results | `run()`, `evaluate()` and optimizers |

R6 objects have reference semantics. Assigning a module to a new name gives
you a second name for the same object, so a change through either name shows
up in both. `mod$copy()` makes an independent module with the same signature,
demos and config, and no traces:

```{r}
#| eval: true
classify <- module(sig)

same <- classify
same$config$temperature <- 0
classify$config$temperature

twin <- classify$copy()
twin$config$temperature <- 1
classify$config$temperature
```

State is what makes a module inspectable. Every `run()` appends a trace with
the timestamp, latency, token counts, cost, model, prompt and response.
`mod$get_traces()` returns them as a tibble with one row per call,
`summarize_traces(mod)` totals them, and `get_last_prompt()` prints the most
recent prompt and response.

## Running, evaluating and compiling

Five functions cover the life of a module:

| Function | What it does | Returns | Changes the module |
|---|---|---|---|
| `run(mod, ...)` | Calls the model for the given inputs | A named list of outputs | Adds traces |
| `run_dataset(mod, data)` | Calls the model once per row | `data` plus a `result` list-column | Adds traces |
| `evaluate(mod, data, metric)` | Runs every row and scores it with `metric` | `mean_score`, `scores`, `predictions` and more | Adds traces |
| `compile(mod, teleprompter, trainset)` | Searches demos or instructions | A new, compiled module | No |
| `optimize_grid(mod, data, metric, parameters = )` | Scores every combination of settings | The module, updated | Yes, in place |

Most of what an optimizer tunes is discrete: which instruction wording, which
examples to show. So optimizers propose candidates, score each with your
metric on your data, and keep the best. `mod$is_compiled()` reports whether a
module came out of that process.
[How optimization works](concepts-optimization-theory.html) describes the
individual optimizers.

## Other module types

`module()` makes a single prediction. Other constructors change how a call is
carried out while keeping the same `run()` and `evaluate()` interface:
`chain_of_thought()` adds a reasoning field, `react()` lets the model call
tools, `best_of_n()` and `refine()` retry against a reward function,
`multi_chain_comparison()` and `ensemble()` combine several answers, and
`pipeline()` chains modules. [Choose a module type](advanced-modules.html)
compares them.

When a step is plain R logic, or needs to call other modules in a way a
pipeline cannot express, wrap a function with `module_fn()`:

```{r}
#| eval: true
count_words <- module_fn(
  "text -> n_words: int",
  function(text) list(n_words = lengths(strsplit(text, "\\s+")))
)
run(count_words, text = "The battery died after a week.")
```

Function modules work with `run()`, `run_dataset()` and `evaluate()`.
`compile()` refuses them, because an optimizer cannot see inside an arbitrary
R function.

## When plain prompting is still fine

Not every LLM call needs a signature, a dataset and an optimizer. A direct
ellmer call is often the better choice when the task is simple and unlikely to
change, when you are still exploring and do not yet know what a good output
looks like, or when the prompt runs once and there is nothing to reuse.

The structure pays off once a call runs repeatedly, has to keep working when
you change models, or needs evidence that a change helped. That evidence does
not always require hand-labeled answers: `metric_contains()` checks for a
fixed pattern, a custom metric can check format or length, and a second model
can grade outputs against a rubric (see
[Evaluate with vitals](vitals-recipes.html)).

The next page, [Metrics and evaluation](concepts-why-metrics-matter.html),
covers how to choose a metric and how far to trust a score.
