---
title: "dsprrr vs. DSPy: Feature Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dsprrr vs. DSPy: Feature Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

dsprrr is an R implementation of [DSPy](https://dspy.ai)'s programming
model, built on [ellmer](https://ellmer.tidyverse.org) and tidyverse
conventions. If you know DSPy, this page tells you what carries over, what
is different, and what is not (yet) available.

## Version baseline

This comparison was checked against DSPy **3.3.0**, released on 2026-08-03.
That release introduces experimental `Flex` and `ReActV2` modules, advances an
experimental typed provider-neutral `LMRequest -> LMResponse` migration path,
and hardens errors, execution limits, and serialization. See the
[DSPy 3.3.0 release notes](https://github.com/stanfordnlp/dspy/releases/tag/3.3.0).

dsprrr adopts the durable contracts that fit R and ellmer. It does not present
an existing prompt optimizer as equivalent to a new DSPy feature when the
execution or safety model differs.

dsprrr is not a line-by-line port. It follows DSPy's concepts —
signatures, modules, metrics, and optimizers ("teleprompters") — while
embracing R idioms: tibbles in and out, S7/R6 objects, and ellmer for all
provider communication.

## Modules

| DSPy | dsprrr | Notes |
|------|--------|-------|
| `dspy.Predict` | `module(sig, type = "predict")` | Core predictor |
| `dspy.ChainOfThought` | `chain_of_thought()`, `with_reasoning()` | Implemented as signature transforms |
| `dspy.ReAct` / experimental `ReActV2` | `module(sig, type = "react")` | Native ellmer turn history, tool-call IDs, parallel calls per assistant turn, enforced iteration limit, then structured finalization; behaviorally aligned, not a port of the Python class |
| `dspy.ProgramOfThought` | `program_of_thought()` | Generates and executes **R** code (not Python); accepts either a caller-owned reused runner or a fresh per-invocation interpreter factory |
| `dspy.CodeAct` | `code_act()` | Hybrid tools + R code execution with an enforced inner tool-call limit; accepts either a caller-owned reused runner or a fresh per-invocation interpreter factory. The built-in runner is trusted-input-only, and sandboxed backends can implement the runner protocol |
| `dspy.BestOfN` | `best_of_n()` | Reward-function-guided retries |
| `dspy.Refine` | `refine()` | Retries with LLM-generated feedback |
| `dspy.MultiChainComparison` | `multi_chain_comparison()` | |
| `dspy.RLM` | `rlm_module()` (experimental) | Inference-time adaptive exploration over an R REPL, with optimizable action and extraction predictors, inherited or separate sub-LMs, replayed recursive values, typed submission repair, and trajectory metadata |
| Experimental `dspy.Flex` | `flex()` | GEPA can optimize a bounded predictor graph or an R `forward()` program. dsprrr requires an explicit fresh interpreter for executable source rather than choosing a default sandbox |
| `dspy.Parallel` / `Module.batch` | `run_dataset()`, `run(..., .concurrency = concurrency_control(...))` | Batch over a data frame; heterogeneous (module, example) fan-out is not yet a dedicated module |
| `dspy.majority` | `ensemble()` with `reduce_majority()` | Plus `reduce_weighted_vote()`, `reduce_best_by_metric()` |
| `dspy.KNN` | `KNNFewShot` teleprompter / KNN module | Bring-your-own vectorizer (e.g., `ragnar::embed_openai()`) |
| Retrieval (custom functions) | `rag_module()` + ragnar | First-class ragnar retriever integration |

For ProgramOfThought, CodeAct, and RLM, choose either a caller-owned `runner`
or a zero-argument `interpreter_factory` that creates a fresh runner per
invocation. Executable Flex accepts only the factory form and requires an
enforced sandbox by default; JSON Flex needs no interpreter. See [Flex:
Optimize the Whole Program](flex-optimization.html) for the practical choice
between those modes.

The one-call `rlm()` helper is the exception to the explicit-binding rule: it
creates a fresh managed `mcp-repl` sandbox factory by default. That path is
intended for compact JSON-compatible context. For trusted rich R objects, opt
into `r_code_runner(persistent = TRUE)` so one callr process stages the context
once and keeps derived values for the invocation.

### RLM parity and intentional differences

dsprrr now matches the DSPy 3.3 RLM execution contracts that determine how an
investigation behaves:

- one factory-owned interpreter belongs to each invocation, and persistent
  backends retain state across its iterations;
- `generate_action` and `extract` are child predictors visible to graph
  optimizers;
- `sub_lm = NULL` inherits the outer LM, while an explicit sub-LM can handle
  focused queries;
- `llm_query()` and `llm_query_batched()` return host-produced values to the
  same generated R evaluation through validated replay;
- invalid typed `SUBMIT()` calls become repairable observations rather than
  silently becoming final output;
- execution observations use a 10,000-character head-and-tail module bound
  after any stricter runner transport limit;
  and
- structured results include the output source, bounded REPL trajectory,
  recursive-call counts, usage, and runner policy.

The remaining RLM differences are intentional and user-visible:

| Boundary | DSPy | dsprrr |
|---|---|---|
| Generated language | Python | R, with signature inputs under `.context` |
| Returned value | Prediction with trajectory and top-level `final_reasoning` | `run()` output plus structured metadata; per-step reasoning is in the bounded trajectory and output source is explicit |
| Interpreter binding | Optional interpreter may be supplied to each `forward()` call | Runner or factory is bound when constructing `rlm_module()`; the one-call `rlm()` helper creates that binding for one invocation |
| Default local execution | DSPy interpreter configuration | `rlm()` creates a fresh managed `mcp-repl` OS sandbox with network disabled and workspace writes allowed |
| Large or rich input | `SandboxSerializable` can define interpreter-specific one-time staging | Persistent trusted callr stages serializable native R objects once; no general public custom-staging protocol yet. Managed MCP requires the final raw-or-compressed JSON-RPC request to fit 7 KB and each encoded control frame to fit 3,000 bytes |
| Typed outputs | Signature adapters validate the configured Python/Pydantic type | RLM validates explicit ellmer string, number, integer, boolean, enum, array, and object types; opaque `TypeJsonSchema` nodes are rejected rather than accepted without validation |
| Stored trajectory | Full interpreter trajectory | Bounded, formatted execution evidence; large raw runner output is not retained |
| Recursive replay | Interpreter-dependent | One nonce-bound, schema-checked ordered ledger returns query/tool values in code, but cannot roll back direct external side effects before a request |
| Batch failures | Per-provider failures become ordered error values | Per-request provider failures become ordered `[ERROR] ...` values; unexpected or batch-infrastructure failures terminate |
| Host tools | Interpreter-defined | Declared RLM tools execute in the host process, outside the guest sandbox; the bridge caps one generated step at 1,000 tool calls |

RLM remains experimental in dsprrr. The execution and result contracts are
tested; the convenience API may still evolve. See [the deterministic release
regression](tutorial-rlm-dsprrr.html) and [the RLM execution
contract](how-rlm-works.html).

One notable difference: DSPy 3.0 *removed* `dspy.Assert`/`dspy.Suggest`
in favor of `BestOfN`/`Refine`. dsprrr keeps both styles: declarative
assertions with retry/backtracking (`with_assertions()`,
`assert_output()`, `suggest_output()`) *and* the `best_of_n()`/`refine()`
wrappers. If you prefer the modern DSPy style, use the wrappers; use
assertions when you want declarative output contracts with automatic
feedback injection.

## Optimizers (teleprompters)

| DSPy | dsprrr | Fidelity notes |
|------|--------|----------------|
| `LabeledFewShot` | `LabeledFewShot` | Equivalent for a root Predict module; nested-predictor graphs, including RLM and Flex wrappers, are rejected rather than receiving mismatched root-task demos |
| `BootstrapFewShot` | `BootstrapFewShot` | Compiles ordinary pipelines jointly; programs containing Flex or RLM are rejected because their runtime predictors cannot receive root-task demos |
| `BootstrapFewShotWithRandomSearch` | `BootstrapFewShotWithRandomSearch` | Random search over ordinary labeled/bootstrap candidates; rejects programs containing Flex or RLM rather than returning a baseline-only no-op |
| `MIPROv2` | `MIPROv2` | Root Predict modules search instruction + demo candidates; nested graphs search child instructions with `max_bootstrapped_demos = 0L` |
| `SIMBA` | `SIMBA` | Adapted: hard-example mining + LLM-generated rules; simplified vs. the full introspective algorithm |
| `GEPA` | `GEPA` | Adapted reflective optimization for instructions and complete Flex sources, with separate train/validation roles, multi-objective selection, lineage, and optional retained outputs. Cached subsample merge acceptance, fine-grained resume, and built-in experiment trackers remain different |
| `COPRO` | `COPRO` | Equivalent (coordinate ascent over instructions) |
| `KNNFewShot` | `KNNFewShot` | Equivalent |
| `Ensemble` | `ensemble()` | Direct module constructor rather than a teleprompter |
| `BetterTogether` | `BetterTogether` | Chains prompt optimizers via strategy strings; does **not** alternate prompt/weight optimization (no finetuning backend) |
| `BootstrapFinetune` | — | Not implemented (planned); dsprrr currently optimizes prompts, not weights |
| `GRPO` (RL via Arbor) | — | Not implemented |
| `BootstrapFewShotWithOptuna`, `AvatarOptimizer`, `InferRules` | — | Niche/legacy in DSPy; not planned |
| — | `GridSearchTeleprompter`, `optimize_grid()` | dsprrr addition: tidymodels-style grid search over module parameters |
| — | `Omni` | dsprrr addition: independent best-of exploration plus a fresh continuation optimizer, with common validation scoring and optional mirai concurrency |
| — | `AutoResearch` | dsprrr addition: persistent research-agent loop over validated, jointly editable module snapshots with sandboxed R analysis |
| — | `MetaHarness` | dsprrr addition: fresh batch proposers plus host-owned frontier selection, lineage, budgets, and checkpoint resume |

For an RLM graph, GEPA can tune both child predictors. MIPROv2 currently tunes
their instructions only and requires `max_bootstrapped_demos = 0L`; nested demo
bootstrapping fails explicitly until predictor-local child evidence is
available. `BootstrapFewShot()`, its random-search variant, and
`LabeledFewShot()` reject programs containing an RLM because task examples do
not match the children’s `state -> ...` signatures.

### GEPA feedback metrics

DSPy's GEPA expects metrics that return a score *and textual feedback*.
dsprrr supports the same protocol:

```{r}
metric <- metric_with_feedback(
  function(prediction, expected) {
    if (identical(prediction$answer, expected)) {
      list(score = 1, feedback = "Correct.")
    } else {
      list(
        score = 0,
        feedback = paste("Wrong: expected", expected, "- check the arithmetic.")
      )
    }
  },
  field = "answer"
)

tp <- GEPA(metric = metric, generations = 5L)
compiled <- compile(tp, mod, trainset, .llm = llm)
```

The feedback for failed examples is injected into GEPA's reflection
prompt, so the reflection LLM learns *why* outputs failed, not just that
they did.

### Trace-aware metrics

DSPy 3.3 makes execution traces available to metrics used by reflective
optimization. dsprrr provides the same durable capability through
`metric_with_trace()`. The wrapped function receives the prediction, expected
row, and a stable trace envelope:

```{r}
metric <- metric_with_trace(
  function(prediction, expected, program_trace) {
    correct <- identical(prediction$answer, expected$answer)
    list(
      score = as.numeric(correct),
      feedback = paste(
        program_trace$status,
        "with",
        length(program_trace$events),
        "trace events"
      )
    )
  },
  field = "answer"
)

result <- evaluate(mod, testset, metric, .llm = llm)
result$traces[[1]][c("row_id", "epoch", "status")]
```

Each trace contains `row_id`, `epoch`, `status`, ordered `events`, and module
`metadata`. With repeated evaluation, `result$epoch_traces` preserves the
row-aligned traces for every epoch. Trace events may contain prompts, inputs,
and model responses, so handle them as potentially sensitive data.

## Signatures and types

| DSPy | dsprrr |
|------|--------|
| `"question -> answer: int"` string signatures | `signature("question -> answer: integer")` |
| Class-based signatures with `InputField`/`OutputField` | `signature(inputs = list(input(...)), output_type = ...)` |
| `Signature.with_instructions()` / `Signature.append_instructions()` | `with_instructions()` / `append_instructions()`; both return a new signature without mutating the original |
| Pydantic-typed outputs | ellmer type objects (`type_string()`, `type_enum()`, `type_object()`, `type_array()`) |
| `dspy.Image`, `dspy.Audio`, `dspy.File` | ellmer `Content` objects (images, PDFs) passed as inputs |
| `dspy.History` | Native ellmer turns preserved in ReAct metadata and traces; not a signature type |
| `dspy.Tool`, `dspy.ToolCalls`, `ToolCallResults` | ellmer `ToolDef`, `ContentToolRequest`, and `ContentToolResult`; IDs remain attached to native turns |
| `dspy.Reasoning` (native reasoning traces) | Not yet first-class; `with_reasoning()` adds a prompted reasoning field |

## Programs and composition

DSPy composes programs as Python classes with multiple predictors.
dsprrr composes pipelines:

```{r}
program <- pipeline(
  mod_retrieve,
  step(mod_answer, map = c(documents = "context")),
  mod_format
)
```

`BootstrapFewShot` compiles pipelines **jointly**, like DSPy: the teacher
pipeline runs end-to-end, final outputs are scored, and each step
harvests demonstrations from passing traces. Other teleprompters
currently optimize a pipeline's steps individually (instruction-level
optimizers operate on single modules).

## Infrastructure

| Capability | DSPy | dsprrr |
|------------|------|--------|
| LM client | `dspy.LM`; experimental typed `LMRequest -> LMResponse` migration boundary in 3.3 | Provider-neutral ellmer `Chat`; `run()` normalizes prompt and content inputs into typed requests, while a complete package-wide invocation record is still planned |
| Configuration | `dspy.configure()` / `dspy.context()` | `dsp_configure()`, `with_lm()`, `local_lm()` |
| Caching | Two-tier memory + disk | Two-tier memory + disk (`configure_cache()`) |
| Async | `acall`/`aforward`, `asyncify` | `run_async()` with promises for ordinary Predict modules and isolated background workflows for factory-backed ProgramOfThought, CodeAct, and RLM. Caller-owned interpreters and specialized streaming remain rejected rather than shared or bypassed |
| Streaming | `streamify()` + `StreamListener` | `run_stream()` + `stream_listener()`; one-shot fallback preserves specialized `forward()` semantics, while direct token streaming is limited to ordinary Predict steps and emits status events per pipeline step |
| Usage tracking | `track_usage` | `get_tokens()`, `get_cost()`, `session_cost()` |
| Parallel evaluation | `Evaluate(num_threads = ...)` | `evaluate(.concurrency = concurrency_control(...))` via mirai or ellmer's native parallelism. Declarative zero/one-step Flex is supported; executable and multi-step Flex currently require sequential rows |
| Saving programs | `save`/`load`; sanitized LM state and explicit unsafe-class opt-in in 3.3 | Versioned whole-program artifacts via `save_program()` / `load_program()` or pins, with registry-backed runtime IDs and explicit trusted opt-in |
| Observability | MLflow autolog, OpenTelemetry callbacks | Traces tibble, `inspect_history()`, `export_traces()`; package-level OpenTelemetry spans are planned on top of ellmer |
| Adapters (Chat/JSON/XML/TwoStep/BAML) | Yes | No adapter layer; ellmer's `chat_structured()` handles structured output |
| Evaluation framework | `dspy.Evaluate`, including trace-aware metrics | `evaluate()`, `eval_program()`, `metric_with_trace()`, plus **vitals** integration |

## What dsprrr has that DSPy doesn't

- **tidymodels integration**: use modules as parsnip engines, tune with
  dials parameters (`temperature`, `top_p`, `reasoning_effort`).
- **vitals integration**: bridge modules and metrics to the vitals
  evaluation framework (`as_vitals_solver()`, `as_dsprrr_metric()`).
- **ragnar integration**: production RAG with `rag_module()` and
  `ragnar_tool()`.
- **Assertions with backtracking**: kept and maintained (removed in
  DSPy 3.0).
- **Grid search compilation**: `optimize_grid()` for explicit,
  tidymodels-style parameter sweeps.

## Known gaps (roadmap)

The RLM differences above are disclosed boundaries. Some are intentional API
choices; others, such as a general public custom-staging protocol, remain parity
work. Remaining package-level gaps are:

In rough priority order, based on the stable DSPy 3.3 runtime:

1. **General custom input staging for interpreter-backed modules**, analogous
   to DSPy's `SandboxSerializable`, beyond the current persistent-callr and
   runner-specific paths.
2. **Remaining Flex gaps**: concurrent executable evaluation, fine-grained
   checkpoint resume, and GEPA's cached subsample merge-acceptance gate remain
   unimplemented. The top-level R `forward()` function, explicit interpreter
   factory, and non-portable R/Python source are intentional API differences,
   not missing parity.
3. **One package-wide invocation/result contract** carrying native turns,
   usage, cost, cache state, timing, and normalized errors across every module.
4. **Package-level OpenTelemetry spans** for module, optimizer, evaluation,
   cache, and tool activity, composed with ellmer's provider telemetry.
5. **Native reasoning-trace capture** as a typed output (analogous to
   `dspy.Reasoning`).
6. **Predictor-local RLM demo evidence** so MIPROv2 can bootstrap child demos;
   its graph mode currently tunes child instructions only, while GEPA selects
   the RLM child components explicitly.
7. **Adapter-style fallbacks** for models with weak structured-output
   support (analogous to `TwoStepAdapter`).
8. **Weight and RL optimization**, after provider-neutral training data,
   reproducibility, cost accounting, and artifact contracts are stable.

If one of these blocks your use case, please
[open an issue](https://github.com/JamesHWade/dsprrr/issues).
