---
title: "dsprrr vs. DSPy: Feature Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dsprrr vs. DSPy: Feature Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

dsprrr is an R implementation of [DSPy](https://dspy.ai)'s programming
model, built on [ellmer](https://ellmer.tidyverse.org) and tidyverse
conventions. If you know DSPy, this page tells you what carries over, what
is different, and what is not (yet) available.

## Version baseline

This comparison was checked against DSPy **3.3.0**, released on 2026-08-03.
That release introduces experimental `Flex` and `ReActV2` modules, advances an
experimental typed provider-neutral `LMRequest -> LMResponse` migration path,
and hardens errors, execution limits, and serialization. See the
[DSPy 3.3.0 release notes](https://github.com/stanfordnlp/dspy/releases/tag/3.3.0).

dsprrr adopts the durable contracts that fit R and ellmer. It does not present
an existing prompt optimizer as equivalent to a new DSPy feature when the
execution or safety model differs.

dsprrr is not a line-by-line port. It follows DSPy's concepts —
signatures, modules, metrics, and optimizers ("teleprompters") — while
embracing R idioms: tibbles in and out, S7/R6 objects, and ellmer for all
provider communication.

## Modules

| DSPy | dsprrr | Notes |
|------|--------|-------|
| `dspy.Predict` | `module(sig, type = "predict")` | Core predictor |
| `dspy.ChainOfThought` | `chain_of_thought()`, `with_reasoning()` | Implemented as signature transforms |
| `dspy.ReAct` / experimental `ReActV2` | `module(sig, type = "react")` | Native ellmer turn history, tool-call IDs, parallel calls per assistant turn, enforced iteration limit, then structured finalization; behaviorally aligned, not a port of the Python class |
| `dspy.ProgramOfThought` | `program_of_thought()` | Generates and executes **R** code (not Python); accepts either a caller-owned reused runner or a fresh per-invocation interpreter factory |
| `dspy.CodeAct` | `code_act()` | Hybrid tools + R code execution with an enforced inner tool-call limit; accepts either a caller-owned reused runner or a fresh per-invocation interpreter factory. The built-in runner is trusted-input-only, and sandboxed backends can implement the runner protocol |
| `dspy.BestOfN` | `best_of_n()` | Reward-function-guided retries |
| `dspy.Refine` | `refine()` | Retries with LLM-generated feedback |
| `dspy.MultiChainComparison` | `multi_chain_comparison()` | |
| `dspy.RLM` | `rlm_module()` | Recursive language models over an R REPL; accepts the 3.3 `max_iters` spelling and either runner lifecycle. Tool names, inputs, sub-LM responses, and authenticated control frames are validated strictly; compacted mcp-repl control responses fail closed |
| Experimental `dspy.Flex` | `flex()` | Dual-mode analogue: a safe versioned JSON graph by default, plus opt-in complete R source in a fresh interpreter with a versioned host bridge and separate predictor/tool budgets. R source defines `forward()` rather than a Python module class, and no default sandbox is silently selected |
| `dspy.Parallel` / `Module.batch` | `run_dataset()`, `run(..., .parallel = TRUE)` | Batch over a data frame; heterogeneous (module, example) fan-out is not yet a dedicated module |
| `dspy.majority` | `ensemble()` with `reduce_majority()` | Plus `reduce_weighted_vote()`, `reduce_best_by_metric()` |
| `dspy.KNN` | `KNNFewShot` teleprompter / KNN module | Bring-your-own vectorizer (e.g., `ragnar::embed_openai()`) |
| Retrieval (custom functions) | `rag_module()` + ragnar | First-class ragnar retriever integration |

For ProgramOfThought, CodeAct, and RLM, `runner` and
`interpreter_factory` are mutually exclusive and one is required. A direct
`runner` is caller-owned: dsprrr reuses the object across invocations and never
closes it, while the backend determines whether execution state persists and
whether reset is available. An `interpreter_factory` is a zero-argument
function called once per invocation; its fresh runner is owned by dsprrr and
closed exactly once when that invocation ends, including on failure. This
mirrors DSPy 3.3's lifecycle while retaining an explicit caller-owned option
for R workflows.

Executable Flex accepts only `interpreter_factory`, never a caller-owned
runner, so each call has isolated bridge state. Its default
`require_sandbox = TRUE` rejects trusted-input-only runners. The declarative
JSON mode remains the resource-free default and needs no interpreter.

One notable difference: DSPy 3.0 *removed* `dspy.Assert`/`dspy.Suggest`
in favor of `BestOfN`/`Refine`. dsprrr keeps both styles: declarative
assertions with retry/backtracking (`with_assertions()`,
`assert_output()`, `suggest_output()`) *and* the `best_of_n()`/`refine()`
wrappers. If you prefer the modern DSPy style, use the wrappers; use
assertions when you want declarative output contracts with automatic
feedback injection.

## Optimizers (teleprompters)

| DSPy | dsprrr | Fidelity notes |
|------|--------|----------------|
| `LabeledFewShot` | `LabeledFewShot` | Equivalent |
| `BootstrapFewShot` | `BootstrapFewShot` | Equivalent; compiles pipelines jointly (demos for every step harvested from end-to-end traces) |
| `BootstrapFewShotWithRandomSearch` | `BootstrapFewShotWithRandomSearch` | Equivalent |
| `MIPROv2` | `MIPROv2` | Discrete Bayesian optimization with UCB over instruction + demo candidates |
| `SIMBA` | `SIMBA` | Adapted: hard-example mining + LLM-generated rules; simplified vs. the full introspective algorithm |
| `GEPA` | `GEPA` | Adapted: reflective mutation and validated complete-program candidates for ordinary and Flex programs. dsprrr separates trainset discovery from valset selection, combines validation-example winners with the multi-metric Pareto front, supports round-robin/budget-atomic-all/custom component selection, and records immediate lineage, attempted merges, per-example scores/winners, and optional retained outputs. Cached subsample merge acceptance, fine-grained resume, and built-in experiment trackers remain different |
| `COPRO` | `COPRO` | Equivalent (coordinate ascent over instructions) |
| `KNNFewShot` | `KNNFewShot` | Equivalent |
| `Ensemble` | `Ensemble` | Equivalent |
| `BetterTogether` | `BetterTogether` | Chains prompt optimizers via strategy strings; does **not** alternate prompt/weight optimization (no finetuning backend) |
| `BootstrapFinetune` | — | Not implemented (planned); dsprrr currently optimizes prompts, not weights |
| `GRPO` (RL via Arbor) | — | Not implemented |
| `BootstrapFewShotWithOptuna`, `AvatarOptimizer`, `InferRules` | — | Niche/legacy in DSPy; not planned |
| — | `GridSearchTeleprompter`, `optimize_grid()` | dsprrr addition: tidymodels-style grid search over module parameters |
| — | `Omni` | dsprrr addition: independent best-of exploration plus a fresh continuation optimizer, with common validation scoring and optional mirai concurrency |
| — | `AutoResearch` | dsprrr addition: persistent research-agent loop over validated, jointly editable module snapshots with sandboxed R analysis |
| — | `MetaHarness` | dsprrr addition: fresh batch proposers plus host-owned frontier selection, lineage, budgets, and checkpoint resume |

### GEPA feedback metrics

DSPy's GEPA expects metrics that return a score *and textual feedback*.
dsprrr supports the same protocol:

```{r}
metric <- metric_with_feedback(
  function(prediction, expected) {
    if (identical(prediction$answer, expected)) {
      list(score = 1, feedback = "Correct.")
    } else {
      list(
        score = 0,
        feedback = paste("Wrong: expected", expected, "- check the arithmetic.")
      )
    }
  },
  field = "answer"
)

tp <- GEPA(metric = metric, generations = 5L)
compiled <- compile(tp, mod, trainset, .llm = llm)
```

The feedback for failed examples is injected into GEPA's reflection
prompt, so the reflection LLM learns *why* outputs failed, not just that
they did.

### Trace-aware metrics

DSPy 3.3 makes execution traces available to metrics used by reflective
optimization. dsprrr provides the same durable capability through
`metric_with_trace()`. The wrapped function receives the prediction, expected
row, and a stable trace envelope:

```{r}
metric <- metric_with_trace(
  function(prediction, expected, program_trace) {
    correct <- identical(prediction$answer, expected$answer)
    list(
      score = as.numeric(correct),
      feedback = paste(
        program_trace$status,
        "with",
        length(program_trace$events),
        "trace events"
      )
    )
  },
  field = "answer"
)

result <- evaluate(mod, testset, metric, .llm = llm)
result$traces[[1]][c("row_id", "epoch", "status")]
```

Each trace contains `row_id`, `epoch`, `status`, ordered `events`, and module
`metadata`. With repeated evaluation, `result$epoch_traces` preserves the
row-aligned traces for every epoch. Trace events may contain prompts, inputs,
and model responses, so handle them as potentially sensitive data.

## Signatures and types

| DSPy | dsprrr |
|------|--------|
| `"question -> answer: int"` string signatures | `signature("question -> answer: integer")` |
| Class-based signatures with `InputField`/`OutputField` | `signature(inputs = list(input(...)), output_type = ...)` |
| `Signature.with_instructions()` / `Signature.append_instructions()` | `with_instructions()` / `append_instructions()`; both return a new signature without mutating the original |
| Pydantic-typed outputs | ellmer type objects (`type_string()`, `type_enum()`, `type_object()`, `type_array()`) |
| `dspy.Image`, `dspy.Audio`, `dspy.File` | ellmer `Content` objects (images, PDFs) passed as inputs |
| `dspy.History` | Native ellmer turns preserved in ReAct metadata and traces; not a signature type |
| `dspy.Tool`, `dspy.ToolCalls`, `ToolCallResults` | ellmer `ToolDef`, `ContentToolRequest`, and `ContentToolResult`; IDs remain attached to native turns |
| `dspy.Reasoning` (native reasoning traces) | Not yet first-class; `with_reasoning()` adds a prompted reasoning field |

## Programs and composition

DSPy composes programs as Python classes with multiple predictors.
dsprrr composes pipelines:

```{r}
program <- mod_retrieve %>>%
  map_inputs(mod_answer, documents = "context") %>>%
  mod_format
```

`BootstrapFewShot` compiles pipelines **jointly**, like DSPy: the teacher
pipeline runs end-to-end, final outputs are scored, and each step
harvests demonstrations from passing traces. Other teleprompters
currently optimize a pipeline's steps individually (instruction-level
optimizers operate on single modules).

## Infrastructure

| Capability | DSPy | dsprrr |
|------------|------|--------|
| LM client | `dspy.LM`; experimental typed `LMRequest -> LMResponse` migration boundary in 3.3 | Provider-neutral ellmer `Chat`; `build_module_request()` normalizes prompt/content input, but a complete package-wide invocation record is still planned |
| Configuration | `dspy.configure()` / `dspy.context()` | `dsp_configure()`, `with_lm()`, `local_lm()` |
| Caching | Two-tier memory + disk | Two-tier memory + disk (`configure_cache()`) |
| Async | `acall`/`aforward`, `asyncify` | `run_async()` with promises for ordinary Predict modules and isolated background workflows for factory-backed ProgramOfThought, CodeAct, and RLM. Caller-owned interpreters and specialized streaming remain rejected rather than shared or bypassed |
| Streaming | `streamify()` + `StreamListener` | `run_stream()` + `stream_listener()`; one-shot fallback preserves specialized `forward()` semantics, while direct token streaming is limited to ordinary Predict steps and emits status events per pipeline step |
| Usage tracking | `track_usage` | `get_tokens()`, `get_cost()`, `session_cost()` |
| Parallel evaluation | `Evaluate(num_threads = ...)` | `evaluate(.parallel = TRUE)` via mirai or ellmer's native parallelism. Declarative zero/one-step Flex is supported; executable and multi-step Flex currently require sequential rows |
| Saving programs | `save`/`load`; sanitized LM state and explicit unsafe-class opt-in in 3.3 | Versioned whole-program artifacts via `save_program()` / `load_program()` or pins, with registry-backed runtime IDs and explicit trusted opt-in |
| Observability | MLflow autolog, OpenTelemetry callbacks | Traces tibble, `inspect_history()`, `export_traces()`; package-level OpenTelemetry spans are planned on top of ellmer |
| Adapters (Chat/JSON/XML/TwoStep/BAML) | Yes | No adapter layer; ellmer's `chat_structured()` handles structured output |
| Evaluation framework | `dspy.Evaluate`, including trace-aware metrics | `evaluate()`, `eval_program()`, `metric_with_trace()`, plus **vitals** integration |

## What dsprrr has that DSPy doesn't

- **tidymodels integration**: use modules as parsnip engines, tune with
  dials parameters (`temperature`, `top_p`, `reasoning_effort`).
- **vitals integration**: bridge modules and metrics to the vitals
  evaluation framework (`as_vitals_solver()`, `as_dsprrr_metric()`).
- **ragnar integration**: production RAG with `rag_module()` and
  `ragnar_tool()`.
- **Assertions with backtracking**: kept and maintained (removed in
  DSPy 3.0).
- **Grid search compilation**: `optimize_grid()` for explicit,
  tidymodels-style parameter sweeps.

## Known gaps (roadmap)

In rough priority order, based on the stable DSPy 3.3 runtime:

1. **Remaining Flex parity**: dsprrr now supports complete interpreter-backed
   R source, deterministic computation, control flow, explicit host tools, the
   DSPy Flex primitive family, separate runtime predictor/tool-call limits, and
   GEPA source search. It intentionally requires an explicit interpreter
   factory rather than selecting a default sandbox, uses a top-level R
   `forward()` function rather than a Python module class, and cannot port
   source between the two.
   Concurrent executable evaluation, fine-grained checkpoint resume, and
   GEPA's cached subsample merge-acceptance gate remain unimplemented.
2. **One package-wide invocation/result contract** carrying native turns,
   usage, cost, cache state, timing, and normalized errors across every module.
3. **Package-level OpenTelemetry spans** for module, optimizer, evaluation,
   cache, and tool activity, composed with ellmer's provider telemetry.
4. **Native reasoning-trace capture** as a typed output (analogous to
   `dspy.Reasoning`).
5. **Joint multi-step optimization for other instruction optimizers**
   (including MIPROv2); demo bootstrapping is already joint and GEPA now
   selects complete-program components explicitly.
6. **Adapter-style fallbacks** for models with weak structured-output
   support (analogous to `TwoStepAdapter`).
7. **Weight and RL optimization**, after provider-neutral training data,
   reproducibility, cost accounting, and artifact contracts are stable.

If one of these blocks your use case, please
[open an issue](https://github.com/JamesHWade/dsprrr/issues).
