---
title: "dsprrr vs. DSPy: Feature Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dsprrr vs. DSPy: Feature Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

dsprrr is an R implementation of [DSPy](https://dspy.ai)'s programming
model, built on [ellmer](https://ellmer.tidyverse.org) and tidyverse
conventions. If you know DSPy, this page tells you what carries over, what
is different, and what is not (yet) available. It reflects DSPy 3.x as of
mid-2026.

dsprrr is not a line-by-line port. It follows DSPy's concepts —
signatures, modules, metrics, and optimizers ("teleprompters") — while
embracing R idioms: tibbles in and out, S7/R6 objects, and ellmer for all
provider communication.

## Modules

| DSPy | dsprrr | Notes |
|------|--------|-------|
| `dspy.Predict` | `module(sig, type = "predict")` | Core predictor |
| `dspy.ChainOfThought` | `chain_of_thought()`, `with_reasoning()` | Implemented as signature transforms |
| `dspy.ReAct` | `module(sig, type = "react")` | Tool-calling agent loop via ellmer tools |
| `dspy.ProgramOfThought` | `program_of_thought()` | Generates and executes **R** code (not Python) |
| `dspy.CodeAct` | `code_act()` | Hybrid tools + R code execution; requires an explicit, opt-in `r_code_runner()` |
| `dspy.BestOfN` | `best_of_n()` | Reward-function-guided retries |
| `dspy.Refine` | `refine()` | Retries with LLM-generated feedback |
| `dspy.MultiChainComparison` | `multi_chain_comparison()` | |
| `dspy.RLM` | `rlm_module()` | Recursive language models over an R REPL |
| `dspy.Parallel` / `Module.batch` | `run_dataset()`, `run(..., .parallel = TRUE)` | Batch over a data frame; heterogeneous (module, example) fan-out is not yet a dedicated module |
| `dspy.majority` | `ensemble()` with `reduce_majority()` | Plus `reduce_weighted_vote()`, `reduce_best_by_metric()` |
| `dspy.KNN` | `KNNFewShot` teleprompter / KNN module | Bring-your-own vectorizer (e.g., `ragnar::embed_openai()`) |
| Retrieval (custom functions) | `rag_module()` + ragnar | First-class ragnar retriever integration |

One notable difference: DSPy 3.0 *removed* `dspy.Assert`/`dspy.Suggest`
in favor of `BestOfN`/`Refine`. dsprrr keeps both styles: declarative
assertions with retry/backtracking (`with_assertions()`,
`assert_output()`, `suggest_output()`) *and* the `best_of_n()`/`refine()`
wrappers. If you prefer the modern DSPy style, use the wrappers; use
assertions when you want declarative output contracts with automatic
feedback injection.

## Optimizers (teleprompters)

| DSPy | dsprrr | Fidelity notes |
|------|--------|----------------|
| `LabeledFewShot` | `LabeledFewShot` | Equivalent |
| `BootstrapFewShot` | `BootstrapFewShot` | Equivalent; compiles pipelines jointly (demos for every step harvested from end-to-end traces) |
| `BootstrapFewShotWithRandomSearch` | `BootstrapFewShotWithRandomSearch` | Equivalent |
| `MIPROv2` | `MIPROv2` | Discrete Bayesian optimization with UCB over instruction + demo candidates |
| `SIMBA` | `SIMBA` | Adapted: hard-example mining + LLM-generated rules; simplified vs. the full introspective algorithm |
| `GEPA` | `GEPA` | Adapted ("GEPA-lite"): reflective mutation + Pareto selection; supports feedback metrics via `metric_with_feedback()`; no per-component selection or inference-time search yet |
| `COPRO` | `COPRO` | Equivalent (coordinate ascent over instructions) |
| `KNNFewShot` | `KNNFewShot` | Equivalent |
| `Ensemble` | `Ensemble` | Equivalent |
| `BetterTogether` | `BetterTogether` | Chains prompt optimizers via strategy strings; does **not** alternate prompt/weight optimization (no finetuning backend) |
| `BootstrapFinetune` | — | Not implemented (planned); dsprrr currently optimizes prompts, not weights |
| `GRPO` (RL via Arbor) | — | Not implemented |
| `BootstrapFewShotWithOptuna`, `AvatarOptimizer`, `InferRules` | — | Niche/legacy in DSPy; not planned |
| — | `GridSearchTeleprompter`, `optimize_grid()` | dsprrr addition: tidymodels-style grid search over module parameters |

### GEPA feedback metrics

DSPy's GEPA expects metrics that return a score *and textual feedback*.
dsprrr supports the same protocol:

```{r}
metric <- metric_with_feedback(
  function(prediction, expected) {
    if (identical(prediction$answer, expected)) {
      list(score = 1, feedback = "Correct.")
    } else {
      list(
        score = 0,
        feedback = paste("Wrong: expected", expected, "- check the arithmetic.")
      )
    }
  },
  field = "answer"
)

tp <- GEPA(metric = metric, generations = 5L)
compiled <- compile(tp, mod, trainset, .llm = llm)
```

The feedback for failed examples is injected into GEPA's reflection
prompt, so the reflection LLM learns *why* outputs failed, not just that
they did.

## Signatures and types

| DSPy | dsprrr |
|------|--------|
| `"question -> answer: int"` string signatures | `signature("question -> answer: integer")` |
| Class-based signatures with `InputField`/`OutputField` | `signature(inputs = list(input(...)), output_type = ...)` |
| Pydantic-typed outputs | ellmer type objects (`type_string()`, `type_enum()`, `type_object()`, `type_array()`) |
| `dspy.Image`, `dspy.Audio`, `dspy.File` | ellmer `Content` objects (images, PDFs) passed as inputs |
| `dspy.History` | Implicit via ellmer `Chat$get_turns()`; not a signature type |
| `dspy.Tool`, `dspy.ToolCalls` | ellmer `ToolDef` via `as_ellmer_tool()` / `register_dsprrr_tool()` |
| `dspy.Reasoning` (native reasoning traces) | Not yet first-class; `with_reasoning()` adds a prompted reasoning field |

## Programs and composition

DSPy composes programs as Python classes with multiple predictors.
dsprrr composes pipelines:

```{r}
program <- mod_retrieve %>>%
  map_inputs(mod_answer, documents = "context") %>>%
  mod_format
```

`BootstrapFewShot` compiles pipelines **jointly**, like DSPy: the teacher
pipeline runs end-to-end, final outputs are scored, and each step
harvests demonstrations from passing traces. Other teleprompters
currently optimize a pipeline's steps individually (instruction-level
optimizers operate on single modules).

## Infrastructure

| Capability | DSPy | dsprrr |
|------------|------|--------|
| LM client | `dspy.LM` (LiteLLM; decoupling in 3.2+) | ellmer `Chat` (100+ providers via ellmer) |
| Configuration | `dspy.configure()` / `dspy.context()` | `dsp_configure()`, `with_lm()`, `local_lm()` |
| Caching | Two-tier memory + disk | Two-tier memory + disk (`configure_cache()`) |
| Async | `acall`/`aforward`, `asyncify` | `run_async()` with promises |
| Streaming | `streamify()` + `StreamListener` | `run_stream()` + `stream_listener()`; token streaming for single string fields, status events per pipeline step |
| Usage tracking | `track_usage` | `get_tokens()`, `get_cost()`, `session_cost()` |
| Parallel evaluation | `Evaluate(num_threads = ...)` | `evaluate(.parallel = TRUE)` via mirai or ellmer's native parallelism |
| Saving programs | `save`/`load`, whole-program serialization | `pin_module_config()` / `restore_module_config()` (pins-based) |
| Observability | MLflow autolog, OpenTelemetry callbacks | Traces tibble, `inspect_history()`, `export_traces()`; MLflow integration planned |
| Adapters (Chat/JSON/XML/TwoStep/BAML) | Yes | No adapter layer; ellmer's `chat_structured()` handles structured output |
| Evaluation framework | `dspy.Evaluate` | `evaluate()`, `eval_program()`, plus **vitals** integration |

## What dsprrr has that DSPy doesn't

- **tidymodels integration**: use modules as parsnip engines, tune with
  dials parameters (`temperature`, `top_p`, `reasoning_effort`).
- **vitals integration**: bridge modules and metrics to the vitals
  evaluation framework (`as_vitals_solver()`, `as_dsprrr_metric()`).
- **ragnar integration**: production RAG with `rag_module()` and
  `ragnar_tool()`.
- **Assertions with backtracking**: kept and maintained (removed in
  DSPy 3.0).
- **Grid search compilation**: `optimize_grid()` for explicit,
  tidymodels-style parameter sweeps.

## Known gaps (roadmap)

In rough priority order:

1. **Weight optimization**: `BootstrapFinetune` / RL-based optimizers.
2. **Native reasoning-trace capture** as a typed output (analogous to
   `dspy.Reasoning`).
3. **Joint multi-step optimization for instruction optimizers** (MIPROv2,
   GEPA per-component selection); demo bootstrapping is already joint.
4. **Adapter-style fallbacks** for models with weak structured-output
   support (analogous to `TwoStepAdapter`).
5. **MLflow / OpenTelemetry observability**.

If one of these blocks your use case, please
[open an issue](https://github.com/JamesHWade/dsprrr/issues).
