---
title: "Evaluation and Metrics in dsprrr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluation and Metrics in dsprrr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(dsprrr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = eval_vignette()
)

# Configure vcr for vignettes
vcr::vcr_configure(
  dir = "_vcr",
  filter_request_headers = c("Authorization", "x-api-key"),
  match_requests_on = c("method", "uri")
)

vcr::setup_knitr()
```

## Introduction

Before you can systematically improve your LLM programs, you need to **evaluate them**. As the DSPy philosophy teaches: *"it's unproductive to launch optimization runs using a poorly designed program or a bad metric."*

This vignette teaches you how to:

1. **Define metrics** that measure what matters for your task
2. **Evaluate modules** systematically on test datasets
3. **Use tidymodels integration** for advanced evaluation workflows
4. **Interpret results** to guide optimization

Think of evaluation as the foundation of the iterative development cycle:

```
Define Task → Build Module → Evaluate → Optimize → Repeat
                                ↑
                        You are here!
```

After mastering evaluation, you'll be ready to move on to [Automatic Prompt Optimization](compilation-optimization.html).

## Setup

```{r setup}
library(dsprrr)
library(ellmer)

# Configure your LLM (we'll use Claude as an example)
llm <- chat_claude(
  api_key = Sys.getenv("ANTHROPIC_API_KEY"),
  model = "claude-3-5-sonnet-20241022"
)
```

## Understanding Metrics

Metrics are functions that evaluate how well your LLM's output matches expected results. A good metric should:

- **Capture what matters** for your specific task
- **Return numeric scores** (0-1 range is conventional)
- **Handle edge cases** gracefully (missing fields, malformed outputs)
- **Be fast to compute** (you'll run it thousands of times)

dsprrr provides several built-in metrics and makes it easy to create custom ones.

### Built-in Metrics

#### Exact Match

The simplest metric checks if outputs match exactly:

```{r metrics-exact}
# Basic exact match
metric <- metric_exact_match()

# Test the metric
metric("positive", "positive") # 1 (TRUE converted to numeric)
metric("positive", "negative") # 0 (FALSE converted to numeric)

# Case-insensitive matching
metric_ignore_case <- metric_exact_match(ignore_case = TRUE)
metric_ignore_case("Positive", "positive") # 1

# Extract specific field from structured output
metric_field <- metric_exact_match(field = "sentiment")
metric_field(
  list(sentiment = "positive", confidence = 0.9),
  list(sentiment = "positive", confidence = 0.8)
) # 1 - only compares sentiment field
```

**When to use exact match:**
- Classification tasks with discrete categories
- Tasks where partial credit doesn't make sense
- Quick sanity checks during development

#### F1 Score

For text generation tasks, F1 score measures token overlap:

```{r metrics-f1}
# Token-based F1 score
metric_f1_basic <- metric_f1()

# Calculate F1 between predictions and expected text
score <- metric_f1_basic(
  "The capital of France is Paris",
  "Paris is the capital city of France"
)
print(score) # ~0.67 (good overlap but not perfect)

# With field extraction
metric_f1_field <- metric_f1(field = "answer")
score <- metric_f1_field(
  list(answer = "The Eiffel Tower", confidence = "high"),
  list(answer = "Eiffel Tower in Paris", other = "data")
)
```

**When to use F1:**
- Open-ended text generation
- Summarization tasks
- Question answering where multiple phrasings are correct

#### Contains Metric

Check if output contains specific patterns:

```{r metrics-contains}
# Check for substring
metric_has_positive <- metric_contains("positive", ignore_case = TRUE)
metric_has_positive("The result is POSITIVE", NULL) # 1
metric_has_positive("The result is negative", NULL) # 0

# Regular expression patterns
metric_has_number <- metric_contains("\\d+", fixed = FALSE)
metric_has_number("The answer is 42", NULL) # 1
metric_has_number("No numbers here", NULL) # 0

# Check specific field
metric_field_contains <- metric_contains("urgent", field = "priority")
metric_field_contains(
  list(priority = "urgent - needs immediate attention"),
  NULL
) # 1
```

**When to use contains:**
- Checking for required keywords or phrases
- Validating that certain information appears in output
- Format validation (e.g., presence of citations, references)

### Custom Metrics

Create your own evaluation logic:

```{r metrics-custom}
# Simple boolean metric
length_check <- metric_custom(
  function(pred, exp) {
    nchar(as.character(pred)) <= 100
  },
  name = "length_limit"
)

# Numeric scoring metric
similarity_metric <- metric_custom(
  function(pred, exp) {
    # Calculate some similarity score (0-1)
    pred_lower <- tolower(as.character(pred))
    exp_lower <- tolower(as.character(exp))

    # Simple character overlap ratio
    common <- sum(utf8ToInt(pred_lower) %in% utf8ToInt(exp_lower))
    total <- max(nchar(pred_lower), nchar(exp_lower))
    if (total == 0) return(1.0)

    common / total
  },
  name = "char_similarity"
)

# Use the custom metric
score <- similarity_metric("Hello World", "hello world")
```

**Pattern: Composite Metrics**

Combine multiple metrics with weights:

```{r metrics-composite}
# Create a weighted composite metric
composite_metric <- function(pred, exp) {
  scores <- c(
    # 30% weight on summary quality
    summary = metric_f1(field = "summary")(pred, exp),
    # 30% weight on correct sentiment
    sentiment = metric_exact_match(field = "sentiment")(pred, exp),
    # 40% weight on correct recommendation
    recommendation = metric_exact_match(field = "recommendation")(pred, exp)
  )

  # Weighted average
  weights <- c(summary = 0.3, sentiment = 0.3, recommendation = 0.4)
  sum(scores * weights)
}

wrapped_metric <- metric_custom(composite_metric, name = "composite_score")
```

### Threshold Metrics

Convert numeric metrics to boolean pass/fail:

```{r metrics-threshold}
# F1 score must be at least 0.8 to pass
f1_threshold <- metric_threshold(
  metric_f1(),
  threshold = 0.8,
  comparison = ">="
)

# Returns TRUE/FALSE instead of numeric score
passes <- f1_threshold(
  "The quick brown fox jumps",
  "The quick brown fox jumps over"
) # FALSE (F1 < 0.8)
```

### Multi-Field Metrics

Evaluate multiple fields simultaneously:

```{r metrics-multifield}
# Check if all specified fields match
metric_all <- metric_field_match(
  c("sentiment", "confidence"),
  require_all = TRUE
)

pred <- list(sentiment = "positive", confidence = "high", extra = "ignored")
exp <- list(sentiment = "positive", confidence = "high", other = "data")
metric_all(pred, exp) # 1 - both fields match

# Check if any field matches
metric_any <- metric_field_match(
  c("answer", "alternate_answer"),
  require_all = FALSE
)
```

## Evaluating Modules

Once you have a metric, use `evaluate_dsp()` to systematically test your module:

```{r evaluation-basic}
#| cassette: evaluation-sentiment-classifier

# Create a sentiment classifier
classifier <- signature(
  "text -> sentiment: enum('positive', 'negative', 'neutral')"
) |>
  module(type = "predict", template = "Analyze sentiment: {text}\n\nSentiment:")

# Prepare test dataset (separate from training!)
test_data <- dsp_trainset(
  text = c(
    "This product is amazing! Best purchase ever!",
    "Terrible experience. Would not recommend.",
    "It's okay, nothing special but works."
  ),
  sentiment = c(
    "positive",
    "negative",
    "neutral"
  )
)

# Evaluate with a specific metric
results <- evaluate_dsp(
  module = classifier,
  dataset = test_data,
  metric = metric_exact_match(field = "sentiment"),
  .llm = llm,
  verbose = TRUE
)

# Examine results
print(results$mean_score)   # e.g., 0.67 (2/3 correct)
print(results$scores)        # Individual scores per example
print(results$n_evaluated)   # 3
print(results$n_errors)      # 0 (no runtime errors)
```

### Understanding Evaluation Results

The `evaluate_dsp()` function returns a rich results object:

```{r evaluation-results, eval = FALSE}
str(results)
# List of 8
#  $ mean_score  : num 0.67     # Average score across examples
#  $ scores      : num [1:3]    # Per-example scores
#  $ predictions : list [1:3]   # Model outputs
#  $ metadata    : list [1:3]   # LLM metadata (tokens, cost, etc.)
#  $ n_evaluated : int 3        # Number of successful evaluations
#  $ n_errors    : int 0        # Number of failures
#  $ errors      : chr [1:0]    # Error messages (if any)
#  $ dataset     : tibble [3×2] # Input dataset with results
```

**Key fields:**
- `mean_score`: Use this for comparing modules or variants
- `scores`: Inspect to find problematic examples
- `predictions`: Debug what the model actually produced
- `metadata`: Track token usage and cost
- `errors`: Diagnose runtime issues

### Multi-Aspect Evaluation

Evaluate multiple aspects of your module's output:

```{r evaluation-multiaspect}
#| cassette: evaluation-document-analyzer

# Define a complex task
doc_analyzer <- signature(
  inputs = list(
    input("document", description = "Document text to analyze")
  ),
  output_type = type_object(
    summary = type_string(description = "Concise summary"),
    sentiment = type_enum(values = c("positive", "negative", "neutral", "mixed")),
    recommendation = type_enum(values = c("approve", "review", "reject"))
  ),
  instructions = "Analyze documents for key information and provide recommendations."
) |>
  module(type = "predict")

# Test data
test_docs <- dsp_trainset(
  document = c(
    "Our Q3 revenue increased by 25% YoY, driven by strong product sales.",
    "Project delays have caused budget overruns of 40%. Team morale is low."
  ),
  sentiment = c("positive", "negative"),
  recommendation = c("approve", "reject")
)

# Evaluate each aspect separately
sentiment_results <- evaluate_dsp(
  module = doc_analyzer,
  dataset = test_docs,
  metric = metric_exact_match(field = "sentiment"),
  .llm = llm,
  verbose = FALSE
)

recommendation_results <- evaluate_dsp(
  module = doc_analyzer,
  dataset = test_docs,
  metric = metric_exact_match(field = "recommendation"),
  .llm = llm,
  verbose = FALSE
)

# Or use a composite metric for overall quality
multi_metric <- metric_field_match(
  c("sentiment", "recommendation"),
  require_all = TRUE  # All fields must match
)

full_results <- evaluate_dsp(
  module = doc_analyzer,
  dataset = test_docs,
  metric = multi_metric,
  .llm = llm,
  verbose = TRUE
)

print(full_results$mean_score)  # Stricter: both fields must match
```

## Tidymodels Integration

dsprrr integrates deeply with the tidymodels ecosystem, allowing you to use familiar tools for evaluation.

### Using Yardstick Metrics

Convert yardstick metrics to dsprrr metrics with `as_dsprrr_metric()`:

```{r yardstick-basics}
#| cassette: evaluation-yardstick
library(yardstick)

# Prepare data with factor levels for yardstick
test_reviews <- tibble::tibble(
  text = c("Loved it", "Awful", "Pretty good", "Fantastic"),
  target = factor(
    c("positive", "negative", "positive", "positive"),
    levels = c("negative", "positive")
  )
)

# Create classifier
classifier <- signature("text -> sentiment") |>
  module(type = "predict", template = "Review: {text}\nSentiment:")

# Convert yardstick metric to dsprrr
metric <- as_dsprrr_metric(
  metric_set(accuracy),
  truth = "target",
  estimate = "prediction",
  # Transform ensures prediction matches factor levels
  transform = function(pred) {
    factor(pred, levels = c("negative", "positive"))
  },
  type = "yardstick"
)

# Evaluate using yardstick
results <- evaluate_dsp(
  module = classifier,
  dataset = test_reviews,
  metric = metric,
  .llm = llm
)

print(results$mean_score)
print(results$metrics)  # Yardstick tibble with .metric, .estimator, .estimate
```

### Using Multiple Yardstick Metrics

You can evaluate with multiple yardstick metrics at once:

```{r yardstick-multiple, eval = FALSE}
# Multi-metric evaluation
multi_metric <- as_dsprrr_metric(
  metric_set(accuracy, precision, recall, f_meas),
  truth = "target",
  estimate = "prediction",
  transform = function(pred) factor(pred, levels = levels(test_reviews$target))
)

results <- evaluate_dsp(
  module = classifier,
  dataset = test_reviews,
  metric = multi_metric,
  .llm = llm
)

# Access all metric results
print(results$metrics)
# A tibble: 4 × 3
#   .metric   .estimator .estimate
#   <chr>     <chr>          <dbl>
# 1 accuracy  binary         0.75
# 2 precision binary         0.80
# 3 recall    binary         0.75
# 4 f_meas    binary         0.77
```

### Cross-Validation with rsample

For robust evaluation, use rsample's resampling:

```{r rsample-cv, eval = FALSE}
library(rsample)

# Create cross-validation folds
folds <- vfold_cv(test_reviews, v = 3)

# The evaluate function works with single datasets
# For cross-validation, you'd typically use this in optimization
# (covered in the compilation-optimization vignette)

# But you can manually evaluate on each fold:
fold_results <- lapply(folds$splits, function(split) {
  assess_data <- assessment(split)

  evaluate_dsp(
    module = classifier,
    dataset = assess_data,
    metric = metric,
    .llm = llm,
    verbose = FALSE
  )
})

# Aggregate results
mean_cv_score <- mean(sapply(fold_results, function(r) r$mean_score))
print(paste("Cross-validated score:", mean_cv_score))
```

## Best Practices for Evaluation

### 1. Always Use Separate Test Data

```{r best-practice-split, eval = FALSE}
# Split your data properly
split_data <- function(data, train_pct = 0.7) {
  n <- nrow(data)
  train_idx <- sample(n, size = floor(n * train_pct))

  list(
    train = data[train_idx, ],
    test = data[-train_idx, ]
  )
}

# Split before any development
splits <- split_data(my_full_dataset)

# Use train for optimization (compilation)
# Use test for final evaluation
```

**Why:** Evaluating on training data gives overly optimistic results. Always hold out a test set.

### 2. Choose the Right Metric

```{r best-practice-metric}
# For classification: exact match
classification_metric <- metric_exact_match(field = "label")

# For generation: F1 or custom similarity
generation_metric <- metric_f1(field = "text")

# For complex tasks: composite metrics
composite <- function(pred, exp) {
  accuracy <- metric_exact_match(field = "category")(pred, exp)
  quality <- metric_f1(field = "description")(pred, exp)

  0.6 * accuracy + 0.4 * quality  # Weighted combination
}
```

**Why:** Different tasks need different metrics. Classification needs exact match; generation needs partial credit.

### 3. Check Metric Distribution

```{r best-practice-distribution, eval = FALSE}
# Evaluate and inspect score distribution
results <- evaluate_dsp(
  module = my_module,
  dataset = test_data,
  metric = my_metric,
  .llm = llm
)

# Check distribution
hist(results$scores)
summary(results$scores)

# Identify problematic examples
low_scoring <- which(results$scores < 0.5)
test_data[low_scoring, ]
results$predictions[low_scoring]
```

**Why:** A high mean score might hide systematic failures on certain types of examples.

### 4. Validate Metric Quality

Before using a metric, test it manually:

```{r best-practice-validate}
# Test your metric with known examples
test_metric <- metric_exact_match(field = "category")

# Should return 1
test_metric(
  list(category = "spam", other = "data"),
  list(category = "spam", different = "fields")
)

# Should return 0
test_metric(
  list(category = "spam"),
  list(category = "ham")
)

# Edge case: missing field
test_metric(
  list(other = "data"),
  list(category = "spam")
)  # Should handle gracefully
```

**Why:** Metric bugs waste time and lead to poor optimization results.

### 5. Ensure Dataset Diversity

```{r best-practice-diversity}
# Check diversity in your evaluation set
check_diversity <- function(dataset, field) {
  values <- dataset[[field]]
  unique_ratio <- length(unique(values)) / length(values)

  if (unique_ratio < 0.8) {
    warning(
      "Low diversity in ", field, ": ",
      round(unique_ratio * 100), "% unique"
    )
  }

  table(values)
}

check_diversity(test_data, "sentiment")

# Balance classes if needed
balance_dataset <- function(data, label_col, max_per_class = NULL) {
  labels <- unique(data[[label_col]])

  balanced <- lapply(labels, function(l) {
    subset_data <- data[data[[label_col]] == l, ]

    if (!is.null(max_per_class) && nrow(subset_data) > max_per_class) {
      subset_data[sample(nrow(subset_data), max_per_class), ]
    } else {
      subset_data
    }
  })

  do.call(rbind, balanced)
}
```

**Why:** Imbalanced or homogeneous test sets don't reflect real-world performance.

## Common Patterns

### Pattern 1: A/B Testing Modules

```{r pattern-ab-testing}
#| cassette: evaluation-ab-test

# Create two module variants
module_a <- signature("text -> summary: string") |>
  module(type = "predict", template = "Summarize concisely: {text}")

module_b <- signature("text -> summary: string") |>
  module(type = "predict", template = "Provide a detailed summary: {text}")

# Evaluate both on same data
test_summaries <- dsp_trainset(
  text = c(
    "Long article text here...",
    "Another article..."
  ),
  summary = c(
    "Expected summary 1",
    "Expected summary 2"
  )
)

results_a <- evaluate_dsp(module_a, test_summaries, metric_f1(field = "summary"), llm)
results_b <- evaluate_dsp(module_b, test_summaries, metric_f1(field = "summary"), llm)

# Compare
cat("Module A:", results_a$mean_score, "\n")
cat("Module B:", results_b$mean_score, "\n")
```

### Pattern 2: Progressive Evaluation

```{r pattern-progressive, eval = FALSE}
# Start small, then scale up
small_test <- test_data[1:10, ]
medium_test <- test_data[1:50, ]
full_test <- test_data

# Quick check with small set
quick_results <- evaluate_dsp(module, small_test, metric, llm)

if (quick_results$mean_score > 0.5) {
  # Looks promising, evaluate on more data
  medium_results <- evaluate_dsp(module, medium_test, metric, llm)

  if (medium_results$mean_score > 0.5) {
    # Final comprehensive evaluation
    final_results <- evaluate_dsp(module, full_test, metric, llm)
  }
}
```

### Pattern 3: Debugging with Detailed Inspection

```{r pattern-debugging, eval = FALSE}
# Evaluate with full details
results <- evaluate_dsp(
  module = classifier,
  dataset = test_data,
  metric = my_metric,
  .llm = llm,
  verbose = TRUE
)

# Find failures
failures <- results$scores == 0
failure_indices <- which(failures)

# Inspect each failure
for (i in failure_indices) {
  cat("\n=== Example", i, "===\n")
  cat("Input:", test_data$text[i], "\n")
  cat("Expected:", test_data$sentiment[i], "\n")
  cat("Predicted:", results$predictions[[i]], "\n")
  cat("Metadata:", str(results$metadata[[i]]), "\n")
}
```

## Troubleshooting

### Issue: Metric Returns Unexpected Values

```{r troubleshoot-metric, eval = FALSE}
# Test your metric in isolation
test_metric <- metric_exact_match(field = "category")

# Try with explicit examples
result <- test_metric(
  list(category = "spam", other = "data"),
  list(category = "spam", different = "fields")
)
print(result)  # Should be 1

# Check that it handles edge cases
result <- test_metric(
  list(wrong_field = "value"),
  list(category = "spam")
)
print(result)  # Should be 0 or NA, not crash
```

### Issue: Evaluation Takes Too Long

```{r troubleshoot-slow, eval = FALSE}
# Use a smaller sample for development
dev_sample <- test_data[sample(nrow(test_data), 20), ]

results <- evaluate_dsp(
  module = classifier,
  dataset = dev_sample,  # Smaller sample
  metric = metric,
  .llm = llm,
  verbose = FALSE  # Turn off progress to save time
)
```

### Issue: High Variance in Scores

```{r troubleshoot-variance, eval = FALSE}
# Run evaluation multiple times to check consistency
runs <- replicate(3, {
  evaluate_dsp(classifier, test_data, metric, llm)$mean_score
}, simplify = TRUE)

print(runs)
sd(runs)  # High SD indicates instability

# Consider:
# 1. Setting temperature = 0 for deterministic outputs
# 2. Using more examples
# 3. Simplifying your metric
```

## Next Steps

Now that you can evaluate your LLM programs systematically, you're ready to **optimize them**.

The [Automatic Prompt Optimization](compilation-optimization.html) vignette shows you how to:

- Use **teleprompters** to systematically improve your modules
- Run **grid search** over different configurations
- Leverage **tidymodels** for sophisticated optimization workflows
- Apply **cross-validation** for robust optimization

Remember: Good evaluation is the foundation of good optimization. Time spent crafting the right metric pays dividends throughout your development cycle.

## Summary

**Key Takeaways:**

1. **Metrics define success** - Choose metrics that capture what matters for your task
2. **Evaluate systematically** - Use `evaluate_dsp()` for consistent testing
3. **Use separate test data** - Never evaluate on training data
4. **Leverage tidymodels** - yardstick and rsample integrate seamlessly
5. **Inspect failures** - Low scores are learning opportunities

**Common Metrics:**
- `metric_exact_match()` - Classification tasks
- `metric_f1()` - Text generation with partial credit
- `metric_contains()` - Checking for required content
- `metric_custom()` - Your own logic
- `as_dsprrr_metric()` - Wrap yardstick metrics

**Next:** [Automatic Prompt Optimization](compilation-optimization.html)
