---
title: "Getting Started with dsprrr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dsprrr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is dsprrr?

dsprrr treats LLM prompts as **programs that can be optimized**, not strings to be tweaked by hand. You declare what you want with *signatures*, wrap them in *modules* for reuse, and let *teleprompters* find the best prompts automatically.

```{r}
library(dsprrr)
library(ellmer)

# Declare what you want
chat_openai() |> dsp("question -> answer", question = "What is 2+2?")
#> "4"
```

That one-liner handles prompt construction, structured output parsing, and type validation. No prompt engineering required.

## Choose Your Path

### New to dsprrr?

Start with the **tutorial sequence**—hands-on lessons that build on each other:
1. [Your First LLM Call](tutorial-hello-world.html) — Make structured calls with `dsp()` (10 min)
2. [Building a Classifier](tutorial-build-classifier.html) — Create reusable modules (20 min)
3. [Extracting Structured Data](tutorial-structured-outputs.html) — Multi-field outputs (25 min)
4. [Improving with Examples](tutorial-improve-with-demos.html) — Few-shot learning (25 min)
5. [Finding Best Configuration](tutorial-optimize-your-module.html) — Grid search (30 min)
6. [Taking to Production](tutorial-deploy-to-production.html) — Save and deploy (30 min)

### Already know the basics?

Jump to what you need:
- **[Quick Reference](cheatsheet.html)** — Signature syntax, module types, metrics
- **[Troubleshooting](troubleshooting.html)** — Fix common issues

### Want to understand the "why"?

Read the conceptual guides:
- **[The DSPy Philosophy](concepts-dspy-philosophy.html)** — Programs, not prompts
- **[Understanding Signatures & Modules](concepts-signatures-modules.html)** — Why S7 and R6
- **[How Optimization Works](concepts-optimization-theory.html)** — Teleprompter theory

### Building something specific?

Check the how-to guides:
- **[Compile & Optimize](compilation-optimization.html)** — Full optimization workflow
- **[Build RAG Pipelines](rag-workflows.html)** — Retrieval-augmented generation
- **[Deploy to Production](orchestration.html)** — pins, targets, and Quarto

## 5-Minute Taste

Here's dsprrr in action—from simple call to optimized module:

```{r}
library(dsprrr)
library(ellmer)

# 1. Quick call
chat <- chat_openai()
chat |> dsp("text -> sentiment: enum('positive', 'negative', 'neutral')",
            text = "Love this product!")
#> "positive"

# 2. Reusable module
classifier <- chat |> as_module("text -> sentiment: enum('positive', 'negative', 'neutral')")
classifier$predict(text = c("Great!", "Awful", "Meh"))
#> c("positive", "negative", "neutral")

# 3. Optimized module
trainset <- dsp_trainset(
  text = c("Amazing!", "Terrible!", "It's okay"),
  sentiment = c("positive", "negative", "neutral")
)

optimized <- compile_module(
  program = classifier,
  teleprompter = LabeledFewShot(k = 2),
  trainset = trainset
)
```

## Prerequisites

Before you begin:

1. **Install R** (4.1 or later)

2. **Install the packages**:
```r
install.packages("pak")
pak::pak("JamesHWade/dsprrr")
pak::pak("tidyverse/ellmer")
```

3. **Set your API key**:
```r
# In your .Renviron file
OPENAI_API_KEY=sk-your-key-here
```

## Learning Path

```
┌─────────────────────────────────────────────────────────────────────┐
│                         TUTORIALS                                    │
│  ┌──────────┐   ┌──────────┐   ┌──────────┐   ┌──────────┐          │
│  │ 1. Hello │ → │ 2. Build │ → │ 3. Struct│ → │ 4. Demos │          │
│  │   World  │   │Classifier│   │ Outputs  │   │          │          │
│  └──────────┘   └──────────┘   └──────────┘   └──────────┘          │
│                                                      ↓               │
│                                ┌──────────┐   ┌──────────┐          │
│                                │ 5.Optim- │ → │ 6. Prod- │          │
│                                │   ize    │   │  uction  │          │
│                                └──────────┘   └──────────┘          │
│                                                      ↓               │
│                         ADVANCED TUTORIALS                           │
│                    ┌────────────────┐   ┌────────────────┐          │
│                    │ Text Adventure │   │  llms.txt Gen  │          │
│                    └────────────────┘   └────────────────┘          │
└─────────────────────────────────────────────────────────────────────┘
                                   ↓
                 ┌─────────────────┴─────────────────┐
                 ↓                                   ↓
        ┌────────────────┐                 ┌────────────────┐
        │   HOW-TO       │                 │   CONCEPTS     │
        │   GUIDES       │                 │                │
        └────────────────┘                 └────────────────┘
                 ↓                                   ↓
        ┌────────────────┐                 ┌────────────────┐
        │   REFERENCE    │ ←───────────────│   (cheatsheet) │
        └────────────────┘                 └────────────────┘
```

Start at the top and work your way down. Each tutorial builds on the previous one.

## What's Next?

Ready to begin? Start with **[Tutorial 1: Your First LLM Call](tutorial-hello-world.html)**.
