---
title: "Getting Started with dsprrr"
output:
  rmarkdown::html_vignette:
    includes:
      in_header: mermaid.html
vignette: >
  %\VignetteIndexEntry{Getting Started with dsprrr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is dsprrr?

dsprrr treats LLM prompts as **programs that can be optimized**, not strings to be tweaked by hand. You declare what you want with *signatures*, wrap them in *modules* for reuse, and let *teleprompters* find the best prompts automatically.

```{r}
library(dsprrr)
library(ellmer)

# Declare what you want
answerer <- module(signature("question -> answer"))
run(answerer, question = "What is 2+2?", .llm = chat_openai())
#> "4"
```

The signature and module handle prompt construction, structured output parsing,
and type validation. No prompt engineering required.

## Choose Your Path

### New to dsprrr?

Start with the **tutorial sequence**—hands-on lessons that build on each other:

1. [Your First LLM Call](tutorial-hello-world.html) — Build and run a typed module (10 min)
2. [Building a Classifier](tutorial-build-classifier.html) — Create reusable modules (20 min)
3. [Extracting Structured Data](tutorial-structured-outputs.html) — Multi-field outputs (25 min)
4. [Improving with Examples](tutorial-improve-with-demos.html) — Few-shot learning (25 min)
5. [Finding Best Configuration](tutorial-optimize-your-module.html) — Grid search (30 min)
6. [Taking to Production](tutorial-deploy-to-production.html) — Save and deploy (30 min)

### Already know the basics?

Jump to what you need:

- **[Quick Reference](cheatsheet.html)** — Signature syntax, module types, metrics
- **[Troubleshooting](troubleshooting.html)** — Fix common issues

### Want to understand the "why"?

Read the conceptual guides:

- **[The DSPy Philosophy](concepts-dspy-philosophy.html)** — Programs, not prompts
- **[Understanding Signatures & Modules](concepts-signatures-modules.html)** — Why S7 and R6
- **[How Optimization Works](concepts-optimization-theory.html)** — Teleprompter theory

### Building something specific?

Check the how-to guides:

- **[Compile & Optimize](compilation-optimization.html)** — Full optimization workflow
- **[Build RAG Pipelines](rag-workflows.html)** — Retrieval-augmented generation
- **[Deploy to Production](orchestration.html)** — pins, targets, and Quarto

## 5-Minute Taste

Here's dsprrr in action—from simple call to optimized module:

```{r}
library(dsprrr)
library(ellmer)

# 1. Reusable typed module
chat <- chat_openai()
classifier <- module(
  signature("text -> sentiment: enum('positive', 'negative', 'neutral')"),
  chat = chat
)
run(classifier, text = "Love this product!")
#> "positive"

# 2. Batch execution
run_dataset(classifier, tibble::tibble(text = c("Great!", "Awful", "Meh")))
#> c("positive", "negative", "neutral")

# 3. Optimized module
trainset <- tibble::tibble(
  text = c("Amazing!", "Terrible!", "It's okay"),
  sentiment = c("positive", "negative", "neutral")
)

optimized <- compile_module(
  program = classifier,
  teleprompter = LabeledFewShot(k = 2),
  trainset = trainset
)
```

## Prerequisites

Before you begin:

1. **Get an API key** from an LLM provider supported by
   [ellmer](https://ellmer.tidyverse.org/)—OpenAI and Anthropic are the most
   common choices. The tutorials use OpenAI.

2. **Install the packages**:
```r
# install.packages("pak")
pak::pak("JamesHWade/dsprrr")
```

3. **Set your API key**:
```r
# In your .Renviron file
OPENAI_API_KEY=sk-your-key-here
```

4. **Verify your setup**:
```r
library(dsprrr)
dsprrr_sitrep()
#> ── Packages ──
#> ✔ ellmer 0.4.0 (OK)
#> ✔ dsprrr 0.1.0
#>
#> ── Default Chat ──
#> ✔ OpenAI (gpt-4o-mini)
#>   Source: Auto-detected from OPENAI_API_KEY
```

`dsprrr_sitrep()` reports which API keys are visible, which chat provider
dsprrr will use by default, and whether your installed packages are
compatible. If something looks wrong, see
[Troubleshooting](troubleshooting.html).

## Learning Path

```{mermaid}
flowchart TB
  subgraph Tutorials
    T1["1. Hello World"] --> T2["2. Build Classifier"] --> T3["3. Structured Outputs"]
    T3 --> T4["4. Demos"] --> T5["5. Optimize"] --> T6["6. Production"]
  end

  T6 --> Adv["Advanced Tutorials"]
  Adv --> TA["Text Adventure"]
  Adv --> LL["llms.txt Gen"]

  TA --> Split["Next Steps"]
  LL --> Split

  Split --> HowTo["How-To Guides"]
  Split --> Concepts["Concepts"]

  HowTo --> Reference["Reference"]
  Concepts --> Cheatsheet["Cheatsheet"]
  Cheatsheet --> Reference
```

Start at the top and work your way down. Each tutorial builds on the previous one.

## What's Next?

Ready to begin? Start with **[Tutorial 1: Your First LLM Call](tutorial-hello-world.html)**.
