---
title: "Generating llms.txt for R Packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating llms.txt for R Packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
# Skip evaluation during R CMD check or CI (cassettes may not match current code)
# VITALS_SHOULD_EVAL can override for cassette recording
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::setup_knitr()
}
```

[llms.txt](https://llmstxt.org/) is a proposed standard for AI-friendly documentation. This tutorial builds a dsprrr pipeline to automatically generate `llms.txt` files for R packages.

*This tutorial is adapted from the [DSPy llms.txt tutorial](https://dspy.ai/tutorials/llms_txt_generation/) by the DSPy team.*

## Why This Matters Beyond llms.txt

The multi-stage pipeline you'll build here is a blueprint for many real-world AI workflows:

- **API documentation generation**: The same pattern—gather metadata, analyze structure, generate prose—applies to any documentation task. Replace R packages with REST APIs, GraphQL schemas, or database tables.

- **Code review automation**: Each stage (understand purpose → analyze structure → identify issues) maps directly to how a code review agent works. The typed S7 results ensure review findings don't get lost between stages.

- **Migration guide creation**: When upgrading dependencies or refactoring APIs, you need the same capabilities: understand what exists, identify patterns, generate actionable guidance.

- **Knowledge base construction**: The pipeline extracts structured knowledge from unstructured sources. Swap packages for internal wikis, Slack channels, or support tickets.

The key insight is **staged analysis with typed handoffs**. Each LLM call has a focused job. S7 classes ensure intermediate results are validated before the next stage sees them. This is more reliable than asking one prompt to do everything.

We'll start with plain lists and functions, then show how S7 adds structure for larger projects.

## What You'll Build

A multi-stage analysis pipeline that:

1. Gathers package metadata (DESCRIPTION, README, exports)
2. Analyzes purpose and key concepts
3. Analyzes code structure
4. Generates usage examples
5. Produces formatted `llms.txt`

```{r setup}
library(dsprrr)
library(ellmer)
library(cli)
```

## Part 1: The Simple Approach

For a quick script or personal use, plain lists work fine.

### Gather Package Info

```{r simple-gather}
gather_package_info <- function(pkg_path = ".") {
  # Read DESCRIPTION
  desc_path <- file.path(pkg_path, "DESCRIPTION")
  if (!file.exists(desc_path)) {
    cli_abort("No DESCRIPTION file found at {.path {pkg_path}}")
  }

  desc <- read.dcf(desc_path)

  # Parse dependencies
  imports <- desc[1, "Imports"] %||% ""
  deps <- if (nzchar(imports)) {
    trimws(strsplit(imports, ",")[[1]])
  } else {
    character()
  }

  # Read README
  readme_path <- file.path(pkg_path, "README.md")
  readme <- if (file.exists(readme_path)) {
    paste(readLines(readme_path, warn = FALSE), collapse = "\n")
  } else {
    ""
  }

  # Get R files
  r_dir <- file.path(pkg_path, "R")
  r_files <- if (dir.exists(r_dir)) {
    list.files(r_dir, pattern = "\\.R$", ignore.case = TRUE)
  } else {
    character()
  }

  # Get exports from NAMESPACE
  ns_path <- file.path(pkg_path, "NAMESPACE")
  exports <- if (file.exists(ns_path)) {
    ns_lines <- readLines(ns_path, warn = FALSE)
    export_lines <- grep("^export\\(", ns_lines, value = TRUE)
    gsub("export\\((.+)\\)", "\\1", export_lines)
  } else {
    character()
  }

  # Check for vignettes
  vignette_dir <- file.path(pkg_path, "vignettes")
  has_vignettes <- dir.exists(vignette_dir) &&
    length(list.files(vignette_dir, pattern = "\\.(Rmd|qmd)$")) > 0

  # Return a simple list
 list(
    name = desc[1, "Package"],
    title = desc[1, "Title"] %||% "",
    description = desc[1, "Description"] %||% "",
    readme = readme,
    r_files = r_files,
    exports = exports,
    dependencies = deps,
    has_vignettes = has_vignettes
  )
}
```

### Define Signatures

Here's where dsprrr comes in—each analysis stage has a clear contract. Signatures define *what* each stage needs and *what* it produces. This separation matters: when a stage fails or produces poor output, you know exactly where to look.

Notice how we use `type_object()` and `type_array()` to define structured outputs. The LLM returns JSON matching this schema, which we can then pass reliably to the next stage:

```{r simple-signatures}
# Stage 1: Analyze purpose
analyze_purpose_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("title", "Package title from DESCRIPTION"),
    input("description_text", "Description from DESCRIPTION"),
    input("readme_excerpt", "First 2000 chars of README"),
    input("exported_functions", "Comma-separated exports")
  ),
  output_type = type_object(
    purpose = type_string("One sentence: what problem does this solve?"),
    key_concepts = type_array(
      type_object(
        term = type_string("Concept name"),
        definition = type_string("One sentence definition")
      ),
      "3-5 core concepts"
    ),
    target_audience = type_string("Who should use this?"),
    prerequisites = type_array(type_string(), "Required knowledge")
  ),
  instructions = "Analyze this R package to extract its core purpose.
Be precise and technical. Focus on what makes it unique."
)

# Stage 2: Analyze structure
analyze_structure_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("r_files", "R files in R/ directory"),
    input("exports", "Exported function names"),
    input("has_vignettes", "Whether package has vignettes"),
    input("dependencies", "Package dependencies")
  ),
  output_type = type_object(
    organization = type_string("How is code organized? (1-2 sentences)"),
    main_files = type_array(
      type_object(
        file = type_string("Filename"),
        purpose = type_string("What it contains")
      ),
      "3-5 most important files"
    ),
    entry_points = type_array(type_string(), "Main functions to start with"),
    patterns = type_string("Notable patterns: S3/S4/R6/S7, tidyeval, etc.")
  ),
  instructions = "Analyze package structure to help developers navigate it.
Identify important files and entry points."
)

# Stage 3: Generate examples
generate_examples_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("purpose", "What the package does"),
    input("entry_points", "Main functions"),
    input("key_concepts", "Core concepts as JSON")
  ),
  output_type = type_object(
    basic = type_string("3-5 line minimal example"),
    intermediate = type_string("5-10 line common workflow"),
    gotchas = type_array(type_string(), "1-3 common mistakes")
  ),
  instructions = "Generate realistic R code examples.
Examples must be syntactically valid R."
)

# Stage 4: Generate final llms.txt
generate_llmstxt_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("purpose", "Package purpose"),
    input("target_audience", "Who uses this"),
    input("key_concepts_json", "JSON of term/definition pairs"),
    input("organization", "Code organization"),
    input("entry_points", "Main functions"),
    input("main_files_json", "JSON of file/purpose pairs"),
    input("basic_example", "Basic usage example"),
    input("intermediate_example", "Intermediate example"),
    input("gotchas", "Common mistakes")
  ),
  output_type = type_string(),
  instructions = "Generate llms.txt in markdown format with sections:
# {pkg_name}, Key Concepts, Quick Start, Common Workflow,
Code Organization, Entry Points, Watch Out For.
Keep it concise - this is reference documentation for AI systems."
)
```

### Create Modules

Each module wraps a signature. We use `type = "chain_of_thought"` for the analysis stages—this asks the LLM to show its work, which improves accuracy for complex analysis tasks. The final `llmstxt` stage just needs to synthesize; no reasoning required.

```{r simple-modules}
create_modules <- function(llm) {
  list(
    purpose = module(analyze_purpose_sig, type = "chain_of_thought"),
    structure = module(analyze_structure_sig, type = "chain_of_thought"),
    examples = module(generate_examples_sig, type = "chain_of_thought"),
    llmstxt = module(generate_llmstxt_sig),
    llm = llm
  )
}
```

### Run the Pipeline

The pipeline chains stages together. Each stage's output feeds the next. This is where the structured signatures pay off—we can confidently pass `purpose$key_concepts` to the examples stage because we know its shape.

```{r simple-pipeline}
analyze_package <- function(pkg_path = ".", llm = chat_openai()) {
  cli_h1("Analyzing package")

  modules <- create_modules(llm)
  pkg_info <- gather_package_info(pkg_path)

  cli_alert_success("Gathered metadata for {.pkg {pkg_info$name}}")

  # Stage 1: Purpose
  cli_alert_info("Analyzing purpose...")
  purpose <- run(
    modules$purpose,
    pkg_name = pkg_info$name,
    title = pkg_info$title,
    description_text = pkg_info$description,
    readme_excerpt = substr(pkg_info$readme, 1, 2000),
    exported_functions = paste(pkg_info$exports, collapse = ", "),
    .llm = modules$llm
  )

  # Stage 2: Structure
  cli_alert_info("Analyzing structure...")
  structure <- run(
    modules$structure,
    pkg_name = pkg_info$name,
    r_files = paste(pkg_info$r_files, collapse = ", "),
    exports = paste(pkg_info$exports, collapse = ", "),
    has_vignettes = pkg_info$has_vignettes,
    dependencies = paste(pkg_info$dependencies, collapse = ", "),
    .llm = modules$llm
  )

  # Stage 3: Examples
  cli_alert_info("Generating examples...")
  examples <- run(
    modules$examples,
    pkg_name = pkg_info$name,
    purpose = purpose$purpose,
    entry_points = paste(structure$entry_points, collapse = ", "),
    key_concepts = jsonlite::toJSON(purpose$key_concepts, auto_unbox = TRUE),
    .llm = modules$llm
  )

  # Stage 4: Final output
  cli_alert_info("Generating llms.txt...")
  llmstxt <- run(
    modules$llmstxt,
    pkg_name = pkg_info$name,
    purpose = purpose$purpose,
    target_audience = purpose$target_audience,
    key_concepts_json = jsonlite::toJSON(purpose$key_concepts, auto_unbox = TRUE),
    organization = structure$organization,
    entry_points = paste(structure$entry_points, collapse = ", "),
    main_files_json = jsonlite::toJSON(structure$main_files, auto_unbox = TRUE),
    basic_example = examples$basic,
    intermediate_example = examples$intermediate,
    gotchas = paste(examples$gotchas, collapse = "; "),
    .llm = modules$llm
  )

  cli_alert_success("Done!")

  # Return everything as a simple list
  list(
    pkg_info = pkg_info,
    purpose = purpose,
    structure = structure,
    examples = examples,
    llmstxt = llmstxt
  )
}
```

### Use It

```{r simple-use}
#| cassette: simple-use
# Find package root (works from vignettes/ or project root)
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."

# Analyze and print
result <- analyze_package(pkg_root)
cat(result$llmstxt)
```

**This works.** For a one-off script, you're done.

---

## Part 2: Adding Structure with S7

The simple approach works for scripts. But as pipelines grow—more stages, more developers, production use—plain lists show their limits:

- **No validation**: What if a stage returns `NULL` for a required field? You won't find out until three stages later when something breaks mysteriously.
- **No documentation**: What fields does each result have? You'll need to trace through the code.
- **Hard to compose**: Passing results between stages is error-prone. Typos in field names silently return `NULL`.

S7 gives you typed containers that catch these problems at construction time:

```{r s7-setup}
library(S7)
```

### S7 Classes for Results

```{r s7-classes}
# Package metadata
PackageInfo <- new_class("PackageInfo",
  properties = list(
    name = class_character,
    title = class_character,
    description = class_character,
    readme = new_property(class_character, default = ""),
    r_files = new_property(class_character, default = character()),
    exports = new_property(class_character, default = character()),
    dependencies = new_property(class_character, default = character()),
    has_vignettes = new_property(class_logical, default = FALSE)
  )
)

# Purpose analysis result
PurposeAnalysis <- new_class("PurposeAnalysis",
  properties = list(
    purpose = class_character,
    key_concepts = class_list,
    target_audience = class_character,
    prerequisites = new_property(class_character, default = character())
  )
)

# Structure analysis result
StructureAnalysis <- new_class("StructureAnalysis",
  properties = list(
    organization = class_character,
    main_files = class_list,
    entry_points = class_character,
    patterns = class_character
  )
)

# Generated examples
Examples <- new_class("Examples",
  properties = list(
    basic = class_character,
    intermediate = class_character,
    gotchas = new_property(class_character, default = character())
  )
)

# Complete analysis
AnalysisResult <- new_class("AnalysisResult",
  properties = list(
    pkg_info = PackageInfo,
    purpose = PurposeAnalysis,
    structure = StructureAnalysis,
    examples = Examples,
    llmstxt = new_property(class_character, default = "")
  )
)
```

Now you get:

- **Type checking**: Can't create a `PurposeAnalysis` without a `purpose`
- **Documentation**: Class definitions show what fields exist
- **IDE support**: Autocomplete works with `@` slots

### Print Methods

```{r s7-print}
method(print, PackageInfo) <- function(x, ...) {
  cli_h3("Package: {x@name}")
  cli_text("{x@title}")
  cli_text("{length(x@exports)} exports, {length(x@r_files)} R files")
  invisible(x)
}

method(print, AnalysisResult) <- function(x, ...) {
  cli_h2("Analysis: {x@pkg_info@name}")
  cli_text("{.strong Purpose:} {x@purpose@purpose}")
  cli_text("{.strong Audience:} {x@purpose@target_audience}")
  cli_text("{.strong Entry points:} {.val {x@structure@entry_points}}")
  invisible(x)
}
```

### Updated Gather Function

```{r s7-gather}
gather_package_info <- function(pkg_path = ".") {
  desc_path <- file.path(pkg_path, "DESCRIPTION")
  if (!file.exists(desc_path)) {
    cli_abort("No DESCRIPTION file found at {.path {pkg_path}}")
  }

  desc <- read.dcf(desc_path)

  imports <- desc[1, "Imports"] %||% ""
  deps <- if (nzchar(imports)) {
    trimws(strsplit(imports, ",")[[1]])
  } else {
    character()
  }

  readme_path <- file.path(pkg_path, "README.md")
  readme <- if (file.exists(readme_path)) {
    paste(readLines(readme_path, warn = FALSE), collapse = "\n")
  } else {
    ""
  }

  r_dir <- file.path(pkg_path, "R")
  r_files <- if (dir.exists(r_dir)) {
    list.files(r_dir, pattern = "\\.R$", ignore.case = TRUE)
  } else {
    character()
  }

  ns_path <- file.path(pkg_path, "NAMESPACE")
  exports <- if (file.exists(ns_path)) {
    ns_lines <- readLines(ns_path, warn = FALSE)
    export_lines <- grep("^export\\(", ns_lines, value = TRUE)
    gsub("export\\((.+)\\)", "\\1", export_lines)
  } else {
    character()
  }

  vignette_dir <- file.path(pkg_path, "vignettes")
  has_vignettes <- dir.exists(vignette_dir) &&
    length(list.files(vignette_dir, pattern = "\\.(Rmd|qmd)$")) > 0

  # Return S7 object instead of list
  PackageInfo(
    name = desc[1, "Package"],
    title = desc[1, "Title"] %||% "",
    description = desc[1, "Description"] %||% "",
    readme = readme,
    r_files = r_files,
    exports = exports,
    dependencies = deps,
    has_vignettes = has_vignettes
  )
}
```

### Stage Functions

Each stage returns a typed S7 object. This is the key improvement over the simple approach: if the LLM returns incomplete data (missing `purpose`, for example), the `PurposeAnalysis()` constructor fails immediately with a clear error. No silent `NULL` propagation:

```{r s7-stages}
analyze_purpose <- function(pkg_info, modules) {
  cli_alert_info("Analyzing purpose and concepts...")

  result <- run(
    modules$purpose,
    pkg_name = pkg_info@name,
    title = pkg_info@title,
    description_text = pkg_info@description,
    readme_excerpt = substr(pkg_info@readme, 1, 2000),
    exported_functions = paste(pkg_info@exports, collapse = ", "),
    .llm = modules$llm
  )

  PurposeAnalysis(
    purpose = result$purpose,
    key_concepts = result$key_concepts,
    target_audience = result$target_audience,
    prerequisites = result$prerequisites %||% character()
  )
}

analyze_structure <- function(pkg_info, modules) {
  cli_alert_info("Analyzing code structure...")

  result <- run(
    modules$structure,
    pkg_name = pkg_info@name,
    r_files = paste(pkg_info@r_files, collapse = ", "),
    exports = paste(pkg_info@exports, collapse = ", "),
    has_vignettes = pkg_info@has_vignettes,
    dependencies = paste(pkg_info@dependencies, collapse = ", "),
    .llm = modules$llm
  )

  StructureAnalysis(
    organization = result$organization,
    main_files = result$main_files,
    entry_points = result$entry_points,
    patterns = result$patterns
  )
}

generate_examples <- function(pkg_info, purpose, structure, modules) {
  cli_alert_info("Generating usage examples...")

  result <- run(
    modules$examples,
    pkg_name = pkg_info@name,
    purpose = purpose@purpose,
    entry_points = paste(structure@entry_points, collapse = ", "),
    key_concepts = jsonlite::toJSON(purpose@key_concepts, auto_unbox = TRUE),
    .llm = modules$llm
  )

  Examples(
    basic = result$basic,
    intermediate = result$intermediate,
    gotchas = result$gotchas %||% character()
  )
}

generate_llmstxt <- function(pkg_info, purpose, structure, examples, modules) {
  cli_alert_info("Generating llms.txt...")

  run(
    modules$llmstxt,
    pkg_name = pkg_info@name,
    purpose = purpose@purpose,
    target_audience = purpose@target_audience,
    key_concepts_json = jsonlite::toJSON(purpose@key_concepts, auto_unbox = TRUE),
    organization = structure@organization,
    entry_points = paste(structure@entry_points, collapse = ", "),
    main_files_json = jsonlite::toJSON(structure@main_files, auto_unbox = TRUE),
    basic_example = examples@basic,
    intermediate_example = examples@intermediate,
    gotchas = paste(examples@gotchas, collapse = "; "),
    .llm = modules$llm
  )
}
```

### Main Function

```{r s7-main}
analyze_package <- function(pkg_path = ".", llm = chat_openai()) {
  cli_h1("Analyzing package")

  modules <- create_modules(llm)

  # Gather info (returns PackageInfo)
  pkg_info <- gather_package_info(pkg_path)
  cli_alert_success("Gathered metadata for {.pkg {pkg_info@name}}")
  print(pkg_info)

  # Run pipeline stages (each returns typed result)
  purpose <- analyze_purpose(pkg_info, modules)
  structure <- analyze_structure(pkg_info, modules)
  examples <- generate_examples(pkg_info, purpose, structure, modules)
  llmstxt <- generate_llmstxt(pkg_info, purpose, structure, examples, modules)

  cli_alert_success("Done!")

  # Return typed result
  AnalysisResult(
    pkg_info = pkg_info,
    purpose = purpose,
    structure = structure,
    examples = examples,
    llmstxt = llmstxt
  )
}
```

### Convenience Functions

Finally, we wrap everything in user-friendly functions. These hide the complexity while preserving access to the full `AnalysisResult` for users who need it:

```{r s7-convenience}
generate_llmstxt_file <- function(pkg_path = ".", output = NULL, llm = chat_openai()) {
  result <- analyze_package(pkg_path, llm)

  output <- output %||% file.path(pkg_path, "llms.txt")
  writeLines(result@llmstxt, output)
  cli_alert_success("Wrote {.file {output}}")

  invisible(result)
}

preview_llmstxt <- function(pkg_path = ".", llm = chat_openai()) {
  result <- analyze_package(pkg_path, llm)
  cat(result@llmstxt)
  invisible(result)
}
```

### Running It

```{r s7-run}
#| cassette: s7-run
# Find package root (works from vignettes/ or project root)
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."

# Analyze current package
result <- analyze_package(pkg_root)

# View the analysis (uses print method)
print(result)

# See the generated llms.txt
cat(result@llmstxt)
```

## Example Output

Running on dsprrr produces:

```markdown
# dsprrr

> DSPy-style LLM programming for R: signatures define I/O, modules
> encapsulate prompts, optimizers improve them automatically.

Data scientists and ML engineers building production LLM applications
who want systematic prompt optimization rather than manual tuning.

## Key Concepts

- **Signature**: Declarative specification of module inputs and outputs
  using arrow notation (`question -> answer`) or explicit types.
- **Module**: Reusable, stateful wrapper around an LLM call with
  configuration and optimization state.
- **Teleprompter**: Optimization strategy that compiles modules by
  adding few-shot examples or refining instructions.
- **Trace**: Record of module execution for debugging and analysis.

## Quick Start

```r
library(dsprrr)
library(ellmer)

mod <- signature("question -> answer") |> module(type = "predict")
run(mod, question = "What is R?", .llm = chat_openai())
```

## Common Workflow

```r
sig <- signature("context, question -> answer",
                 instructions = "Answer based only on context.")
mod <- module(sig, type = "predict")

trainset <- dsp_trainset(
  context = c("R is for statistics.", "Python is general-purpose."),
  question = c("What is R for?", "Describe Python."),
  answer = c("Statistics", "General-purpose programming")
)

optimized <- compile(LabeledFewShot(k = 2), mod, trainset = trainset)
evaluate(optimized, testset, metric = metric_exact_match())
```

## Code Organization

Core abstractions in signature.R (S7) and module-base.R (R6).
Module variants in separate files. Optimization in teleprompter.R.

### Important Files

- **signature.R**: S7 Signature class with string parser
- **module-base.R**: R6 Module base class
- **module-predict.R**: PredictModule for text generation
- **teleprompter.R**: LabeledFewShot, GEPA, MIPROv2
- **run.R**: run() and run_dataset() generics

## Entry Points

- `signature()`: Define input/output contract
- `module()`: Create module from signature
- `run()`: Execute module with inputs
- `compile()`: Optimize with teleprompter
- `dsp()`: Quick one-liner for Chat objects

## Watch Out For

- Modules are stateful—clone before modifying shared instances
- `run()` requires `.llm` unless you've called `set_default_chat()`
- Complex outputs need `type_object()` from ellmer
```

## When to Use Which

| Approach | Use When |
|----------|----------|
| **Plain lists** | One-off scripts, quick prototypes |
| **S7 classes** | Reusable pipelines, packages, need validation |

## Adapting This Pattern

The staged pipeline pattern adapts to many documentation and analysis tasks:

| Application | Gather Stage | Analysis Stages | Output Stage |
|------------|--------------|-----------------|--------------|
| **API docs** | Parse OpenAPI spec | Analyze endpoints, group by resource | Generate markdown reference |
| **Changelogs** | Parse git commits, issues | Categorize changes, identify breaking | Generate release notes |
| **Code review** | Diff files, parse AST | Check style, find bugs, assess complexity | Generate review comments |
| **Test generation** | Parse function signatures | Identify edge cases, dependencies | Generate test cases |
| **Migration guides** | Diff API versions | Identify breaking changes, patterns | Generate upgrade steps |

The key is the same: define clear signatures for each stage, use structured outputs to pass data between stages, and let S7 classes enforce the contracts.
