---
title: "Generating llms.txt for R Packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating llms.txt for R Packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

[llms.txt](https://llmstxt.org/) is a proposed standard for AI-friendly documentation. This tutorial builds a dsprrr pipeline to automatically generate `llms.txt` files for R packages—useful for helping AI assistants understand your codebase.

## What You'll Build

A multi-stage pipeline that:

1. Gathers package metadata (DESCRIPTION, README, exports)
2. Analyzes the package's purpose and key concepts
3. Analyzes the code structure and organization
4. Generates example usage snippets
5. Produces a formatted `llms.txt` file

```{r setup}
library(dsprrr)
library(ellmer)
```

## Step 1: Define the Signatures

Each analysis stage has a clear input/output contract:

```{r signatures}
# Stage 1: Analyze the package's purpose and concepts
analyze_pkg_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("title", "Package title from DESCRIPTION"),
    input("description_text", "Description field from DESCRIPTION"),
    input("readme_excerpt", "First 2000 chars of README"),
    input("exported_functions", "Comma-separated list of exported functions")
  ),
  output_type = type_object(
    purpose = type_string("One sentence: what problem does this package solve?"),
    key_concepts = type_array(
      type_object(
        term = type_string("Concept name"),
        definition = type_string("One sentence definition")
      ),
      "3-5 core concepts users must understand"
    ),
    target_audience = type_string("Who should use this package?"),
    prerequisites = type_array(type_string(), "Required knowledge or packages")
  ),
  instructions = "Analyze this R package to extract its core purpose and concepts.
Be precise and technical. Focus on what makes this package unique.
Key concepts should be domain terms or abstractions the package introduces."
)

# Stage 2: Analyze code structure
analyze_structure_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("r_files", "List of .R files in R/ directory"),
    input("exported_functions", "Exported function names"),
    input("has_vignettes", "Whether package has vignettes"),
    input("dependencies", "Package dependencies from DESCRIPTION")
  ),
  output_type = type_object(
    organization = type_string("How is the code organized? (1-2 sentences)"),
    main_files = type_array(
      type_object(
        file = type_string("Filename"),
        purpose = type_string("What this file contains")
      ),
      "3-5 most important files to understand"
    ),
    entry_points = type_array(type_string(), "Main functions users should start with"),
    internal_patterns = type_string("Notable patterns: S3/S4/R6/S7, NSE, etc.")
  ),
  instructions = "Analyze the package structure to help developers navigate it.
Identify the most important files and entry points. Note any design patterns."
)

# Stage 3: Generate usage examples
generate_examples_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("purpose", "What the package does"),
    input("entry_points", "Main functions to demonstrate"),
    input("key_concepts", "Core concepts as JSON")
  ),
  output_type = type_object(
    basic_example = type_string("3-5 line minimal example showing core usage"),
    intermediate_example = type_string("5-10 line example showing common workflow"),
    gotchas = type_array(type_string(), "1-3 common mistakes or surprising behaviors")
  ),
  instructions = "Generate realistic R code examples for this package.
Examples must be syntactically valid R. Use realistic variable names.
The basic example should work standalone. Intermediate can build on it."
)

# Stage 4: Generate the final llms.txt
generate_llmstxt_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("purpose", "Package purpose"),
    input("target_audience", "Who uses this"),
    input("key_concepts_json", "JSON array of term/definition objects"),
    input("organization", "Code organization description"),
    input("entry_points", "Main functions"),
    input("main_files_json", "JSON array of file/purpose objects"),
    input("basic_example", "Basic usage example"),
    input("intermediate_example", "Intermediate example"),
    input("gotchas", "Common mistakes")
  ),
  output_type = type_string(),
  instructions = "Generate a complete llms.txt file with this exact structure:

# {pkg_name}

> {one-line purpose}

{target_audience}

## Key Concepts

{bulleted list of term: definition}

## Quick Start

```r
{basic_example}
```

## Common Workflow

```r
{intermediate_example}
```

## Code Organization

{organization}

### Important Files

{bulleted list of file: purpose}

## Entry Points

{bulleted list of main functions with brief description}

## Watch Out For

{bulleted list of gotchas}

Keep it concise. This is reference documentation for AI systems."
)
```

## Step 2: The Analysis Pipeline

Create an R6 class that orchestrates the analysis:

```{r pipeline}
PackageAnalyzer <- R6::R6Class("PackageAnalyzer",
  public = list(
    analyze_pkg = NULL,
    analyze_structure = NULL,
    generate_examples = NULL,
    generate_llmstxt = NULL,
    llm = NULL,

    initialize = function(llm) {
      self$llm <- llm

      # Use chain-of-thought for better analysis
      self$analyze_pkg <- analyze_pkg_sig |>
        with_reasoning() |>
        module(type = "predict")

      self$analyze_structure <- analyze_structure_sig |>
        with_reasoning() |>
        module(type = "predict")

      self$generate_examples <- generate_examples_sig |>
        with_reasoning() |>
        module(type = "predict")

      self$generate_llmstxt <- generate_llmstxt_sig |>
        module(type = "predict")
    },

    analyze = function(pkg_path = ".") {
      cli::cli_h1("Analyzing package")

      # Gather package info
      pkg_info <- private$gather_pkg_info(pkg_path)
      cli::cli_alert_success("Gathered package metadata")

      # Stage 1: Analyze purpose and concepts
      cli::cli_alert_info("Analyzing purpose and concepts...")
      pkg_analysis <- run(
        self$analyze_pkg,
        pkg_name = pkg_info$name,
        title = pkg_info$title,
        description_text = pkg_info$description,
        readme_excerpt = substr(pkg_info$readme, 1, 2000),
        exported_functions = paste(pkg_info$exports, collapse = ", "),
        .llm = self$llm
      )

      # Stage 2: Analyze structure
      cli::cli_alert_info("Analyzing code structure...")
      structure_analysis <- run(
        self$analyze_structure,
        pkg_name = pkg_info$name,
        r_files = paste(pkg_info$r_files, collapse = ", "),
        exported_functions = paste(pkg_info$exports, collapse = ", "),
        has_vignettes = pkg_info$has_vignettes,
        dependencies = paste(pkg_info$deps, collapse = ", "),
        .llm = self$llm
      )

      # Stage 3: Generate examples
      cli::cli_alert_info("Generating usage examples...")
      examples <- run(
        self$generate_examples,
        pkg_name = pkg_info$name,
        purpose = pkg_analysis$purpose,
        entry_points = paste(structure_analysis$entry_points, collapse = ", "),
        key_concepts = jsonlite::toJSON(pkg_analysis$key_concepts, auto_unbox = TRUE),
        .llm = self$llm
      )

      # Stage 4: Generate llms.txt
      cli::cli_alert_info("Generating llms.txt...")
      llmstxt <- run(
        self$generate_llmstxt,
        pkg_name = pkg_info$name,
        purpose = pkg_analysis$purpose,
        target_audience = pkg_analysis$target_audience,
        key_concepts_json = jsonlite::toJSON(pkg_analysis$key_concepts, auto_unbox = TRUE),
        organization = structure_analysis$organization,
        entry_points = paste(structure_analysis$entry_points, collapse = ", "),
        main_files_json = jsonlite::toJSON(structure_analysis$main_files, auto_unbox = TRUE),
        basic_example = examples$basic_example,
        intermediate_example = examples$intermediate_example,
        gotchas = paste(examples$gotchas, collapse = "; "),
        .llm = self$llm
      )

      cli::cli_alert_success("Done!")

      list(
        pkg_info = pkg_info,
        purpose = pkg_analysis,
        structure = structure_analysis,
        examples = examples,
        llmstxt = llmstxt
      )
    }
  ),

  private = list(
    gather_pkg_info = function(pkg_path) {
      # Read DESCRIPTION
      desc_path <- file.path(pkg_path, "DESCRIPTION")
      if (!file.exists(desc_path)) {
        cli::cli_abort("No DESCRIPTION file found at {.path {pkg_path}}")
      }

      desc <- read.dcf(desc_path)
      pkg_name <- desc[1, "Package"]
      title <- desc[1, "Title"]
      description <- desc[1, "Description"]
      deps <- strsplit(desc[1, "Imports"] %||% "", ",\\s*")[[1]]

      # Read README
      readme_path <- file.path(pkg_path, "README.md")
      readme <- if (file.exists(readme_path)) {
        paste(readLines(readme_path, warn = FALSE), collapse = "\n")
      } else {
        ""
      }

      # Get R files
      r_dir <- file.path(pkg_path, "R")
      r_files <- if (dir.exists(r_dir)) {
        list.files(r_dir, pattern = "\\.R$", ignore.case = TRUE)
      } else {
        character()
      }

      # Get exports from NAMESPACE
      ns_path <- file.path(pkg_path, "NAMESPACE")
      exports <- if (file.exists(ns_path)) {
        ns_lines <- readLines(ns_path, warn = FALSE)
        export_lines <- grep("^export\\(", ns_lines, value = TRUE)
        gsub("export\\((.+)\\)", "\\1", export_lines)
      } else {
        character()
      }

      # Check for vignettes
      vignette_dir <- file.path(pkg_path, "vignettes")
      has_vignettes <- dir.exists(vignette_dir) &&
        length(list.files(vignette_dir, pattern = "\\.(Rmd|qmd)$")) > 0

      list(
        name = pkg_name,
        title = title,
        description = description,
        readme = readme,
        r_files = r_files,
        exports = exports,
        deps = deps,
        has_vignettes = has_vignettes
      )
    }
  )
)
```

## Step 3: Running the Pipeline

```{r run}
# Create analyzer
analyzer <- PackageAnalyzer$new(llm = chat_openai())

# Analyze current package
result <- analyzer$analyze(".")

# View the generated llms.txt
cat(result$llmstxt)

# Save to file
writeLines(result$llmstxt, "llms.txt")
```

## Example Output

Running this on dsprrr produces:

```markdown
# dsprrr

> DSPy-style LLM programming for R: signatures define I/O, modules encapsulate prompts, optimizers improve them automatically.

Data scientists and ML engineers building production LLM applications in R who want systematic prompt optimization rather than manual tuning.

## Key Concepts

- **Signature**: Declarative specification of module inputs and outputs using arrow notation (`question -> answer`) or explicit type definitions.
- **Module**: Reusable, stateful wrapper around an LLM call with configuration (temperature, template) and optimization state (demos, traces).
- **Teleprompter**: Optimization strategy that compiles modules by adding few-shot examples, refining instructions, or searching hyperparameters.
- **Trace**: Record of a module execution including inputs, outputs, tokens, and cost for debugging and analysis.

## Quick Start

```r
library(dsprrr)
library(ellmer)

mod <- signature("question -> answer") |> module(type = "predict")
run(mod, question = "What is R?", .llm = chat_openai())
```

## Common Workflow

```r
# Define task
sig <- signature("context, question -> answer",
                 instructions = "Answer based only on context.")
mod <- module(sig, type = "predict")

# Prepare labeled data
trainset <- dsp_trainset(
  context = c("R is for statistics.", "Python is general-purpose."),
  question = c("What is R for?", "Describe Python."),
  answer = c("Statistics", "General-purpose programming")
)

# Optimize with few-shot examples
optimized <- compile(LabeledFewShot(k = 2), mod, trainset = trainset)

# Evaluate
evaluate(optimized, testset, metric = metric_exact_match())
```

## Code Organization

Core abstractions in signature.R (S7) and module-base.R (R6). Module variants (predict, react, multichain) in separate files. Optimization infrastructure in optimizer-*.R and teleprompter.R.

### Important Files

- **signature.R**: S7 Signature class with string parser
- **module-base.R**: R6 Module base class with forward(), run(), trace methods
- **module-predict.R**: PredictModule for text generation
- **teleprompter.R**: LabeledFewShot, GridSearchTeleprompter, GEPA
- **run.R**: run() and run_dataset() generics

## Entry Points

- `signature()`: Define input/output contract
- `module()`: Create module from signature
- `run()`: Execute module with inputs
- `compile()`: Optimize module with teleprompter
- `dsp()`: Quick one-liner for Chat objects

## Watch Out For

- Modules are stateful R6 objects—clone before modifying shared instances
- `run()` requires `.llm` argument unless you've set a default with `set_default_chat()`
- Signatures use ellmer types internally—complex outputs need `type_object()`
```

## Batch Processing

Generate llms.txt for multiple packages:

```{r batch}
library(promises)
library(future)
plan(multisession, workers = 4)

pkg_paths <- c(
  "~/projects/pkg1",
  "~/projects/pkg2",
  "~/projects/pkg3"
)

# Analyze in parallel
results <- lapply(pkg_paths, function(path) {
  future::future({
    analyzer <- PackageAnalyzer$new(llm = chat_openai())
    analyzer$analyze(path)
  })
})

# Collect results
for (i in seq_along(results)) {
  result <- future::value(results[[i]])
  output_path <- file.path(pkg_paths[i], "llms.txt")
  writeLines(result$llmstxt, output_path)
  cli::cli_alert_success("Wrote {.file {output_path}}")
}
```

## Optimization

Improve the generator with examples of high-quality llms.txt files:

```{r optimize}
# Training data from well-documented packages
training <- dsp_trainset(
  pkg_name = c("dplyr", "ggplot2"),
  purpose = c(
    "Grammar of data manipulation with composable verbs",
    "Grammar of graphics for declarative visualization"
  ),
  # ... fill in expected outputs for each stage ...
  expected_llmstxt = c(
    "# dplyr\n\n> Grammar of data manipulation...",
    "# ggplot2\n\n> Declarative graphics layer by layer..."
  )
)

# Compile with labeled examples
analyzer$generate_llmstxt <- compile(
  LabeledFewShot(k = 2),
  analyzer$generate_llmstxt,
  trainset = training
)

# Or use GEPA for instruction optimization (when available)
# analyzer$generate_llmstxt <- compile(
#   GEPA(metric = llmstxt_quality_metric),
#   analyzer$generate_llmstxt,
#   trainset = training
# )
```

## Integration with pkgdown

Add llms.txt generation to your pkgdown build:

```{r pkgdown-hook}
# In _pkgdown.yml, add a custom article or hook
# Or create a script in inst/scripts/generate-llmstxt.R

generate_llmstxt_for_pkgdown <- function(pkg_path = ".") {
  analyzer <- PackageAnalyzer$new(llm = chat_openai())
  result <- analyzer$analyze(pkg_path)

  # Write to docs/ for pkgdown site
  docs_dir <- file.path(pkg_path, "docs")
  if (!dir.exists(docs_dir)) dir.create(docs_dir)

  writeLines(result$llmstxt, file.path(docs_dir, "llms.txt"))

  # Also write to package root
  writeLines(result$llmstxt, file.path(pkg_path, "llms.txt"))

  invisible(result)
}
```

## What This Demonstrates

- **Multi-stage pipelines**: Four modules chained, each building on previous outputs
- **Structured extraction**: Rich `type_object()` definitions for reliable parsing
- **Chain of thought**: `with_reasoning()` for better analytical outputs
- **Practical utility**: Generates real, usable documentation
- **R6 encapsulation**: Pipeline state managed cleanly

The same pipeline pattern applies to:

- API documentation generation
- Changelog summarization
- Code review automation
- Test case generation
- Migration guide creation
