---
title: "Generating llms.txt for R Packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating llms.txt for R Packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

[llms.txt](https://llmstxt.org/) is a proposed standard for providing LLM-friendly documentation. This tutorial shows how to use dsprrr to automatically generate an `llms.txt` file for any R package.

```{r setup}
library(dsprrr)
library(ellmer)
```

## The Pipeline

Generating good documentation requires multiple analysis steps:

1. **Analyze the package** — Extract purpose, key functions, and concepts
2. **Analyze the structure** — Understand the file organization
3. **Generate the llms.txt** — Combine everything into the final document

Each step is a separate module, allowing us to optimize them independently.

## Step 1: Package Analysis

First, we analyze the DESCRIPTION and README to understand what the package does:
```{r analyze-package}
analyze_pkg_sig <- signature(
  inputs = list(
    input("description", description = "Contents of DESCRIPTION file"),
    input("readme", description = "Contents of README.md"),
    input("exported_functions", description = "List of exported function names")
  ),
  output_type = type_object(
    purpose = type_string("One sentence describing what the package does"),
    key_functions = type_array(
      type_object(
        name = type_string(),
        description = type_string("One sentence")
      ),
      description = "The 5 most important functions"
    ),
    concepts = type_array(type_string(), description = "Key concepts a user should understand"),
    dependencies = type_array(type_string(), description = "Critical dependencies and why they matter")
  ),
  instructions = "Analyze this R package. Be concise and precise. Focus on what users need to know."
)

analyze_pkg <- analyze_pkg_sig |>
  with_reasoning() |>
  module(type = "predict")
```

## Step 2: Structure Analysis

Understand how the package is organized:

```{r analyze-structure}
analyze_structure_sig <- signature(
  inputs = list(
    input("file_tree", description = "List of files in the package"),
    input("r_files", description = "Names of files in R/ directory")
  ),
  output_type = type_object(
    organization = type_string("How the code is organized (1-2 sentences)"),
    entry_points = type_array(type_string(), description = "Main files to start reading"),
    test_coverage = type_string("Brief note on testing approach")
  ),
  instructions = "Analyze the package structure. Help someone navigate the codebase."
)

analyze_structure <- analyze_structure_sig |>
  module(type = "predict")
```

## Step 3: Generate llms.txt

Combine the analyses into the final document:

```{r generate}
generate_sig <- signature(
  inputs = list(
    input("pkg_name", description = "Package name"),
    input("purpose", description = "What the package does"),
    input("key_functions", description = "JSON array of important functions"),
    input("concepts", description = "Key concepts as comma-separated list"),
    input("organization", description = "How code is organized"),
    input("entry_points", description = "Where to start reading")
  ),
  output_type = type_string(),
  instructions = "Generate an llms.txt file following this format:

# {pkg_name}

> {one-line description}

## Purpose
{2-3 sentences on what this package does and why}

## Key Concepts
{bullet list of concepts}

## Main Functions
{bullet list with function name and description}

## Code Organization
{how to navigate the codebase}

## Quick Start
{3-5 line code example}

Be concise. This is reference documentation for LLMs, not a tutorial."
)

generate_llms_txt <- generate_sig |>
  module(type = "predict")
```

## Helper Functions

We need utilities to gather package information:

```{r helpers}
# Read package files
read_pkg_files <- function(pkg_path = ".") {
  list(
    description = readLines(file.path(pkg_path, "DESCRIPTION")) |>
      paste(collapse = "\n"),
    readme = if (file.exists(file.path(pkg_path, "README.md"))) {
      readLines(file.path(pkg_path, "README.md")) |> paste(collapse = "\n")
    } else "",
    file_tree = list.files(pkg_path, recursive = TRUE),
    r_files = list.files(file.path(pkg_path, "R"), pattern = "\\.R$"),
    exported = tryCatch({
      ns <- readLines(file.path(pkg_path, "NAMESPACE"))
      ns[grepl("^export\\(", ns)] |>
        gsub("export\\((.+)\\)", "\\1", x = _)
    }, error = function(e) character())
  )
}
```

## Running the Pipeline

Now we can generate llms.txt for any package:

```{r run-pipeline}
generate_llms_txt_for_pkg <- function(pkg_path = ".", llm = chat_openai()) {
  # Gather info
  pkg <- read_pkg_files(pkg_path)
  pkg_name <- gsub("Package: ", "", grep("^Package:", pkg$description, value = TRUE))

  # Step 1: Analyze package
  analysis <- run(
    analyze_pkg,
    description = pkg$description,
    readme = substr(pkg$readme, 1, 4000),  # Truncate long READMEs
    exported_functions = paste(pkg$exported, collapse = ", "),
    .llm = llm
  )

  # Step 2: Analyze structure
  structure <- run(
    analyze_structure,
    file_tree = paste(pkg$file_tree, collapse = "\n"),
    r_files = paste(pkg$r_files, collapse = ", "),
    .llm = llm
  )

  # Step 3: Generate llms.txt
  result <- run(
    generate_llms_txt,
    pkg_name = pkg_name,
    purpose = analysis$purpose,
    key_functions = jsonlite::toJSON(analysis$key_functions, auto_unbox = TRUE),
    concepts = paste(analysis$concepts, collapse = ", "),
    organization = structure$organization,
    entry_points = paste(structure$entry_points, collapse = ", "),
    .llm = llm
  )

  result
}

# Generate for current package
llms_txt <- generate_llms_txt_for_pkg(".")
cat(llms_txt)
```

## Example Output

Running this on dsprrr produces:

```markdown
# dsprrr

> DSPy-style LLM programming for R with signatures, optimization, and tracing.

## Purpose
dsprrr extends ellmer with declarative signatures that define LLM input/output
contracts, modules that encapsulate prompts as reusable components, and
optimizers that automatically improve prompts using labeled data.

## Key Concepts
- Signatures: Declare inputs and outputs with `signature("question -> answer")`
- Modules: Wrap signatures with configuration and state
- Optimization: Compile modules with teleprompters like LabeledFewShot
- Tracing: Every LLM call is recorded for debugging

## Main Functions
- `signature()`: Define input/output contract
- `module()`: Create a reusable LLM component
- `run()`: Execute a module with inputs
- `compile()`: Optimize a module with a teleprompter
- `dsp()`: Quick one-liner for Chat objects

## Code Organization
Core types in signature.R and module-base.R. Module variants (predict, react)
in separate files. Optimization in teleprompter.R and optimizer-*.R files.

## Quick Start
library(dsprrr)
library(ellmer)
mod <- signature("question -> answer") |> module(type = "predict")
run(mod, question = "What is R?", .llm = chat_openai())
```

## Optimizing the Pipeline

Once you have examples of good llms.txt files, optimize the generator:

```{r optimize}
# Collect examples of well-written llms.txt
training_data <- dsp_trainset(
  pkg_name = c("dplyr", "ggplot2"),
  purpose = c(
    "A grammar of data manipulation with verbs like filter, select, mutate",
    "A grammar of graphics for declarative plot creation"
  ),
  # ... other fields ...
  expected_output = c(
    "# dplyr\n\n> Fast, consistent tools for data manipulation...",
    "# ggplot2\n\n> Create elegant graphics layer by layer..."
  )
)

# Optimize with GEPA (when available) or LabeledFewShot
generate_llms_txt_optimized <- compile(
  LabeledFewShot(k = 2),
  generate_llms_txt,
  trainset = training_data
)
```

## What This Demonstrates

This tutorial shows:

- **Multi-step pipelines**: Breaking complex tasks into focused modules
- **Structured outputs**: Using `type_object()` for reliable extraction
- **Chain of thought**: `with_reasoning()` improves analysis quality
- **Practical utility**: Generate real documentation from code

The same pipeline pattern works for:
- API documentation generation
- Code review summaries
- Changelog generation
- Test case generation

## Batch Processing

Generate llms.txt for multiple packages:
```{r batch}
pkg_paths <- c("~/packages/pkg1", "~/packages/pkg2", "~/packages/pkg3")

results <- lapply(pkg_paths, function(path) {
  tryCatch(
    generate_llms_txt_for_pkg(path),
    error = function(e) paste("Error:", e$message)
  )
})

# Write to files
Map(function(txt, path) {
  writeLines(txt, file.path(path, "llms.txt"))
}, results, pkg_paths)
```

## Next Steps

- Add validation with `metric_contains()` to ensure required sections exist
- Use `run_async()` for parallel package processing
- Integrate with pkgdown for automatic generation during site builds

See `vignette("compilation-optimization")` for more on optimization strategies.
