---
title: "Generate an llms.txt file"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate an llms.txt file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(Sys.getenv("NOT_CRAN"), "false")
)
should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  # Cassettes replay in order; a cache hit on the second run's first request
  # would shift every later response.
  dsprrr::configure_cache(enable_memory = FALSE, enable_disk = FALSE)

  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

[llms.txt](https://llmstxt.org/) is a proposed Markdown file at the root of a
website that tells language models what the site is for and where to look.
This project builds a dsprrr program that writes one for an R package, runs
the code examples it produces, and turns that check into a metric you can
optimize the program against. It is adapted from the
[DSPy llms.txt tutorial](https://dspy.ai/tutorials/llms_txt_generation/).

If your package has a pkgdown site, you already have an llms.txt. Since
pkgdown 2.2.0, `pkgdown::build_site()` writes one that joins your home page
with the reference and article indexes, and adds a Markdown copy of every
page. That file lists everything on your site. It can't tell a model which
functions a newcomer needs or which mistakes people make. This project writes
that shorter, curated summary, and because a program writes it, you can
measure it and improve it.

```{r setup}
#| eval: true
#| message: false
library(dsprrr)
library(ellmer)
```

## Gather the facts

Everything the model sees comes from files R can read without a model: the
DESCRIPTION, the README, the file names in `R/`, and the exports in
`NAMESPACE`.

```{r gather}
#| eval: true
gather_package_info <- function(pkg_path = ".") {
  path <- function(...) file.path(pkg_path, ...)
  file_lines <- function(file) {
    if (!file.exists(path(file))) return(character())
    readLines(path(file), warn = FALSE)
  }
  desc <- read.dcf(path("DESCRIPTION"))
  # read.dcf() only returns the fields a package actually has
  field <- function(name) {
    if (name %in% colnames(desc)) unname(desc[1, name]) else ""
  }
  exports <- grep("^export\\(", file_lines("NAMESPACE"), value = TRUE)

  list(
    name = field("Package"),
    title = field("Title"),
    description = field("Description"),
    readme = paste(file_lines("README.md"), collapse = "\n"),
    r_files = list.files(path("R"), pattern = "\\.R$", ignore.case = TRUE),
    exports = gsub("export\\((.+)\\)", "\\1", exports),
    dependencies = trimws(strsplit(field("Imports"), ",")[[1]]),
    has_vignettes = length(list.files(path("vignettes"), "\\.(Rmd|qmd)$")) > 0
  )
}

# "." from the package root, ".." from the vignettes folder
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."
info <- gather_package_info(pkg_root)
info$title
lengths(info[c("r_files", "exports", "dependencies")])
```

## Four stages

The generator makes four calls, each with its own signature:

| Stage | Module | Reads | Writes |
|---|---|---|---|
| Purpose | `chain_of_thought()` | title, description, README excerpt, exports | purpose, key concepts, audience, prerequisites |
| Structure | `chain_of_thought()` | file names, exports, dependencies | organization, main files, entry points, patterns |
| Examples | `chain_of_thought()` | purpose, entry points, key concepts | a basic and an intermediate example, gotchas |
| Write | `module()` | all of the above | the llms.txt text |

The output types are ellmer types, so each stage returns fields the next one
can use without parsing prose:

```{r signatures}
#| eval: true
analyze_purpose_sig <- signature(
  inputs = list(
    input("pkg_name", description = "Package name"),
    input("title", description = "Package title from DESCRIPTION"),
    input("description_text", description = "Description from DESCRIPTION"),
    input("readme_excerpt", description = "First 2000 chars of README"),
    input("exported_functions", description = "Comma-separated exports")
  ),
  output_type = type_object(
    purpose = type_string("One sentence: what problem does this solve?"),
    key_concepts = type_array(
      type_object(
        term = type_string("Concept name"),
        definition = type_string("One sentence definition")
      ),
      "3-5 core concepts"
    ),
    target_audience = type_string("Who should use this?"),
    prerequisites = type_array(type_string(), "Required knowledge")
  ),
  instructions = "Analyze this R package to extract its core purpose.
Be precise and technical. Focus on what makes it unique."
)

analyze_structure_sig <- signature(
  inputs = list(
    input("pkg_name", description = "Package name"),
    input("r_files", description = "R files in R/ directory"),
    input("exports", description = "Exported function names"),
    input("has_vignettes", description = "Whether package has vignettes"),
    input("dependencies", description = "Package dependencies")
  ),
  output_type = type_object(
    organization = type_string("How is code organized? (1-2 sentences)"),
    main_files = type_array(
      type_object(
        file = type_string("Filename"),
        purpose = type_string("What it contains")
      ),
      "3-5 most important files"
    ),
    entry_points = type_array(type_string(), "Main functions to start with"),
    patterns = type_string("Notable patterns: S3/S4/R6/S7, tidyeval, etc.")
  ),
  instructions = "Analyze package structure to help developers navigate it.
Identify important files and entry points."
)

generate_examples_sig <- signature(
  inputs = list(
    input("pkg_name", description = "Package name"),
    input("purpose", description = "What the package does"),
    input("entry_points", description = "Main functions"),
    input("key_concepts", description = "Core concepts as JSON")
  ),
  output_type = type_object(
    basic = type_string("3-5 line minimal example"),
    intermediate = type_string("5-10 line common workflow"),
    gotchas = type_array(type_string(), "1-3 common mistakes")
  ),
  instructions = "Generate realistic R code examples.
Examples must be syntactically valid R."
)

generate_llmstxt_sig <- signature(
  inputs = list(
    input("pkg_name", description = "Package name"),
    input("purpose", description = "Package purpose"),
    input("target_audience", description = "Who uses this"),
    input("key_concepts_json", description = "JSON of term/definition pairs"),
    input("organization", description = "Code organization"),
    input("entry_points", description = "Main functions"),
    input("main_files_json", description = "JSON of file/purpose pairs"),
    input("basic_example", description = "Basic usage example"),
    input("intermediate_example", description = "Intermediate example"),
    input("gotchas", description = "Common mistakes")
  ),
  output_type = type_string(),
  instructions = "Generate llms.txt in markdown format with sections:
# {pkg_name}, Key Concepts, Quick Start, Common Workflow,
Code Organization, Entry Points, Watch Out For.
Keep it concise - this is reference documentation for AI systems."
)
```

The three analysis stages use `chain_of_thought()`, which adds a `reasoning`
output and asks the model to think before it answers. The writer only
formats, so it is a plain `module()`. Printing a module shows exactly what it
will ask for:

```{r modules}
#| eval: true
purpose_mod <- chain_of_thought(analyze_purpose_sig)
structure_mod <- chain_of_thought(analyze_structure_sig)
examples_mod <- chain_of_thought(generate_examples_sig)
writer_mod <- module(generate_llmstxt_sig)

examples_mod
```

## Run it

`analyze_package()` passes each stage's fields to the next. `run()` treats an
input longer than one as a batch, one call per element, so vectors and
tables are collapsed to single strings on the way:

```{r analyze}
analyze_package <- function(info, llm = chat_openai(model = "gpt-6-luna")) {
  # One chat for all four stages, so later stages also see earlier turns.
  purpose <- run(
    purpose_mod,
    pkg_name = info$name,
    title = info$title,
    description_text = info$description,
    readme_excerpt = substr(info$readme, 1, 2000),
    exported_functions = paste(info$exports, collapse = ", "),
    .llm = llm
  )
  structure <- run(
    structure_mod,
    pkg_name = info$name,
    r_files = paste(info$r_files, collapse = ", "),
    exports = paste(info$exports, collapse = ", "),
    has_vignettes = info$has_vignettes,
    dependencies = paste(info$dependencies, collapse = ", "),
    .llm = llm
  )
  key_concepts <- jsonlite::toJSON(purpose$key_concepts, auto_unbox = TRUE)
  entry_points <- paste(structure$entry_points, collapse = ", ")

  examples <- run(
    examples_mod,
    pkg_name = info$name,
    purpose = purpose$purpose,
    entry_points = entry_points,
    key_concepts = key_concepts,
    .llm = llm
  )
  run(
    writer_mod,
    pkg_name = info$name,
    purpose = purpose$purpose,
    target_audience = purpose$target_audience,
    key_concepts_json = key_concepts,
    organization = structure$organization,
    entry_points = entry_points,
    main_files_json = jsonlite::toJSON(structure$main_files, auto_unbox = TRUE),
    basic_example = examples$basic,
    intermediate_example = examples$intermediate,
    gotchas = paste(examples$gotchas, collapse = "; "),
    .llm = llm
  )
}
```

The writer's output type is a bare `type_string()`, so `run()` returns the
text itself. Here is the file gpt-4.1 wrote for dsprrr when this article was
recorded, in January 2026; the code now uses gpt-6-luna, whose file will read
differently:

```{r simple-use}
#| cassette: simple-use
llms_txt <- analyze_package(info)
cat(llms_txt)
```

## Check the examples

The prose is plausible. The Common Workflow code is not: `module()` has no
`task` argument, `compile()` needs a teleprompter, `optimize_grid()` has no
`trainset` argument, and `evaluate()` takes `data`, not `dataset`. Look back
at the printed `examples_mod`: the examples stage gets function names and a
purpose statement, but no argument lists, so the model guesses them.

Reading every generated example for mistakes like these doesn't scale.
Running them does. `run_block()` evaluates one code block in a fresh R
process, in a temporary directory and without API keys, so model-written
code stays away from your session and your credentials:

```{r check}
code_blocks <- function(md) {
  blocks <- regmatches(md, gregexpr("(?s)```r\\n.*?\\n```", md, perl = TRUE))
  gsub("^```r\\n|\\n```$", "", blocks[[1]])
}

run_block <- function(code) {
  callr::r(
    function(code) {
      setwd(tempdir())
      tryCatch(
        {
          eval(parse(text = code), envir = new.env())
          NA_character_
        },
        error = function(e) gsub("\\s+", " ", rlang::cnd_header(e))
      )
    },
    args = list(code = code),
    env = c(OPENAI_API_KEY = "", ANTHROPIC_API_KEY = "", GOOGLE_API_KEY = "")
  )
}

# One entry per code block: NA if it ran, otherwise the error
example_problems <- function(md) {
  problems <- vapply(code_blocks(md), run_block, "", USE.NAMES = FALSE)
  # Stopping at the model call counts as running: there is no chat here.
  problems[grepl("^No default Chat", problems)] <- NA
  problems
}

example_problems(llms_txt)
```

The Quick Start runs as far as `run()`, which is as far as it can get
without a model. The Common Workflow stops at its first mistake.

One sample says little, so here is a second run of the same program:

```{r s7-run}
#| cassette: s7-run
second_run <- analyze_package(info)
example_problems(second_run)
```

Different text, same kind of failure: this time the intermediate example
passes a whole tibble to `run()` (`run(qa_mod, test_data)`), which expects
named inputs. Rerunning until an example happens to work is not a fix. The
program needs to show the model how the functions are called, and you need
a score that tells you whether the change worked.

## Make it one program

`analyze_package()` is ordinary R wrapped around four modules, so dsprrr
can't evaluate or compile it as a whole. A `pipeline()` can be: it runs like
a module, `evaluate()` scores it on a dataset, and `BootstrapFewShot()`
compiles all of its steps together. Pipelines are linear, though. Each step
receives only the previous step's outputs, plus fixed values passed to
`step()` (see [Chain modules into pipelines](chaining-modules.html)).

So the redesign has two steps. The first reads the facts and writes notes,
including a quick start, and it now also gets every exported function with
its arguments. The second turns the notes into Markdown:

```{r pipeline}
#| eval: true
usage_lines <- function(pkg) {
  ns <- asNamespace(pkg)
  fns <- Filter(\(f) is.function(ns[[f]]), sort(getNamespaceExports(pkg)))
  vapply(fns, \(f) paste0(f, "(", toString(names(formals(ns[[f]]))), ")"), "")
}

notes <- chain_of_thought(signature(
  inputs = list(
    input("pkg_name", description = "Package name"),
    input("description_text", description = "Description from DESCRIPTION"),
    input("usage", description = "Exported functions and their arguments")
  ),
  output_type = type_object(
    purpose = type_string("One sentence: what problem does this solve?"),
    key_concepts = type_array(type_string(), "3-5 core concepts"),
    quick_start = type_string("5-10 lines of R that run as written"),
    gotchas = type_array(type_string(), "1-3 common mistakes")
  ),
  instructions = "Summarize this R package for other language models.
Call only functions listed in `usage`, with the arguments listed there."
))

write_up <- module(signature(
  "purpose, key_concepts, quick_start, gotchas -> llms_txt",
  instructions = "Write the body of an llms.txt file in Markdown:
a one-paragraph summary, then Key Concepts, Quick Start (the code unchanged,
in an R code block) and Watch Out For."
))

# select drops the reasoning field before it reaches the writer
llms_program <- pipeline(
  step(notes, select = c("purpose", "key_concepts", "quick_start", "gotchas")),
  write_up
)
```

The package name stops at the first step, so R adds the heading:

```{r pipeline-run}
#| eval: false
llm <- chat_openai(model = "gpt-6-luna")

out <- run(
  llms_program,
  pkg_name = "dsprrr",
  description_text = info$description,
  usage = paste(usage_lines("dsprrr"), collapse = "\n"),
  .llm = llm
)
writeLines(c("# dsprrr", "", out$llms_txt), "llms-summary.txt")
```

To publish the result with a pkgdown site, put it in `pkgdown/assets/`, whose
files are copied to the site root. Keep a name other than `llms.txt`, or
pkgdown's own file will replace it.

## Measure it and improve it

The share of code blocks that run is a metric. It needs no reference answer,
so a training set is just a few packages described the same way:

```{r measure}
#| eval: false
package_facts <- function(pkg) {
  tibble::tibble(
    pkg_name = pkg,
    description_text = utils::packageDescription(pkg)$Description,
    usage = paste(usage_lines(pkg), collapse = "\n")
  )
}
packages <- do.call(rbind, lapply(c("glue", "jsonlite", "withr"), package_facts))

examples_run <- function(prediction, expected) {
  problems <- example_problems(prediction$llms_txt)
  if (length(problems) == 0) 0 else mean(is.na(problems))
}

evaluate(llms_program, packages, metric = examples_run, .llm = llm)$mean_score
```

`compile()` can then improve the program without anyone editing a prompt.
`BootstrapFewShot()` runs the pipeline on each package, keeps the runs whose
examples all ran (`metric_threshold = 1`), and adds their notes and
write-ups to both steps as demonstrations. It warns that `packages` has no
output column, which is expected here because the metric doesn't compare
against one:

```{r compile}
#| eval: false
tuned <- compile(
  llms_program,
  BootstrapFewShot(
    metric = examples_run,
    metric_threshold = 1,
    max_bootstrapped_demos = 2L
  ),
  packages,
  .llm = llm
)

dsprrr_facts <- package_facts("dsprrr")
evaluate(tuned, dsprrr_facts, metric = examples_run, .llm = llm)$mean_score
```

`compile()` returns a new program and leaves `llms_program` unchanged, so you
can score both on packages that were not in the training set before choosing.
[Compile and optimize](compilation-optimization.html) covers the other
optimizers and how to choose between them.
