---
title: "Generating llms.txt for R Packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating llms.txt for R Packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

[llms.txt](https://llmstxt.org/) is a proposed standard for AI-friendly documentation. This tutorial builds a dsprrr pipeline to automatically generate `llms.txt` files for R packages, using S7 for typed, immutable data structures.

## What You'll Build

A multi-stage analysis pipeline that:

1. Gathers package metadata (DESCRIPTION, README, exports)
2. Analyzes purpose and key concepts
3. Analyzes code structure
4. Generates usage examples
5. Produces formatted `llms.txt`

```{r setup}
library(dsprrr)
library(ellmer)
library(S7)
library(cli)
```

## Step 1: S7 Classes for Analysis Results

Define typed containers for each analysis stage:

```{r s7-classes}
# Package metadata gathered from files
PackageInfo <- new_class("PackageInfo",
  properties = list(
    name = class_character,
    title = class_character,
    description = class_character,
    readme = new_property(class_character, default = ""),
    r_files = new_property(class_character, default = character()),
    exports = new_property(class_character, default = character()),
    dependencies = new_property(class_character, default = character()),
    has_vignettes = new_property(class_logical, default = FALSE)
  )
)

# Result of purpose/concept analysis
PurposeAnalysis <- new_class("PurposeAnalysis",
  properties = list(
    purpose = class_character,
    key_concepts = class_list,  # list of list(term, definition)
    target_audience = class_character,
    prerequisites = new_property(class_character, default = character())
  )
)

# Result of structure analysis
StructureAnalysis <- new_class("StructureAnalysis",
  properties = list(
    organization = class_character,
    main_files = class_list,  # list of list(file, purpose)
    entry_points = class_character,
    patterns = class_character
  )
)

# Generated examples
Examples <- new_class("Examples",
  properties = list(
    basic = class_character,
    intermediate = class_character,
    gotchas = new_property(class_character, default = character())
  )
)

# Complete analysis result
AnalysisResult <- new_class("AnalysisResult",
  properties = list(
    pkg_info = PackageInfo,
    purpose = PurposeAnalysis,
    structure = StructureAnalysis,
    examples = Examples,
    llmstxt = new_property(class_character, default = "")
  )
)
```

## Step 2: Print Methods

```{r print-methods}
method(print, PackageInfo) <- function(x, ...) {
  cli_h3("Package: {x@name}")
  cli_text("{x@title}")
  cli_text("{length(x@exports)} exports, {length(x@r_files)} R files")
  invisible(x)
}

method(print, AnalysisResult) <- function(x, ...) {
  cli_h2("Analysis: {x@pkg_info@name}")
  cli_text("{.strong Purpose:} {x@purpose@purpose}")
  cli_text("{.strong Audience:} {x@purpose@target_audience}")
  cli_text("{.strong Entry points:} {.val {x@structure@entry_points}}")
  invisible(x)
}
```

## Step 3: Signatures for Each Analysis Stage

```{r signatures}
# Stage 1: Analyze purpose and concepts
analyze_purpose_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("title", "Package title from DESCRIPTION"),
    input("description_text", "Description from DESCRIPTION"),
    input("readme_excerpt", "First 2000 chars of README"),
    input("exported_functions", "Comma-separated exports")
  ),
  output_type = type_object(
    purpose = type_string("One sentence: what problem does this solve?"),
    key_concepts = type_array(
      type_object(
        term = type_string("Concept name"),
        definition = type_string("One sentence definition")
      ),
      "3-5 core concepts"
    ),
    target_audience = type_string("Who should use this?"),
    prerequisites = type_array(type_string(), "Required knowledge")
  ),
  instructions = "Analyze this R package to extract its core purpose.
Be precise and technical. Focus on what makes it unique.
Key concepts should be domain terms the package introduces."
)

# Stage 2: Analyze code structure
analyze_structure_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("r_files", "R files in R/ directory"),
    input("exports", "Exported function names"),
    input("has_vignettes", "Whether package has vignettes"),
    input("dependencies", "Package dependencies")
  ),
  output_type = type_object(
    organization = type_string("How is code organized? (1-2 sentences)"),
    main_files = type_array(
      type_object(
        file = type_string("Filename"),
        purpose = type_string("What it contains")
      ),
      "3-5 most important files"
    ),
    entry_points = type_array(type_string(), "Main functions to start with"),
    patterns = type_string("Notable patterns: S3/S4/R6/S7, tidyeval, etc.")
  ),
  instructions = "Analyze package structure to help developers navigate it.
Identify important files and entry points. Note design patterns."
)

# Stage 3: Generate examples
generate_examples_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("purpose", "What the package does"),
    input("entry_points", "Main functions"),
    input("key_concepts", "Core concepts as JSON")
  ),
  output_type = type_object(
    basic = type_string("3-5 line minimal example"),
    intermediate = type_string("5-10 line common workflow"),
    gotchas = type_array(type_string(), "1-3 common mistakes")
  ),
  instructions = "Generate realistic R code examples.
Examples must be syntactically valid R. Use realistic names.
Basic example should work standalone."
)

# Stage 4: Generate final llms.txt
generate_llmstxt_sig <- signature(
  inputs = list(
    input("pkg_name", "Package name"),
    input("purpose", "Package purpose"),
    input("target_audience", "Who uses this"),
    input("key_concepts_json", "JSON of term/definition pairs"),
    input("organization", "Code organization"),
    input("entry_points", "Main functions"),
    input("main_files_json", "JSON of file/purpose pairs"),
    input("basic_example", "Basic usage example"),
    input("intermediate_example", "Intermediate example"),
    input("gotchas", "Common mistakes")
  ),
  output_type = type_string(),
  instructions = "Generate llms.txt with this exact structure:

# {pkg_name}

> {one-line purpose}

{target_audience}

## Key Concepts

{bulleted list: **term**: definition}

## Quick Start

```r
{basic_example}
```

## Common Workflow

```r
{intermediate_example}
```

## Code Organization

{organization}

### Important Files

{bulleted list: **file**: purpose}

## Entry Points

{bulleted list of functions with brief description}

## Watch Out For

{bulleted list of gotchas}

Keep it concise. This is reference documentation for AI systems."
)
```

## Step 4: Create Modules

```{r modules}
create_analyzer_modules <- function(llm) {
  list(
    purpose = analyze_purpose_sig |> with_reasoning() |> module(type = "predict"),
    structure = analyze_structure_sig |> with_reasoning() |> module(type = "predict"),
    examples = generate_examples_sig |> with_reasoning() |> module(type = "predict"),
    llmstxt = generate_llmstxt_sig |> module(type = "predict"),
    llm = llm
  )
}
```

## Step 5: Package Info Gathering

```{r gather-info}
gather_package_info <- function(pkg_path = ".") {
  # Read DESCRIPTION
  desc_path <- file.path(pkg_path, "DESCRIPTION")
  if (!file.exists(desc_path)) {
    cli_abort("No DESCRIPTION file found at {.path {pkg_path}}")
  }

  desc <- read.dcf(desc_path)
  pkg_name <- desc[1, "Package"]
  title <- desc[1, "Title"] %||% ""
  description <- desc[1, "Description"] %||% ""

  # Parse dependencies
  imports <- desc[1, "Imports"] %||% ""
  deps <- if (nzchar(imports)) {
    trimws(strsplit(imports, ",")[[1]])
  } else {
    character()
  }

  # Read README
  readme_path <- file.path(pkg_path, "README.md")
  readme <- if (file.exists(readme_path)) {
    paste(readLines(readme_path, warn = FALSE), collapse = "\n")
  } else {
    ""
  }

  # Get R files
  r_dir <- file.path(pkg_path, "R")
  r_files <- if (dir.exists(r_dir)) {
    list.files(r_dir, pattern = "\\.R$", ignore.case = TRUE)
  } else {
    character()
  }

  # Get exports from NAMESPACE
  ns_path <- file.path(pkg_path, "NAMESPACE")
  exports <- if (file.exists(ns_path)) {
    ns_lines <- readLines(ns_path, warn = FALSE)
    export_lines <- grep("^export\\(", ns_lines, value = TRUE)
    gsub("export\\((.+)\\)", "\\1", export_lines)
  } else {
    character()
  }

  # Check for vignettes
  vignette_dir <- file.path(pkg_path, "vignettes")
  has_vignettes <- dir.exists(vignette_dir) &&
    length(list.files(vignette_dir, pattern = "\\.(Rmd|qmd)$")) > 0

  PackageInfo(
    name = pkg_name,
    title = title,
    description = description,
    readme = readme,
    r_files = r_files,
    exports = exports,
    dependencies = deps,
    has_vignettes = has_vignettes
  )
}
```

## Step 6: Analysis Pipeline

Each stage takes the accumulated state and returns updated state:

```{r pipeline}
# Stage 1: Analyze purpose
analyze_purpose <- function(pkg_info, modules) {
  cli_alert_info("Analyzing purpose and concepts...")

  result <- run(
    modules$purpose,
    pkg_name = pkg_info@name,
    title = pkg_info@title,
    description_text = pkg_info@description,
    readme_excerpt = substr(pkg_info@readme, 1, 2000),
    exported_functions = paste(pkg_info@exports, collapse = ", "),
    .llm = modules$llm
  )

  PurposeAnalysis(
    purpose = result$purpose,
    key_concepts = result$key_concepts,
    target_audience = result$target_audience,
    prerequisites = result$prerequisites %||% character()
  )
}

# Stage 2: Analyze structure
analyze_structure <- function(pkg_info, modules) {
  cli_alert_info("Analyzing code structure...")

  result <- run(
    modules$structure,
    pkg_name = pkg_info@name,
    r_files = paste(pkg_info@r_files, collapse = ", "),
    exports = paste(pkg_info@exports, collapse = ", "),
    has_vignettes = pkg_info@has_vignettes,
    dependencies = paste(pkg_info@dependencies, collapse = ", "),
    .llm = modules$llm
  )

  StructureAnalysis(
    organization = result$organization,
    main_files = result$main_files,
    entry_points = result$entry_points,
    patterns = result$patterns
  )
}

# Stage 3: Generate examples
generate_examples <- function(pkg_info, purpose_analysis, structure_analysis, modules) {
  cli_alert_info("Generating usage examples...")

  result <- run(
    modules$examples,
    pkg_name = pkg_info@name,
    purpose = purpose_analysis@purpose,
    entry_points = paste(structure_analysis@entry_points, collapse = ", "),
    key_concepts = jsonlite::toJSON(purpose_analysis@key_concepts, auto_unbox = TRUE),
    .llm = modules$llm
  )

  Examples(
    basic = result$basic,
    intermediate = result$intermediate,
    gotchas = result$gotchas %||% character()
  )
}

# Stage 4: Generate llms.txt
generate_llmstxt <- function(pkg_info, purpose, structure, examples, modules) {
  cli_alert_info("Generating llms.txt...")

  run(
    modules$llmstxt,
    pkg_name = pkg_info@name,
    purpose = purpose@purpose,
    target_audience = purpose@target_audience,
    key_concepts_json = jsonlite::toJSON(purpose@key_concepts, auto_unbox = TRUE),
    organization = structure@organization,
    entry_points = paste(structure@entry_points, collapse = ", "),
    main_files_json = jsonlite::toJSON(structure@main_files, auto_unbox = TRUE),
    basic_example = examples@basic,
    intermediate_example = examples@intermediate,
    gotchas = paste(examples@gotchas, collapse = "; "),
    .llm = modules$llm
  )
}
```

## Step 7: Main Analysis Function

```{r analyze}
analyze_package <- function(pkg_path = ".", llm = chat_openai()) {
  cli_h1("Analyzing package")

  modules <- create_analyzer_modules(llm)

  # Gather info
  pkg_info <- gather_package_info(pkg_path)
  cli_alert_success("Gathered metadata for {.pkg {pkg_info@name}}")
  print(pkg_info)

  # Run pipeline stages
  purpose <- analyze_purpose(pkg_info, modules)
  structure <- analyze_structure(pkg_info, modules)
  examples <- generate_examples(pkg_info, purpose, structure, modules)
  llmstxt <- generate_llmstxt(pkg_info, purpose, structure, examples, modules)

  cli_alert_success("Done!")

  AnalysisResult(
    pkg_info = pkg_info,
    purpose = purpose,
    structure = structure,
    examples = examples,
    llmstxt = llmstxt
  )
}
```

## Step 8: Convenience Functions

```{r convenience}
# Analyze and write to file
generate_llmstxt_file <- function(pkg_path = ".", output = NULL, llm = chat_openai()) {
  result <- analyze_package(pkg_path, llm)

  output <- output %||% file.path(pkg_path, "llms.txt")
  writeLines(result@llmstxt, output)
  cli_alert_success("Wrote {.file {output}}")

  invisible(result)
}

# Just view the llms.txt content
preview_llmstxt <- function(pkg_path = ".", llm = chat_openai()) {
  result <- analyze_package(pkg_path, llm)
  cat(result@llmstxt)
  invisible(result)
}
```

## Running the Pipeline

```{r run}
# Analyze current package
result <- analyze_package(".")

# View the analysis
print(result)

# See the generated llms.txt
cat(result@llmstxt)

# Write to file
writeLines(result@llmstxt, "llms.txt")

# Or use the convenience function
generate_llmstxt_file(".", output = "llms.txt")
```

## Example Output

Running on dsprrr produces:

```markdown
# dsprrr

> DSPy-style LLM programming for R: signatures define I/O, modules
> encapsulate prompts, optimizers improve them automatically.

Data scientists and ML engineers building production LLM applications
who want systematic prompt optimization rather than manual tuning.

## Key Concepts

- **Signature**: Declarative specification of module inputs and outputs
  using arrow notation (`question -> answer`) or explicit types.
- **Module**: Reusable, stateful wrapper around an LLM call with
  configuration and optimization state.
- **Teleprompter**: Optimization strategy that compiles modules by
  adding few-shot examples or refining instructions.
- **Trace**: Record of module execution for debugging and analysis.

## Quick Start

```r
library(dsprrr)
library(ellmer)

mod <- signature("question -> answer") |> module(type = "predict")
run(mod, question = "What is R?", .llm = chat_openai())
```

## Common Workflow

```r
sig <- signature("context, question -> answer",
                 instructions = "Answer based only on context.")
mod <- module(sig, type = "predict")

trainset <- dsp_trainset(
  context = c("R is for statistics.", "Python is general-purpose."),
  question = c("What is R for?", "Describe Python."),
  answer = c("Statistics", "General-purpose programming")
)

optimized <- compile(LabeledFewShot(k = 2), mod, trainset = trainset)
evaluate(optimized, testset, metric = metric_exact_match())
```

## Code Organization

Core abstractions in signature.R (S7) and module-base.R (R6).
Module variants in separate files. Optimization in teleprompter.R.

### Important Files

- **signature.R**: S7 Signature class with string parser
- **module-base.R**: R6 Module base class
- **module-predict.R**: PredictModule for text generation
- **teleprompter.R**: LabeledFewShot, GEPA, MIPROv2
- **run.R**: run() and run_dataset() generics

## Entry Points

- `signature()`: Define input/output contract
- `module()`: Create module from signature
- `run()`: Execute module with inputs
- `compile()`: Optimize with teleprompter
- `dsp()`: Quick one-liner for Chat objects

## Watch Out For

- Modules are stateful—clone before modifying shared instances
- `run()` requires `.llm` unless you've called `set_default_chat()`
- Complex outputs need `type_object()` from ellmer
```

## Batch Processing

```{r batch}
analyze_multiple <- function(pkg_paths, llm = chat_openai()) {
  results <- lapply(pkg_paths, function(path) {
    tryCatch(
      analyze_package(path, llm),
      error = function(e) {
        cli_alert_danger("Failed: {.path {path}}: {e$message}")
        NULL
      }
    )
  })

  # Filter successes
  Filter(Negate(is.null), results)
}

# Analyze and write all
write_all_llmstxt <- function(pkg_paths, llm = chat_openai()) {
  results <- analyze_multiple(pkg_paths, llm)

  for (result in results) {
    output <- file.path(
      dirname(result@pkg_info@name),
      "llms.txt"
    )
    # In practice, you'd track the original path
    writeLines(result@llmstxt, output)
  }

  invisible(results)
}
```

## Optimization

Improve the generator with high-quality examples:

```{r optimize}
# Training data from well-documented packages
training <- dsp_trainset(
  pkg_name = c("dplyr", "ggplot2"),
  purpose = c(
    "Grammar of data manipulation with composable verbs",
    "Grammar of graphics for declarative visualization"
  )
  # ... other fields as needed
)

# Compile individual modules
modules <- create_analyzer_modules(chat_openai())

modules$purpose <- compile(
  LabeledFewShot(k = 2),
  modules$purpose,
  trainset = purpose_training
)

# Or use GEPA for instruction optimization
# modules$llmstxt <- compile(
#   GEPA(metric = llmstxt_quality),
#   modules$llmstxt,
#   trainset = training
# )
```

## Integration with pkgdown

```{r pkgdown}
# Add to pkgdown workflow
# In inst/scripts/generate-llmstxt.R:

generate_for_pkgdown <- function(pkg_path = ".") {
  result <- analyze_package(pkg_path)

  # Write to package root
  writeLines(result@llmstxt, file.path(pkg_path, "llms.txt"))

  # Also write to docs/ for pkgdown site
  docs_dir <- file.path(pkg_path, "docs")
  if (dir.exists(docs_dir)) {
    writeLines(result@llmstxt, file.path(docs_dir, "llms.txt"))
  }

  invisible(result)
}
```

## What This Demonstrates

- **S7 for results**: Typed containers for each analysis stage
- **Multi-stage pipelines**: Four modules, each building on previous
- **Structured extraction**: `type_object()` for reliable parsing
- **Chain of thought**: `with_reasoning()` for better analysis
- **Practical utility**: Generates real, usable documentation

The same pipeline pattern applies to:

- API documentation generation
- Changelog summarization
- Code review automation
- Test case generation
- Migration guide creation
