---
title: "Models, providers and streaming"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Models, providers and streaming}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
has_dials <- requireNamespace("dials", quietly = TRUE)
```

dsprrr sends every request through an [ellmer](https://ellmer.tidyverse.org)
`Chat`, so it works with any provider ellmer supports. This page shows how to
pick the Chat a call uses, set provider parameters, call reasoning models, send
images, expose a module as a tool, change timeouts, and get results
asynchronously or as a stream.

```{r setup}
#| eval: true
#| message: false
library(dsprrr)
```

## Choose the Chat for a call

Each call uses the first Chat it finds, in this order:

| Where the Chat comes from | How to set it | Applies to |
|---|---|---|
| The `.llm` argument | `run(mod, ..., .llm = chat)` | That call |
| The module's own Chat | `module(sig, chat = chat)`, or `set_module_lm(program, chat)` for every module in a pipeline | Every call on that module |
| A scoped Chat | `with_lm(chat, { ... })`, or `local_lm(chat)` inside a function | Code in that block or function |
| The default Chat | `set_default_chat(chat)` or `dsp_configure()` | The session |

If none is set, dsprrr creates a default Chat from the first API key it finds:
`OPENAI_API_KEY`, then `ANTHROPIC_API_KEY`, then `GOOGLE_API_KEY`.
`dsprrr_sitrep()` shows the active default.

```{r}
set_default_chat(ellmer::chat_openai(model = "gpt-6-luna"))
classify <- module(signature("text -> sentiment: enum('positive', 'negative')"))
reviews <- c("Arrived broken.", "Works great.")

run(classify, text = reviews) # default Chat

claude <- ellmer::chat_anthropic(model = "claude-sonnet-4-5")
run(classify, text = reviews, .llm = claude) # this call only
with_lm(claude, run(classify, text = reviews)) # every call in the block
```

A module's own Chat takes precedence over `with_lm()`. `dsp_configure()`
builds a Chat from a provider name, e.g.
`dsp_configure(provider = "openai", model = "gpt-6-luna")`, applies its
`temperature` argument through `ellmer::params()`, and passes other arguments
to the ellmer constructor.

A Chat keeps its conversation. Each single `run()` call adds its question and
answer to the Chat it used (the `.llm` argument, the module's own Chat, a
scoped Chat or the default), and later calls on that Chat send the whole
history. Batch rows, `run_dataset()` and `evaluate()` start each row from a
copy of the history at that moment and add nothing to it; so does a module
whose `config` sets request parameters. Over a long session, input tokens
grow, earlier exchanges can influence answers, and cache keys change because
they include the history. To keep calls independent, create a new Chat for the
task or clear one with `chat$set_turns(list())`.

## Set provider parameters

Put sampling and output settings on the Chat with `ellmer::params()`. ellmer
translates them into each provider's field names (for OpenAI, `max_tokens` is
sent as `max_output_tokens` and `reasoning_effort` as `reasoning.effort`):

```{r}
chat <- ellmer::chat_openai(
  model = "gpt-6-luna",
  params = ellmer::params(
    reasoning_effort = "none",
    temperature = 0,
    max_tokens = 500
  )
)
```

A module can also carry parameters in its `config`, as in
`module(sig, config = list(reasoning_effort = "low"))`; `optimize_grid()`
applies the values it tries this way. Standard parameters go through
`ellmer::params()` and are translated the same way; other names are sent as
you give them.

## Reasoning models

Reasoning models work through a problem before they answer. gpt-6-luna
reasons by default; its `reasoning_effort` can be `"none"`, `"low"`,
`"medium"` (the default), `"high"`, `"xhigh"` or `"max"`. Three things change
when you use one:

- `temperature` and `top_p` apply only when reasoning is off. With gpt-6-luna,
  set `reasoning_effort = "none"` before you set or sweep them, and leave them
  unset otherwise. dsprrr sends whatever you set; it does not adjust
  parameters for reasoning models.
- Set the amount of thinking with `reasoning_effort`, on the Chat
  (`ellmer::params(reasoning_effort = )`) or in a module's `config`, including
  a grid tried by `optimize_grid()`. Either way it reaches OpenAI as
  `reasoning.effort`.
- The reasoning is not part of the result. `run()` returns the signature's
  output fields; use `chain_of_thought()` if you want a `reasoning` field.

```{r}
solver <- module(signature("problem -> answer: int"))
thinker <- ellmer::chat_openai(
  model = "gpt-6-luna",
  params = ellmer::params(reasoning_effort = "low")
)
run(
  solver,
  problem = "A farmer has 17 sheep. All but 9 run away. How many are left?",
  .llm = thinker
)
```

`is_reasoning_model()` guesses from the name alone. It matches OpenAI's
o-series, `gpt-5` and `gpt-6` names and any name containing "reasoning":

```{r reasoning-names}
#| eval: true
models <- c("gpt-6-luna", "o4-mini", "claude-sonnet-4-5")
vapply(models, is_reasoning_model, logical(1))
```

Its only use in dsprrr is `module_parameters()`, which suggests tunable
parameters and swaps `temperature` and `top_p` for `reasoning_effort` when the
name looks like a reasoning model:

```{r reasoning-params}
#| eval: !expr has_dials
qa <- module(signature("question -> answer"))
knobs <- c("temperature", "top_p", "reasoning_effort")
module_parameters(qa, model = "claude-sonnet-4-5", include = knobs)
module_parameters(qa, model = "gpt-6-luna", include = knobs)
```

## Images and other content

Pass ellmer content objects as inputs. dsprrr writes the other inputs into the
prompt text and sends each content object after it:

```{r}
read_receipt <- module(signature("receipt -> merchant: str, total: float"))
chat <- ellmer::chat_openai(model = "gpt-6-luna")

run(
  read_receipt,
  receipt = ellmer::content_image_url("https://example.com/receipts/0901.jpg"),
  .llm = chat
)
```

A list of content objects is a batch, one call per element:

```{r}
receipts <- list(
  ellmer::content_image_url("https://example.com/receipts/0901.jpg"),
  ellmer::content_image_url("https://example.com/receipts/0903.jpg")
)
totals <- run(read_receipt, receipt = receipts, .llm = chat)
totals[[1]]$total
```

The result has one named list per image.
`ellmer::content_image_file()` and `ellmer::content_pdf_url()` build other
kinds of content; the model must accept the kind you send.

## Use a module as a tool

`as_ellmer_tool()` wraps a module as an ellmer tool, so a chat can call it in
the middle of a conversation. The tool runs the module on its own Chat
(`.llm`), not on the conversation that calls it:

```{r}
sentiment <- module(
  signature("text -> sentiment: enum('positive', 'negative', 'neutral')")
)
sentiment_tool <- as_ellmer_tool(
  sentiment,
  name = "classify_sentiment",
  description = "Classify the sentiment of a customer review.",
  output = "json",
  .llm = ellmer::chat_openai(model = "gpt-6-luna")
)

assistant <- ellmer::chat_openai(model = "gpt-6-luna")
assistant$register_tool(sentiment_tool)
assistant$chat("Is this review positive? 'I love this product!'")
```

`output = "json"` hands the result back as JSON text. The default, `"auto"`,
returns an R list, which ellmer 0.5.0 converts with a deprecation warning.

## Timeouts and retries

ellmer applies both. Set them with options before you make calls:

```{r}
options(
  ellmer_timeout_s = 600, # default 300 seconds
  ellmer_max_tries = 5 # default 3
)
```

`run_dataset(.return_format = "structured")` records a failed row in the
`.error` column and carries on. Per-row time limits for batches come from
`concurrency_control()`; see [Run dsprrr in pipelines](orchestration.html).

## Async calls

`run_async()` returns a promise that resolves to the named list `run()` would
return:

```{r}
library(promises)
qa <- module(signature("question -> answer"))

run_async(
  qa,
  question = "What is the capital of Peru?",
  .llm = ellmer::chat_openai(model = "gpt-6-luna")
) |>
  then(\(result) cat(result$answer, "\n"))
```

Promises resolve when R is idle: at the console, in Shiny, or in a script after
`while (!later::loop_empty()) later::run_now()`. `run_async()` calls the
provider directly, so it skips the response cache and records no trace. It
accepts modules built with `module()` or `chain_of_thought()`; `react()` and
most other specialized modules raise an error. Give each concurrent call its
own Chat, because ellmer reads each answer from the Chat's last turn and calls
sharing a Chat can swap answers. For many inputs, a batch `run()` with
`.concurrency` is simpler.

## Streaming

`run_stream()` passes output to callbacks as it arrives. Attach a
`stream_listener()` to an output field:

```{r}
writer <- module(signature("topic -> essay"))

run_stream(
  writer,
  topic = "Why rivers meander",
  .llm = ellmer::chat_openai(model = "gpt-6-luna"),
  listeners = stream_listener("essay", \(chunk) cat(chunk))
)
```

When the signature has one string output and the coro package is installed,
dsprrr asks for plain text and hands each chunk to the listener; the full text
becomes the field's value. Otherwise the module runs normally and the listener
fires once with the finished value. `run_stream()` returns the result
invisibly, skips the cache and records no trace. An `on_status` callback
receives step and field events, one set per step for a pipeline.

`stream_async()` returns a coro async generator of text chunks, not a promise.
Read it inside `coro::async()`:

```{r}
library(coro)
chunks <- stream_async(
  writer,
  topic = "Why rivers meander",
  .llm = ellmer::chat_openai(model = "gpt-6-luna")
)
print_all <- async(function() {
  for (chunk in await_each(chunks)) cat(chunk)
})
print_all()
```

To run many calls at once and track what they cost, continue with
[Run dsprrr in pipelines](orchestration.html).
