---
title: "Run dsprrr in pipelines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run dsprrr in pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
has_pins <- requireNamespace("pins", quietly = TRUE)
# Run the evaluated examples in a scratch directory
workdir <- tempfile("dsprrr-pipeline-")
dir.create(workdir)
knitr::opts_knit$set(root.dir = workdir)
```

This page is about running dsprrr outside an interactive session. It shows how
to start a [targets](https://docs.ropensci.org/targets/) pipeline and a Quarto
report from dsprrr's templates, check inputs with `validate_workflow()`, keep
evaluation results in a pins board, and control concurrency and cost in batch
runs. Saving and restoring the module itself is covered in
[Tutorial 6: Save and reuse a module](tutorial-deploy-to-production.html).

```{r setup}
#| eval: true
#| message: false
library(dsprrr)
```

## Start from the templates

`use_dsprrr_template()` copies a targets pipeline (`"targets"`), a Quarto
report (`"quarto"`) or both (`"all"`) into a folder. Existing files are kept
unless `overwrite = TRUE`.

```{r templates}
#| eval: true
use_dsprrr_template("all", path = "sentiment-pipeline")
```

`_targets.R` is a complete sentiment-classification pipeline. Replace the
example data and module with your own; the other targets wire them together:

| Target | What it does |
|---|---|
| `train_data`, `test_data` | Six training and three test reviews with a `target` column |
| `module_definition` | A sentiment module with instructions and a template |
| `llm_client` | The Chat, rebuilt on every run (see below) |
| `optimized_module` | A copy of the module tuned by `optimize_grid()` over `temperature` |
| `evaluation_results` | `evaluate()` on the test data |
| `pins_board`, `pinned_config`, `pinned_evaluation` | A versioned `pins::board_folder("pins")` holding the optimized module and the evaluation |
| `pinned_traces` | Meant to pin traces; it writes nothing for now, because the stored module has no recorded traces |
| `summary_stats`, `summary_json` | A summary list, also written to `outputs/summary.json` |

`llm_client` calls Claude Sonnet 4.5 through `ellmer::chat_anthropic()`. Set
the `DSPRRR_MODEL` environment variable for another Claude model. For any other
Chat, set a factory near the top of `_targets.R`; it receives the model name:

```{r}
options(dsprrr.targets.llm_factory = function(model) {
  ellmer::chat_openai(model = "gpt-6-luna")
})
```

From the pipeline's folder, run it and read results with targets:

```{r}
library(targets)
tar_make()
tar_read(evaluation_results)
```

On later runs targets rebuilds `llm_client`, sees the same Chat, and skips
every target whose code and inputs haven't changed: edit the test data and only
the evaluation and what depends on it run again. Because `llm_client` is always
rebuilt, `tar_outdated()` lists everything downstream of it even when
`tar_make()` would skip those targets.

## Render the report

`report.qmd` reads pins from the board named by its `pins_board_path`
parameter. Its `module_name`, `eval_name` and `traces_name` parameters default
to the pin names the pipeline writes. After `tar_make()`, render it with the
Quarto CLI, overriding parameters with `-P`:

```{bash}
quarto render report.qmd -P eval_name:sentiment-eval-results
```

Treat the report as a starting point. It needs R 4.4 or later (it uses `%||%`
without loading rlang), and its evaluation section (mean score, score
distribution, predictions) is the part that works today. The module section
expects an older pin layout and prints empty fields, and the trace sections
stay empty while `pinned_traces` writes nothing. To render the report as part
of the pipeline, uncomment the `tar_quarto()` target (from tarchetypes) at the
end of `_targets.R`.

## Check inputs before a run

`validate_workflow()` makes no LLM calls. It checks that `module` is a dsprrr
module with at least one input, that `data` has a column for every signature
input, and that `board` is a pins board. It prints a summary and invisibly
returns a list with `valid` and the individual `checks`:

```{r validate}
#| eval: !expr has_pins
classify <- module(
  signature("text -> sentiment: enum('positive', 'negative', 'neutral')")
)
reviews <- tibble::tibble(
  text = c("Great value", "Arrived broken", "It's a chair"),
  sentiment = c("positive", "negative", "neutral")
)
board <- pins::board_folder("sentiment-pipeline/pins", versioned = TRUE)

validate_workflow(classify, data = reviews, board = board)
```

A missing column fails the check:

```{r validate-fail}
#| eval: !expr has_pins
check <- validate_workflow(classify, data = tibble::tibble(review = "Great"))
check$valid
```

It does not look at column types, the Chat or API keys, and it does not read or
write the board. Stop a script early with
`if (!check$valid) stop("Workflow validation failed")`.

## Keep evaluation results

`pin_vitals_log()` stores an `evaluate()` result (mean score, per-row scores,
predictions) with basic module information. Despite the name, it doesn't
accept a vitals `Task`; vitals writes its own logs (see
[Evaluate with vitals](vitals-recipes.html)).

```{r}
result <- evaluate(
  classify,
  reviews,
  metric = metric_exact_match(field = "sentiment"),
  .llm = ellmer::chat_openai(model = "gpt-6-luna")
)
pin_vitals_log(board, "sentiment-eval", result, module = classify)

pins::pin_versions(board, "sentiment-eval")
pins::pin_read(board, "sentiment-eval")$mean_score
```

Each write to a versioned board adds a version; pass an id from
`pin_versions()` to `pin_read(version = )` to read an older one. For shared
storage, use another pins board, such as
`pins::board_s3("my-bucket", prefix = "dsprrr/")` or `pins::board_connect()`.

## Control concurrency and cost

Batch calls run one row at a time unless you pass a `concurrency_control()` to
`.concurrency` in `run()`, `run_dataset()`, `evaluate()` or
`as_vitals_task()`:

```{r concurrency}
#| eval: true
control <- concurrency_control(backend = "ellmer", max_active = 8L, max_errors = 5L)
control
```

| Backend | How rows run | Limits |
|---|---|---|
| `"sequential"` (default) | One at a time | Reuses cached responses |
| `"ellmer"` | Concurrently in this R session, through `ellmer::parallel_chat_structured()` | No finite timeouts; skips the response cache |
| `"mirai"` | In background R processes | The only backend for `task_timeout` and `total_timeout`; the Chat must come from the module or the default, not `.llm`; skips the response cache |
| `"auto"` | Mirai when a timeout is set; otherwise sequential for `max_active = 1`, else ellmer when the Chat supports it, then mirai, then sequential | |

`max_errors` stops scheduling new rows after that many failures, and `cancel`
decides whether running mirai tasks are stopped at a limit. In structured
results, each row's metadata records the backend that ran it.

```{r}
results <- run(
  classify,
  text = reviews$text,
  .llm = ellmer::chat_openai(model = "gpt-6-luna"),
  .concurrency = control,
  .return_format = "structured"
)

get_cost(results)$total # dollars for this batch
summarize_traces(classify) # tokens, cost and latency recorded on this module
```

`get_cost()` reads structured `run()` results and `evaluate()` results; an
unknown price stays `NA` rather than counting as free. `session_cost()` sums
dsprrr's prompt history, which keeps the last 100 calls (raise the limit with
`options(dsprrr.prompt_history_max = )`). Calls made on the ellmer backend
currently enter that history without tokens or cost, so its totals read `NA`
after such a batch. To keep a module's call records, `pin_trace(board, name,
module)` pins them; add `include_prompts = TRUE` and `include_outputs = TRUE`
to keep the text as well.

For the settings that apply to each Chat, such as timeouts and retries, see
[Models, providers and streaming](models-and-providers.html).
