---
title: "RAG Workflows with ragnar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RAG Workflows with ragnar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(dsprrr)
library(ellmer)
# library(ragnar)  # Install from: pak::pak("tidyverse/ragnar")
```

## Introduction

Retrieval-Augmented Generation (RAG) combines document retrieval with LLM generation to produce grounded, factual responses. dsprrr provides native integration with ragnar, the tidyverse RAG package.

## RAG Architecture

A typical RAG workflow:

1. **Index**: Process documents and create embeddings
2. **Retrieve**: Find relevant documents for a query
3. **Generate**: Use retrieved context to generate responses

dsprrr's `RAGModule` handles steps 2 and 3 automatically.

## Creating a RAG Module

### With a ragnar Store

```{r rag-store}
# First, create a ragnar store with your documents
# library(ragnar)
#
# documents <- c(
#   "R is a programming language for statistical computing and graphics.",
#   "Python is widely used for machine learning and data science.",
#   "Julia is designed for high-performance numerical computing.",
#   "dsprrr brings DSPy-style programming to R."
# )
#
# store <- ragnar_store_create(
#   documents = documents,
#   embedding_fn = embed_openai()
# )

# Create a RAG module
# mod <- rag_module(
#   signature = "question, relevant_context -> answer",
#   store = store,
#   k = 3  # Retrieve top 3 documents
# )

# Run queries
# result <- run(mod, question = "What is dsprrr?")
```

### With a Custom Retriever

For custom retrieval logic, provide a retriever function:

```{r custom-retriever}
# Custom retriever function
my_retriever <- function(query, k = 5) {
  # Your custom retrieval logic here
  # Could be: database query, API call, local search, etc.
  c(
    "Document 1: Relevant information...",
    "Document 2: More context..."
  )
}

# Create module with custom retriever
mod <- rag_module(
  signature = "query, relevant_context -> response",
  retriever = my_retriever,
  k = 5
)
```

## RAG Module Configuration

### Context Field Naming

Control how retrieved context appears in prompts:

```{r context-format}
mod <- rag_module(
  signature = "question, context -> answer",
  store = store,
  context_format = "context"  # Match signature field name
)
```

### Adjusting Retrieval

```{r retrieval-config}
# Retrieve more documents for complex questions
mod_detailed <- rag_module(
  signature = "question, relevant_context -> detailed_answer",
  store = store,
  k = 10  # More context for detailed responses
)

# Retrieve fewer for focused answers
mod_focused <- rag_module(
  signature = "question, relevant_context -> brief_answer",
  store = store,
  k = 2  # Minimal context
)
```

## Creating Search Tools for ReAct Agents

For agentic workflows, create search tools that can be called by LLMs:

```{r ragnar-tool}
# Create a search tool from a ragnar store
# search_tool <- ragnar_tool(
#   store = store,
#   k = 5,
#   name = "search_knowledge",
#   description = "Search the knowledge base for relevant information"
# )

# Use with a ReAct module
# react_mod <- module(
#   signature = "question -> answer",
#   type = "react",
#   tools = list(search_tool)
# )

# The agent can now search for information
# result <- run(react_mod, question = "What programming languages are best for data science?")
```

### Creating Tools Directly from Documents

For quick setup, create a tool directly from documents:

```{r create-search-tool}
docs <- c(
  "The capital of France is Paris.",
  "The capital of Germany is Berlin.",
  "The capital of Italy is Rome."
)

# Create tool in one step (requires ragnar and embedding function)
# search_tool <- create_search_tool(
#   documents = docs,
#   embedding_fn = ragnar::embed_openai(),
#   k = 2,
#   name = "lookup_capitals"
# )
```

## RAG Best Practices

### Document Preparation

```{r doc-prep}
# Good: Chunk documents appropriately
chunks <- c(
  "Chapter 1, Section 1: Introduction to R...",
  "Chapter 1, Section 2: Data types in R...",
  "Chapter 2, Section 1: Functions in R..."
)

# Include metadata in chunks for better retrieval
# store <- ragnar_store_create(
#   documents = data.frame(
#     text = chunks,
#     source = c("ch1.1", "ch1.2", "ch2.1"),
#     topic = c("intro", "datatypes", "functions")
#   ),
#   embedding_fn = embed_openai()
# )
```

### Query Optimization

```{r query-tips}
# The query field should match typical question patterns
# Good signature - clear query field
mod <- rag_module(
  signature = "question, relevant_context -> answer",
  store = store
)

# The module looks for these field names for retrieval:
# "query", "question", "text", "input", "prompt"
```

### Handling No Results

```{r no-results}
# RAGModule handles empty results gracefully
# When no documents are retrieved, context is:
# "No relevant context found."

# Your signature instructions should handle this:
mod <- rag_module(
  signature(
    "question, relevant_context -> answer",
    instructions = "Answer the question using the provided context. If no relevant context is found, say 'I don't have information about that.'"
  ),
  store = store
)
```

## Tracing RAG Calls

RAG modules include retrieval information in traces:

```{r rag-tracing}
# Run a query
# result <- run(mod, question = "What is R?")

# Inspect the trace
# trace <- get_last_trace()
# trace$retrieved_context  # What was retrieved
# trace$query             # The query used for retrieval
```

## Combining RAG with Optimization

RAG modules support optimization like any other module:

```{r rag-optimize, eval = FALSE}
# Optimize retrieval parameters
train_data <- tibble::tibble(
  question = c("What is R?", "What is Python?"),
  expected = c("statistical computing", "machine learning")
)

optimize_grid(
  mod,
  data = train_data,
  metric = metric_contains(),  # Check if answer contains expected
  parameters = list(
    k = c(2, 5, 10)  # Try different retrieval counts
  )
)
```

## Summary

Key RAG integration features:

1. **`rag_module()`**: Create modules that auto-retrieve context
2. **`ragnar_tool()`**: Create search tools for agents
3. **`create_search_tool()`**: Quick tool creation from documents
4. **Custom retrievers**: Plug in any retrieval backend
5. **Tracing**: Full visibility into retrieved context
6. **Optimization**: Tune retrieval parameters like k

For more on ragnar, see the [ragnar documentation](https://tidyverse.github.io/ragnar/).
