---
title: "Building a Text Adventure Game"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Text Adventure Game}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```
This tutorial builds a complete text-based adventure game using dsprrr. The AI handles narrative generation, NPC dialogue, and action resolution while you control the game framework and player state.

## What You'll Build

- Dynamic scene generation with branching narratives
- AI-powered NPC conversations that remember context
- Action resolution with skill checks and consequences
- Inventory and character progression
- Save/load game functionality

```{r setup}
library(dsprrr)
library(ellmer)
library(R6)
library(cli)
```

## Step 1: Game Data Structures

First, define the player and game context using R6 classes:

```{r player-class}
Player <- R6::R6Class("Player",
  public = list(
    name = NULL,
    health = 100L,
    level = 1L,
    experience = 0L,
    inventory = character(),
    skills = list(
      strength = 10L,
      intelligence = 10L,
      charisma = 10L,
      stealth = 10L
    ),

    initialize = function(name, skills = NULL) {
      self$name <- name
      if (!is.null(skills)) self$skills <- skills
    },

    add_item = function(item) {
      self$inventory <- c(self$inventory, item)
      cli::cli_alert_success("Added {.val {item}} to inventory!")
    },

    remove_item = function(item) {
      if (item %in% self$inventory) {
        self$inventory <- setdiff(self$inventory, item)
        return(TRUE)
      }
      FALSE
    },

    gain_experience = function(amount) {
      self$experience <- self$experience + amount
      old_level <- self$level
      self$level <- 1L + (self$experience %/% 100L)
      if (self$level > old_level) {
        cli::cli_alert_success("Level up! You are now level {self$level}!")
      }
    },

    take_damage = function(amount) {
      self$health <- max(0L, self$health - amount)
      cli::cli_alert_danger("You took {amount} damage! Health: {self$health}/100")
    },

    heal = function(amount) {
      self$health <- min(100L, self$health + amount)
      cli::cli_alert_success("Healed {amount} HP! Health: {self$health}/100")
    }
  )
)

GameContext <- R6::R6Class("GameContext",
  public = list(
    current_location = "Village Square",
    story_progress = 0L,
    visited_locations = character(),
    npcs_met = character(),
    completed_quests = character(),
    recent_actions = character(),
    game_flags = list(),

    add_flag = function(flag, value = TRUE) {
      self$game_flags[[flag]] <- value
    },

    has_flag = function(flag) {
      isTRUE(self$game_flags[[flag]])
    },

    record_action = function(action) {
      self$recent_actions <- c(tail(self$recent_actions, 4), action)
    }
  )
)
```

## Step 2: AI Signatures

Define clear input/output contracts for each AI task:

```{r signatures}
# Scene generation: describe where we are and what's possible
scene_sig <- signature(
  inputs = list(
    input("location", "Current location name"),
    input("player_info", "Player stats and inventory summary"),
    input("story_progress", "Current story progress level (0-10)"),
    input("recent_actions", "Last few player actions"),
    input("visited_before", "Whether player has visited this location")
  ),
  output_type = type_object(
    description = type_string("2-3 sentence vivid scene description"),
    actions = type_array(type_string(), "4-6 possible actions"),
    npcs = type_array(type_string(), "NPCs present (empty if none)"),
    items = type_array(type_string(), "Visible/discoverable items"),
    atmosphere = type_enum(c("peaceful", "tense", "mysterious", "dangerous"))
  ),
  instructions = "You are a game master for a fantasy text adventure.
Create immersive scenes with interesting choices. Vary the options based on
location and story progress. Include at least one NPC interaction and one
exploration option when appropriate. If the player has visited before,
acknowledge familiarity but note any changes."
)

# NPC dialogue: handle conversations
dialogue_sig <- signature(
  inputs = list(
    input("npc_name", "Name of the NPC"),
    input("npc_role", "NPC's role (merchant, elder, guard, etc.)"),
    input("player_says", "What the player said or did"),
    input("relationship", "Previous interactions (new/friendly/hostile)"),
    input("context", "Current story context and location")
  ),
  output_type = type_object(
    response = type_string("NPC's spoken dialogue (1-3 sentences)"),
    action = type_string("Physical action or gesture the NPC makes"),
    mood = type_enum(c("friendly", "neutral", "suspicious", "hostile", "fearful")),
    offers_quest = type_boolean(),
    quest_hint = type_string("Brief quest description if offered, else empty"),
    reveals_info = type_string("Important lore or hints revealed, else empty"),
    gives_item = type_string("Item given to player, else 'none'")
  ),
  instructions = "You are an NPC in a fantasy RPG. Stay in character.
Merchants want to trade, elders share wisdom, guards are dutiful.
Build relationships over multiple interactions. Offer quests that
fit the story. Reveal information gradually to maintain mystery."
)

# Action resolution: determine outcomes
action_sig <- signature(
  inputs = list(
    input("action", "What the player is attempting"),
    input("relevant_skill", "Which skill applies (strength/intelligence/charisma/stealth)"),
    input("skill_value", "Player's skill value (10 is average, 15+ is skilled)"),
    input("difficulty", "How hard is this (easy/medium/hard/very_hard)"),
    input("context", "Relevant circumstances that might help or hinder"),
    input("has_useful_item", "Whether player has an item that could help")
  ),
  output_type = type_object(
    success = type_boolean(),
    outcome = type_string("2-3 sentence description of what happened"),
    damage_taken = type_integer("Damage to player (0 if none)"),
    item_gained = type_string("Item found/earned, else 'none'"),
    item_lost = type_string("Item consumed/lost, else 'none'"),
    experience = type_integer("XP earned (0-25 based on difficulty)"),
    unlocks = type_string("New location or story element unlocked, else 'none'"),
    consequence = type_string("Lasting effect on game state, else 'none'")
  ),
  instructions = "Resolve actions fairly but dramatically. Higher skills improve
odds but don't guarantee success. Create interesting failures that advance
the story. Hard actions should have bigger rewards. Useful items can
turn failures into successes or reduce negative outcomes."
)
```

## Step 3: Game AI Module

Create an R6 class that wraps the AI modules:

```{r game-ai}
GameAI <- R6::R6Class("GameAI",
  public = list(
    scene_mod = NULL,
    dialogue_mod = NULL,
    action_mod = NULL,
    llm = NULL,

    initialize = function(llm) {
      self$llm <- llm

      # Use chain-of-thought for better reasoning
      self$scene_mod <- scene_sig |>
        with_reasoning() |>
        module(type = "predict")

      self$dialogue_mod <- dialogue_sig |>
        with_reasoning() |>
        module(type = "predict")

      self$action_mod <- action_sig |>
        with_reasoning() |>
        module(type = "predict")
    },

    generate_scene = function(player, context) {
      player_info <- sprintf(
        "Level %d %s | HP: %d/100 | STR:%d INT:%d CHA:%d STL:%d | Items: %s",
        player$level, player$name, player$health,
        player$skills$strength, player$skills$intelligence,
        player$skills$charisma, player$skills$stealth,
        if (length(player$inventory) > 0)
          paste(player$inventory, collapse = ", ") else "none"
      )

      run(
        self$scene_mod,
        location = context$current_location,
        player_info = player_info,
        story_progress = as.character(context$story_progress),
        recent_actions = paste(context$recent_actions, collapse = " -> "),
        visited_before = context$current_location %in% context$visited_locations,
        .llm = self$llm
      )
    },

    handle_dialogue = function(npc_name, npc_role, player_says, context) {
      relationship <- if (npc_name %in% context$npcs_met) "returning" else "new"

      run(
        self$dialogue_mod,
        npc_name = npc_name,
        npc_role = npc_role,
        player_says = player_says,
        relationship = relationship,
        context = sprintf("Location: %s, Story progress: %d/10",
                         context$current_location, context$story_progress),
        .llm = self$llm
      )
    },

    resolve_action = function(action, player, context, difficulty = "medium") {
      # Determine relevant skill
      skill <- if (grepl("fight|attack|break|force", action, ignore.case = TRUE)) {
        "strength"
      } else if (grepl("sneak|hide|steal|pick", action, ignore.case = TRUE)) {
        "stealth"
      } else if (grepl("persuade|convince|talk|charm", action, ignore.case = TRUE)) {
        "charisma"
      } else {
        "intelligence"
      }

      # Check for useful items
      useful_items <- c("rope", "torch", "lockpick", "sword", "potion", "map", "key")
      has_useful <- any(tolower(player$inventory) %in% useful_items)

      run(
        self$action_mod,
        action = action,
        relevant_skill = skill,
        skill_value = as.character(player$skills[[skill]]),
        difficulty = difficulty,
        context = sprintf("Location: %s, %s",
                         context$current_location,
                         paste(tail(context$recent_actions, 2), collapse = ", ")),
        has_useful_item = has_useful,
        .llm = self$llm
      )
    }
  )
)
```

## Step 4: Display Functions

Create a nice console UI using cli:

```{r display}
display_header <- function() {
  cli::cli_h1("{.emph MYSTIC REALM ADVENTURE}")
  cli::cli_rule()
}

display_status <- function(player) {
  cli::cli_div(theme = list(span.stat = list(color = "cyan")))
  cli::cli_text("{.strong {player$name}} | Level {player$level} | XP: {player$experience}")
  cli::cli_text("HP: {.stat {player$health}/100} | STR:{.stat {player$skills$strength}} INT:{.stat {player$skills$intelligence}} CHA:{.stat {player$skills$charisma}} STL:{.stat {player$skills$stealth}}")
  if (length(player$inventory) > 0) {
    cli::cli_text("Inventory: {.val {player$inventory}}")
  }
  cli::cli_end()
  cli::cli_rule()
}

display_scene <- function(scene, context) {
  # Atmosphere indicator
  atmo_emoji <- switch(scene$atmosphere,
    peaceful = "\U0001F343",    # leaf
    tense = "\U0001F441",       # eye
    mysterious = "\U0001F319",  # moon
    dangerous = "\U0001F525"    # fire
  )

  cli::cli_h2("{atmo_emoji} {context$current_location}")
  cli::cli_text(scene$description)
  cat("\n")

  if (length(scene$npcs) > 0 && !identical(scene$npcs, "")) {
    cli::cli_text("{.strong NPCs:} {.val {scene$npcs}}")
  }
  if (length(scene$items) > 0 && !identical(scene$items, "")) {
    cli::cli_text("{.strong Items:} {.emph {scene$items}}")
  }
  cat("\n")

  # Display actions as numbered list
  cli::cli_h3("What do you do?")
  actions <- c(scene$actions, "Check inventory", "Save game", "Quit")
  for (i in seq_along(actions)) {
    cli::cli_text("{i}. {actions[i]}")
  }
  cat("\n")

  actions
}

display_dialogue <- function(npc_name, response) {
  cat("\n")
  cli::cli_alert_info("{.strong {npc_name}}: \"{response$response}\"")
  if (nzchar(response$action)) {
    cli::cli_text("{.emph ({response$action})}")
  }
  if (response$offers_quest && nzchar(response$quest_hint)) {
    cli::cli_alert_warning("Quest available: {response$quest_hint}")
  }
  if (nzchar(response$reveals_info)) {
    cli::cli_alert("You learned: {response$reveals_info}")
  }
  if (response$gives_item != "none") {
    cli::cli_alert_success("Received: {.val {response$gives_item}}")
  }
  cat("\n")
}

display_action_result <- function(result, player) {
  cat("\n")
  if (result$success) {
    cli::cli_alert_success(result$outcome)
  } else {
    cli::cli_alert_danger(result$outcome)
  }

  if (result$damage_taken > 0) {
    player$take_damage(result$damage_taken)
  }
  if (result$item_gained != "none") {
    player$add_item(result$item_gained)
  }
  if (result$item_lost != "none") {
    player$remove_item(result$item_lost)
    cli::cli_alert_warning("Lost: {.val {result$item_lost}}")
  }
  if (result$experience > 0) {
    player$gain_experience(result$experience)
    cli::cli_alert_info("+{result$experience} XP")
  }
  if (nzchar(result$unlocks) && result$unlocks != "none") {
    cli::cli_alert_success("Unlocked: {.val {result$unlocks}}")
  }
  cat("\n")
}
```

## Step 5: Save/Load System

```{r save-load}
save_game <- function(player, context, filename = "savegame.json") {
  save_data <- list(
    player = list(
      name = player$name,
      health = player$health,
      level = player$level,
      experience = player$experience,
      inventory = player$inventory,
      skills = player$skills
    ),
    context = list(
      current_location = context$current_location,
      story_progress = context$story_progress,
      visited_locations = context$visited_locations,
      npcs_met = context$npcs_met,
      completed_quests = context$completed_quests,
      game_flags = context$game_flags
    )
  )
  jsonlite::write_json(save_data, filename, auto_unbox = TRUE, pretty = TRUE)
  cli::cli_alert_success("Game saved to {.file {filename}}")
}

load_game <- function(filename = "savegame.json") {
  if (!file.exists(filename)) {
    cli::cli_alert_danger("Save file not found: {.file {filename}}")
    return(NULL)
  }

  save_data <- jsonlite::read_json(filename)

  player <- Player$new(
    name = save_data$player$name,
    skills = save_data$player$skills
  )
  player$health <- save_data$player$health
  player$level <- save_data$player$level
  player$experience <- save_data$player$experience
  player$inventory <- unlist(save_data$player$inventory) %||% character()

 context <- GameContext$new()
  context$current_location <- save_data$context$current_location
  context$story_progress <- save_data$context$story_progress
  context$visited_locations <- unlist(save_data$context$visited_locations) %||% character()
  context$npcs_met <- unlist(save_data$context$npcs_met) %||% character()
  context$completed_quests <- unlist(save_data$context$completed_quests) %||% character()
  context$game_flags <- save_data$context$game_flags

  cli::cli_alert_success("Game loaded!")
  list(player = player, context = context)
}
```

## Step 6: Character Creation

```{r character-creation}
create_character <- function() {
  display_header()
  cli::cli_h2("Character Creation")

  name <- readline("Enter your character's name: ")

  cli::cli_text("\nYou have {.strong 10 bonus points} to distribute among skills.")
  cli::cli_text("Base skills start at 10.\n")

  skills <- list(strength = 10L, intelligence = 10L, charisma = 10L, stealth = 10L)
  points <- 10L

  for (skill in names(skills)) {
    if (points > 0) {
      cli::cli_text("Points remaining: {.val {points}}")
      input <- readline(sprintf("Add to %s (0-%d): ", skill, points))
      add <- min(points, max(0L, as.integer(input) %||% 0L))
      skills[[skill]] <- skills[[skill]] + add
      points <- points - add
    }
  }

  player <- Player$new(name = name, skills = skills)

  cli::cli_alert_success("\nWelcome to Mystic Realm, {.strong {name}}!")
  cli::cli_text("Your adventure begins...\n")

  player
}
```

## Step 7: Main Game Loop

```{r game-loop}
# NPC role database
npc_roles <- list(
  "Village Elder" = "wise leader who knows ancient lore",
  "Merchant" = "trader interested in rare items and gold",
  "Guard" = "dutiful protector, suspicious of strangers",
  "Mysterious Stranger" = "hooded figure with unknown motives",
  "Blacksmith" = "crafts weapons and armor, appreciates quality materials",
  "Healer" = "kind herbalist who can restore health",
  "Thief" = "cunning rogue with information about hidden things"
)

play_game <- function(llm = chat_openai()) {
  # Initialize
  ai <- GameAI$new(llm)
  player <- create_character()
  context <- GameContext$new()

  playing <- TRUE

  while (playing && player$health > 0) {
    # Clear screen (optional)
    # cat("\014")

    display_header()
    display_status(player)

    # Generate scene
    scene <- ai$generate_scene(player, context)
    actions <- display_scene(scene, context)

    # Get player input
    choice <- as.integer(readline("Choose (number): "))

    if (is.na(choice) || choice < 1 || choice > length(actions)) {
      cli::cli_alert_warning("Invalid choice, try again.")
      next
    }

    chosen <- actions[choice]
    context$record_action(chosen)

    # Handle special actions
    if (chosen == "Check inventory") {
      if (length(player$inventory) == 0) {
        cli::cli_alert_info("Your inventory is empty.")
      } else {
        cli::cli_h3("Inventory")
        for (item in player$inventory) {
          cli::cli_li("{item}")
        }
      }
      readline("Press Enter to continue...")
      next
    }

    if (chosen == "Save game") {
      save_game(player, context)
      readline("Press Enter to continue...")
      next
    }

    if (chosen == "Quit") {
      if (readline("Save before quitting? (y/n): ") == "y") {
        save_game(player, context)
      }
      playing <- FALSE
      next
    }

    # Check if talking to NPC
    npc_target <- NULL
    for (npc in scene$npcs) {
      if (grepl(tolower(npc), tolower(chosen), fixed = TRUE) ||
          grepl("talk|speak|approach", tolower(chosen))) {
        npc_target <- npc
        break
      }
    }

    if (!is.null(npc_target)) {
      # Handle NPC dialogue
      role <- npc_roles[[npc_target]] %||% "mysterious character"
      response <- ai$handle_dialogue(npc_target, role, chosen, context)
      display_dialogue(npc_target, response)

      # Track NPC meeting
      if (!(npc_target %in% context$npcs_met)) {
        context$npcs_met <- c(context$npcs_met, npc_target)
      }

      # Handle item gift
      if (response$gives_item != "none") {
        player$add_item(response$gives_item)
      }
    } else {
      # Resolve action
      # Determine difficulty based on action
      difficulty <- if (grepl("fight|battle|defeat", chosen, ignore.case = TRUE)) {
        "hard"
      } else if (grepl("sneak|steal|pick lock", chosen, ignore.case = TRUE)) {
        "medium"
      } else if (grepl("look|examine|search", chosen, ignore.case = TRUE)) {
        "easy"
      } else {
        "medium"
      }

      result <- ai$resolve_action(chosen, player, context, difficulty)
      display_action_result(result, player)

      # Handle location changes
      if (grepl("go|travel|enter|head", chosen, ignore.case = TRUE)) {
        # Extract new location from action or unlock
        new_loc <- if (nzchar(result$unlocks) && result$unlocks != "none") {
          result$unlocks
        } else {
          # Try to extract from action
          sub(".*(?:go to|enter|head to) (?:the )?", "", chosen, ignore.case = TRUE)
        }
        if (nzchar(new_loc) && new_loc != chosen) {
          context$visited_locations <- c(context$visited_locations, context$current_location)
          context$current_location <- tools::toTitleCase(new_loc)
          context$story_progress <- context$story_progress + 1L
          cli::cli_alert_info("Traveled to {.strong {context$current_location}}")
        }
      }
    }

    readline("Press Enter to continue...")
  }

  # Game over
  if (player$health <= 0) {
    cli::cli_h1("GAME OVER")
    cli::cli_text("You have fallen in your quest...")
  } else {
    cli::cli_text("Thanks for playing!")
  }
}
```

## Playing the Game

```{r play}
# Start a new game
play_game(llm = chat_openai())

# Or load a saved game
# saved <- load_game()
# if (!is.null(saved)) {
#   player <- saved$player
#   context <- saved$context
#   # Continue playing...
# }
```

## Example Session

```
─ MYSTIC REALM ADVENTURE ──────────────────────────────

Aria | Level 1 | XP: 0
HP: 100/100 | STR:12 INT:14 CHA:12 STL:12
Inventory: torch
────────────────────────────────────────────────────────

── 🍃 Village Square ──

The morning sun warms the cobblestones of Willowbrook's
central square. A weathered fountain burbles beside a
massive oak tree, its branches heavy with age. Merchants
call out their wares while villagers hurry about their
business.

NPCs: Village Elder, Merchant
Items: strange medallion, fresh bread

── What do you do? ──

1. Approach the Village Elder
2. Browse the merchant's wares
3. Examine the strange medallion by the fountain
4. Head toward the forest path
5. Rest on the bench under the oak
6. Check inventory
7. Save game
8. Quit

Choose (number): 1

ℹ Village Elder: "Ah, a new face in Willowbrook. I sense
  purpose in your step, young one. Perhaps you seek the
  ancient ruins north of here?"
  (The elder strokes his long white beard thoughtfully)
! Quest available: Investigate the Northern Ruins
ℹ You learned: The ruins are said to hold a powerful
  artifact, but none who entered have returned.

Press Enter to continue...
```

## Optimization

Once you've collected examples of good AI responses, optimize the modules:

```{r optimize}
# Example training data for scene generation
scene_training <- dsp_trainset(
  location = c("Dark Forest", "Ancient Ruins"),
  # ... other inputs ...
  description = c(
    "Twisted trees block out the sun, their branches like grasping fingers...",
    "Crumbling stone pillars rise from the mist, covered in glowing runes..."
  )
  # ... other outputs ...
)

# Compile with GEPA for better instructions (when available)
# ai$scene_mod <- compile(GEPA(metric = my_metric), ai$scene_mod, scene_training)
```

## What This Demonstrates

- **Module composition**: Three specialized AI modules working together
- **Chain of thought**: `with_reasoning()` for better narrative decisions
- **State management**: R6 classes for mutable game state
- **Structured outputs**: Rich type definitions ensure consistent responses
- **Interactive systems**: Dynamic AI responses to player choices

The same patterns apply to chatbots, interactive assistants, customer support, and any system with multi-turn AI interactions.
