---
title: "Building a Text Adventure Game"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Text Adventure Game}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This tutorial builds a text-based adventure game to demonstrate how dsprrr modules compose into interactive systems. You'll create an AI game master that generates scenes, handles dialogue, and resolves player actions.

```{r setup}
library(dsprrr)
library(ellmer)
```

## The Game Structure

We'll build three modules that work together:

1. **SceneGenerator** — Describes the current location and available actions
2. **DialogueHandler** — Manages NPC conversations
3. **ActionResolver** — Determines outcomes of player choices

## Defining the Signatures

Each module needs a clear input/output contract:

```{r signatures}
# Scene generation: where are we and what can we do?
scene_sig <- signature(

  "location, inventory, history -> description, actions, npcs_present",
  instructions = "You are a game master for a fantasy text adventure.
Describe the scene vividly but concisely (2-3 sentences).
List 3-5 possible actions as a comma-separated string.
List any NPCs present by name, or 'none' if empty."
)

# Dialogue: talking to NPCs
dialogue_sig <- signature(
  "npc_name, player_says, context -> npc_response, mood, gives_item",
  instructions = "You are an NPC in a fantasy game.
Respond in character (1-2 sentences).
Mood should be: friendly, neutral, suspicious, or hostile.
gives_item should be 'none' or the name of an item if the NPC gives something."
)

# Action resolution: what happens when the player does something?
action_sig <- signature(
  "action, location, inventory, difficulty -> success: boolean, outcome, inventory_change",
  instructions = "Resolve the player's action.
Be fair but create drama. Not everything should succeed.
outcome is a 1-2 sentence description of what happens.
inventory_change is 'none', '+item_name' to add, or '-item_name' to remove."
)
```

## Creating the Modules

Wrap each signature in a module with chain-of-thought reasoning for better decisions:

```{r modules}
# Add reasoning to make better narrative choices
scene_mod <- scene_sig |>
  with_reasoning() |>
  module(type = "predict")

dialogue_mod <- dialogue_sig |>
  with_reasoning() |>
  module(type = "predict")

action_mod <- action_sig |>
  with_reasoning() |>
  module(type = "predict")
```

## Game State

Track the player's progress with a simple list:
```{r state}
new_game <- function() {
  list(
    location = "village_square",
    inventory = c("torch", "bread"),
    history = "You arrived at the village seeking the ancient artifact.",
    visited = character(),
    npcs_met = character()
  )
}

game <- new_game()
```

## The Game Loop

Now wire it together into a playable game:
```{r game-loop}
play_turn <- function(game, player_input, llm) {
  # Generate the current scene
  scene <- run(
    scene_mod,
    location = game$location,
    inventory = paste(game$inventory, collapse = ", "),
    history = game$history,
    .llm = llm
  )

  cat("\n", scene$description, "\n\n")
  cat("You can: ", scene$actions, "\n")
  if (scene$npcs_present != "none") {
    cat("You see: ", scene$npcs_present, "\n")
  }
  cat("Inventory: ", paste(game$inventory, collapse = ", "), "\n\n")

 # Handle player input
  if (grepl("^talk ", player_input, ignore.case = TRUE)) {
    npc <- sub("^talk ", "", player_input, ignore.case = TRUE)
    response <- run(
      dialogue_mod,
      npc_name = npc,
      player_says = "Hello",
      context = paste("Location:", game$location, "- History:", game$history),
      .llm = llm
    )
    cat("\n", npc, " says: \"", response$npc_response, "\"\n", sep = "")
    if (response$gives_item != "none") {
      game$inventory <- c(game$inventory, response$gives_item)
      cat("(Received: ", response$gives_item, ")\n", sep = "")
    }
  } else {
    # Resolve the action
    result <- run(
      action_mod,
      action = player_input,
      location = game$location,
      inventory = paste(game$inventory, collapse = ", "),
      difficulty = "medium",
      .llm = llm
    )

    if (result$success) {
      cat("\nSuccess! ", result$outcome, "\n")
    } else {
      cat("\nFailed. ", result$outcome, "\n")
    }

    # Update inventory
    if (result$inventory_change != "none") {
      if (grepl("^\\+", result$inventory_change)) {
        item <- sub("^\\+", "", result$inventory_change)
        game$inventory <- c(game$inventory, item)
        cat("(Added to inventory: ", item, ")\n", sep = "")
      } else if (grepl("^-", result$inventory_change)) {
        item <- sub("^-", "", result$inventory_change)
        game$inventory <- setdiff(game$inventory, item)
        cat("(Removed from inventory: ", item, ")\n", sep = "")
      }
    }
  }

  # Update history
  game$history <- paste(game$history, player_input, sep = " -> ")

  game
}
```

## Playing the Game

```{r play}
llm <- chat_openai()  # or chat_anthropic(), chat_google_gemini()

game <- new_game()

# Play a few turns
game <- play_turn(game, "look around", llm)
game <- play_turn(game, "talk to merchant", llm)
game <- play_turn(game, "go to the forest", llm)
game <- play_turn(game, "search for tracks", llm)
```

Example output:

```
The village square bustles with morning activity. A weathered merchant
tends his cart of curious goods, while the path north leads to a dark forest.

You can: look around, talk to merchant, examine cart, go to forest, rest
You see: merchant
Inventory: torch, bread

> talk to merchant

merchant says: "Ah, a traveler! Looking for the artifact too, eh?
Take this map—you'll need it in the forest."
(Received: old_map)

> go to forest

Success! You follow the winding path into the forest. The trees grow thick
and the light dims. Your torch flickers.
```

## Adding Optimization

Once you have labeled examples of good game responses, optimize the modules:

```{r optimize}
# Collect some example scenes you like
training_scenes <- dsp_trainset(
  location = c("forest_clearing", "dungeon_entrance"),
  inventory = c("torch, sword", "torch, rope, key"),
  history = c("Fought wolves in the forest", "Found the hidden dungeon"),
  description = c(
    "Moonlight filters through ancient oaks into a peaceful clearing. A moss-covered stone altar stands at the center.",
    "A massive iron door blocks your path, its surface covered in strange runes that seem to pulse with faint light."
  ),
  actions = c(
    "examine altar, rest, continue north, search undergrowth, climb tree",
    "try the key, read runes, search for another entrance, knock, go back"
  ),
  npcs_present = c("none", "none")
)

# Compile with few-shot examples
scene_mod_optimized <- compile(
  LabeledFewShot(k = 2),
  scene_mod,
  trainset = training_scenes
)
```

## What This Demonstrates

This game showcases several dsprrr patterns:

- **Signature composition**: Three modules with distinct responsibilities
- **Chain of thought**: `with_reasoning()` helps the LLM make better narrative decisions
- **State management**: Game state flows through modules without coupling them
- **Interactive systems**: Modules respond to dynamic user input

The same patterns apply to chatbots, interactive assistants, and any system where LLM outputs feed back into subsequent calls.

## Next Steps

- Add location transitions with a `locations` map
- Implement combat with health tracking
- Use `optimize_grid()` to tune temperature for more creative vs consistent responses
- Add tool use for dice rolls: `module(type = "react", tools = list(roll_dice))`

See `vignette("advanced-modules")` for more module patterns.
