---
title: "Building a Text Adventure Game"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Text Adventure Game}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
# Skip evaluation during R CMD check or CI (cassettes may not match current code)
# VITALS_SHOULD_EVAL can override for cassette recording
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::setup_knitr()
}
```

This tutorial builds a text-based adventure game using dsprrr. The AI handles narrative generation, NPC dialogue, and action resolution while you control the game framework and player state.

## Why This Matters Beyond Games

The patterns in this tutorial directly apply to production AI systems:

- **Customer service agents**: Like NPCs, they maintain conversation context, respond appropriately to user mood, and escalate when needed. The `dialogue_sig` pattern—tracking relationship history, detecting sentiment, and generating contextual responses—is exactly how you'd build an intelligent support bot.

- **Interactive documentation**: Imagine a technical assistant that remembers what you've tried, suggests next steps based on your skill level, and adapts explanations to your context. That's the `scene_sig` pattern: generate options based on user state and history.

- **Workflow orchestration**: The action resolution system—evaluating whether an action succeeds based on context, applying consequences, and updating state—maps directly to approval workflows, form validation, and multi-step processes.

The game framing makes these patterns concrete and testable. Once you understand how to build a coherent NPC that remembers past conversations, building a customer service agent that does the same is straightforward.

*This tutorial is adapted from the [DSPy text adventure tutorial](https://dspy.ai/tutorials/ai_text_game/) by the DSPy team.*

We'll start with the simplest approach—plain lists—then show how S7 adds structure as the project grows.

## What You'll Build

- Dynamic scene generation with branching narratives
- AI-powered NPC conversations that remember context
- Action resolution with skill checks and consequences
- Inventory and character progression
- Save/load game functionality

```{r setup}
library(dsprrr)
library(ellmer)
library(cli)
```

## Part 1: The Simple Approach

For most R projects, plain lists and functions are all you need.

### Game State as Lists

```{r simple-state}
# Player is just a list
create_player <- function(name) {
  list(
    name = name,
    health = 100L,
    level = 1L,
    experience = 0L,
    inventory = character(),
    skills = list(
      strength = 10L,
      intelligence = 10L,
      charisma = 10L,
      stealth = 10L
    )
  )
}

# Game context is also a list
create_context <- function() {
  list(
    location = "Village Square",
    story_progress = 0L,
    visited = character(),
    npcs_met = character(),
    recent_actions = character()
  )
}

# Bundle into game state
create_game <- function(player) {
  list(
    player = player,
    context = create_context(),
    running = TRUE
  )
}
```

### Helper Functions

```{r simple-helpers}
# Modify player state
add_item <- function(player, item) {
  player$inventory <- c(player$inventory, item)
  cli_alert_success("Added {.val {item}} to inventory!")
  player
}

take_damage <- function(player, amount) {
  player$health <- max(0L, player$health - amount)
  cli_alert_danger("Took {amount} damage! HP: {player$health}/100")
  player
}

heal <- function(player, amount) {
  player$health <- min(100L, player$health + amount)
  cli_alert_success("Healed {amount} HP! HP: {player$health}/100")
  player
}

gain_xp <- function(player, amount) {
  player$experience <- player$experience + amount
  new_level <- 1L + (player$experience %/% 100L)
  if (new_level > player$level) {
    cli_alert_success("Level up! Now level {new_level}!")
    player$level <- new_level
  }
  player
}

is_alive <- function(player) {
  player$health > 0L
}

# Modify context
visit_location <- function(context, location) {
  context$visited <- unique(c(context$visited, context$location))
  context$location <- location
  context$story_progress <- context$story_progress + 1L
  cli_alert_info("Traveled to {.strong {location}}")
  context
}

record_action <- function(context, action) {
  context$recent_actions <- c(tail(context$recent_actions, 4L), action)
  context
}
```

### AI Signatures

This is where dsprrr comes in—define clear contracts for each AI task. Instead of writing prompt strings, we declare what information goes in and what structure comes out. This separation matters: when you want to improve the AI's scene descriptions, you tweak the signature's instructions rather than hunting through string templates.

Each signature serves a distinct purpose in the game loop:

**Scene generation** takes the current game state and produces a rich description with available actions. Notice the `atmosphere` enum—constraining outputs to specific values makes downstream logic predictable.

```{r simple-signatures}
scene_sig <- signature(
  inputs = list(
    input("location", description = "Current location name"),
    input("player_info", description = "Player stats summary"),
    input("story_progress", description = "Story progress 0-10"),
    input("recent_actions", description = "Last few player actions"),
    input("visited_before", description = "Whether location was visited")
  ),
  output_type = type_object(
    description = type_string("2-3 sentence scene description"),
    actions = type_array(type_string(), "4-6 possible actions"),
    npcs = type_array(type_string(), "NPCs present"),
    items = type_array(type_string(), "Visible items"),
    atmosphere = type_enum(c("peaceful", "tense", "mysterious", "dangerous"))
  ),
  instructions = "You are a game master for a fantasy text adventure.
Create immersive scenes with interesting choices."
)
```

**NPC dialogue** handles conversations. The `relationship` input ("new" or "returning") lets the AI adjust tone—a merchant who recognizes you might offer better prices. The structured output ensures we can programmatically handle items changing hands.

```{r}
dialogue_sig <- signature(
  inputs = list(
    input("npc_name", description = "NPC name"),
    input("npc_role", description = "NPC role and personality"),
    input("player_says", description = "Player's action or words"),
    input("relationship", description = "new or returning"),
    input("context", description = "Location and story context")
  ),
  output_type = type_object(
    response = type_string("NPC dialogue (1-3 sentences)"),
    mood = type_enum(c("friendly", "neutral", "suspicious", "hostile")),
    gives_item = type_string("Item given, or 'none'")
  ),
  instructions = "You are an NPC in a fantasy RPG. Stay in character."
)
```

**Action resolution** determines outcomes. Rather than rolling dice, we give the AI context (skill level, difficulty) and let it generate both the success/failure boolean and a narrative explanation. The structured output (`damage_taken`, `item_gained`, `experience`) lets us update game state automatically.

```{r}
action_sig <- signature(
  inputs = list(
    input("action", description = "What player attempts"),
    input("relevant_skill", description = "Which skill applies"),
    input("skill_value", description = "Player's skill value"),
    input("difficulty", description = "easy/medium/hard"),
    input("context", description = "Relevant circumstances")
  ),
  output_type = type_object(
    success = type_boolean(),
    outcome = type_string("2-3 sentence result"),
    damage_taken = type_integer("Damage to player"),
    item_gained = type_string("Item earned, or 'none'"),
    experience = type_integer("XP earned 0-25")
  ),
  instructions = "Resolve actions fairly but dramatically."
)
```

### Create Modules

Each signature becomes a module. We use `chain_of_thought()`, which adds a "reasoning" field to the output—the model explains its thinking before giving the final answer. This improves output quality, especially for complex decisions like whether an action should succeed.

```{r simple-modules}
create_modules <- function(llm) {
  list(
    scene = chain_of_thought(scene_sig),
    dialogue = chain_of_thought(dialogue_sig),
    action = chain_of_thought(action_sig),
    llm = llm
  )
}
```

### Game Logic

The core pattern: take game state, generate AI content, apply effects, repeat. Each function handles one concern—`player_summary` formats state for the AI, `detect_skill` maps player intent to game mechanics, and the `generate_*`/`resolve_*` functions orchestrate LLM calls.

```{r simple-logic}
player_summary <- function(player) {
  sprintf(
    "Level %d %s | HP:%d | STR:%d INT:%d CHA:%d STL:%d | Items: %s",
    player$level, player$name, player$health,
    player$skills$strength, player$skills$intelligence,
    player$skills$charisma, player$skills$stealth,
    if (length(player$inventory) > 0)
      paste(player$inventory, collapse = ", ")
    else "none"
  )
}

detect_skill <- function(action) {
  action_lower <- tolower(action)
  if (grepl("fight|attack|force", action_lower)) "strength"
  else if (grepl("sneak|hide|steal", action_lower)) "stealth"
  else if (grepl("persuade|convince|charm", action_lower)) "charisma"
  else "intelligence"
}

generate_scene <- function(game, modules) {
  run(
    modules$scene,
    location = game$context$location,
    player_info = player_summary(game$player),
    story_progress = as.character(game$context$story_progress),
    recent_actions = paste(game$context$recent_actions, collapse = " -> "),
    visited_before = game$context$location %in% game$context$visited,
    .llm = modules$llm
  )
}

resolve_action <- function(game, action, modules) {
  skill <- detect_skill(action)

  result <- run(
    modules$action,
    action = action,
    relevant_skill = skill,
    skill_value = as.character(game$player$skills[[skill]]),
    difficulty = "medium",
    context = game$context$location,
    .llm = modules$llm
  )

  # Apply effects
  if (result$damage_taken > 0) {
    game$player <- take_damage(game$player, result$damage_taken)
  }
  if (result$item_gained != "none" && nzchar(result$item_gained)) {
    game$player <- add_item(game$player, result$item_gained)
  }
  if (result$experience > 0) {
    game$player <- gain_xp(game$player, result$experience)
  }

  if (result$success) {
    cli_alert_success(result$outcome)
  } else {
    cli_alert_danger(result$outcome)
  }

  game
}
```

### Save/Load

```{r simple-save}
save_game <- function(game, filename = "savegame.json") {
  jsonlite::write_json(game, filename, auto_unbox = TRUE, pretty = TRUE)
  cli_alert_success("Saved to {.file {filename}}")
  game
}

load_game <- function(filename = "savegame.json") {
  if (!file.exists(filename)) {
    cli_alert_danger("Save file not found")
    return(NULL)
  }
  game <- jsonlite::read_json(filename, simplifyVector = TRUE)
  cli_alert_success("Loaded from {.file {filename}}")
  game
}
```

### Minimal Game Loop

```{r simple-loop}
play_game <- function(llm = chat_openai()) {
  modules <- create_modules(llm)

  name <- readline("Character name: ")
  game <- create_game(create_player(name))

  repeat {
    scene <- generate_scene(game, modules)

    cli_h2("{game$context$location}")
    cli_text(scene$description)
    cat("\n")

    actions <- c(scene$actions, "Save", "Quit")
    for (i in seq_along(actions)) {
      cli_text("{i}. {actions[i]}")
    }

    choice <- as.integer(readline("Choose: "))
    chosen <- actions[choice]

    if (chosen == "Save") {
      game <- save_game(game)
      next
    }
    if (chosen == "Quit") break

    game$context <- record_action(game$context, chosen)
    game <- resolve_action(game, chosen, modules)

    if (!is_alive(game$player)) {
      cli_h1("GAME OVER")
      break
    }
  }

  invisible(game)
}
```

**This works.** For a quick prototype or personal project, you're done.

---

## Part 2: Adding Structure with S7

As projects grow, plain lists show their limits:

- No validation (what if health goes negative?)
- No documentation (which fields exist?)
- No type safety (is `level` an integer or character?)
- Hard to extend (adding new player types?)

S7 solves these while staying functional:

```{r s7-setup}
library(S7)
```

### S7 Classes for Game State

```{r s7-classes}
Player <- new_class("Player",
  properties = list(
    name = class_character,
    health = new_property(class_integer, default = 100L),
    level = new_property(class_integer, default = 1L),
    experience = new_property(class_integer, default = 0L),
    inventory = new_property(class_character, default = character()),
    skills = new_property(class_list, default = list(
      strength = 10L,
      intelligence = 10L,
      charisma = 10L,
      stealth = 10L
    ))
  ),
  validator = function(self) {
    if (self@health < 0L) return("health cannot be negative")
    if (self@health > 100L) return("health cannot exceed 100")
    if (self@level < 1L) return("level must be at least 1")
    NULL
  }
)

GameContext <- new_class("GameContext",
  properties = list(
    location = new_property(class_character, default = "Village Square"),
    story_progress = new_property(class_integer, default = 0L),
    visited = new_property(class_character, default = character()),
    npcs_met = new_property(class_character, default = character()),
    recent_actions = new_property(class_character, default = character()),
    flags = new_property(class_list, default = list())
  )
)

GameState <- new_class("GameState",
  properties = list(
    player = Player,
    context = GameContext,
    running = new_property(class_logical, default = TRUE)
  )
)
```

Now you get:

- **Validation**: `Player(name = "X", health = -10L)` throws an error
- **Documentation**: Properties are self-describing
- **Type safety**: Can't accidentally assign wrong types
- **Tooling**: IDE autocomplete works with `@` slots

### S7 Generics for State Transitions

Instead of plain functions, use generics for extensibility:

```{r s7-generics}
add_item <- new_generic("add_item", "player")
method(add_item, Player) <- function(player, item) {
  player@inventory <- c(player@inventory, item)
  cli_alert_success("Added {.val {item}} to inventory!")
  player
}

take_damage <- new_generic("take_damage", "player")
method(take_damage, Player) <- function(player, amount) {
  player@health <- max(0L, player@health - as.integer(amount))
  cli_alert_danger("Took {amount} damage! HP: {player@health}/100")
  player
}

heal <- new_generic("heal", "player")
method(heal, Player) <- function(player, amount) {
  player@health <- min(100L, player@health + as.integer(amount))
  cli_alert_success("Healed {amount} HP! HP: {player@health}/100")
  player
}

gain_xp <- new_generic("gain_xp", "player")
method(gain_xp, Player) <- function(player, amount) {
  player@experience <- player@experience + as.integer(amount)
  new_level <- 1L + (player@experience %/% 100L)
  if (new_level > player@level) {
    cli_alert_success("Level up! Now level {new_level}!")
    player@level <- new_level
  }
  player
}

is_alive <- new_generic("is_alive", "player")
method(is_alive, Player) <- function(player) {
  player@health > 0L
}

visit_location <- new_generic("visit_location", "context")
method(visit_location, GameContext) <- function(context, location) {
  context@visited <- unique(c(context@visited, context@location))
  context@location <- location
  context@story_progress <- context@story_progress + 1L
  cli_alert_info("Traveled to {.strong {location}}")
  context
}

record_action <- new_generic("record_action", "context")
method(record_action, GameContext) <- function(context, action) {
  context@recent_actions <- c(tail(context@recent_actions, 4L), action)
  context
}

meet_npc <- new_generic("meet_npc", "context")
method(meet_npc, GameContext) <- function(context, npc) {
  context@npcs_met <- unique(c(context@npcs_met, npc))
  context
}

set_flag <- new_generic("set_flag", "context")
method(set_flag, GameContext) <- function(context, flag, value = TRUE) {
  context@flags[[flag]] <- value
  context
}

has_flag <- new_generic("has_flag", "context")
method(has_flag, GameContext) <- function(context, flag) {
  isTRUE(context@flags[[flag]])
}
```

### Print Methods

```{r s7-print}
method(print, Player) <- function(x, ...) {
  cli_h3("{x@name}")
  cli_text("Level {x@level} | HP: {x@health}/100 | XP: {x@experience}")
  cli_text(
    "STR:{x@skills$strength} INT:{x@skills$intelligence} ",
    "CHA:{x@skills$charisma} STL:{x@skills$stealth}"
  )
  if (length(x@inventory) > 0) {
    cli_text("Inventory: {.val {x@inventory}}")
  }
  invisible(x)
}

method(print, GameState) <- function(x, ...) {
  cli_rule("{.emph MYSTIC REALM ADVENTURE}")
  print(x@player)
  cli_rule()
  invisible(x)
}
```

### Richer Signatures

With more structure, we can add richer AI interactions:

```{r s7-signatures}
scene_sig <- signature(
  inputs = list(
    input("location", description = "Current location name"),
    input("player_info", description = "Player stats summary"),
    input("story_progress", description = "Story progress 0-10"),
    input("recent_actions", description = "Last few player actions"),
    input("visited_before", description = "Whether location was visited")
  ),
  output_type = type_object(
    description = type_string("2-3 sentence scene description"),
    actions = type_array(type_string(), "4-6 possible actions"),
    npcs = type_array(type_string(), "NPCs present"),
    items = type_array(type_string(), "Visible items"),
    atmosphere = type_enum(c("peaceful", "tense", "mysterious", "dangerous"))
  ),
  instructions = "You are a game master for a fantasy text adventure.
Create immersive scenes with interesting choices. Include NPC interactions
and exploration options. If visited before, note familiarity."
)

dialogue_sig <- signature(
  inputs = list(
    input("npc_name", description = "NPC name"),
    input("npc_role", description = "NPC role and personality"),
    input("player_says", description = "Player's action or words"),
    input("relationship", description = "new or returning"),
    input("context", description = "Location and story context")
  ),
  output_type = type_object(
    response = type_string("NPC dialogue (1-3 sentences)"),
    action = type_string("Physical gesture or action"),
    mood = type_enum(c("friendly", "neutral", "suspicious", "hostile")),
    offers_quest = type_boolean(),
    quest_hint = type_string("Quest description if offered"),
    reveals_info = type_string("Lore revealed"),
    gives_item = type_string("Item given, or 'none'")
  ),
  instructions = "You are an NPC in a fantasy RPG. Stay in character.
Build relationships over interactions. Reveal information gradually."
)

action_sig <- signature(
  inputs = list(
    input("action", description = "What player attempts"),
    input("relevant_skill", description = "Which skill applies"),
    input("skill_value", description = "Player's skill value"),
    input("difficulty", description = "easy/medium/hard/very_hard"),
    input("context", description = "Relevant circumstances"),
    input("has_useful_item", description = "Whether player has helpful item")
  ),
  output_type = type_object(
    success = type_boolean(),
    outcome = type_string("2-3 sentence result"),
    damage_taken = type_integer("Damage to player"),
    item_gained = type_string("Item earned, or 'none'"),
    item_lost = type_string("Item lost, or 'none'"),
    experience = type_integer("XP earned 0-25"),
    unlocks = type_string("New location/element unlocked"),
    consequence = type_string("Lasting effect on game")
  ),
  instructions = "Resolve actions fairly but dramatically. Higher skills
improve odds. Create interesting failures. Useful items help."
)
```

### Full Game Functions

```{r s7-game-functions}
player_summary <- function(player) {
  sprintf(
    "Level %d %s | HP:%d | STR:%d INT:%d CHA:%d STL:%d | Items: %s",
    player@level, player@name, player@health,
    player@skills$strength, player@skills$intelligence,
    player@skills$charisma, player@skills$stealth,
    if (length(player@inventory) > 0)
      paste(player@inventory, collapse = ", ")
    else "none"
  )
}

detect_skill <- function(action) {
  action_lower <- tolower(action)
  if (grepl("fight|attack|break|force|lift", action_lower)) "strength"
  else if (grepl("sneak|hide|steal|pick|shadow", action_lower)) "stealth"
  else if (grepl("persuade|convince|charm|talk|negotiate", action_lower)) "charisma"
  else "intelligence"
}

detect_difficulty <- function(action) {
  action_lower <- tolower(action)
  if (grepl("fight|battle|defeat|dangerous", action_lower)) "hard"
  else if (grepl("look|examine|rest|wait", action_lower)) "easy"
  else "medium"
}

has_useful_item <- function(player) {
  useful <- c("torch", "rope", "sword", "key", "map", "potion", "lockpick")
  any(tolower(player@inventory) %in% useful)
}

npc_role <- function(npc_name) {
  roles <- list(
    "Village Elder" = "wise leader who knows ancient lore",
    "Merchant" = "trader interested in rare items",
    "Guard" = "dutiful protector, suspicious of strangers",
    "Mysterious Stranger" = "hooded figure with unknown motives",
    "Blacksmith" = "crafts weapons, appreciates quality materials",
    "Healer" = "kind herbalist who restores health"
  )
  roles[[npc_name]] %||% "mysterious character"
}
```

### Scene and Dialogue Handling

```{r s7-scene-dialogue}
generate_scene <- function(game, modules) {
  run(
    modules$scene,
    location = game@context@location,
    player_info = player_summary(game@player),
    story_progress = as.character(game@context@story_progress),
    recent_actions = paste(game@context@recent_actions, collapse = " -> "),
    visited_before = game@context@location %in% game@context@visited,
    .llm = modules$llm
  )
}

display_scene <- function(scene, context) {
  emoji <- switch(scene$atmosphere,
    peaceful = "\U0001F343",
    tense = "\U0001F441",
    mysterious = "\U0001F319",
    dangerous = "\U0001F525"
  )

  cli_h2("{emoji} {context@location}")
  cli_text(scene$description)
  cat("\n")

  if (length(scene$npcs) > 0 && !identical(scene$npcs, "")) {
    cli_text("{.strong NPCs:} {.val {scene$npcs}}")
  }
  if (length(scene$items) > 0 && !identical(scene$items, "")) {
    cli_text("{.strong Items:} {.emph {scene$items}}")
  }
  cat("\n")

  cli_h3("What do you do?")
  all_actions <- c(scene$actions, "Check inventory", "Save game", "Quit")
  for (i in seq_along(all_actions)) {
    cli_text("{i}. {all_actions[i]}")
  }
  cat("\n")

  all_actions
}

handle_dialogue <- function(game, npc, action, modules) {
  relationship <- if (npc %in% game@context@npcs_met) "returning" else "new"

  response <- run(
    modules$dialogue,
    npc_name = npc,
    npc_role = npc_role(npc),
    player_says = action,
    relationship = relationship,
    context = sprintf("Location: %s, Progress: %d/10",
                      game@context@location, game@context@story_progress),
    .llm = modules$llm
  )

  cat("\n")
  cli_alert_info("{.strong {npc}}: \"{response$response}\"")
  if (nzchar(response$action)) cli_text("{.emph ({response$action})}")
  if (response$offers_quest && nzchar(response$quest_hint)) {
    cli_alert_warning("Quest available: {response$quest_hint}")
  }
  if (nzchar(response$reveals_info)) {
    cli_alert("You learned: {response$reveals_info}")
  }
  cat("\n")

  game@context <- meet_npc(game@context, npc)

  if (response$gives_item != "none" && nzchar(response$gives_item)) {
    game@player <- add_item(game@player, response$gives_item)
  }

  game
}
```

### Action Resolution

```{r s7-actions}
resolve_action <- function(game, action, modules) {
  skill <- detect_skill(action)

  result <- run(
    modules$action,
    action = action,
    relevant_skill = skill,
    skill_value = as.character(game@player@skills[[skill]]),
    difficulty = detect_difficulty(action),
    context = game@context@location,
    has_useful_item = has_useful_item(game@player),
    .llm = modules$llm
  )

  cat("\n")
  if (result$success) {
    cli_alert_success(result$outcome)
  } else {
    cli_alert_danger(result$outcome)
  }

  if (result$damage_taken > 0) {
    game@player <- take_damage(game@player, result$damage_taken)
  }
  if (result$item_gained != "none" && nzchar(result$item_gained)) {
    game@player <- add_item(game@player, result$item_gained)
  }
  if (result$experience > 0) {
    game@player <- gain_xp(game@player, result$experience)
    cli_alert_info("+{result$experience} XP")
  }
  if (nzchar(result$unlocks) && result$unlocks != "none") {
    cli_alert_success("Unlocked: {.val {result$unlocks}}")
  }

  if (grepl("go|travel|enter|head", action, ignore.case = TRUE) && result$success) {
    destination <- if (nzchar(result$unlocks) && result$unlocks != "none") {
      result$unlocks
    } else {
      sub(".*(?:go to|enter|head to) (?:the )?", "", action, ignore.case = TRUE)
    }
    if (nzchar(destination) && destination != action) {
      game@context <- visit_location(game@context, tools::toTitleCase(destination))
    }
  }

  cat("\n")
  game
}
```

### Save/Load with S7

```{r s7-save-load}
save_game <- function(game, filename = "savegame.json") {
  save_data <- list(
    player = list(
      name = game@player@name,
      health = game@player@health,
      level = game@player@level,
      experience = game@player@experience,
      inventory = as.list(game@player@inventory),
      skills = game@player@skills
    ),
    context = list(
      location = game@context@location,
      story_progress = game@context@story_progress,
      visited = as.list(game@context@visited),
      npcs_met = as.list(game@context@npcs_met),
      flags = game@context@flags
    )
  )
  jsonlite::write_json(save_data, filename, auto_unbox = TRUE, pretty = TRUE)
  cli_alert_success("Saved to {.file {filename}}")
  game
}

load_game <- function(filename = "savegame.json") {
  if (!file.exists(filename)) {
    cli_alert_danger("Save file not found: {.file {filename}}")
    return(NULL)
  }

  data <- jsonlite::read_json(filename)

  player <- Player(
    name = data$player$name,
    health = as.integer(data$player$health),
    level = as.integer(data$player$level),
    experience = as.integer(data$player$experience),
    inventory = as.character(unlist(data$player$inventory) %||% character()),
    skills = lapply(data$player$skills, as.integer)
  )

  context <- GameContext(
    location = data$context$location,
    story_progress = as.integer(data$context$story_progress),
    visited = as.character(unlist(data$context$visited) %||% character()),
    npcs_met = as.character(unlist(data$context$npcs_met) %||% character()),
    flags = data$context$flags %||% list()
  )

  cli_alert_success("Loaded from {.file {filename}}")
  GameState(player = player, context = context)
}
```

### Character Creation

```{r s7-character}
create_character <- function() {
  cli_h1("{.emph MYSTIC REALM ADVENTURE}")
  cli_h2("Character Creation")

  name <- readline("Enter your character's name: ")

  cli_text("\nYou have {.strong 10 bonus points} to distribute.")
  cli_text("Base skills start at 10.\n")

  skills <- list(strength = 10L, intelligence = 10L, charisma = 10L, stealth = 10L)
  points <- 10L

  for (skill in names(skills)) {
    if (points > 0L) {
      cli_text("Points remaining: {.val {points}}")
      input <- readline(sprintf("Add to %s (0-%d): ", skill, points))
      add <- min(points, max(0L, as.integer(input) %||% 0L))
      skills[[skill]] <- skills[[skill]] + add
      points <- points - add
    }
  }

  player <- Player(name = name, skills = skills)

  cli_alert_success("\nWelcome, {.strong {name}}!")
  cli_text("Your adventure begins...\n")

  GameState(player = player, context = GameContext())
}
```

### Full Game Loop

```{r s7-game-loop}
play_game <- function(llm = chat_openai()) {
  modules <- create_modules(llm)
  game <- create_character()

  repeat {
    print(game)

    scene <- generate_scene(game, modules)
    actions <- display_scene(scene, game@context)

    choice <- as.integer(readline("Choose (number): "))

    if (is.na(choice) || choice < 1 || choice > length(actions)) {
      cli_alert_warning("Invalid choice")
      next
    }

    chosen <- actions[choice]
    game@context <- record_action(game@context, chosen)

    if (chosen == "Check inventory") {
      if (length(game@player@inventory) == 0) {
        cli_alert_info("Your inventory is empty.")
      } else {
        cli_h3("Inventory")
        for (item in game@player@inventory) cli_li("{item}")
      }
      readline("Press Enter...")
      next
    }

    if (chosen == "Save game") {
      game <- save_game(game)
      readline("Press Enter...")
      next
    }

    if (chosen == "Quit") {
      if (readline("Save first? (y/n): ") == "y") {
        game <- save_game(game)
      }
      game@running <- FALSE
      break
    }

    npc_target <- NULL
    for (npc in scene$npcs) {
      if (grepl(tolower(npc), tolower(chosen), fixed = TRUE)) {
        npc_target <- npc
        break
      }
    }

    if (!is.null(npc_target)) {
      game <- handle_dialogue(game, npc_target, chosen, modules)
    } else {
      game <- resolve_action(game, chosen, modules)
    }

    if (!is_alive(game@player)) {
      cli_h1("GAME OVER")
      cli_text("You have fallen...")
      break
    }

    readline("Press Enter...")
  }

  cli_text("Thanks for playing!")
  invisible(game)
}
```

## When to Use Which

| Approach | Use When |
|----------|----------|
| **Plain lists** | Prototypes, scripts, personal projects |
| **S7 classes** | Shared code, packages, complex state, need validation |

Both work with dsprrr. The AI signatures and modules are the same—only the state management differs.
