---
title: "Building a Text Adventure Game"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Text Adventure Game}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This tutorial builds a complete text-based adventure game using dsprrr and S7. The AI handles narrative generation, NPC dialogue, and action resolution while you control the game framework and player state.

## What You'll Build

- Dynamic scene generation with branching narratives
- AI-powered NPC conversations that remember context
- Action resolution with skill checks and consequences
- Inventory and character progression
- Save/load game functionality

```{r setup}
library(dsprrr)
library(ellmer)
library(S7)
library(cli)
```

## Step 1: S7 Classes for Game State

Define immutable, typed game state using S7:

```{r s7-classes}
# Player class with typed properties and validation
Player <- new_class("Player",
 properties = list(
    name = class_character,
    health = new_property(class_integer, default = 100L),
    level = new_property(class_integer, default = 1L),
    experience = new_property(class_integer, default = 0L),
    inventory = new_property(class_character, default = character()),
    skills = new_property(class_list, default = list(
      strength = 10L,
      intelligence = 10L,
      charisma = 10L,
      stealth = 10L
    ))
  ),
  validator = function(self) {
    if (self@health < 0L) return("health cannot be negative")
    if (self@health > 100L) return("health cannot exceed 100")
    if (self@level < 1L) return("level must be at least 1")
    NULL
  }
)

# Game context tracks world state
GameContext <- new_class("GameContext",
  properties = list(
    location = new_property(class_character, default = "Village Square"),
    story_progress = new_property(class_integer, default = 0L),
    visited = new_property(class_character, default = character()),
    npcs_met = new_property(class_character, default = character()),
    recent_actions = new_property(class_character, default = character()),
    flags = new_property(class_list, default = list())
  )
)

# Bundle everything into game state
GameState <- new_class("GameState",
  properties = list(
    player = Player,
    context = GameContext,
    running = new_property(class_logical, default = TRUE)
  )
)
```

## Step 2: S7 Generics for State Transitions

Define pure functions that return modified state:

```{r s7-generics}
# Player state transitions
add_item <- new_generic("add_item", "player")
method(add_item, Player) <- function(player, item) {
  player@inventory <- c(player@inventory, item)
  cli_alert_success("Added {.val {item}} to inventory!")
 player
}

remove_item <- new_generic("remove_item", "player")
method(remove_item, Player) <- function(player, item) {
  if (item %in% player@inventory) {
    player@inventory <- setdiff(player@inventory, item)
    cli_alert_warning("Removed {.val {item}} from inventory")
  }
 player
}

take_damage <- new_generic("take_damage", "player")
method(take_damage, Player) <- function(player, amount) {
  player@health <- max(0L, player@health - as.integer(amount))
  cli_alert_danger("Took {amount} damage! HP: {player@health}/100")
  player
}

heal <- new_generic("heal", "player")
method(heal, Player) <- function(player, amount) {
  player@health <- min(100L, player@health + as.integer(amount))
  cli_alert_success("Healed {amount} HP! HP: {player@health}/100")
  player
}

gain_xp <- new_generic("gain_xp", "player")
method(gain_xp, Player) <- function(player, amount) {
  player@experience <- player@experience + as.integer(amount)
  new_level <- 1L + (player@experience %/% 100L)
  if (new_level > player@level) {
    cli_alert_success("Level up! Now level {new_level}!")
    player@level <- new_level
  }
  player
}

is_alive <- new_generic("is_alive", "player")
method(is_alive, Player) <- function(player) {
  player@health > 0L
}

# Context state transitions
record_action <- new_generic("record_action", "context")
method(record_action, GameContext) <- function(context, action) {
  context@recent_actions <- c(tail(context@recent_actions, 4L), action)
  context
}

visit_location <- new_generic("visit_location", "context")
method(visit_location, GameContext) <- function(context, location) {
  context@visited <- unique(c(context@visited, context@location))
  context@location <- location
  context@story_progress <- context@story_progress + 1L
  cli_alert_info("Traveled to {.strong {location}}")
  context
}

meet_npc <- new_generic("meet_npc", "context")
method(meet_npc, GameContext) <- function(context, npc) {
  context@npcs_met <- unique(c(context@npcs_met, npc))
  context
}

set_flag <- new_generic("set_flag", "context")
method(set_flag, GameContext) <- function(context, flag, value = TRUE) {
  context@flags[[flag]] <- value
  context
}

has_flag <- new_generic("has_flag", "context")
method(has_flag, GameContext) <- function(context, flag) {
  isTRUE(context@flags[[flag]])
}
```

## Step 3: Print Methods

```{r print-methods}
method(print, Player) <- function(x, ...) {
  cli_h3("{x@name}")
  cli_text("Level {x@level} | HP: {x@health}/100 | XP: {x@experience}")
  cli_text(
    "STR:{x@skills$strength} INT:{x@skills$intelligence} ",
    "CHA:{x@skills$charisma} STL:{x@skills$stealth}"
  )
  if (length(x@inventory) > 0) {
    cli_text("Inventory: {.val {x@inventory}}")
  }
  invisible(x)
}

method(print, GameState) <- function(x, ...) {
  cli_rule("{.emph MYSTIC REALM ADVENTURE}")
  print(x@player)
  cli_rule()
  invisible(x)
}
```

## Step 4: AI Signatures

Define clear input/output contracts for each AI task:

```{r signatures}
scene_sig <- signature(
  inputs = list(
    input("location", "Current location name"),
    input("player_info", "Player stats summary"),
    input("story_progress", "Story progress 0-10"),
    input("recent_actions", "Last few player actions"),
    input("visited_before", "Whether location was visited")
  ),
  output_type = type_object(
    description = type_string("2-3 sentence scene description"),
    actions = type_array(type_string(), "4-6 possible actions"),
    npcs = type_array(type_string(), "NPCs present"),
    items = type_array(type_string(), "Visible items"),
    atmosphere = type_enum(c("peaceful", "tense", "mysterious", "dangerous"))
  ),
  instructions = "You are a game master for a fantasy text adventure.
Create immersive scenes with interesting choices. Include NPC interactions
and exploration options. If visited before, note familiarity."
)

dialogue_sig <- signature(
  inputs = list(
    input("npc_name", "NPC name"),
    input("npc_role", "NPC role and personality"),
    input("player_says", "Player's action or words"),
    input("relationship", "new or returning"),
    input("context", "Location and story context")
  ),
  output_type = type_object(
    response = type_string("NPC dialogue (1-3 sentences)"),
    action = type_string("Physical gesture or action"),
    mood = type_enum(c("friendly", "neutral", "suspicious", "hostile")),
    offers_quest = type_boolean(),
    quest_hint = type_string("Quest description if offered"),
    reveals_info = type_string("Lore revealed"),
    gives_item = type_string("Item given, or 'none'")
  ),
  instructions = "You are an NPC in a fantasy RPG. Stay in character.
Build relationships over interactions. Reveal information gradually."
)

action_sig <- signature(
  inputs = list(
    input("action", "What player attempts"),
    input("relevant_skill", "Which skill applies"),
    input("skill_value", "Player's skill value"),
    input("difficulty", "easy/medium/hard/very_hard"),
    input("context", "Relevant circumstances"),
    input("has_useful_item", "Whether player has helpful item")
  ),
  output_type = type_object(
    success = type_boolean(),
    outcome = type_string("2-3 sentence result"),
    damage_taken = type_integer("Damage to player"),
    item_gained = type_string("Item earned, or 'none'"),
    item_lost = type_string("Item lost, or 'none'"),
    experience = type_integer("XP earned 0-25"),
    unlocks = type_string("New location/element unlocked"),
    consequence = type_string("Lasting effect on game")
  ),
  instructions = "Resolve actions fairly but dramatically. Higher skills
improve odds. Create interesting failures. Useful items help."
)
```

## Step 5: Module Container

Use a simple list for modules (they're already dsprrr objects):

```{r modules}
create_modules <- function(llm) {
  list(
    scene = scene_sig |> with_reasoning() |> module(type = "predict"),
    dialogue = dialogue_sig |> with_reasoning() |> module(type = "predict"),
    action = action_sig |> with_reasoning() |> module(type = "predict"),
    llm = llm
  )
}
```

## Step 6: Pure Functions for Game Logic

```{r game-logic}
# Format player info for AI
player_summary <- function(player) {
 sprintf(
    "Level %d %s | HP:%d | STR:%d INT:%d CHA:%d STL:%d | Items: %s",
    player@level, player@name, player@health,
    player@skills$strength, player@skills$intelligence,
    player@skills$charisma, player@skills$stealth,
    if (length(player@inventory) > 0)
      paste(player@inventory, collapse = ", ")
    else "none"
  )
}

# Detect which skill applies to an action
detect_skill <- function(action) {
  action_lower <- tolower(action)
  if (grepl("fight|attack|break|force|lift", action_lower)) "strength"
  else if (grepl("sneak|hide|steal|pick|shadow", action_lower)) "stealth"
  else if (grepl("persuade|convince|charm|talk|negotiate", action_lower)) "charisma"
  else "intelligence"
}

# Detect action difficulty
detect_difficulty <- function(action) {
  action_lower <- tolower(action)
  if (grepl("fight|battle|defeat|dangerous", action_lower)) "hard"
  else if (grepl("look|examine|rest|wait", action_lower)) "easy"
  else "medium"
}

# Check for useful items
has_useful_item <- function(player) {
  useful <- c("torch", "rope", "sword", "key", "map", "potion", "lockpick")
  any(tolower(player@inventory) %in% useful)
}

# NPC role lookup
npc_role <- function(npc_name) {
  roles <- list(
    "Village Elder" = "wise leader who knows ancient lore",
    "Merchant" = "trader interested in rare items",
    "Guard" = "dutiful protector, suspicious of strangers",
    "Mysterious Stranger" = "hooded figure with unknown motives",
    "Blacksmith" = "crafts weapons, appreciates quality materials",
    "Healer" = "kind herbalist who restores health",
    "Thief" = "cunning rogue with secret knowledge"
  )
  roles[[npc_name]] %||% "mysterious character"
}
```

## Step 7: Scene Generation

```{r generate-scene}
generate_scene <- function(game, modules) {
  run(
    modules$scene,
    location = game@context@location,
    player_info = player_summary(game@player),
    story_progress = as.character(game@context@story_progress),
    recent_actions = paste(game@context@recent_actions, collapse = " -> "),
    visited_before = game@context@location %in% game@context@visited,
    .llm = modules$llm
  )
}

display_scene <- function(scene, context) {
  # Atmosphere emoji
  emoji <- switch(scene$atmosphere,
    peaceful = "\U0001F343",
    tense = "\U0001F441",
    mysterious = "\U0001F319",
    dangerous = "\U0001F525"
  )

  cli_h2("{emoji} {context@location}")
  cli_text(scene$description)
  cat("\n")

  if (length(scene$npcs) > 0 && !identical(scene$npcs, "")) {
    cli_text("{.strong NPCs:} {.val {scene$npcs}}")
  }
  if (length(scene$items) > 0 && !identical(scene$items, "")) {
    cli_text("{.strong Items:} {.emph {scene$items}}")
  }
  cat("\n")

  # Numbered action list
  cli_h3("What do you do?")
  all_actions <- c(scene$actions, "Check inventory", "Save game", "Quit")
  for (i in seq_along(all_actions)) {
    cli_text("{i}. {all_actions[i]}")
  }
  cat("\n")

  all_actions
}
```

## Step 8: Dialogue Handling

```{r dialogue}
handle_dialogue <- function(game, npc, action, modules) {
  relationship <- if (npc %in% game@context@npcs_met) "returning" else "new"

  response <- run(
    modules$dialogue,
    npc_name = npc,
    npc_role = npc_role(npc),
    player_says = action,
    relationship = relationship,
    context = sprintf("Location: %s, Progress: %d/10",
                      game@context@location, game@context@story_progress),
    .llm = modules$llm
  )

  # Display
  cat("\n")
  cli_alert_info("{.strong {npc}}: \"{response$response}\"")
  if (nzchar(response$action)) cli_text("{.emph ({response$action})}")
  if (response$offers_quest && nzchar(response$quest_hint)) {
    cli_alert_warning("Quest available: {response$quest_hint}")
  }
  if (nzchar(response$reveals_info)) {
    cli_alert("You learned: {response$reveals_info}")
  }
  cat("\n")

  # Update state immutably
  game@context <- meet_npc(game@context, npc)

  if (response$gives_item != "none" && nzchar(response$gives_item)) {
    game@player <- add_item(game@player, response$gives_item)
  }

  game
}
```

## Step 9: Action Resolution

```{r action-resolution}
resolve_action <- function(game, action, modules) {
  skill <- detect_skill(action)

  result <- run(
    modules$action,
    action = action,
    relevant_skill = skill,
    skill_value = as.character(game@player@skills[[skill]]),
    difficulty = detect_difficulty(action),
    context = game@context@location,
    has_useful_item = has_useful_item(game@player),
    .llm = modules$llm
  )

  # Display result
  cat("\n")
  if (result$success) {
    cli_alert_success(result$outcome)
  } else {
    cli_alert_danger(result$outcome)
  }

  # Apply state changes immutably using pipes
  if (result$damage_taken > 0) {
    game@player <- take_damage(game@player, result$damage_taken)
  }
  if (result$item_gained != "none" && nzchar(result$item_gained)) {
    game@player <- add_item(game@player, result$item_gained)
  }
  if (result$item_lost != "none" && nzchar(result$item_lost)) {
    game@player <- remove_item(game@player, result$item_lost)
  }
  if (result$experience > 0) {
    game@player <- gain_xp(game@player, result$experience)
    cli_alert_info("+{result$experience} XP")
  }
  if (nzchar(result$unlocks) && result$unlocks != "none") {
    cli_alert_success("Unlocked: {.val {result$unlocks}}")
  }

  # Handle movement
  if (grepl("go|travel|enter|head", action, ignore.case = TRUE) && result$success) {
    destination <- if (nzchar(result$unlocks) && result$unlocks != "none") {
      result$unlocks
    } else {
      sub(".*(?:go to|enter|head to) (?:the )?", "", action, ignore.case = TRUE)
    }
    if (nzchar(destination) && destination != action) {
      game@context <- visit_location(game@context, tools::toTitleCase(destination))
    }
  }

  cat("\n")
  game
}
```

## Step 10: Save/Load with S7

```{r save-load}
save_game <- function(game, filename = "savegame.json") {
  save_data <- list(
    player = list(
      name = game@player@name,
      health = game@player@health,
      level = game@player@level,
      experience = game@player@experience,
      inventory = as.list(game@player@inventory),
      skills = game@player@skills
    ),
    context = list(
      location = game@context@location,
      story_progress = game@context@story_progress,
      visited = as.list(game@context@visited),
      npcs_met = as.list(game@context@npcs_met),
      flags = game@context@flags
    )
  )
  jsonlite::write_json(save_data, filename, auto_unbox = TRUE, pretty = TRUE)
  cli_alert_success("Saved to {.file {filename}}")
  game
}

load_game <- function(filename = "savegame.json") {
  if (!file.exists(filename)) {
    cli_alert_danger("Save file not found: {.file {filename}}")
    return(NULL)
  }

  data <- jsonlite::read_json(filename)

  player <- Player(
    name = data$player$name,
    health = as.integer(data$player$health),
    level = as.integer(data$player$level),
    experience = as.integer(data$player$experience),
    inventory = as.character(unlist(data$player$inventory) %||% character()),
    skills = lapply(data$player$skills, as.integer)
  )

  context <- GameContext(
    location = data$context$location,
    story_progress = as.integer(data$context$story_progress),
    visited = as.character(unlist(data$context$visited) %||% character()),
    npcs_met = as.character(unlist(data$context$npcs_met) %||% character()),
    flags = data$context$flags %||% list()
  )

  cli_alert_success("Loaded from {.file {filename}}")
  GameState(player = player, context = context)
}
```

## Step 11: Character Creation

```{r character-creation}
create_character <- function() {
  cli_h1("{.emph MYSTIC REALM ADVENTURE}")
  cli_h2("Character Creation")

  name <- readline("Enter your character's name: ")

  cli_text("\nYou have {.strong 10 bonus points} to distribute.")
  cli_text("Base skills start at 10.\n")

  skills <- list(strength = 10L, intelligence = 10L, charisma = 10L, stealth = 10L)
  points <- 10L

  for (skill in names(skills)) {
    if (points > 0L) {
      cli_text("Points remaining: {.val {points}}")
      input <- readline(sprintf("Add to %s (0-%d): ", skill, points))
      add <- min(points, max(0L, as.integer(input) %||% 0L))
      skills[[skill]] <- skills[[skill]] + add
      points <- points - add
    }
  }

  player <- Player(name = name, skills = skills)

  cli_alert_success("\nWelcome, {.strong {name}}!")
  cli_text("Your adventure begins...\n")

  GameState(player = player, context = GameContext())
}
```

## Step 12: Main Game Loop

```{r game-loop}
play_game <- function(llm = chat_openai()) {
  modules <- create_modules(llm)
  game <- create_character()

  repeat {
    print(game)

    # Generate and display scene
    scene <- generate_scene(game, modules)
    actions <- display_scene(scene, game@context)

    # Get player choice
    choice <- as.integer(readline("Choose (number): "))

    if (is.na(choice) || choice < 1 || choice > length(actions)) {
      cli_alert_warning("Invalid choice")
      next
    }

    chosen <- actions[choice]
    game@context <- record_action(game@context, chosen)

    # Handle special actions
    if (chosen == "Check inventory") {
      if (length(game@player@inventory) == 0) {
        cli_alert_info("Your inventory is empty.")
      } else {
        cli_h3("Inventory")
        for (item in game@player@inventory) cli_li("{item}")
      }
      readline("Press Enter...")
      next
    }

    if (chosen == "Save game") {
      game <- save_game(game)
      readline("Press Enter...")
      next
    }

    if (chosen == "Quit") {
      if (readline("Save first? (y/n): ") == "y") {
        game <- save_game(game)
      }
      game@running <- FALSE
      break
    }

    # Check for NPC interaction
    npc_target <- NULL
    for (npc in scene$npcs) {
      if (grepl(tolower(npc), tolower(chosen), fixed = TRUE)) {
        npc_target <- npc
        break
      }
    }

    if (!is.null(npc_target)) {
      game <- handle_dialogue(game, npc_target, chosen, modules)
    } else {
      game <- resolve_action(game, chosen, modules)
    }

    # Check for death
    if (!is_alive(game@player)) {
      cli_h1("GAME OVER")
      cli_text("You have fallen...")
      break
    }

    readline("Press Enter...")
  }

  cli_text("Thanks for playing!")
  invisible(game)
}
```

## Running the Game

```{r play}
# New game
play_game(llm = chat_openai())

# Load saved game
# game <- load_game()
# if (!is.null(game)) play_game_from(game, chat_openai())
```

## Example Session

```
── MYSTIC REALM ADVENTURE ─────────────────────────────

── Aria ──
Level 1 | HP: 100/100 | XP: 0
STR:12 INT:14 CHA:12 STL:12
Inventory: torch
────────────────────────────────────────────────────────

── 🍃 Village Square ──

The morning sun warms the cobblestones of Willowbrook's
central square. A weathered fountain burbles beside a
massive oak, its branches heavy with age.

NPCs: Village Elder, Merchant
Items: strange medallion, fresh bread

── What do you do? ──

1. Approach the Village Elder
2. Browse the merchant's wares
3. Examine the strange medallion
4. Head toward the forest path
5. Rest on the bench
6. Check inventory
7. Save game
8. Quit

Choose (number): 1

ℹ Village Elder: "Ah, a new face. I sense purpose in
  your step, young one. Perhaps you seek the ancient
  ruins north of here?"
  (strokes his long white beard)
! Quest available: Investigate the Northern Ruins
ℹ You learned: None who entered the ruins have returned.

Press Enter...
```

## What This Demonstrates

- **S7 for state**: Typed properties, validation, immutable updates
- **Generics for transitions**: `add_item()`, `take_damage()`, etc.
- **Composition over inheritance**: Simple lists for module collections
- **Pipe-friendly**: State flows through pure functions
- **dsprrr integration**: Signatures, modules, `with_reasoning()`

The same patterns work for chatbots, customer support systems, and any stateful AI application.
