---
title: "Use dsprrr with tidymodels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use dsprrr with tidymodels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
has_dials <- requireNamespace("dials", quietly = TRUE)
has_parsnip <- requireNamespace("parsnip", quietly = TRUE)
```

This page shows how dsprrr fits with tidymodels today: score module output
with yardstick, and build grids for `optimize_grid()` with dials. It also
explains what the parsnip model `llm_predict()` does, and why it can't be
fitted yet.

```{r setup}
#| eval: true
#| message: false
library(dsprrr)
```

## Score predictions with yardstick

Run the module over a data frame with `run_dataset()`, which keeps every column
and adds a `result` list column. Turn the predicted field into a factor with
the truth's levels, and any yardstick metric works:

```{r}
library(yardstick)

reviews <- tibble::tibble(
  text = c("Arrived broken.", "Works great.", "It's a chair."),
  sentiment = factor(c("negative", "positive", "neutral"))
)
classify <- module(
  signature("text -> sentiment: enum('negative', 'neutral', 'positive')")
)

scored <- run_dataset(
  classify,
  reviews,
  .llm = ellmer::chat_openai(model = "gpt-4.1-mini")
)
scored$.pred_class <- factor(
  vapply(
    scored$result,
    \(r) if (is.list(r)) r$sentiment else NA_character_,
    character(1)
  ),
  levels = levels(scored$sentiment)
)

accuracy(scored, truth = sentiment, estimate = .pred_class)
conf_mat(scored, truth = sentiment, estimate = .pred_class)
```

A row whose call failed holds `NA` in `result`, hence the `is.list()` check.
Setting `levels` matters: yardstick refuses a prediction factor whose levels
differ from the truth's, which happens whenever some class is never predicted.

## Build grids with dials

`temperature()`, `top_p()` and `reasoning_effort()` are dials parameter
objects. `temperature()` and `top_p()` default to the range 0 to 1:

```{r dials-params}
#| eval: !expr has_dials
temperature()
reasoning_effort()
dials::grid_regular(temperature(), levels = 3)
```

`optimize_grid()` takes such a grid directly, or a `dials::parameters()` set to
build one from. It tries each row on the data and keeps the best settings in
the module:

```{r}
optimize_grid(
  classify,
  data = reviews,
  metric = metric_exact_match(field = "sentiment"),
  grid = dials::grid_regular(temperature(), levels = 3),
  .llm = ellmer::chat_openai(model = "gpt-4.1-mini")
)
module_trials(classify)
```

`module_parameters(classify)` suggests a parameter set for a module. Don't
sweep `temperature` for reasoning models; see
[Models, providers and streaming](models-and-providers.html).

## The parsnip model

When parsnip is loaded, dsprrr registers a model type, `llm_predict()`, with
the engine `"dsprrr"` for classification and regression:

```{r parsnip-spec}
#| eval: !expr has_parsnip
library(parsnip)

spec <- llm_predict(
  mode = "classification",
  signature = "text -> sentiment: enum('negative', 'neutral', 'positive')",
  temperature = 0
) |>
  set_engine("dsprrr")
spec
```

Fitting this model does not learn from the training data. The engine's fit
step builds a zero-shot module from `signature` (or, without one, from the
predictor names and the outcome's distinct values, such as
`text -> output: enum('negative', 'positive')`), puts `temperature` and
`top_p` in the module's config, and records the predictor names. The
training rows are not used as examples. `predict()` then runs the module on
`new_data` with the default Chat, so call `set_default_chat()` first. It
returns `.pred_class` for classification, as a factor with only the levels that
were predicted, or `.pred` for regression.

With parsnip 1.6.0, `fit()` and `fit_xy()` on this specification currently
stop with "no applicable method for 'filter' applied to an object of class
"NULL"", because the engine registers no predictor encoding. The same failure
stops `tune::tune_grid()` and workflows, which fit models through these
functions.
Until that is fixed, the module and `run_dataset()` approach at the top of this
page gives the same zero-shot predictions.

For evaluation with graded or repeated samples, see
[Evaluate with vitals](vitals-recipes.html).
