---
title: "Troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r}
#| include: false
#| eval: true
# Build this page as a session with no API keys and no default chat, so every
# error below is reproduced offline and no request reaches a provider.
key_vars <- c("OPENAI_API_KEY", "ANTHROPIC_API_KEY", "GOOGLE_API_KEY")
saved_keys <- Sys.getenv(key_vars, unset = NA)
Sys.unsetenv(key_vars)
saved_options <- options(dsprrr.default_chat = NULL)
dsprrr::clear_default_chat()
```

Find the message you got, read why it happens, and apply the fix below it. The
messages on this page are real: the page was built in an R session without API
keys, and each error was produced with `try()` while rendering it.

## Check your setup with `dsprrr_sitrep()`

`dsprrr_sitrep()` reports package versions, the default chat, which API keys
are set, and the state of the response cache. This is its output in the
session that built this page:

```{r}
#| eval: true
#| message: false
library(dsprrr)
```

```{r}
#| eval: true
dsprrr_sitrep()
```

```{r}
#| include: false
#| eval: true
# Keep the examples below out of the disk cache.
configure_cache(enable_disk = FALSE)
```

## No chat or API key

### "No default Chat available"

```{r}
#| eval: true
qa <- module(signature("question -> answer"))
try(run(qa, question = "What is the capital of France?"))
```

`run()` needs an ellmer chat. Pass one with `.llm`, or set a default once per
session:

```{r}
chat <- ellmer::chat_openai(model = "gpt-4.1-mini")
run(qa, question = "What is the capital of France?", .llm = chat)

set_default_chat(chat)
run(qa, question = "What is the capital of France?")
```

When `OPENAI_API_KEY`, `ANTHROPIC_API_KEY` or `GOOGLE_API_KEY` is set, dsprrr
creates a default chat for that provider on first use. With none of them set,
`dsp_configure()` cannot guess either:

```{r}
#| eval: true
try(dsp_configure())
```

### "Can't find env var `OPENAI_API_KEY`"

```{r}
#| eval: true
chat <- ellmer::chat_openai(model = "gpt-4.1-mini")
try(run(qa, question = "What is the capital of France?", .llm = chat))
```

ellmer creates the chat without checking for a key, so the error appears at
the first request. The warning comes from dsprrr's cache, which also needs the
key to tell accounts apart; it goes away once the key is set.

Store the key in `~/.Renviron` (`usethis::edit_r_environ()` opens it) as a line
such as `OPENAI_API_KEY=sk-...`, then restart R. Each ellmer provider reads its
own variable, for example `ANTHROPIC_API_KEY` for `ellmer::chat_anthropic()`.
A key that is set but invalid or revoked reaches the provider and fails with an
HTTP 401 error instead.

### "Unknown provider"

```{r}
#| eval: true
try(dsp_configure(provider = "mistral"))
```

`dsp_configure()` builds chats for three providers. For any other provider
ellmer supports, create the chat yourself and make it the default:

```{r}
set_default_chat(ellmer::chat_mistral())
```

[Models, providers and streaming](models-and-providers.html) covers provider
setup in detail.

## Signature errors

### "Invalid arrow in signature" and other parsing errors

```{r}
#| eval: true
try(signature("question => answer"))
try(signature("question answer"))
try(signature("question -> reasoning -> answer"))
```

A signature has exactly one `->`, with inputs on the left and outputs on the
right, each separated by commas. To have the model reason before answering,
use `chain_of_thought()`, which adds the `reasoning` output for you:

```{r}
signature("question -> reasoning, answer")
chain_of_thought("question -> answer")
```

### "Unknown type"

```{r}
#| eval: true
try(signature("review -> score: integr"))
```

The types are `str`, `int`, `float`, `bool`, `enum(...)` or `Literal[...]`,
`Optional[...]`, `list[...]` and `dict[...]`. The
[Quick reference](cheatsheet.html#define-signatures) has the full grammar.

### Instructions or bounds are ignored

```{r}
#| eval: true
sig <- signature("question -> answer", instrctions = "Answer in one word.")
sig@instructions
```

`signature()` silently drops arguments it does not recognize, so the misspelled
`instrctions` never reaches the prompt and the default instruction is used.
Bounds are dropped the same way: `score: number[0, 100]` is an unbounded
number, so check ranges in R or with `with_assertions()`.

## Module errors

### "`module()` creates standard prediction modules and does not accept ..."

```{r}
#| eval: true
try(module(signature("question -> answer"), type = "react"))
try(module(signature("question -> answer"), temperature = 0))
```

`module()` builds one kind of module, a single prediction. The error has class
`dsprrr_module_argument_error`. The old `type` argument became separate
constructors:

| Instead of | Use |
|------------|-----|
| `module(sig, type = "react")` | `react(sig, tools = list(...))` |
| `module(sig, type = "chain_of_thought")` | `chain_of_thought(sig)` |
| `module(sig, type = "multichain")` | `multi_chain_comparison(sig)` |
| `module(sig, type = "program_of_thought")` | `program_of_thought(sig, interpreter_factory = r_code_runner)` |
| `module(sig, type = "codeact")` | `code_act(sig, tools = list(...), interpreter_factory = r_code_runner)` |
| `module(sig, type = "rlm")` | `rlm_module(sig, interpreter_factory = ...)` |
| `module(sig, type = "flex")` | `flex(sig)` |
| `module(sig, temperature = 0)` | `module(sig, config = list(temperature = 0))` |

### "First argument must be a Signature object"

```{r}
#| eval: true
try(module("question -> answer"))
```

`module()` needs a `Signature` object: `module(signature("question -> answer"))`.
The other constructors, such as `chain_of_thought()` and `react()`, accept the
string directly.

## Errors when calling `run()`

### "Missing required inputs"

```{r}
#| eval: true
context_qa <- module(signature("context, question -> answer"))
try(run(context_qa, question = "Who maintains ggplot2?"))
```

Pass every input the signature names. A misspelled input name ends up here
too; compare "Signature expects" with "You provided".

### "Unknown dot-prefixed argument"

```{r}
#| eval: true
try(run(qa, question = "What is the capital of France?", .temperature = 0))
try(run(qa, question = "What is the capital of France?", .parallel = TRUE))
```

Arguments that start with a dot configure the call. `run()` accepts `.llm`,
`.cache`, `.concurrency`, `.return_format`, `.progress`, `.verbose`,
`.show_prompt` and `.trace_context`; anything else is rejected rather than
sent to the model as an input. Set temperature on the chat
(`ellmer::params(temperature = 0)`) or the module
(`config = list(temperature = 0)`), and run batches in parallel with
`.concurrency = concurrency_control(max_active = 4L)`.

### "Extra input not declared in the signature"

```{r}
#| eval: true
try(run(qa, question = "What is the capital of France?", tone = "formal", .llm = chat))
```

This is a warning, not an error: the call goes on to the model (on this page
it then stops at the missing API key). The extra value is added to the prompt
and is available to a custom `template`, but demos, optimizers and type checks
ignore it. If the model should use the value, declare it in the signature:
`signature("question, tone -> answer")`.

### "Predict templates use single-brace placeholders"

```{r}
#| eval: true
templated <- module(
  signature("question -> answer"),
  template = "Answer briefly.\n\nQuestion: {{question}}"
)
try(run(templated, question = "What is the capital of France?", .llm = chat))
```

Templates are glue strings. Write each placeholder with one pair of braces:

```{r}
templated <- module(
  signature("question -> answer"),
  template = "Answer briefly.\n\nQuestion: {question}"
)
```

## Evaluation and optimization errors

### "`@k must be <integer>, not <double>`"

```{r}
#| eval: true
try(LabeledFewShot(k = 3))
```

Teleprompter settings are typed, and whole numbers need an `L`:
`LabeledFewShot(k = 3L)`, `BootstrapFewShot(max_bootstrapped_demos = 4L)`,
`GEPA(generations = 5L)`.

### "Can't tell which field this metric should compare"

```{r}
#| eval: true
keyword_baseline <- module_fn("text -> sentiment", function(text) {
  positive <- grepl("love|great", text, ignore.case = TRUE)
  list(sentiment = if (positive) "positive" else "negative")
})
reviews <- tibble::tibble(
  text = c("I love it", "Arrived broken", "Great value"),
  label = c("positive", "negative", "positive")
)
result <- evaluate(keyword_baseline, reviews, metric = metric_exact_match())
result$mean_score
result$n_metric_errors
```

`evaluate()` calls the metric with the module's output and the whole data row.
`metric_exact_match()` and `metric_f1()` compare one output field with the
data column of the same name. Without `field` they look for exactly one name
the output and the data share, and fail when there is none (here the column is
called `label`) or more than one. Failed rows count as 0 in `mean_score`, so a
score of 0 can mean the metric failed on every row; check `n_errors`. Rename
the column and name the field, or write a metric that reads both:

```{r}
#| eval: true
reviews$sentiment <- reviews$label
evaluate(
  keyword_baseline,
  reviews,
  metric = metric_exact_match(field = "sentiment")
)$mean_score

label_matches <- function(prediction, expected) {
  prediction$sentiment == expected$label
}
evaluate(keyword_baseline, reviews, metric = label_matches)$mean_score
```

### "Callable modules created with `module_fn()` do not support optimization"

```{r}
#| eval: true
try(compile(keyword_baseline, LabeledFewShot(k = 2L), reviews))
```

`compile()` improves prompts, and a `module_fn()` program has none. Use
`module_fn()` for baselines and evaluation, and compile a prediction module
such as `module(signature("text -> sentiment"))`.

## Provider errors

### "LLM call failed" and "Failed to process item"

How a provider error (a bad key, a rate limit, a timeout, a server error)
reaches you depends on how the module was called:

| Call | What you see |
|------|--------------|
| `run()` with one input | the provider's error, unchanged, as above |
| `run()` with a vector, `run_dataset()`, `evaluate()` | a warning "Failed to process item N: ..." for each failed row, and `NA` as its result |
| pipelines and wrappers such as `best_of_n()`, `refine()` and `ensemble()` | "LLM call failed: ...", inside the wrapper's own error |

```{r}
#| eval: true
questions <- tibble::tibble(question = c("What is 2 + 2?", "What is 3 + 3?"))
answers <- run_dataset(qa, questions, .llm = chat, .return_format = "structured")
answers$.error

two_steps <- qa %>>% module(signature("answer -> summary"))
try(run(two_steps, question = "What is 2 + 2?", .llm = chat))
```

With `.return_format = "structured"`, `run_dataset()` keeps each row's message
in `.error`. `evaluate()` counts failed calls in `n_run_errors`, lists them in
`run_errors`, and scores those rows 0. A `react()` agent reports "LLM call
failed in ReAct loop".

### Rate limits (HTTP 429)

ellmer retries requests that fail with a transient error, such as HTTP 429 or
503, or a dropped connection. It tries three times by default, waiting longer
after each failure. Allow more tries:

```{r}
options(ellmer_max_tries = 6)
```

dsprrr sends batch requests one at a time unless you pass `.concurrency`. If
you did, lower the number of requests in flight:

```{r}
run_dataset(qa, questions, .llm = chat, .concurrency = concurrency_control(max_active = 2L))
```

### Timeouts

ellmer waits up to 300 seconds for a response. Long outputs from reasoning
models can take longer; raise the limit:

```{r}
options(ellmer_timeout_s = 600)
```

To stop slow rows in a batch instead, set a per-row limit. Finite timeouts need
the mirai backend, which runs the module in worker processes, so attach the
chat to the module instead of passing `.llm`:

```{r}
timed_qa <- module(signature("question -> answer"), chat = chat)
run_dataset(
  timed_qa,
  questions,
  .concurrency = concurrency_control(backend = "mirai", max_active = 2L, task_timeout = 120)
)
```

## Cache problems

### The same answer every time ("Using cached LLM responses")

dsprrr caches responses in memory and on disk. A request identical to an
earlier one (same model, parameters, prompt and output type) returns the
stored answer, and the first cache hit in a session prints "Using cached LLM
responses". That is usually what you want, but not when you need a fresh sample
or are measuring how answers vary.

```{r}
run(qa, question = "Suggest a name for a cat.", .llm = chat, .cache = FALSE) # this call
clear_cache() # forget everything cached
configure_cache(enable = FALSE) # stop caching for this session
```

In CI, set the environment variable `DSPRRR_CACHE_ENABLED=false`. To sample
several different answers on purpose, use `best_of_n()`, which keeps its
attempts apart in the cache.

### "Disk caching is unavailable at ..."

```{r}
#| eval: true
shared_dir <- file.path(tempdir(), "shared-cache")
dir.create(shared_dir)
Sys.chmod(shared_dir, mode = "0755") # readable by other accounts
configure_cache(disk_path = shared_dir)
try(run(qa, question = "What is the capital of France?", .llm = chat))
```

The disk cache stores prompts and responses, so dsprrr uses it only when you
own the directory, the directory has mode `0700` and every file in it has mode
`0600`. Otherwise it warns once, with the reason and the `chmod` command that
fixes it, and keeps only a memory cache for the session. (The error after the
warning is the missing API key.) This happens with directories copied from
elsewhere, created under a permissive umask, or shared between accounts. On
Windows, dsprrr cannot check these permissions and relies on the ACLs of the
per-user cache directory.

Run the `chmod` command from the warning, or restrict the directory from R:

```{r}
cache_dir <- tools::R_user_dir("dsprrr", "cache") # or Sys.getenv("DSPRRR_CACHE_PATH")
Sys.chmod(cache_dir, mode = "0700")
Sys.chmod(list.files(cache_dir, full.names = TRUE), mode = "0600")
```

Alternatively, point `configure_cache(disk_path = )` at a new directory, which
dsprrr creates with private permissions, or turn the disk tier off with
`configure_cache(enable_disk = FALSE)`. For a cache that only trusted accounts
can write, `configure_cache(disk_private = FALSE)` skips the check.
`dsprrr_sitrep()` shows the result.

## The answer is wrong or empty

When a call succeeds but the answer is off, look at what the model received:

```{r}
get_last_prompt() # prompt, response, model, tokens and cost of the last call
inspect_history(n = 5) # a tibble of the last five calls
export_traces(qa, include_prompts = TRUE, include_outputs = TRUE) # every call of one module
```

Check that each input appears in the prompt, that your instructions made it in,
and that the output types are as strict as you meant. If you are still stuck,
open an issue at <https://github.com/JamesHWade/dsprrr/issues> with a minimal
example and the output of `dsprrr_sitrep()`.

```{r}
#| include: false
#| eval: true
options(saved_options)
restore_keys <- saved_keys[!is.na(saved_keys)]
if (length(restore_keys) > 0) do.call(Sys.setenv, as.list(restore_keys))
```
