---
title: "Tutorial 1: Your First LLM Call"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 1: Your First LLM Call}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

In this tutorial, you'll make your first structured LLM call with dsprrr. By the end, you'll be able to ask questions, get typed responses, and understand why signatures are powerful.

**Time**: 10-15 minutes

## What You'll Build

A working question-answering system that returns structured data—not just raw text.

## Prerequisites

- R installed
- An OpenAI API key (set as `OPENAI_API_KEY` environment variable)
- Install the packages:

```r
install.packages("pak")
pak::pak("JamesHWade/dsprrr")
pak::pak("tidyverse/ellmer")
```

## Step 1: Load the Packages

```{r setup}
library(dsprrr)
library(ellmer)
```

You should see no errors. If you do, check that your API key is set correctly.

## Step 2: Create a Chat Connection

Connect to OpenAI:

```{r create-chat}
chat <- chat_openai()
```

This creates a chat object you'll use for all your LLM calls.

## Step 3: Your First Structured Call

Define a typed module and run it:

```{r first-call}
#| cassette: true
qa <- module(signature("question -> answer"))
run(qa, question = "What is the capital of France?", .llm = chat)
```

You should see `"Paris"` returned. Let's break down what happened:

- `"question -> answer"` is a **signature**—it declares one input (`question`) and one output (`answer`)
- dsprrr handled all the prompt engineering for you
- The result came back as structured data, not raw text

## Step 4: Try Different Questions

The same signature works for any question:
```{r more-questions}
#| cassette: true
run(qa, question = "What is 7 * 8?", .llm = chat)

run(qa, question = "Who wrote Romeo and Juliet?", .llm = chat)
```

Notice how you get clean, direct answers—no extra prose.

## Step 5: Add Output Types

So far, answers have been strings. But what if you want a number or a specific choice?

### Getting a Number

```{r number-output}
#| cassette: true
math <- module(signature("math_problem -> result: number"))
run(math, math_problem = "What is 15% of 200?", .llm = chat)
```

The `: number` after `result` tells dsprrr you want a numeric value, not a string.

### Getting a Choice (Enum)

```{r enum-output}
#| cassette: true
sentiment <- module(
  signature("text -> sentiment: enum('positive', 'negative', 'neutral')")
)
run(sentiment, text = "I absolutely loved this movie!", .llm = chat)
```

The LLM must pick from exactly those three options. Try changing the text to see different results:

```{r enum-variations}
#| cassette: true
run(sentiment, text = "This was a complete waste of time.", .llm = chat)

run(sentiment, text = "It was okay, I guess.", .llm = chat)
```

### Getting True/False

```{r bool-output}
#| cassette: true
truth <- module(signature("statement -> is_true: bool"))
run(truth, statement = "The Earth orbits the Sun.", .llm = chat)

run(truth, statement = "Cats are larger than elephants.", .llm = chat)
```

## Step 6: Multiple Inputs

Signatures can have multiple inputs. Separate them with commas:

```{r multiple-inputs}
#| cassette: true
contextual_qa <- module(signature("context, question -> answer"))
run(
  contextual_qa,
  context = "R was created in 1993 by Ross Ihaka and Robert Gentleman at the University of Auckland.",
  question = "When was R created?",
  .llm = chat
)
```

Now the LLM uses your context to answer the question:

```{r context-example}
#| cassette: true
run(
  contextual_qa,
  context = "The bakery opens at 7am and closes at 6pm. They sell croissants for $3 each.",
  question = "How much do croissants cost?",
  .llm = chat
)
```

## Step 7: Adding Instructions

You can guide the LLM's behavior with instructions:

```{r with-instructions}
#| cassette: true
one_word <- module(
  signature("question -> answer", instructions = "Answer in exactly one word.")
)
run(one_word, question = "What color is the sky on a clear day?", .llm = chat)

pirate <- module(
  signature("question -> answer", instructions = "Answer like a pirate.")
)
run(pirate, question = "What is the capital of France?", .llm = chat)
```

## What You Learned

In this tutorial, you:

1. Built and ran your first typed module
2. Used signatures to declare inputs and outputs
3. Added output types: `string`, `number`, `bool`, `enum()`
4. Combined multiple inputs
5. Added instructions to guide behavior

## What's Different from Raw LLM Calls?

Without dsprrr, you'd write prompts like:
```
You are a helpful assistant. The user will ask a question.
Respond with just the answer, nothing else.
User: What is the capital of France?
```

With dsprrr, you just declare `"question -> answer"` and the framework handles the rest. This becomes powerful when you need to:

- Optimize prompts automatically
- Chain multiple LLM calls together
- Get consistent, typed outputs

## Next Steps

Ready to build something reusable? Continue to:

- **[Tutorial 2: Building a Classifier](tutorial-build-classifier.html)** — Create a module you can use repeatedly
- **[Quick Reference](cheatsheet.html)** — Look up signature syntax
- **[The DSPy Philosophy](concepts-dspy-philosophy.html)** — Understand *why* signatures work this way
