---
title: "Tutorial 2: Building a Reusable Classifier"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 2: Building a Reusable Classifier}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

In [Tutorial 1](tutorial-hello-world.html), you made one-off LLM calls with `dsp()`. But what if you need to classify hundreds of texts? Creating a new call each time is tedious and slow.

In this tutorial, you'll build a **reusable module**—a classifier you can use over and over.

**Time**: 20-25 minutes

## What You'll Build

A sentiment classifier that:
- Processes single texts or batches
- Remembers its configuration
- Can be saved and reused

## Prerequisites

- Completed [Tutorial 1](tutorial-hello-world.html)
- `OPENAI_API_KEY` set in your environment

```{r setup}
library(dsprrr)
library(ellmer)
```

## Step 1: The Problem with `dsp()`

In Tutorial 1, you classified sentiment like this:

```{r dsp-repeated, eval=FALSE}
chat <- chat_openai()
chat |> dsp("text -> sentiment: enum('positive', 'negative', 'neutral')", text = "Great!")
chat |> dsp("text -> sentiment: enum('positive', 'negative', 'neutral')", text = "Awful")
chat |> dsp("text -> sentiment: enum('positive', 'negative', 'neutral')", text = "Meh")
```

This works, but you're repeating the signature every time. If you want to change the signature, you have to change it everywhere.

## Step 2: Create a Reusable Module

The `as_module()` function wraps a signature into a reusable object:

```{r create-module}
chat <- chat_openai()

classifier <- chat |>
  as_module("text -> sentiment: enum('positive', 'negative', 'neutral')")

classifier
```

Now `classifier` is an object you can use repeatedly.

## Step 3: Classify Single Texts

Use the `$predict()` method to classify:

```{r predict-single}
#| cassette: true
classifier$predict(text = "I absolutely loved this movie!")
```

Try a few more:

```{r predict-more}
#| cassette: true
classifier$predict(text = "This was a complete waste of time.")

classifier$predict(text = "It was okay, I guess.")

classifier$predict(text = "The service was terrible but the food was amazing.")
```

## Step 4: Batch Processing

Here's where modules shine. Process multiple texts at once by passing a vector:

```{r batch-vector}
#| cassette: true
reviews <- c(
  "Best purchase I've ever made!",
  "Broke after one day. Total garbage.",
  "Does what it says. Nothing special.",
  "Exceeded all my expectations!",
  "Would not recommend to anyone."
)

classifier$predict(text = reviews)
```

All five classifications came back in a single call. Much more efficient than five separate calls.

## Step 5: The Full Control Approach

`as_module()` is convenient, but sometimes you need more control. The `signature()` + `module()` approach gives you that:

```{r full-control}
# Define the signature separately
sig <- signature(
  "text -> sentiment: enum('positive', 'negative', 'neutral')",
  instructions = "Classify the overall sentiment. If mixed, choose the dominant emotion."
)

sig
```

Now create a module from the signature:

```{r create-module-full}
classifier2 <- module(sig, type = "predict")

classifier2
```

## Step 6: Running with `run()`

With the full control approach, use `run()` to execute:

```{r run-single}
#| cassette: true
run(classifier2, text = "This is fantastic!", .llm = chat)
```

Notice you pass the chat object via `.llm`. This gives you flexibility—you can use different LLMs for different calls.

Batch processing works the same way:

```{r run-batch}
#| cassette: true
run(
  classifier2,
  text = c("Love it!", "Hate it!", "It's fine"),
  .llm = chat
)
```

## Step 7: Working with Data Frames

Real data often comes in data frames. Use `run_dataset()`:

```{r dataset-processing}
#| cassette: true
library(tibble)

reviews_df <- tibble(
  id = 1:4,
  text = c(
    "Absolutely wonderful experience!",
    "Never buying from them again.",
    "Solid product, fair price.",
    "Changed my life for the better."
  )
)

results <- run_dataset(classifier2, reviews_df, .llm = chat)
results
```

The results include your original columns plus the classification.

## Step 8: Adding Descriptions

Make your inputs more informative with descriptions:

```{r with-descriptions}
#| cassette: true
sig <- signature(
  inputs = list(
    input("review_text", description = "Customer review to classify")
  ),
  output_type = type_enum(values = c("positive", "negative", "neutral")),
  instructions = "Classify the customer sentiment."
)

detailed_classifier <- module(sig, type = "predict")

run(
  detailed_classifier,
  review_text = "Five stars! Would buy again!",
  .llm = chat
)
```

Descriptions help the LLM understand what it's working with.

## Step 9: Checking Your Work

Modules track their calls. See what happened:

```{r trace-summary}
#| eval: false
classifier2$trace_summary()
```

This shows you how many calls were made and the token costs.

## What You Learned

In this tutorial, you:

1. Created reusable modules with `as_module()`
2. Used `$predict()` for single and batch processing
3. Built modules with full control using `signature()` + `module()`
4. Processed data frames with `run_dataset()`
5. Added input descriptions for clarity
6. Checked your work with `trace_summary()`

## When to Use Each Approach

| Approach | Best For |
|----------|----------|
| `dsp()` | Quick one-off calls, exploration |
| `as_module()` | Simple reusable modules, prototyping |
| `signature()` + `module()` | Production code, optimization workflows |

## The Module Advantage

Why bother with modules when `dsp()` works?

1. **Reusability**: Define once, use everywhere
2. **Efficiency**: Batch processing reduces API calls
3. **Configuration**: Change settings in one place
4. **Optimization**: Modules can be improved with training data (covered in [Tutorial 4](tutorial-improve-with-demos.html))
5. **Tracing**: Track what happened for debugging

## Next Steps

Your classifier works, but can it handle more complex outputs? Continue to:

- **[Tutorial 3: Extracting Structured Data](tutorial-structured-outputs.html)** — Get multiple fields and nested structures
- **[Quick Reference](cheatsheet.html)** — Module types and methods
- **[Understanding Signatures & Modules](concepts-signatures-modules.html)** — Why S7 for signatures, R6 for modules
