---
title: "Tutorial 4: Improve with examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 4: Improve with examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

A module can return the right shape and still pick the wrong answer. By the end
of this tutorial you can measure how often a support-ticket classifier is
right, add worked examples (demos) to its prompt by hand or with
`LabeledFewShot`, and check with `evaluate()` whether they helped.

```{r setup}
#| eval: true
library(dsprrr)
library(ellmer)

chat <- chat_openai(model = "gpt-6-luna")
```

The answers on this page were recorded with `gpt-5-mini` and are replayed when the
site is built, so `gpt-6-luna` may word them differently.

## A ticket classifier

The classifier sorts support tickets into four categories:

```{r baseline-classifier}
#| eval: true
sig <- signature(
  "ticket -> category: enum('billing', 'technical', 'shipping', 'general')",
  instructions = "Classify the customer support ticket."
)

classifier <- module(sig)
```

```{r baseline-test}
#| cassette: true
run(classifier, ticket = "My package hasn't arrived yet", .llm = chat)

run(classifier, ticket = "I was charged twice for my order", .llm = chat)

run(classifier, ticket = "The app keeps crashing when I try to login", .llm = chat)
```

Three right answers on tickets you picked yourself say little about accuracy.
For that you need tickets with known answers.

## Measure a baseline

Label some tickets with the category you expect. Treat them as a test set:
they are for scoring the classifier, not for showing to it.

```{r tickets}
#| eval: true
tickets <- tibble::tribble(
  ~ticket,                                  ~category,
  "I was charged twice for the same item",  "billing",
  "How do I update my payment method?",     "billing",
  "The website won't load on my phone",     "technical",
  "My password reset email never arrived",  "technical",
  "When will my order ship?",               "shipping",
  "Can I change my delivery address?",      "shipping",
  "The product arrived damaged",            "shipping",
  "I need a refund for my subscription",    "billing",
  "How do I contact customer service?",     "general",
  "What are your business hours?",          "general"
)
```

Run the classifier on every ticket and compare its answers with the labels.
`run_dataset()` returns your columns plus a `result` column holding the
module's answer for each row. Take the predictions from `result`: the
`category` column in the output is your label, copied from the input.

```{r baseline-accuracy}
#| cassette: true
baseline_results <- run_dataset(classifier, tickets, .llm = chat)

predicted <- vapply(baseline_results$result, \(r) r$category, character(1))
mean(predicted == tickets$category)

# The tickets it got wrong
tickets[predicted != tickets$category, ]
```

In the recorded run, gpt-5-mini labels all ten tickets correctly, so the list
of misses is empty.
That is the check to make before adding demos. Demos lengthen every prompt, and
they can only help on tickets the model gets wrong, so with this baseline there
is nothing for them to fix. When your baseline does miss tickets, read the
misses: they tell you which kinds of ticket need an example.

The rest of this tutorial adds demos anyway, to show how, and measures them on
the same ten tickets.

## Add demos by hand

A demo is a worked example: an input and the output you want for it. Pass a
list of them to `module()`:

```{r manual-demos}
#| eval: true
classifier_with_demos <- module(
  sig,
  demos = list(
    list(
      inputs = list(ticket = "I was charged twice for the same item"),
      output = list(category = "billing")
    ),
    list(
      inputs = list(ticket = "The app crashes on startup"),
      output = list(category = "technical")
    ),
    list(
      inputs = list(ticket = "My package is late"),
      output = list(category = "shipping")
    )
  )
)
```

```{r test-with-demos}
#| cassette: true
run(classifier_with_demos, ticket = "I need a receipt for my purchase", .llm = chat)

run(classifier_with_demos, ticket = "The button doesn't respond when clicked", .llm = chat)
```

dsprrr puts the demos in the prompt, ahead of the new ticket. This is the
prompt for the second call:

```{r demo-prompt}
cat(get_last_prompt()$prompt)
```

## Sample demos with `LabeledFewShot`

`LabeledFewShot` builds demos from the rows of a labeled data frame instead.
dsprrr calls objects like this teleprompters: `compile()` applies one to a
module and returns a new module, leaving the original unchanged. Here the rows
come from `tickets`, which is a shortcut with a cost, explained below.

```{r labeled-fewshot}
#| eval: true
compiled <- classifier |> compile(LabeledFewShot(k = 3L), tickets)

best_demos(compiled, as_tibble = TRUE)
```

`LabeledFewShot` does not look for good examples. It samples `k` rows at
random, with a fixed seed (`seed = 123L` by default) so that you get the same
rows each time. This sample has a technical, a general and a billing ticket,
and no shipping ticket, so check what it picked before you rely on it.

## Compare the versions

`evaluate()` runs a module on every row of a data frame and scores each answer
with a metric. `metric_exact_match(field = "category")` scores 1 when the
returned `category` equals the row's `category` column and 0 otherwise, so the
mean score is accuracy:

```{r final-comparison}
#| cassette: true
metric <- metric_exact_match(field = "category")

versions <- list(
  "no demos" = classifier,
  "hand-written demos" = classifier_with_demos,
  "LabeledFewShot demos" = compiled
)

vapply(
  versions,
  \(m) evaluate(m, tickets, metric = metric, .llm = chat)$mean_score,
  numeric(1)
)
```

All three versions score 1. On this task, the recorded model does as well without demos
as with them.

The comparison also favors the demo versions, because some of the scored
tickets sit in their prompts with the answers attached: the first hand-written
demo is ticket 1 word for word, and `LabeledFewShot` drew tickets 2, 3 and 10.
On your own data, draw demos from a training set and score every version on
tickets outside it. Make that test set large enough to trust, too: with ten
tickets, one answer moves accuracy by 10 points.

Every demo adds tokens to every call. If you try several values of `k`, keep
the smallest one that scores as well as the larger ones.

Next, [Tutorial 5](tutorial-optimize-your-module.html) splits the data into
training and test sets and lets dsprrr search over instructions and demos.
