---
title: "Tutorial 6: Save and reuse a module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 6: Save and reuse a module}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

Compiling a module, as in Tutorials 4 and 5, produces demos and instructions
that you do not want to recompute in every R session. By the end of this
tutorial you can save a compiled module's configuration to a
[pins](https://pins.rstudio.com) board, restore it as a working module, and
export the traces the module records as it runs.

```{r setup}
library(dsprrr)
library(ellmer)
library(pins)

chat <- chat_openai(model = "gpt-6-luna")
```

The answers on this page were recorded with `gpt-4.1` and are replayed when the
site is built, so `gpt-6-luna` may word them differently.

## Compile a module

This is the sentiment classifier from Tutorial 5, compiled with
`LabeledFewShot` on nine short reviews:

```{r create-optimized}
#| cassette: true
sig <- signature(
  "review -> sentiment: enum('positive', 'negative', 'neutral')",
  instructions = "Classify the sentiment of this product review."
)

trainset <- tibble::tribble(
  ~review,            ~sentiment,
  "Love it!",         "positive",
  "Hate it!",         "negative",
  "It's okay.",       "neutral",
  "Amazing!",         "positive",
  "Terrible!",        "negative",
  "Meh.",             "neutral",
  "Best ever!",       "positive",
  "Worst purchase!",  "negative",
  "Fine I guess.",    "neutral"
)

classifier <- module(sig) |> compile(LabeledFewShot(k = 3L), trainset)

run(classifier, review = "This product is fantastic!", .llm = chat)
```

## Save the configuration

`pin_module_config()` writes the module's signature, demos, settings and
optimization results to a pins board. The board here is a temporary folder,
which R deletes when the session ends:

```{r save-config}
board <- board_temp()

pin_module_config(board, "sentiment-classifier", classifier)
```

For real use, pick a board that outlives the session:

```{r persistent-board}
#| eval: false
board <- board_local() # a folder in your user data directory
board <- board_folder("pins") # or a folder in your project
```

Shared boards such as `board_connect()`, `board_s3()` and `board_gcs()` work the
same way and let other people load your module. Each call to
`pin_module_config()` with the same name saves a new version:
`pin_versions(board, "sentiment-classifier")` lists them, and
`pin_read(board, "sentiment-classifier", version = )` reads an older one.

## Restore it

In a new session, connect to the same board, read the pin and rebuild the
module:

```{r restore-config}
config <- pin_read(board, "sentiment-classifier")
restored <- restore_module_config(config)

identical(restored$demos, classifier$demos)
```

The chat is not part of the saved configuration, so pass `.llm` when you run the
restored module:

```{r run-restored}
#| eval: false
run(restored, review = "Worst product ever!", .llm = chat)
```

`restore_module_config()` reads only the artifact format of the installed
dsprrr (version 6 in the message above), so keep the code that compiled the
module: if a dsprrr upgrade changes the format, compile and pin again. A module
that holds R functions or tools also needs a `registry` when you save and
restore it; see `pin_module_config()`. To use a saved module in a targets
pipeline or a Quarto report, see [Run dsprrr in pipelines](orchestration.html).

## Keep a reviewed version

A pin keeps every version you save. To also record which version a person
approved for production, and what it was compiled from, keep the program in
[graft](https://jameshwade.github.io/graft/). Save the whole program with
`save_program()`, store it with its training examples as evidence, and accept
it for a purpose:

```{r graft-promote}
#| eval: false
store <- graft::graft_store("programs", create = TRUE)
id <- program_artifact_id(classifier)
path <- save_program(classifier, tempfile(fileext = ".rds"))

examples <- graft::graft_save(
  store,
  as.character(jsonlite::toJSON(trainset)),
  id = "sentiment:trainset",
  media_type = "application/json"
)
program <- graft::graft_save_file(
  store,
  path,
  id = "sentiment:program",
  dependencies = examples
)
graft::graft_accept(
  store,
  program,
  stream = "sentiment-classifier",
  expected = NULL,
  key = id,
  actor = "reviewer",
  reason = "Checked its evaluation on held-out reviews",
  purpose = "production"
)
```

In production, load whichever program is accepted now:

```{r graft-load}
#| eval: false
recall <- graft::graft_recall(
  store,
  stream = "sentiment-classifier",
  purpose = "production",
  eligible = TRUE
)
path <- tempfile(fileext = ".rds")
writeBin(recall@roots[[1]]@bytes, path)
production <- load_program(path)

identical(program_artifact_id(production), id)
```

The ID from `program_artifact_id()` is also recorded in every call's metadata
and trace, so a trace points to the exact program that was reviewed. graft
keeps each acceptance, and any later withdrawal, in the stream's history.

## Save and inspect traces

A module records a trace for each call it makes, with token counts, latency,
cost and model. Run the module a few times, then pin its traces next to the
configuration:

```{r save-traces}
#| cassette: true
run(classifier, review = "Great product!", .llm = chat)
run(classifier, review = "Not worth the money.", .llm = chat)
run(classifier, review = "Does the job.", .llm = chat)

pin_trace(board, "sentiment-traces", classifier)
```

`export_traces()` returns the same data as a tibble with one row per call,
including the call made right after compiling:

```{r analyze-traces}
export_traces(classifier)
```

The tibble leaves out prompts and responses, so you can share it without
exposing the text your module handled. Add them with `include_prompts = TRUE`
and `include_outputs = TRUE`; `pin_trace()` takes the same two arguments.
`summarize_traces(classifier)` totals the tokens, cost and latency.

That completes the tutorials; [Compile and optimize](compilation-optimization.html)
and [Choose an optimizer](advanced-optimization.html) take optimization further.
