---
title: "Tutorial 5: Optimize a module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 5: Optimize a module}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

[Tutorial 4](tutorial-improve-with-demos.html) compared versions of a module
by hand. dsprrr can run that comparison for you: give it candidate
configurations, labeled data and a metric, and it scores every candidate and
keeps the best. By the end of this tutorial you can search instruction variants
with `optimize_grid()`, search instructions and demos together with
`GridSearchTeleprompter`, and check the result on reviews the search never saw.

```{r setup}
#| eval: true
library(dsprrr)
library(ellmer)

chat <- chat_openai(model = "gpt-5-mini")
```

## A classifier and two datasets

The task is product-review sentiment:

```{r setup-problem}
#| eval: true
sig <- signature(
  "review -> sentiment: enum('positive', 'negative', 'neutral')",
  instructions = "Classify the sentiment of this product review."
)
```

The search scores candidates on `trainset`. `testset` stays out of the search,
so the final check measures how the chosen configuration handles new reviews
rather than how well it fits the training ones:

```{r create-data}
#| eval: true
trainset <- tibble::tribble(
  ~review,                                              ~sentiment,
  "Absolutely love this product! Best purchase ever.",  "positive",
  "Complete waste of money. Broke after one day.",      "negative",
  "It's okay. Does what it says.",                      "neutral",
  "Exceeded all my expectations!",                      "positive",
  "Terrible quality. Very disappointed.",               "negative",
  "Nothing special, but it works.",                     "neutral",
  "Amazing! Would buy again.",                          "positive",
  "Don't bother. Total junk.",                          "negative",
  "Decent for the price.",                              "neutral",
  "Fantastic quality and fast shipping!",               "positive"
)

testset <- tibble::tribble(
  ~review,                                     ~sentiment,
  "Great value for money!",                    "positive",
  "Stopped working after a week.",             "negative",
  "Average product, average price.",           "neutral",
  "Couldn't be happier with this purchase!",   "positive"
)
```

## Measure the baseline

Score the module as it is before searching, so you have a number to beat:

```{r baseline-comparison}
#| cassette: true
baseline <- module(sig)

baseline_results <- evaluate(
  baseline,
  testset,
  metric = metric_exact_match(field = "sentiment"),
  .llm = chat
)

baseline_results$mean_score
baseline_results$scores
```

gpt-5-mini labels all four test reviews correctly, so here a search can at best
match the baseline. Four reviews are too few to rank configurations anyway:
each is worth 25 percentage points, and one changed answer can reverse a
comparison. The datasets in this tutorial are small to keep it cheap; when the
choice matters, test on dozens to hundreds of labeled rows.

## Search instruction variants

`optimize_grid()` runs every candidate on `data`, scores it with `metric`, and
leaves the module set to the best candidate. `parameters` lists the values to
try. Each `instructions_suffix` is appended to the signature's instructions, so
this grid compares the original instructions with two longer versions:

```{r grid-instructions}
#| eval: false
classifier <- module(sig)

optimize_grid(
  classifier,
  data = trainset,
  metric = metric_exact_match(field = "sentiment"),
  parameters = list(
    instructions_suffix = c(
      "",
      "Consider the overall tone.",
      "Focus on how satisfied the customer is."
    )
  ),
  .llm = chat
)

module_trials(classifier)
```

Each candidate runs on all ten training reviews, 30 calls in total.
`optimize_grid()` changes `classifier` in place. `module_trials()` returns a
one-row summary with the best score, the best parameters and a `trials`
list-column; `optimization_result(classifier)$trials` has one row per
candidate, with its score and cost.

When `parameters` names several settings, `optimize_grid()` tries every
combination, so the number of calls is the product of the numbers of values,
times the number of rows. Settings you can search for a Predict module include:

| Setting | What it changes |
|---|---|
| `instructions` | Replaces the signature's instructions. |
| `instructions_suffix` | Appends text to the signature's instructions. |
| `template` | The glue string that lays out the inputs, such as `"Review: {review}"`. |
| `temperature`, `top_p` | Sampling settings sent to the model. Reasoning models such as gpt-5-mini do not support them. |

## Search instructions and demos together

`GridSearchTeleprompter` combines an instruction search with the demos of
Tutorial 4: it attaches `k` demos sampled from the training data, scores each
instruction variant with those demos in place, and keeps the best variant. Like
other teleprompters it works through `compile()`, which returns a new module
and leaves the one you pass in unchanged:

```{r grid-teleprompter}
#| eval: false
variants <- tibble::tibble(
  id = c("concise", "analytical", "empathetic"),
  instructions_suffix = c(
    "Respond with just the sentiment.",
    "Analyze the language carefully before deciding.",
    "Consider how the customer is feeling."
  )
)

tp <- GridSearchTeleprompter(
  variants = variants,
  metric = metric_exact_match(field = "sentiment"),
  k = 2L
)

set.seed(2026)
optimized <- compile(module(sig), tp, trainset, .llm = chat)

optimization_result(optimized)$best_params$id
```

Unless you pass `valset` to `compile()`, `GridSearchTeleprompter` scores the
variants on a random 20% of `trainset` (two reviews here) and draws the demos
from the rest. `set.seed()` makes that split repeatable. Two reviews cannot
tell three variants apart, so give it a larger `valset` when you have the data.

## Check the result on the test set

```{r test-evaluation}
#| eval: false
test_results <- evaluate(
  optimized,
  testset,
  metric = metric_exact_match(field = "sentiment"),
  .llm = chat
)

test_results$mean_score
```

Compare `test_results$mean_score` with `baseline_results$mean_score`. A
configuration that wins on the training reviews but not on the test set has
fitted those particular reviews rather than the task.

Each candidate costs a full pass over its data: the instruction grid above
makes 30 calls and the teleprompter 6. The `total_cost` column of
`optimization_result(classifier)$trials` records what each candidate cost.

Next, [Tutorial 6](tutorial-deploy-to-production.html) saves a compiled module
so that a later session can use it without repeating the search.
