---
title: "Tutorial 6: Taking to Production"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 6: Taking to Production}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::setup_knitr()
}
```

You've built and optimized a module through Tutorials 1-5. Now what? You need to:
- Save the optimized configuration
- Load it in production
- Track what happens in deployment
- Monitor for issues

This tutorial shows you how.

**Time**: 30-35 minutes

## What You'll Build

A production-ready module with:
- Persistent configuration (survives R restarts)
- Execution traces for debugging
- Validation before deployment

## Prerequisites

- Completed [Tutorial 5](tutorial-optimize-your-module.html)
- `OPENAI_API_KEY` set in your environment

```{r setup}
library(dsprrr)
library(ellmer)
library(pins)
library(tibble)

chat <- chat_openai()
```

## Step 1: Create an Optimized Module

Let's start with a trained classifier:

```{r create-optimized}
sig <- signature(
  "review -> sentiment: enum('positive', 'negative', 'neutral')",
  instructions = "Classify the sentiment of this product review."
)

trainset <- dsp_trainset(
  review = c(
    "Love it!", "Hate it!", "It's okay.",
    "Amazing!", "Terrible!", "Meh.",
    "Best ever!", "Worst purchase!", "Fine I guess."
  ),
  sentiment = c(
    "positive", "negative", "neutral",
    "positive", "negative", "neutral",
    "positive", "negative", "neutral"
  )
)

# Optimize
classifier <- compile_module(
  program = module(sig, type = "predict"),
  teleprompter = LabeledFewShot(k = 3),
  trainset = trainset
)

# Verify it works
run(classifier, review = "This product is fantastic!", .llm = chat)
```

## Step 2: Save Configuration with Pins

The `pins` package provides persistent storage. Save your module configuration:

```{r save-config}
# Create a local board (folder on disk)
board <- board_folder(tempdir())

# Save the module configuration
pin_module_config(classifier, board, "sentiment-classifier")
```

Your optimized configuration—including demos, parameters, and instructions—is now saved.

## Step 3: List Saved Modules

See what's stored:

```{r list-pins}
board |> pin_list()

# Get metadata
board |> pin_meta("sentiment-classifier")
```

## Step 4: Restore in a New Session

Imagine you restart R or deploy to a different machine:

```{r restore-config}
# Later, in a new session...
restored <- restore_module_config(board, "sentiment-classifier")

# It works immediately
run(restored, review = "Worst product ever!", .llm = chat)
```

The restored module has all the optimization work preserved.

## Step 5: Version Your Modules

Pins automatically version your saves:

```{r versioning}
# Make some changes
improved <- compile_module(
  program = restored,
  teleprompter = LabeledFewShot(k = 4),  # Try more examples
  trainset = trainset
)

# Save again - creates new version
pin_module_config(improved, board, "sentiment-classifier")

# List versions
board |> pin_versions("sentiment-classifier")
```

## Step 6: Roll Back to Previous Version

If a new version performs worse:

```{r rollback}
# Get specific version
versions <- board |> pin_versions("sentiment-classifier")

# Restore the first version
if (nrow(versions) > 1) {
  original <- restore_module_config(
    board,
    "sentiment-classifier",
    version = versions$version[1]
  )
}
```

## Step 7: Save Execution Traces

Track what your module does in production:

```{r save-traces}
# Run some predictions
run(classifier, review = "Great product!", .llm = chat)
run(classifier, review = "Not worth the money.", .llm = chat)
run(classifier, review = "Does the job.", .llm = chat)

# Save traces for analysis
pin_trace(classifier, board, "sentiment-traces")
```

## Step 8: Analyze Traces

Load and examine traces:

```{r analyze-traces}
# Export as tibble
traces <- export_traces(classifier, format = "tibble")
traces

# Summary statistics
classifier$trace_summary()
```

Traces include:
- Input/output for each call
- Token usage
- Latency
- Errors (if any)

## Step 9: Validate Before Deployment

Use `validate_workflow()` to check everything is ready:
```{r validate}
# Check module is properly configured
validation <- validate_workflow(
  module = classifier,
  board = board,
  pin_name = "sentiment-classifier"
)

validation
```

This checks:
- Module has been optimized
- Configuration is saved
- Signature is valid

## Step 10: Production Patterns

Here's a complete production workflow:

```{r production-workflow, eval=FALSE}
# === DEVELOPMENT ===
# 1. Build and optimize
dev_module <- module(sig, type = "predict")
dev_module$optimize_grid(
  devset = trainset,
  metric = metric_exact_match(field = "sentiment"),
  parameters = list(temperature = c(0, 0.3, 0.7))
)

# 2. Compile with best settings + demos
optimized <- compile_module(
  program = dev_module,
  teleprompter = LabeledFewShot(k = 3),
  trainset = trainset
)

# 3. Evaluate on test data
evaluate(optimized, testset, metric = metric_exact_match(field = "sentiment"))

# 4. Save if good enough
prod_board <- board_s3("my-bucket")  # Or board_connect(), board_folder()
pin_module_config(optimized, prod_board, "sentiment-v1")

# === PRODUCTION ===
# 1. Load the saved configuration
prod_module <- restore_module_config(prod_board, "sentiment-v1")

# 2. Use it
result <- run(prod_module, review = customer_review, .llm = chat_openai())

# 3. Periodically save traces for monitoring
pin_trace(prod_module, prod_board, "sentiment-traces")
```

## Step 11: Different Storage Backends

Pins supports multiple backends:

```{r backends, eval=FALSE}
# Local folder
board_folder("path/to/folder")

# Posit Connect
board_connect()

# AWS S3
board_s3("bucket-name")

# Azure
board_azure("container-name")

# Google Cloud
board_gcs("bucket-name")
```

Choose based on your deployment environment.

## Step 12: Monitoring in Production

Set up regular checks:

```{r monitoring, eval=FALSE}
# Daily: Check trace summary
module$trace_summary()

# Weekly: Evaluate on held-out samples
evaluate(module, weekly_sample, metric = metric_exact_match())

# Monthly: Compare to baseline
# If accuracy drops, investigate or retrain
```

## What You Learned

In this tutorial, you:

1. Saved module configurations with `pin_module_config()`
2. Restored modules with `restore_module_config()`
3. Used versioning for safe updates
4. Rolled back to previous versions
5. Saved and analyzed execution traces
6. Validated modules before deployment
7. Learned production workflow patterns

## The Production Checklist

Before deploying:

- [ ] Module is optimized on representative data
- [ ] Evaluated on held-out test data
- [ ] Configuration saved to persistent storage
- [ ] Versioning enabled for rollback
- [ ] Trace logging configured
- [ ] Monitoring plan in place

## Where to Go From Here

Congratulations! You've completed the core tutorial sequence. You now know how to:
- Make structured LLM calls
- Build reusable modules
- Extract complex data
- Improve with examples
- Optimize parameters
- Deploy to production

### Advanced Tutorials

Build complete applications:
- **[Text Adventure Game](text-adventure.html)** — Interactive AI with state management
- **[Generate llms.txt](llms-txt.html)** — Multi-stage documentation pipeline

### How-To Guides

Solve specific problems:
- **[Build RAG Pipelines](rag-workflows.html)** — Retrieval-augmented generation
- **[Evaluate with Vitals](vitals-integration.html)** — Rigorous evaluation
- **[Production Orchestration](orchestration.html)** — targets and Quarto integration

### Concepts

Understand the "why":
- **[The DSPy Philosophy](concepts-dspy-philosophy.html)** — Programs, not prompts
- **[How Optimization Works](concepts-optimization-theory.html)** — Teleprompter theory
