---
title: "Tutorial 2: Build a classifier"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 2: Build a classifier}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

You will build a sentiment classifier and run it three ways: on one text, on a
vector of texts, and on a data frame with `run_dataset()`. It uses the
signature syntax from [Tutorial 1](tutorial-hello-world.html).

```{r setup}
#| eval: true
#| message: false
library(dsprrr)
library(ellmer)

chat <- chat_openai(model = "gpt-4.1")
```

## Define the classifier

Start from the sentiment signature in Tutorial 1 and print the module:

```{r create-module}
#| eval: true
classifier <- module(
  signature("text -> sentiment: enum('positive', 'negative', 'neutral')")
)

classifier
```

The printout shows the inputs, the output type and the instructions that start
every prompt. This signature has no instructions of its own, so dsprrr wrote a
default from the field names.

## Classify one text

```{r predict-single}
#| cassette: true
run(classifier, text = "I absolutely loved this movie!", .llm = chat)
```

```{r predict-more}
#| cassette: true
run(classifier, text = "This was a complete waste of time.", .llm = chat)

run(classifier, text = "It was okay, I guess.", .llm = chat)

run(
  classifier,
  text = "The service was terrible but the food was amazing.",
  .llm = chat
)
```

The mixed review came back `neutral`. Whether that is right depends on what
you need the labels for; instructions, below, let you decide.

## Classify a vector of texts

Pass a vector and `run()` returns a list with one result per element, in the
same order. Each result is the named list you saw above, so
`purrr::map_chr()` can pull out the labels:

```{r batch-vector}
#| cassette: true
reviews <- c(
  "Best purchase I've ever made!",
  "Broke after one day. Total garbage.",
  "Does what it says. Nothing special.",
  "Exceeded all my expectations!",
  "Would not recommend to anyone."
)

sentiments <- run(classifier, text = reviews, .llm = chat)
purrr::map_chr(sentiments, "sentiment")
```

Each element is still its own request to the model, and by default they are
sent one after another. `concurrency_control()` lets several run at once.

## Add instructions

Instructions replace the default line. Use them to settle cases the labels
leave open, such as mixed reviews:

```{r add-instructions}
sig <- signature(
  "text -> sentiment: enum('positive', 'negative', 'neutral')",
  instructions = "Classify the overall sentiment. If mixed, choose the dominant emotion."
)

classifier2 <- module(sig)
```

The new module takes the same inputs, one at a time or as a vector:

```{r run-single}
#| cassette: true
run(classifier2, text = "This is fantastic!", .llm = chat)
```

```{r run-batch}
#| cassette: true
run(
  classifier2,
  text = c("Love it!", "Hate it!", "It's fine"),
  .llm = chat
)
```

This is the full return value of a vectorized call: a list of named lists.

## Classify a data frame

`run_dataset()` takes a data frame with a column for each input. It returns the
data frame with a `result` list-column added, and other columns, like `id`,
come through unchanged:

```{r dataset-processing}
#| cassette: true
reviews_df <- tibble::tibble(
  id = 1:4,
  text = c(
    "Absolutely wonderful experience!",
    "Never buying from them again.",
    "Solid product, fair price.",
    "Changed my life for the better."
  )
)

results <- run_dataset(classifier2, reviews_df, .llm = chat)
results$sentiment <- purrr::map_chr(results$result, "sentiment")
results
```

Like `run()`, `run_dataset()` sends one request per row. The module records
every call, and `summarize_traces(classifier2)` reports how many requests it
made and the tokens they used.

## Describe the inputs

A signature can also be built from `input()` objects and an ellmer type. An
input's description becomes its heading in the prompt, in place of the
input's name:

```{r with-descriptions}
#| cassette: true
sig <- signature(
  inputs = list(
    input("review_text", description = "Customer review to classify")
  ),
  output_type = type_enum(values = c("positive", "negative", "neutral")),
  instructions = "Classify the customer sentiment."
)

detailed_classifier <- module(sig)

run(
  detailed_classifier,
  review_text = "Five stars! Would buy again!",
  .llm = chat
)
```

The output type here is a bare `type_enum()` rather than a named field, so
`run()` returns the label itself instead of a list.

Next, [Tutorial 3](tutorial-structured-outputs.html) returns several fields
at once, including lists and nested records.
