---
title: "Tutorial 4: Improving with Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 4: Improving with Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

In [Tutorial 3](tutorial-structured-outputs.html), you built extractors that return structured data. But how do you make them more accurate? The answer: **show the LLM examples of correct behavior**.

This technique—called few-shot learning—is one of the most powerful ways to improve LLM performance.

**Time**: 25-30 minutes

## What You'll Build

A classifier that learns from examples, with measurable accuracy improvement.

## Prerequisites

- Completed [Tutorial 3](tutorial-structured-outputs.html)
- `OPENAI_API_KEY` set in your environment

```{r setup}
library(dsprrr)
library(ellmer)
library(tibble)

chat <- chat_openai()
```

## Step 1: The Problem

Let's build a customer support ticket classifier:

```{r baseline-classifier}
sig <- signature(
 "ticket -> category: enum('billing', 'technical', 'shipping', 'general')",
  instructions = "Classify the customer support ticket."
)

classifier <- module(sig, type = "predict")
```

Test it:

```{r baseline-test}
#| cassette: true
run(classifier, ticket = "My package hasn't arrived yet", .llm = chat)

run(classifier, ticket = "I was charged twice for my order", .llm = chat)

run(classifier, ticket = "The app keeps crashing when I try to login", .llm = chat)
```

This works, but how do we know it's accurate? And how do we improve it?

## Step 2: Create Training Data

First, let's create labeled examples using `dsp_trainset()`:

```{r training-data}
trainset <- dsp_trainset(
  ticket = c(
    "I was charged twice for the same item",
    "How do I update my payment method?",
    "The website won't load on my phone",
    "My password reset email never arrived",
    "When will my order ship?",
    "Can I change my delivery address?",
    "The product arrived damaged",
    "I need a refund for my subscription",
    "How do I contact customer service?",
    "What are your business hours?"
  ),
  category = c(
    "billing",
    "billing",
    "technical",
    "technical",
    "shipping",
    "shipping",
    "shipping",
    "billing",
    "general",
    "general"
  )
)

trainset
```

The `dsp_trainset()` function creates a properly formatted tibble where input columns are matched with expected output columns.

## Step 3: Measure Baseline Accuracy

Before improving, let's measure current performance:

```{r baseline-accuracy}
#| cassette: true
baseline_results <- run_dataset(classifier, trainset, .llm = chat)
baseline_results
```

Calculate accuracy:

```{r calculate-accuracy}
correct <- sum(baseline_results$category == trainset$category)
total <- nrow(trainset)
baseline_accuracy <- correct / total

cat("Baseline accuracy:", scales::percent(baseline_accuracy), "\n")
cat("Correct:", correct, "out of", total, "\n")
```

## Step 4: Add Manual Demonstrations

The simplest way to improve: show the LLM examples. Add them via the `demos` parameter:

```{r manual-demos}
classifier_with_demos <- module(
  sig,
  type = "predict",
  demos = list(
    list(
      inputs = list(ticket = "I was charged twice for the same item"),
      output = list(category = "billing")
    ),
    list(
      inputs = list(ticket = "The app crashes on startup"),
      output = list(category = "technical")
    ),
    list(
      inputs = list(ticket = "My package is late"),
      output = list(category = "shipping")
    )
  )
)
```

Each demo has `inputs` (a list of input values) and `output` (a list of expected outputs).

Test the improved classifier:

```{r test-with-demos}
#| cassette: true
run(classifier_with_demos, ticket = "I need a receipt for my purchase", .llm = chat)

run(classifier_with_demos, ticket = "The button doesn't respond when clicked", .llm = chat)
```

## Step 5: Measure Improvement

Run on the same test set:

```{r improved-accuracy}
#| cassette: true
improved_results <- run_dataset(classifier_with_demos, trainset, .llm = chat)

correct_improved <- sum(improved_results$category == trainset$category)
improved_accuracy <- correct_improved / total

cat("Baseline accuracy:", scales::percent(baseline_accuracy), "\n")
cat("Improved accuracy:", scales::percent(improved_accuracy), "\n")
cat("Improvement:", scales::percent(improved_accuracy - baseline_accuracy), "\n")
```

## Step 6: Automatic Demo Selection with LabeledFewShot

Manually picking demos is tedious. The `LabeledFewShot` teleprompter automatically selects good examples from your training data:

```{r labeled-fewshot}
# Create a teleprompter that selects 3 examples
teleprompter <- LabeledFewShot(k = 3L)

# Compile the module with the teleprompter
compiled <- compile_module(
  program = classifier,
  teleprompter = teleprompter,
  trainset = trainset
)
```

The `compile_module()` function takes your module, a teleprompter strategy, and training data. It returns an improved module with automatically selected demonstrations.

## Step 7: Check What Was Selected

See which examples the teleprompter chose:

```{r check-demos}
# The compiled module now has demos
compiled$config$demos
```

## Step 8: Compare All Three Versions

Let's see how each version performs:

```{r final-comparison}
#| cassette: true
# Run all three on the test set
results_baseline <- run_dataset(classifier, trainset, .llm = chat)
results_manual <- run_dataset(classifier_with_demos, trainset, .llm = chat)
results_compiled <- run_dataset(compiled, trainset, .llm = chat)

# Calculate accuracies
acc_baseline <- mean(results_baseline$category == trainset$category)
acc_manual <- mean(results_manual$category == trainset$category)
acc_compiled <- mean(results_compiled$category == trainset$category)

comparison <- tibble(
  Version = c("Baseline (no demos)", "Manual demos (3)", "LabeledFewShot (3)"),
  Accuracy = scales::percent(c(acc_baseline, acc_manual, acc_compiled))
)

comparison
```

## Step 9: Using Built-in Metrics

dsprrr provides metrics for common evaluation tasks. Use `metric_exact_match()` for classification:

```{r builtin-metrics}
#| cassette: true
metric <- metric_exact_match(field = "category")

# Evaluate the compiled module
eval_result <- evaluate(compiled, trainset, metric = metric, .llm = chat)

eval_result
```

The `evaluate()` function returns detailed results including:
- `mean_score`: Overall accuracy
- `scores`: Per-example scores
- `predictions`: What the model predicted

## Step 10: Varying the Number of Examples

More examples aren't always better. Let's compare:

```{r vary-k}
#| cassette: true
results <- list()

for (k in c(1L, 2L, 3L, 4L, 5L)) {
  tp <- LabeledFewShot(k = k)
  compiled_k <- compile_module(classifier, tp, trainset)

  eval_k <- evaluate(compiled_k, trainset, metric = metric, .llm = chat)

  results[[as.character(k)]] <- tibble(
    k = k,
    accuracy = eval_k$mean_score
  )
}

do.call(rbind, results)
```

You'll often find that 2-4 examples works best. Too many can confuse the LLM or exceed context limits.

## What You Learned

In this tutorial, you:

1. Created labeled training data with `dsp_trainset()`
2. Measured baseline accuracy before improving
3. Added manual demonstrations with the `demos` parameter
4. Used `LabeledFewShot` for automatic demo selection
5. Compiled modules with `compile_module()`
6. Evaluated with `metric_exact_match()`
7. Experimented with different numbers of examples

## Why Demos Work

LLMs are excellent at pattern matching. When you show examples of correct behavior:
- They understand the **format** you want
- They learn **edge cases** specific to your domain
- They pick up on **subtle distinctions** between categories

This is called "in-context learning"—the LLM learns from the examples in the prompt without any weight updates.

## The Few-Shot Trade-off

| More Examples | Fewer Examples |
|---------------|----------------|
| Better pattern coverage | Lower token costs |
| More edge cases shown | Faster inference |
| Risk of context overflow | May miss edge cases |

Start with 2-3 examples, measure, and adjust.

## Next Steps

You've improved your module with examples. But what about other parameters—temperature, instructions, templates? Continue to:

- **[Tutorial 5: Finding Best Configuration](tutorial-optimize-your-module.html)** — Grid search over parameters
- **[Quick Reference](cheatsheet.html)** — All metrics and teleprompters
- **[How Optimization Works](concepts-optimization-theory.html)** — The theory behind teleprompters
