---
title: "Tutorial 1: Your first LLM call"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 1: Your first LLM call}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

By the end of this tutorial you can describe a task as a signature, run it
against a language model, and get the answer back as an R value of the type
you asked for: a string, a number, a label or `TRUE`/`FALSE`.

## Set up

Install dsprrr from GitHub. pak installs ellmer from CRAN along with it.

```r
# install.packages("pak")
pak::pak("JamesHWade/dsprrr")
```

The tutorials use OpenAI. Add `OPENAI_API_KEY=<your key>` to your `.Renviron`
file and restart R; `dsprrr_sitrep()` then reports which API keys it can see.

```{r setup}
library(dsprrr)
library(ellmer)
```

dsprrr sends requests through an ellmer chat object. The output on this page
comes from `gpt-4.1`:

```{r create-chat}
chat <- chat_openai(model = "gpt-4.1")
```

Any other ellmer chat, such as `chat_anthropic()`, works the same way.

## Ask a question

A signature lists a task's inputs and outputs: `"question -> answer"` has one
input, `question`, and one output, `answer`. `module()` turns the signature
into something you can run:

```{r first-call}
#| cassette: true
qa <- module(signature("question -> answer"))
run(qa, question = "What is the capital of France?", .llm = chat)
```

`run()` returns a named list with one element per output, so the reply is in
`$answer`. This is the prompt dsprrr wrote from the signature:

```{r first-prompt}
cat(get_last_prompt()$prompt)
```

Along with the prompt, dsprrr sends a schema asking for an object with one
string field, `answer`, and ellmer parses the reply into R. With ellmer alone
you would write the prompt and the schema yourself. Because dsprrr builds both
from the signature, it can also change the prompt for you, which is how later
tutorials improve a module.

The same module answers any question:

```{r more-questions}
#| cassette: true
run(qa, question = "What is 7 * 8?", .llm = chat)

run(qa, question = "Who wrote Romeo and Juliet?", .llm = chat)
```

The signature fixes the shape of the result, a string named `answer`, but not
its wording: the model chose to reply in full sentences. Output types and
instructions give you more say over the value.

## Choose an output type

Add a type after an output's name. With `number` you get a number you can
compute with:

```{r number-output}
#| cassette: true
math <- module(signature("math_problem -> result: number"))
run(math, math_problem = "What is 15% of 200?", .llm = chat)
```

`enum()` lists the allowed labels, and the schema sent to the model permits
only those values:

```{r enum-output}
#| cassette: true
sentiment <- module(
  signature("text -> sentiment: enum('positive', 'negative', 'neutral')")
)
run(sentiment, text = "I absolutely loved this movie!", .llm = chat)
```

```{r enum-variations}
#| cassette: true
run(sentiment, text = "This was a complete waste of time.", .llm = chat)

run(sentiment, text = "It was okay, I guess.", .llm = chat)
```

`bool` returns `TRUE` or `FALSE`:

```{r bool-output}
#| cassette: true
truth <- module(signature("statement -> is_true: bool"))
run(truth, statement = "The Earth orbits the Sun.", .llm = chat)

run(truth, statement = "Cats are larger than elephants.", .llm = chat)
```

An output without a type is a string. The [quick reference](cheatsheet.html)
lists the other types, such as `int` and lists.

## Pass several inputs

Separate inputs with commas. A common pattern is to pass in text for the model
to answer from:

```{r multiple-inputs}
#| cassette: true
contextual_qa <- module(signature("context, question -> answer"))
run(
  contextual_qa,
  context = "R was created in 1993 by Ross Ihaka and Robert Gentleman at the University of Auckland.",
  question = "When was R created?",
  .llm = chat
)
```

```{r context-example}
#| cassette: true
run(
  contextual_qa,
  context = "The bakery opens at 7am and closes at 6pm. They sell croissants for $3 each.",
  question = "How much do croissants cost?",
  .llm = chat
)
```

## Add instructions

`instructions` replaces the default first line of the prompt ("Given the
fields ...") with your own description of the task:

```{r with-instructions}
#| cassette: true
one_word <- module(
  signature("question -> answer", instructions = "Answer in exactly one word.")
)
run(one_word, question = "What color is the sky on a clear day?", .llm = chat)

pirate <- module(
  signature("question -> answer", instructions = "Answer like a pirate.")
)
run(pirate, question = "What is the capital of France?", .llm = chat)
```

Passing `.llm` on every call keeps these examples explicit.
`set_default_chat(chat)` makes a chat the default, so you can leave `.llm` out.

Next, [Tutorial 2](tutorial-build-classifier.html) builds a sentiment
classifier and runs it on one text, a vector of texts and a data frame.
