---
title: "Tutorial 3: Extracting Structured Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 3: Extracting Structured Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::setup_knitr()
}
```

In [Tutorial 2](tutorial-build-classifier.html), you built a classifier that returns a single value. But real extraction tasks need multiple fields: names *and* dates, sentiment *and* confidence, entities *and* relationships.

In this tutorial, you'll extract complex, structured data from text.

**Time**: 25-30 minutes

## What You'll Build

An entity extractor that pulls structured information from news articles and emails.

## Prerequisites

- Completed [Tutorial 2](tutorial-build-classifier.html)
- `OPENAI_API_KEY` set in your environment

```{r setup}
library(dsprrr)
library(ellmer)
library(tibble)

chat <- chat_openai()
```

## Step 1: Multiple Output Fields

So far you've seen single outputs like `-> answer` or `-> sentiment`. Add more outputs with commas:

```{r multiple-outputs}
sig <- signature("text -> sentiment, confidence: number")

extractor <- module(sig, type = "predict")

run(extractor, text = "This product is absolutely fantastic!", .llm = chat)
```

You get back both `sentiment` and `confidence` in a structured result.

Try another:

```{r multiple-outputs-2}
run(extractor, text = "It was okay, nothing special.", .llm = chat)
```

Notice the confidence is lower for ambiguous text.

## Step 2: Typed Multiple Outputs

Add types to each output field:

```{r typed-multiple}
sig <- signature(
  "review -> sentiment: enum('positive', 'negative', 'neutral'), stars: int, summary: string"
)

analyzer <- module(sig, type = "predict")

result <- run(
  analyzer,
  review = "I've been using this blender for 6 months now. It's incredibly powerful and easy to clean. The only downside is it's quite loud. Overall, I'm very happy with it.",
  .llm = chat
)

result
```

You get sentiment, a star rating, *and* a summary—all typed correctly.

## Step 3: Complex Structures with `type_object()`

For nested or complex data, use ellmer's type system directly:

```{r type-object-basic}
sig <- signature(
  inputs = list(
    input("article", description = "News article to analyze")
  ),
  output_type = type_object(
    headline = type_string("A concise headline"),
    sentiment = type_enum(values = c("positive", "negative", "neutral")),
    word_count = type_integer()
  ),
  instructions = "Analyze the news article."
)

article_analyzer <- module(sig, type = "predict")
```

Test it with a news snippet:

```{r type-object-test}
article <- "
Scientists at MIT announced a breakthrough in solar panel efficiency today.
The new panels can convert 47% of sunlight to electricity, nearly double
the current commercial standard. The technology uses a novel layered
approach that captures more of the light spectrum. Researchers expect
commercial applications within 3-5 years.
"

run(article_analyzer, article = article, .llm = chat)
```

## Step 4: Arrays of Values

Extract lists of items with `type_array()`:

```{r arrays}
sig <- signature(
  inputs = list(
    input("text", description = "Text to extract entities from")
  ),
  output_type = type_object(
    people = type_array(type_string(), description = "Names of people mentioned"),
    organizations = type_array(type_string(), description = "Organizations mentioned"),
    locations = type_array(type_string(), description = "Places mentioned")
  ),
  instructions = "Extract named entities from the text."
)

entity_extractor <- module(sig, type = "predict")
```

Test with a news article:

```{r arrays-test}
news <- "
Apple CEO Tim Cook met with President Biden at the White House yesterday
to discuss manufacturing jobs. Cook announced that Apple will invest
$430 billion in the United States over the next five years, creating
20,000 new jobs. The meeting also included Treasury Secretary Janet Yellen
and Commerce Secretary Gina Raimondo.
"

result <- run(entity_extractor, text = news, .llm = chat)
result
```

Access the arrays directly:
```{r arrays-access}
result$people
result$organizations
```

## Step 5: Nested Objects

For hierarchical data, nest `type_object()` calls:

```{r nested}
sig <- signature(
  inputs = list(
    input("email", description = "Email message to parse")
  ),
  output_type = type_object(
    sender = type_object(
      name = type_string(),
      email = type_string()
    ),
    subject = type_string(),
    priority = type_enum(values = c("low", "normal", "high", "urgent")),
    action_items = type_array(type_string()),
    requires_response = type_boolean()
  ),
  instructions = "Parse the email and extract key information."
)

email_parser <- module(sig, type = "predict")
```

Test with an email:

```{r nested-test}
email <- "
From: Sarah Johnson <sarah.johnson@techcorp.com>
Subject: Q4 Budget Review - Action Required

Hi team,

Please review the attached Q4 budget proposal by Friday. We need to:
1. Confirm department allocations
2. Identify any cost-saving opportunities
3. Submit final numbers to finance

This is time-sensitive as the board meeting is next Monday.

Thanks,
Sarah
"

result <- run(email_parser, email = email, .llm = chat)
result
```

Access nested fields:
```{r nested-access}
result$sender$name
result$action_items
result$priority
```

## Step 6: Building an Email Triage System

Let's combine what you've learned into a practical system:

```{r triage-system}
sig <- signature(
  inputs = list(
    input("email", description = "Email to triage")
  ),
  output_type = type_object(
    category = type_enum(
      values = c("meeting", "task", "fyi", "urgent", "spam"),
      description = "Email category"
    ),
    summary = type_string("One-sentence summary"),
    action_required = type_boolean(),
    suggested_response = type_enum(
      values = c("reply_now", "reply_later", "forward", "archive", "delete"),
      description = "Recommended action"
    )
  ),
  instructions = "Triage the email for inbox management."
)

triage <- module(sig, type = "predict")
```

Process a batch of emails:

```{r triage-batch}
emails <- tibble(
  id = 1:3,
  email = c(
    "Meeting tomorrow at 3pm to discuss Q1 results. Please confirm attendance.",
    "FYI - The office will be closed on Monday for the holiday.",
    "URGENT: Server down! Need immediate assistance to restore services."
  )
)

results <- run_dataset(triage, emails, .llm = chat)
results
```

## Step 7: Handling Optional Fields

Some fields might not always be present. Make them nullable:

```{r optional-fields}
sig <- signature(
  inputs = list(
    input("text", description = "Text that may mention a date")
  ),
  output_type = type_object(
    has_date = type_boolean(),
    date = type_string("Date in YYYY-MM-DD format, or null if no date mentioned") |>
      (\(x) { x$nullable <- TRUE; x })(),
    confidence = type_number()
  ),
  instructions = "Extract date information if present."
)

date_extractor <- module(sig, type = "predict")

run(date_extractor, text = "Let's meet next Tuesday", .llm = chat)
run(date_extractor, text = "Great weather today!", .llm = chat)
```

## What You Learned

In this tutorial, you:

1. Extracted multiple output fields with comma notation
2. Added types to each field
3. Built complex structures with `type_object()`
4. Extracted lists with `type_array()`
5. Created nested/hierarchical outputs
6. Built a practical email triage system
7. Handled optional fields

## The Structure Advantage

Structured extraction is powerful because:

- **Guaranteed types**: Numbers come back as numbers, not strings
- **Predictable shape**: Your downstream code knows exactly what to expect
- **Validation**: The LLM must conform to your schema
- **Composability**: Results plug directly into R data structures

## Next Steps

Your extractor works, but can it be improved? Continue to:

- **[Tutorial 4: Improving with Examples](tutorial-improve-with-demos.html)** — Add demonstrations to improve accuracy
- **[Quick Reference](cheatsheet.html)** — All output types at a glance
- **[How Optimization Works](concepts-optimization-theory.html)** — Why examples improve LLM performance
