---
title: "Integration with vitals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration with vitals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = TRUE}
library(dsprrr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = eval_vignette()
)

# Configure vcr for vignettes
vcr::vcr_configure(
  dir = "_vcr",
  filter_request_headers = c("Authorization", "x-api-key"),
  match_requests_on = c("method", "uri")
)

vcr::setup_knitr()
```

```{r setup}
library(dsprrr)
library(vitals)
library(ellmer)
library(tibble)
```

## Overview

`dsprrr` and `vitals` form a complementary ecosystem for LLM application development in R:

- **dsprrr**: Focuses on building and optimizing LLM programs
- **vitals**: Focuses on evaluating LLM applications
- **ellmer**: Provides the common foundation for LLM API interactions

Together, they offer a complete workflow: **Design → Build → Optimize → Evaluate**.

**Note:** This vignette shows advanced vitals integration. For getting started with evaluation, see:

- [Evaluation and Metrics](evaluation-metrics.html) - Core evaluation concepts and metrics
- [Automatic Prompt Optimization](compilation-optimization.html) - Using evaluation for optimization

## Converting dsprrr Modules to vitals Solvers

The new API features in dsprrr make integration with vitals seamless. Here's how to use a dsprrr module as a vitals solver:

### Method 1: Direct Integration with as_vitals_solver()

Thanks to `as_vitals_solver()`, dsprrr modules work seamlessly with vitals. The function automatically handles vitals' calling convention (which passes a vector from `dataset$input`) by converting it to the correct format for your module:

```{r vitals-solver}
#| cassette: true
# Create a dsprrr module
sentiment_module <- signature(
  "text -> sentiment: enum('positive', 'negative', 'neutral')"
) |>
  module(type = "predict")

# Convert to vitals solver - it automatically maps dataset$input to "text"
task <- Task$new(
  dataset = tibble(
    input = c("I love this!", "This is terrible"),
    target = c("positive", "negative")
  ),
  solver = as_vitals_solver(sentiment_module, .llm = chat_openai()),
  scorer = model_graded_qa()
)

# Run the solver
task$solve()

# View results
task$get_samples()
```

### Method 2: Using run_dataset() (New Feature)

The new `run_dataset()` method makes it even easier to process tibbles:

```{r}
# Process a dataset directly
dataset <- tibble(
  text = c("Great product!", "Waste of money", "It's okay"),
  expected = c("positive", "negative", "neutral")
)

# Use the new run_dataset() method
results <- sentiment_module |>
  run_dataset(dataset, .llm = chat_openai())

# results now has all original columns plus 'result' column
print(results)
#> # A tibble: 3 × 3
#>   text            expected result
#>   <chr>           <chr>    <list>
#> 1 Great product!  positive <list [1]>
#> 2 Waste of money  negative <list [1]>
#> 3 It's okay       neutral  <list [1]>
```

## Using vitals Scorers in dsprrr Optimization

Convert vitals scorers to dsprrr metrics for use in optimization:

```{r}
# Create a metric from vitals scorer
# (vitals provides model-graded scorers returning "C"/"I" style outputs)

teleprompter <- GridSearchTeleprompter(
  variants = tibble(
    id = c("brief", "detailed"),
    instructions_mod = c(
      "Be concise",
      "Provide detailed analysis"
    )
  ),
  metric = as_dsprrr_metric(model_graded_qa())
)
```

## Complete Workflow Example

Here's an end-to-end example showing the full integration:

```{r, eval=FALSE}
# Note: This example demonstrates the workflow but requires more complex setup
# For a working example, see Method 1 above

# 1. Design with dsprrr
qa_module <- signature(
  "question -> answer"
) |>
  module(type = "predict")

# 2. Prepare dataset with vitals-compatible structure
qa_dataset <- tibble(
  input = c(
    "Why is the sky blue?",
    "At what temperature does water freeze?"
  ),
  target = c(
    "Due to Rayleigh scattering",
    "0°C or 32°F"
  )
)

# 3. Optimize with dsprrr
optimized_module <- compile(
  qa_module,
  teleprompter = LabeledFewShot(k = 2L),
  trainset = qa_dataset
)

# 4. Evaluate with vitals
eval_task <- Task$new(
  dataset = qa_dataset,
  solver = as_vitals_solver(optimized_module, .llm = chat_openai()),
  scorer = detect_match()
)

eval_task$solve()
eval_task$get_samples()
```

## Module as Function

The new `as_function()` feature makes modules even more natural to use:

```{r, eval=FALSE}
# Note: as_function() is planned for future release
# Convert module to function
classify <- as_function(sentiment_module, .llm = chat_openai())

# Use like any R function
result <- classify(text = "Amazing!")
print(result$sentiment)
#> [1] "positive"

# Works great in pipelines
dataset |>
  mutate(
    sentiment = map_chr(text, ~ classify(text = .x)$sentiment)
  )
```

## Performance Considerations

The new batch processing features provide significant performance benefits:

```{r, eval=FALSE}
# Process 100 items in parallel (future feature!)
large_dataset <- tibble(
  text = sample_texts[1:100]
)

# Automatic parallel processing with progress bar
results <- sentiment_module |>
  run_dataset(
    large_dataset,
    .llm = chat_openai(),
    .parallel = TRUE,
    .progress = TRUE
  )
```

## Summary

The integration between dsprrr and vitals is now seamless thanks to:

1. **Batch processing** in `run()` - handles vectors automatically
2. **Structured returns** - includes metadata needed for evaluation
3. **Dataset support** - `run_dataset()` works directly with tibbles
4. **Function interface** - modules can be called like functions

These features make it easy to:
- Use dsprrr modules as vitals solvers without adapters
- Share datasets between packages
- Combine optimization (dsprrr) with evaluation (vitals)
- Build complete LLM application workflows in R

The packages are stronger together while remaining independently valuable!
