---
title: "Evaluate with vitals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluate with vitals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
# Skip evaluation during R CMD check or CI (cassettes may not match current code)
# VITALS_SHOULD_EVAL can override for cassette recording
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval && (
  nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
  identical(Sys.getenv("NOT_CRAN"), "false")
)

should_eval <- !in_check && dsprrr:::eval_vignette()
has_vitals <- requireNamespace("vitals", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

[vitals](https://vitals.tidyverse.org) is Posit's package for evaluating LLM
applications: a `Task` runs a solver over a dataset, grades each answer with a
scorer, and writes a log you can browse. This page shows how to evaluate a
dsprrr module as a vitals solver, choose a scorer, read the results, and use
vitals scorers as dsprrr metrics. The outputs below are recorded responses from
`gpt-4o-mini`.

```{r setup}
#| eval: !expr has_vitals
library(dsprrr)
library(vitals)
library(ellmer)
library(tibble)
```

## Evaluate a module

`as_vitals_task()` turns a module and a data frame into a vitals `Task`. The
data needs one column per signature input plus a `target` column with the
expected answer:

```{r quick-start}
#| cassette: quick-start

classifier <- signature("input -> label: enum('positive', 'negative')") |>
  module()

test_data <- tibble(
  input = c(
    "I love this product!",
    "Terrible experience, waste of money"
  ),
  target = c("positive", "negative")
)

task <- as_vitals_task(
  module = classifier,
  dataset = test_data,
  scorer = detect_match(),
  name = "sentiment",
  dir = tempdir(),
  .llm = chat_openai(model = "gpt-4o-mini")
)

task$eval()
task$get_samples()
```

`$eval()` runs the module on every row, scores the answers and writes a log to
`dir`; `vitals_view(dir)` opens the log viewer. In `$get_samples()`, `result`
holds the module's output. When the signature has one string or enum output,
that is the plain value; otherwise it is all output fields as JSON. `score` is
`C` (correct) or `I` (incorrect). Leave out `scorer` and you get
`model_graded_qa()`.

## Evaluate on held-out data

To check a compiled module, compile on training rows and evaluate on rows it
has not seen. `LabeledFewShot(k = 2L)` adds two randomly sampled training rows
to the prompt as examples:

```{r train-test}
#| cassette: train-test

qa_data <- tibble(
  question = c(
    "What is the capital of France?",
    "Who wrote Romeo and Juliet?",
    "What is 2 + 2?",
    "What color is the sky?",
    "Who painted the Mona Lisa?",
    "What is the largest planet?"
  ),
  target = c(
    "Paris",
    "Shakespeare",
    "4",
    "Blue",
    "Leonardo da Vinci",
    "Jupiter"
  )
)

set.seed(42)
train_idx <- sample(nrow(qa_data), 4)
trainset <- qa_data[train_idx, ]
testset <- qa_data[-train_idx, ]

qa_module <- signature("question -> answer") |>
  module()

optimized <- compile(
  qa_module,
  LabeledFewShot(k = 2L),
  trainset = trainset
)

test_task <- as_vitals_task(
  module = optimized,
  dataset = testset,
  scorer = detect_match(),
  name = "qa-test",
  dir = tempdir(),
  .llm = chat_openai(model = "gpt-4o-mini")
)

test_task$eval()

test_task$get_samples()[c("target", "result", "score")]
```

The model answered "William Shakespeare" and still passed. `detect_match()`
looks for the target at the end of the answer by default
(`location = "end"`), ignoring case, so "Shakespeare wrote it" would have
failed. Use `location = "exact"`, `"begin"` or `"any"` when that matters.

## Choose a scorer

| Scorer | Marks an answer correct when | Good for |
|---|---|---|
| `detect_match()` | the target appears at the end of the answer (see `location`) | short factual answers and labels |
| `detect_includes()` | the target appears anywhere in the answer | a word or phrase that must be present |
| `detect_pattern()` | text captured by a regular expression matches the target | pulling a number or code out of longer text |
| `model_graded_qa()` | a grader model judges that the answer meets the target | open-ended answers |

The examples above use `detect_match()`. For explanations, `model_graded_qa()`
asks a model whether the answer meets the target, treated as a criterion.
Without a `scorer_chat` it grades with a fresh copy of the solver's Chat, here
`gpt-4o-mini` grading itself. `chain_of_thought()` adds a `reasoning` output,
so `result` is JSON with both fields and the grader sees the reasoning too:

```{r scorer-graded}
#| cassette: scorer-graded

explainer <- signature("topic -> explanation") |>
  chain_of_thought()

task <- as_vitals_task(
  module = explainer,
  dataset = tibble(
    topic = c("Why is the sky blue?", "How do plants make food?"),
    target = c(
      "Light scattering in atmosphere",
      "Photosynthesis using sunlight"
    )
  ),
  scorer = model_graded_qa(),
  name = "explanations",
  dir = tempdir(),
  .llm = chat_openai(model = "gpt-4o-mini")
)

task$eval()
task$get_samples()
```

`detect_includes()` checks that the target text appears anywhere. It is
literal: the prompt below never says which language to use, the model wrote
Python (`def add_numbers`), and the first row fails because "function" is
missing:

```{r scorer-pattern}
#| cassette: scorer-pattern

coder <- signature("task -> code") |> module()

task <- as_vitals_task(
  module = coder,
  dataset = tibble(
    task = c(
      "Write a function to add two numbers",
      "Create a loop that prints 1 to 5"
    ),
    target = c("function", "for")
  ),
  scorer = detect_includes(),
  name = "code-gen",
  dir = tempdir(),
  .llm = chat_openai(model = "gpt-4o-mini")
)

task$eval()
task$get_samples()
```

## Read the results

`$get_samples()` returns one row per sample. `input` is a list column holding
each row's signature inputs, so pull fields out of it to build a readable
table:

```{r analyzing}
#| cassette: analyzing

sentiment <- signature(
  "text -> sentiment: enum('positive', 'negative', 'neutral')"
) |>
  module()

dataset <- tibble(
  text = c(
    "Best purchase ever!",
    "Complete garbage",
    "It's okay I guess",
    "Amazing quality!",
    "Never buying again"
  ),
  target = c("positive", "negative", "neutral", "positive", "negative")
)

task <- as_vitals_task(
  module = sentiment,
  dataset = dataset,
  scorer = detect_match(),
  name = "sentiment-analysis",
  dir = tempdir(),
  .llm = chat_openai(model = "gpt-4o-mini")
)

task$eval()

samples <- task$get_samples()
tibble(
  text = vapply(samples$input, \(x) x$text, character(1)),
  target = samples$target,
  result = samples$result,
  score = samples$score
)

mean(samples$score == "C")
```

Filter on `score == "I"` to list the failures.

## Compare two modules

Run the same data through each variant and compare the share of correct
answers. Here a plain module and a chain-of-thought module answer two word
problems, graded by `model_graded_qa()`:

```{r comparing}
#| cassette: comparing

sig <- signature("question -> answer")
basic <- module(sig)
cot <- chain_of_thought(sig)

test_data <- tibble(
  question = c(
    "If a train travels 60 mph for 2 hours, how far does it go?",
    "What is 15% of 80?"
  ),
  target = c("120 miles", "12")
)

llm <- chat_openai(model = "gpt-4o-mini")

results <- list()
for (name in c("basic", "cot")) {
  mod <- if (name == "basic") basic else cot

  task <- as_vitals_task(
    module = mod,
    dataset = test_data,
    scorer = model_graded_qa(),
    name = paste0("math-", name),
    dir = tempdir(),
    .llm = llm
  )
  task$eval()

  results[[name]] <- mean(task$get_samples()$score == "C")
}

tibble(variant = names(results), accuracy = unlist(results))
```

Both variants get both problems right, so two questions can't separate them.
Use enough rows that a difference would show.

## Repeat samples with epochs

A model can answer the same prompt differently from run to run. `epochs`
repeats every row so you can see how stable the scores are. Pass
`.cache = FALSE` (forwarded to `run_dataset()`) so each repeat makes a new
request; with dsprrr's response cache on, the repeats would replay the first
answer:

```{r epochs}
#| cassette: epochs

classifier <- signature(
  "text -> category: enum('tech', 'sports', 'politics')"
) |>
  module()

test_data <- tibble(
  text = c(
    "New iPhone released today",
    "Lakers win championship",
    "Senate passes new bill"
  ),
  target = c("tech", "sports", "politics")
)

task <- as_vitals_task(
  module = classifier,
  dataset = test_data,
  scorer = detect_match(),
  name = "news-classification",
  epochs = 3L,
  dir = tempdir(),
  .llm = chat_openai(model = "gpt-4o-mini"),
  .cache = FALSE
)

task$eval()

scores <- task$get_samples()
nrow(scores)
table(scores$target, scores$score)
```

Three rows times three epochs gives nine samples, and each headline got the
same label every time.

For larger datasets, add
`.concurrency = concurrency_control(backend = "ellmer", max_active = 4L)` to
`as_vitals_task()` to run rows in parallel. Concurrent rows skip dsprrr's
response cache.

## Build the Task yourself

`as_vitals_solver()` wraps a module as a vitals solver, for when you want to
call `Task$new()` directly. A module whose only input is named `input` reads
vitals' `input` column as is; for other or multiple inputs, use
`as_vitals_task()`, which packs them into that column for you.

```{r}
#| eval: false
triage <- module(
  signature("input -> priority: enum('low', 'medium', 'high')")
)
tickets <- tibble(
  input = c("Checkout fails for every customer", "Typo on the About page"),
  target = c("high", "low")
)

task <- Task$new(
  dataset = tickets,
  solver = as_vitals_solver(triage, .llm = chat_openai(model = "gpt-4o-mini")),
  scorer = detect_match(location = "exact"),
  dir = tempdir()
)
task$eval()
```

## Use vitals scorers as dsprrr metrics

`metric_detect_match()`, `metric_detect_includes()` and
`metric_detect_pattern()` wrap the matching vitals scorers as dsprrr metrics,
so `evaluate()`, `optimize_grid()` and `compile()` can use them. They take the
expected answer from the data's `target` column, return 1 or 0, and work with
modules that have a single output field:

```{r scorer-metrics}
#| eval: !expr has_vitals
ends_with_target <- metric_detect_match()
exact_target <- metric_detect_match(location = "exact")
row <- tibble(question = "Who wrote Hamlet?", target = "Shakespeare")

ends_with_target(list(answer = "William Shakespeare"), row)
exact_target(list(answer = "William Shakespeare"), row)
```

In an evaluation, pass the metric like any other:

```{r}
#| eval: false
evaluate(
  qa_module,
  testset,
  metric = metric_detect_match(),
  .llm = chat_openai(model = "gpt-4o-mini")
)
```

`as_dsprrr_metric()` adapts other vitals-style scorers the same way.
`metric_model_graded_qa()` and `metric_model_graded_fact()` exist too, but they
currently show the grader the prediction as R list syntax and no question
unless the data has an `input` column; for model grading, use
`model_graded_qa()` in a vitals Task as shown above. For how metrics drive
optimization, see [Metrics and evaluation](concepts-why-metrics-matter.html).
