---
title: "Integration with vitals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration with vitals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
# This vignette shows API integration patterns but doesn't evaluate
# because it requires vitals, API keys, and complex setup.
# When VCR cassettes are recorded, evaluation can be enabled.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(dsprrr)
library(vitals)
library(ellmer)
library(tibble)
```

## Overview

`dsprrr` and `vitals` form a complementary ecosystem for LLM application development in R:

- **dsprrr**: Focuses on building and optimizing LLM programs
- **vitals**: Focuses on evaluating LLM applications
- **ellmer**: Provides the common foundation for LLM API interactions

Together, they offer a complete workflow: **Design → Build → Optimize → Evaluate**.

## Converting dsprrr Modules to vitals Solvers

The new API features in dsprrr make integration with vitals seamless. Here's how to use a dsprrr module as a vitals solver:

### Method 1: Direct Integration (with new API)

Thanks to the batch processing support and structured returns, dsprrr modules can now work directly with vitals:

```{r}
# Create a dsprrr module (using 'input' to match vitals conventions)
sentiment_module <- signature(
  "input -> sentiment: enum('positive', 'negative', 'neutral')"
) |>
  module(type = "predict")

# Create vitals task with dsprrr's helper
task <- Task$new(
  dataset = tibble(
    input = c("I love this!", "This is terrible"),
    target = c("positive", "negative")
  ),
  solver = as_vitals_solver(sentiment_module, llm = chat_openai()),
  scorer = model_graded_qa()
)

# Evaluate
task$eval()
```

### Method 2: Using run_dataset() (New Feature)

The new `run_dataset()` method makes it even easier to process tibbles:

```{r}
# Process a dataset directly
dataset <- tibble(
  text = c("Great product!", "Waste of money", "It's okay"),
  expected = c("positive", "negative", "neutral")
)

# Use the new run_dataset() method
results <- sentiment_module |>
  run_dataset(dataset, .llm = chat_openai())

# results now has all original columns plus 'result' column
print(results)
#> # A tibble: 3 × 3
#>   text            expected result
#>   <chr>           <chr>    <list>
#> 1 Great product!  positive <list [1]>
#> 2 Waste of money  negative <list [1]>
#> 3 It's okay       neutral  <list [1]>
```

## Using vitals Scorers in dsprrr Optimization

Convert vitals scorers to dsprrr metrics for use in optimization:

```{r}
# Create a metric from vitals scorer
# (vitals provides model-graded scorers returning "C"/"I" style outputs)

teleprompter <- GridSearchTeleprompter(
  variants = tibble(
    id = c("brief", "detailed"),
    instructions_mod = c(
      "Be concise",
      "Provide detailed analysis"
    )
  ),
  metric = as_dsprrr_metric(model_graded_qa())
)
```

## Complete Workflow Example

Here's an end-to-end example showing the full integration:

```{r}
# 1. Design with dsprrr
qa_module <- signature(
  inputs = list(
    input("context", description = "Background information"),
    input("question", description = "Question to answer")
  ),
  output_type = ellmer::type_object(
    answer = ellmer::type_string(),
    confidence = ellmer::type_number(minimum = 0, maximum = 1)
  ),
  instructions = "Answer the question based on the context"
) |>
  module(type = "predict")

# 2. Prepare dataset
qa_dataset <- tibble(
  context = c(
    "The sky is blue due to Rayleigh scattering",
    "Water freezes at 0°C or 32°F"
  ),
  question = c(
    "Why is the sky blue?",
    "At what temperature does water freeze?"
  ),
  target = c(
    "Due to Rayleigh scattering",
    "0°C or 32°F"
  )
)

# 3. Optimize with dsprrr (using vitals metrics)
optimized_module <- compile(
  qa_module,
  teleprompter = LabeledFewShot(k = 2),
  trainset = qa_dataset
)

# 4. Evaluate with vitals
eval_task <- Task$new(
  dataset = qa_dataset,
  solver = function(inputs, ...) {
    results <- run_dataset(
      optimized_module,
      tibble(
        context = inputs,
        question = inputs # Extract from combined input
      ),
      .return_format = "structured"
    )

    list(
      result = sapply(results$result, function(r) r$answer),
      solver_chat = results$.chat
    )
  },
  scorer = detect_match()
)

eval_task$eval()
```

## Module as Function

The new `as_function()` feature makes modules even more natural to use:

```{r}
# Convert module to function
classify <- as_function(sentiment_module, .llm = chat_openai())

# Use like any R function
result <- classify(text = "Amazing!")
print(result$sentiment)
#> [1] "positive"

# Works great in pipelines
dataset |>
  mutate(
    sentiment = map_chr(text, ~ classify(text = .x)$sentiment)
  )
```

## Performance Considerations

The new batch processing features provide significant performance benefits:

```{r}
# Process 100 items in parallel (new feature!)
large_dataset <- tibble(
  text = sample_texts[1:100]
)

# Automatic parallel processing with progress bar
results <- sentiment_module |>
  run_dataset(
    large_dataset,
    .llm = chat_openai(),
    .parallel = TRUE, # Default
    .progress = TRUE # Default
  )
```

## Summary

The integration between dsprrr and vitals is now seamless thanks to:

1. **Batch processing** in `run()` - handles vectors automatically
2. **Structured returns** - includes metadata needed for evaluation
3. **Dataset support** - `run_dataset()` works directly with tibbles
4. **Function interface** - modules can be called like functions

These features make it easy to:
- Use dsprrr modules as vitals solvers without adapters
- Share datasets between packages
- Combine optimization (dsprrr) with evaluation (vitals)
- Build complete LLM application workflows in R

The packages are stronger together while remaining independently valuable!
