---
title: "dsprrr Experiment Report"
subtitle: "LLM Module Optimization & Evaluation Results"
author: "Generated by dsprrr"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
params:
  pins_board_path: "pins"
  module_name: "sentiment-classifier"
  eval_name: "sentiment-eval-results"
  traces_name: "sentiment-eval-traces"
---

```{r setup}
#| include: false
library(dsprrr)
library(tibble)
library(dplyr)
library(ggplot2)

# Check for optional packages
has_pins <- requireNamespace("pins", quietly = TRUE)
has_targets <- requireNamespace("targets", quietly = TRUE)

# Theme for plots
theme_set(theme_minimal(base_size = 12))
```

## Overview
This report summarizes the results of a dsprrr LLM experiment, including:

- Module configuration and optimization results
- Evaluation metrics and performance
- Execution traces and resource usage

```{r load-data}
#| label: load-pinned-data
# Load data from pins board (if available)
if (has_pins) {
  board <- pins::board_folder(params$pins_board_path)

  # Try to load each pinned artifact
  module_config <- tryCatch(
    pins::pin_read(board, params$module_name),
    error = function(e) NULL
  )

  eval_results <- tryCatch(
    pins::pin_read(board, params$eval_name),
    error = function(e) NULL
  )

  traces_data <- tryCatch(
    pins::pin_read(board, params$traces_name),
    error = function(e) NULL
  )
} else {
  module_config <- NULL
  eval_results <- NULL
  traces_data <- NULL
}

# Alternative: Load from targets cache
if (has_targets && is.null(eval_results)) {
  eval_results <- tryCatch(
    targets::tar_read(evaluation_results),
    error = function(e) NULL
  )
}
```

## Module Configuration
```{r module-config}
#| label: display-module-config
if (!is.null(module_config)) {
  cat("**Module Type:**", module_config$metadata$module_type, "\n\n")
  cat("**Created:**", format(module_config$metadata$created_at), "\n\n")
  cat("**dsprrr Version:**", module_config$metadata$dsprrr_version, "\n\n")

  # Signature
  cat("### Signature\n\n")
  cat("**Inputs:**\n")
  for (inp in module_config$signature$inputs) {
    cat("-", inp$name)
    if (!is.null(inp$description) && inp$description != "") {
      cat(":", inp$description)
    }
    cat("\n")
  }
  cat("\n**Instructions:**\n\n> ", module_config$signature$instructions, "\n")

  # Optimization status
  if (isTRUE(module_config$optimization$compiled)) {
    cat("\n### Optimization Results\n\n")
    cat("**Status:** Compiled\n\n")
    cat("**Best Score:**", round(module_config$optimization$best_score, 4), "\n\n")
    cat("**Number of Trials:**", module_config$optimization$n_trials, "\n\n")

    if (!is.null(module_config$optimization$best_params)) {
      cat("**Best Parameters:**\n\n")
      for (param in names(module_config$optimization$best_params)) {
        cat("-", param, "=",
            as.character(module_config$optimization$best_params[[param]]), "\n")
      }
    }
  } else {
    cat("\n*Module was not optimized*\n")
  }
} else {
  cat("*Module configuration not available*\n")
}
```

## Evaluation Results

```{r eval-metrics}
#| label: evaluation-metrics
if (!is.null(eval_results)) {
  # Display key metrics
  metrics_df <- tibble(
    Metric = c("Mean Score", "Samples Evaluated", "Errors"),
    Value = c(
      round(eval_results$mean_score %||% NA, 4),
      eval_results$n_evaluated %||% NA,
      eval_results$n_errors %||% 0
    )
  )

  knitr::kable(metrics_df, align = c("l", "r"))
} else {
  cat("*Evaluation results not available*\n")
}
```

```{r score-distribution}
#| label: score-distribution
#| fig-cap: "Distribution of scores across evaluated samples"
if (!is.null(eval_results) && !is.null(eval_results$scores)) {
  scores_df <- tibble(score = unlist(eval_results$scores))

  if (nrow(scores_df) > 0) {
    ggplot(scores_df, aes(x = factor(score))) +
      geom_bar(fill = "steelblue", alpha = 0.8) +
      labs(
        x = "Score",
        y = "Count",
        title = "Score Distribution"
      ) +
      theme(plot.title = element_text(hjust = 0.5))
  }
}
```

## Execution Traces

```{r traces-summary}
#| label: traces-summary
if (!is.null(traces_data)) {
  summary <- traces_data$summary

  cat("### Usage Summary\n\n")
  cat("- **Total Requests:**", summary$n_traces, "\n")
  cat("- **Total Tokens:**", format(summary$total_tokens, big.mark = ","), "\n")
  cat("- **Total Cost:** $", format(summary$total_cost, digits = 4), "\n")
  cat("- **Total Time:**", round(summary$total_latency_ms / 1000, 2), "seconds\n\n")

  cat("### Performance\n\n")
  cat("- **Avg Latency:**", round(summary$avg_latency_ms, 1), "ms\n")
  cat("- **Avg Tokens/Request:**", round(summary$avg_tokens_per_request, 1), "\n")
} else {
  cat("*Trace data not available*\n")
}
```

```{r latency-plot}
#| label: latency-over-time
#| fig-cap: "Request latency over time"
if (!is.null(traces_data) && nrow(traces_data$traces) > 0) {
  traces_df <- traces_data$traces

  if ("latency_ms" %in% names(traces_df) && "timestamp" %in% names(traces_df)) {
    ggplot(traces_df, aes(x = timestamp, y = latency_ms)) +
      geom_line(alpha = 0.5) +
      geom_point(size = 2, alpha = 0.7) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
      labs(
        x = "Time",
        y = "Latency (ms)",
        title = "Request Latency Over Time"
      ) +
      theme(plot.title = element_text(hjust = 0.5))
  }
}
```

```{r token-usage}
#| label: token-usage
#| fig-cap: "Token usage per request"
if (!is.null(traces_data) && nrow(traces_data$traces) > 0) {
  traces_df <- traces_data$traces

  if (all(c("input_tokens", "output_tokens") %in% names(traces_df))) {
    token_df <- traces_df |>
      mutate(request_id = row_number()) |>
      tidyr::pivot_longer(
        cols = c(input_tokens, output_tokens),
        names_to = "type",
        values_to = "tokens"
      ) |>
      mutate(type = gsub("_tokens", "", type))

    ggplot(token_df, aes(x = request_id, y = tokens, fill = type)) +
      geom_col(position = "stack", alpha = 0.8) +
      scale_fill_manual(
        values = c("input" = "steelblue", "output" = "coral"),
        labels = c("Input", "Output")
      ) +
      labs(
        x = "Request",
        y = "Tokens",
        fill = "Type",
        title = "Token Usage by Request"
      ) +
      theme(
        plot.title = element_text(hjust = 0.5),
        legend.position = "bottom"
      )
  }
}
```

## Detailed Results

```{r predictions-table}
#| label: predictions-table
if (!is.null(eval_results) && !is.null(eval_results$predictions)) {
  cat("### Predictions\n\n")

  # Create a summary table
  pred_df <- tibble(
    Index = seq_along(eval_results$predictions),
    Prediction = as.character(eval_results$predictions),
    Score = unlist(eval_results$scores)
  )

  # Show first 20 rows
  head(pred_df, 20)
}
```

## Reproducibility

```{r session-info}
#| label: session-info
cat("**Report Generated:**", format(Sys.time()), "\n\n")

if (!is.null(module_config)) {
  cat("**Original R Version:**", module_config$metadata$r_version %||% "Unknown", "\n\n")
}

cat("**Current Session:**\n\n")
sessionInfo()
```

---

*This report was generated using [dsprrr](https://github.com/jameshwade/dsprrr) - Declarative Self-improving Language Programs for R*
