---
title: "Agentic optimization harnesses"
description: >
  Run persistent or fresh-per-iteration research agents against a trusted
  dsprrr evaluator, with Posit mcp-repl sandboxing, budgets, lineage, and
  checkpoints.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Agentic optimization harnesses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
options(conflicts.policy = list(warn = FALSE))
```

Most dsprrr teleprompters own both candidate generation and selection.
`AutoResearch()` and `MetaHarness()` split those jobs: an LLM agent proposes
experiments, while a trusted R evaluator applies validated edits, runs the
program, enforces budgets, and keeps the best result. This page shows how to
run both harnesses on a two-step pipeline, keep the agent's code in a sandbox,
budget and resume a run, and read what happened.

Both harnesses can jointly edit the instructions and templates of every
optimizable module in a program. That makes them useful when the problem is
not one weak prompt but the contract between several pipeline stages. They do
not change the program's structure or run optimizer-written module source. To
let an optimizer change the implementation itself (predictor structure,
deterministic R, or tools), use the experimental `flex()`; see [Flex: optimize
a whole program](flex-optimization.html).

## Choose the search owner

| Harness | Who owns the loop | Agent memory | Candidates |
|---|---|---|---|
| `AutoResearch()` | One research agent | Persistent across experiments | One per `propose` action |
| `MetaHarness()` | The R outer loop | Fresh proposer each iteration | A batch per iteration |

`AutoResearch()` fits when one researcher should pursue a line of inquiry,
use earlier results as working memory, branch, and decide when to stop.
`MetaHarness()` fits when you want independent proposals, explicit frontier
selection, and less dependence on one conversation's hidden state.

With either harness, dsprrr:

1. evaluates the unmodified program as a baseline;
2. gives every candidate a content fingerprint and a parent id;
3. rejects unknown module paths and unsupported fields;
4. skips candidates identical to one already evaluated;
5. charges every evaluation to the `optimizer_control()` budget; and
6. returns the best scored program, including when a budget stops the run.

## The example program

The examples optimize a two-step pipeline that answers questions from a short
context. The first step quotes evidence, the second answers from it, so the
two instructions have to agree on a format. The data has eight rows, which is
enough to show the shape; a real run needs many more.

```{r}
#| eval: true
library(dsprrr)

find_evidence <- module(signature(
  "context, question -> evidence",
  instructions = "Restate the question, then quote the context sentences that answer it."
))
answer_from_evidence <- module(signature(
  "evidence -> answer",
  instructions = "Answer the restated question from the quoted sentences only."
))
qa_program <- pipeline(find_evidence, answer_from_evidence)

examples <- tibble::tribble(
  ~context, ~question, ~answer,
  "The Oslo office opened in 2019 and has 42 staff.", "When did the Oslo office open?", "2019",
  "Refunds are processed within 5 business days.", "How many business days do refunds take?", "5",
  "The Lisbon office has 18 staff.", "When did the Lisbon office open?", "unknown",
  "The API allows 600 requests per minute.", "How many requests per minute does the API allow?", "600",
  "Maria Lopez leads the data platform team.", "Who leads the data platform team?", "Maria Lopez",
  "Backups run nightly at 02:00 UTC.", "How long are backups kept?", "unknown",
  "Support tickets are answered within 4 hours.", "How fast are support tickets answered?", "4 hours",
  "The Berlin office moved to Mitte in 2021.", "How many staff work in the Berlin office?", "unknown"
)
trainset <- examples[1:4, ]
valset <- examples[5:8, ]
```

## Keep code execution sandboxed

The agent can ask to run R code to analyze the program, the frontier, and
the training rows. The harnesses require an operating-system sandbox for that
code by default. `mcp_repl_runner()` provides one by launching Posit's
open-source [mcp-repl](https://github.com/posit-dev/mcp-repl), a persistent R
session behind one MCP tool. Install the executable and the R client:

```sh
uv tool install posit-mcp-repl   # or: pipx install posit-mcp-repl
Rscript -e 'install.packages("mcptools")'
```

Then create the runner:

```{r}
runner <- mcp_repl_runner(
  sandbox = "workspace-write",
  timeout = 30
)
```

The managed sandbox disables network access and limits writes to the
workspace and temporary paths. The session stays alive across calls, so an
agent can load a package or build an analysis once and keep working with it;
`runner$reset()` starts a clean session. Code, results and context have size
limits, listed in [How RLM works](how-rlm-works.html#limits).

A runner built around your own `repl = function(input, timeout_ms)` is
different: dsprrr did not launch that server and cannot vouch for its
isolation, so the runner reports `sandboxed = FALSE` and the harnesses reject
it. `r_code_runner()` is rejected for the same reason: its callr process keeps
your user's permissions and is not a security boundary.

Set `sandbox = FALSE` on a harness only to remove the sandbox action entirely.
Any supplied runner is then ignored; the agent can still propose edits but
cannot run code.

## AutoResearch: one persistent research session

Use separate chats for the program under test and the research agent. That
keeps task traces, model choice, and cost attribution apart.

```{r}
library(ellmer)

task_chat <- chat_openai(model = "gpt-4.1-mini")
research_chat <- chat_anthropic(model = "claude-sonnet-4-5")

metric <- metric_with_feedback(
  function(prediction, expected) {
    correct <- identical(prediction$answer, expected$answer)
    list(
      score = as.numeric(correct),
      feedback = if (correct) {
        "Correct."
      } else {
        paste0(
          "Expected '", expected$answer,
          "' but got '", prediction$answer, "'."
        )
      }
    )
  },
  field = "answer"
)

research <- AutoResearch(
  metric = metric,
  max_iterations = 16L,
  patience = 5L,
  target_score = 0.95,
  max_context_examples = 20L,
  max_feedback_examples = 8L,
  seed = 42L
)

compiled <- compile(
  qa_program,
  research,
  trainset,
  valset = valset,
  .llm = task_chat,
  .agent_llm = research_chat,
  runner = runner,
  objective = paste(
    "Answer from supplied evidence.",
    "Do not invent facts when the context is insufficient."
  ),
  .cache = FALSE
)
```

The harnesses also accept `metric_with_trace()` metrics, which can score a row
from its `status`, ordered `events`, and module `metadata`. Candidate selection
stays in the trusted R evaluator either way.

On each turn the agent must choose one action:

| Action | What the agent does |
|---|---|
| `sandbox` | Runs R code against a `.context` holding the objective, the editable program manifest, visible training rows, and the current frontier |
| `propose` | Branches from a known candidate and submits one candidate: complete replacement instructions or templates at one or more module paths |
| `finish` | Stops, when another experiment is unlikely to help |

The agent decides the research policy, but it cannot write into the module or
the evaluator directly. dsprrr re-validates and applies every candidate.

## Meta-Harness: fresh proposers, trusted selection

Meta-Harness creates a fresh proposer chat for each outer iteration, so the
proposer reasons from the persisted evidence rather than from earlier
conversation. Pass an ellmer `Chat`; dsprrr clones it and clears its turns
before each iteration:

```{r}
harness <- MetaHarness(
  metric = metric,
  max_iterations = 8L,
  max_candidates_per_iteration = 4L,
  frontier_size = 8L,
  patience = 3L,
  seed = 42L
)

compiled <- compile(
  qa_program,
  harness,
  trainset,
  valset = valset,
  .llm = task_chat,
  .agent_llm = research_chat,
  runner = runner,
  objective = "Improve factual accuracy without increasing abstention errors."
)
```

Meta-Harness rejects agents that are not ellmer chats, or that cannot be
cloned, because reusing them could carry hidden conversation state across
iterations.

Within an iteration, the proposer may request a few sandbox steps before it
submits its batch. The outer loop then:

1. builds each proposal from its declared parent;
2. rejects invalid or duplicate candidates;
3. evaluates every unique candidate that fits the remaining budget;
4. updates the scored frontier; and
5. checkpoints the new lineage before starting a fresh proposer.

The agent can inspect and hypothesize, but the scorer, the budget, and the
accepted state stay with the host.

## Joint pipeline edits

Candidates address modules by the paths that `named_parameters()` reports.
They stay stable within one program:

```{r}
#| eval: true
names(named_parameters(qa_program))
```

One candidate may replace fields at several paths. Here, a candidate can
change the evidence step to write `NONE` when the context does not answer the
question, and change the answer step to reply `unknown` when it sees `NONE`.
The candidate ledger stores the complete resulting snapshot, not only the
patch, so fingerprints, resumes, and comparisons are exact.

Compiled and frozen modules are boundaries: the harnesses do not edit them or
anything inside them. To let the agent edit an inner component, change the
program before compiling rather than asking the agent to get around the
boundary.

## Budget every kind of work

The shared optimizer ledger counts baseline and candidate evaluations as
trials, and tracks program calls and tokens during evaluation, proposer calls
and tokens, metric calls, cost, elapsed time, and errors:

```{r}
control <- optimizer_control(
  max_trials = 25L,
  max_metric_calls = 200L,
  max_provider_calls = 250L,
  max_total_tokens = 500000L,
  max_cost = 10,
  max_elapsed_seconds = 1800,
  checkpoint_path = "meta-harness-checkpoint.rds"
)

compiled <- compile(
  qa_program,
  harness,
  trainset,
  valset = valset,
  .llm = task_chat,
  .agent_llm = research_chat,
  runner = runner,
  control = control
)
```

A finite provider, token, or cost cap stops the run when a chat cannot report
its usage, rather than running unmetered. A stopped run returns its best
partial program and says why it stopped:

```{r}
optimization_result(compiled)$budget
optimization_result(compiled)$stop_reason
optimization_result(compiled)$status
```

## Resume and audit a run

Both harnesses support optimizer checkpoints. Resume with the same program,
data, metric, and harness configuration; resource limits may be raised:

```{r}
resumed <- compile(
  qa_program,
  harness,
  trainset,
  valset = valset,
  .llm = task_chat,
  .agent_llm = research_chat,
  runner = runner,
  control = optimizer_control(
    max_trials = 40L,
    max_cost = 20,
    checkpoint_path = "meta-harness-checkpoint.rds",
    resume = TRUE
  )
)
```

The candidate ledger is attached to the returned module:

```{r}
candidates <- optimization_result(resumed)$trials
candidates[, c(
  "id",
  "parent_id",
  "name",
  "iteration",
  "score",
  "selected",
  "status"
)]

# Complete per-candidate program state and evaluation feedback:
candidates$snapshot[[1]]
candidates$feedback[[1]]

optimization_result(resumed)$extensions$meta_harness$frontier_ids
optimization_result(resumed)$extensions$meta_harness$events
optimization_result(resumed)$stop_reason
```

With `log_dir` in the teleprompter or `optimizer_control()`, evaluations are
also written through `TrialLog`.

Resuming restores candidate lineage, evaluation evidence, budgets, and
random-number state. It does not restore the agent's conversation or the
mcp-repl session; the resumed run uses the chat and runner you pass.

## Compose harnesses with Omni

The harnesses are ordinary teleprompters, so `Omni()` can run them as
explorers. Pass their agent and runner through the per-explorer arguments:

```{r}
omni <- Omni(
  metric = metric,
  explorers = list(
    research = AutoResearch(metric = metric, max_iterations = 6L),
    meta = MetaHarness(metric = metric, max_iterations = 4L)
  ),
  continuation = BootstrapFewShotWithRandomSearch(metric = metric)
)

compiled <- compile(
  qa_program,
  omni,
  trainset,
  valset = valset,
  .llm = task_chat,
  explorer_compile_args = list(
    research = list(.agent_llm = research_chat, runner = runner),
    meta = list(.agent_llm = research_chat, runner = runner)
  )
)
```

Do not use a harness as the continuation. Unless the seed wins, the
continuation receives an explorer's compiled program, which the harnesses
refuse to edit; Omni then records the error and keeps the exploration winner.
Keep `parallel = FALSE` (the default) with harness explorers: parallel mode
copies explorer arguments into separate worker processes, and a runner's
mcp-repl connection belongs to your session. See [Compose optimizers with
Omni](omni-meta-optimization.html) for the rest of the Omni workflow.

## Inspiration and differences

The design credits four open projects:

- Andrej Karpathy's
  [`autoresearch`](https://github.com/karpathy/autoresearch), for the explicit
  baseline, experiment, score, keep-or-revert loop;
- the AutoResearch and Meta-Harness engines of [GEPA
  optimize-anything](https://github.com/gepa-ai/gepa), for a common trusted
  evaluation boundary and optimizer composition;
- the [Meta-Harness paper](https://arxiv.org/abs/2603.28052), for the
  fresh-proposer outer loop; and
- Posit's [`mcp-repl`](https://github.com/posit-dev/mcp-repl), for persistent
  R execution with OS-level sandboxing.

dsprrr does not launch a coding-agent CLI, expose an HTTP evaluator, or ask
an agent to edit arbitrary files. It uses provider-neutral ellmer chats, typed
actions, complete module snapshots, the package's optimizer ledger, and
checkpoints. Benchmark these harnesses on your own task and budget; results
reported by the source projects do not automatically transfer to a different
model, metric, or program.
