---
title: "Agentic Optimization with AutoResearch and Meta-Harness"
description: >
  Run persistent or frontier-driven research agents against a trusted dsprrr
  evaluator, with Posit mcp-repl sandboxing, budgets, lineage, and checkpoints.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Agentic Optimization with AutoResearch and Meta-Harness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Most dsprrr teleprompters own both candidate generation and selection.
`AutoResearch()` and `MetaHarness()` split those responsibilities differently:
an LLM proposes experiments, while a trusted R evaluator applies validated
edits, runs the program, enforces budgets, and preserves the best result.

Both harnesses can jointly edit the instructions and templates of every
optimizable leaf in a module graph. This makes them useful when the search
problem is not "find one better prompt," but "repair the contract among several
pipeline stages."

## Choose the Search Owner

| Harness | Who owns the loop? | Agent memory | Candidate shape |
|---|---|---|---|
| `AutoResearch()` | One research agent | Persistent across experiments | One candidate per evaluate action |
| `MetaHarness()` | Trusted R outer loop | Fresh proposer each iteration | A bounded candidate batch |

`AutoResearch()` is the better fit when a researcher should pursue a line of
inquiry, use earlier results as working memory, branch, and decide when to stop.
`MetaHarness()` is the better fit when you want independent proposals,
explicit frontier selection, and less dependence on one conversation's hidden
state.

In both cases, dsprrr:

1. evaluates the unmodified program as a baseline;
2. gives every candidate a content fingerprint and parent id;
3. rejects unknown module paths and unsupported fields;
4. deduplicates equivalent program snapshots;
5. evaluates candidates through `optimizer_control()`; and
6. returns the best scored program, including when a budget stops the run.

## Keep Code Execution Sandboxed

The harnesses require an operating-system-sandboxed runner by default.
`mcp_repl_runner()` connects to Posit's open-source
[`mcp-repl`](https://github.com/posit-dev/mcp-repl), which provides a
persistent R session behind one MCP `repl(input, timeout_ms)` tool.

Install the executable and the R MCP client:

```sh
pipx install posit-mcp-repl
# or: uv tool install posit-mcp-repl
```

```{r}
install.packages("mcptools")
```

Then create the runner:

```{r}
runner <- mcp_repl_runner(
  sandbox = "workspace-write",
  timeout = 30
)
```

The default mcp-repl policy uses OS primitives, disables network access, and
limits writes to the workspace and runtime temp paths. The REPL remains alive
across calls, so an agent can load a package or construct an analysis once and
iterate on it. `runner$reset()` starts a clean session.

`r_code_runner()` is intentionally rejected while sandboxing is enabled. Its
callr subprocess is useful for trusted code, but it retains the host user's
permissions and is not a security boundary.

Set `sandbox = FALSE` on a harness only to remove the sandbox action entirely.
Any supplied runner is then ignored; the agent can still propose program edits
but cannot execute code.

## AutoResearch: One Persistent Research Session

Use separate chats for the program under test and the research agent. That
separates task traces, model choice, and cost attribution.

```{r}
library(dsprrr)
library(ellmer)

task_chat <- chat_openai(model = "gpt-4.1-mini")
research_chat <- chat_anthropic(model = "claude-sonnet-4-6")

metric <- metric_with_feedback(
  function(prediction, expected) {
    correct <- identical(as.character(prediction), expected$answer)
    list(
      score = as.numeric(correct),
      feedback = if (correct) {
        "Correct."
      } else {
        paste("Expected", expected$answer, "but received", prediction)
      }
    )
  },
  field = "answer"
)

research <- AutoResearch(
  metric = metric,
  max_iterations = 16L,
  patience = 5L,
  target_score = 0.95,
  max_context_examples = 20L,
  max_feedback_examples = 8L,
  seed = 42L
)

compiled <- compile(
  research,
  qa_program,
  trainset,
  valset = valset,
  .llm = task_chat,
  .agent_llm = research_chat,
  runner = runner,
  objective = paste(
    "Answer from supplied evidence.",
    "Do not invent facts when the context is insufficient."
  ),
  .cache = FALSE
)
```

On each turn the agent must choose one typed action:

- `sandbox`: run bounded R analysis against a context containing the editable
  program manifest, visible training rows, and current frontier;
- `propose`: branch from a known candidate and submit one or more complete
  instruction or template replacements; or
- `finish`: stop when another evaluation is unlikely to improve the result.

The agent owns the research policy, but it cannot write directly into the
module or evaluator. dsprrr re-validates and applies every candidate.

## Meta-Harness: Fresh Proposers, Trusted Selection

Meta-Harness creates a fresh proposer chat for each outer iteration. The
proposer sees persisted evidence rather than relying on earlier conversation
state:

```{r}
harness <- MetaHarness(
  metric = metric,
  max_iterations = 8L,
  max_candidates_per_iteration = 4L,
  frontier_size = 8L,
  patience = 3L,
  seed = 42L
)

compiled <- compile(
  harness,
  qa_program,
  trainset,
  valset = valset,
  .llm = task_chat,
  .agent_llm = research_chat,
  runner = runner,
  objective = "Improve factual accuracy without increasing abstention errors."
)
```

Within an iteration, the proposer may request a few sandbox steps before it
submits its batch. The outer loop then:

1. materializes each proposal from its declared parent;
2. rejects invalid or duplicate snapshots;
3. evaluates every unique candidate that fits the remaining budget;
4. updates the scored frontier; and
5. checkpoints the new lineage before starting a fresh proposer.

This is a useful boundary for coding-agent experiments: the agent can inspect
and hypothesize, but the scorer, budget, and accepted state remain host-owned.

## Joint Pipeline Edits

Candidate paths come from `named_parameters()` and remain stable within one
program graph:

```{r}
names(named_parameters(qa_pipeline, boundaries = "respect"))
#> [1] "$/steps/1" "$/steps/2" "$/steps/3"
```

A single candidate may replace fields at several paths. For example, a
retriever stage can promise a citation-bearing intermediate format while the
answer stage is changed to consume exactly that format. The candidate ledger
stores the complete resulting snapshot, not only the patch, so fingerprints,
resume behavior, and comparisons are deterministic.

Protected graph boundaries are respected. If you intentionally need to expose
an inner component, change the module graph boundary before compiling rather
than asking the agent to bypass it.

## Budget Every Kind of Work

The shared optimizer ledger accounts for baseline and candidate evaluations as
trials, program calls and tokens during evaluation, proposer calls and tokens,
metric calls, cost, elapsed time, and errors:

```{r}
control <- optimizer_control(
  max_trials = 25L,
  max_metric_calls = 200L,
  max_provider_calls = 250L,
  max_total_tokens = 500000L,
  max_cost = 10,
  max_elapsed_seconds = 1800,
  checkpoint_path = "meta-harness-checkpoint.rds"
)

compiled <- compile(
  harness,
  qa_program,
  trainset,
  valset = valset,
  .llm = task_chat,
  .agent_llm = research_chat,
  runner = runner,
  control = control
)
```

A finite provider, token, or cost cap fails closed when a custom chat cannot
report verified usage. A stopped run returns its best partial program and a
typed stop reason:

```{r}
compiled$config$optimizer$budget_summary
compiled$config$optimizer$stop_reason
compiled$config$optimizer$partial
```

## Resume and Audit a Run

Both harnesses support optimizer checkpoints. Resume with the same program,
data, metric, and harness configuration; resource limits may be raised:

```{r}
resumed <- compile(
  harness,
  qa_program,
  trainset,
  valset = valset,
  .llm = task_chat,
  .agent_llm = research_chat,
  runner = runner,
  control = optimizer_control(
    max_trials = 40L,
    max_cost = 20,
    checkpoint_path = "meta-harness-checkpoint.rds",
    resume = TRUE
  )
)
```

The candidate ledger is attached to the returned module:

```{r}
candidates <- resumed$config$optimizer$candidates
candidates[, c(
  "id",
  "parent_id",
  "name",
  "iteration",
  "score",
  "selected",
  "status"
)]

# Complete per-candidate program state and evaluation feedback:
candidates$snapshot[[1]]
candidates$feedback[[1]]

resumed$config$optimizer$frontier_ids
resumed$config$optimizer$events
resumed$config$optimizer$termination
```

With `log_dir` in the teleprompter or `optimizer_control()`, evaluations are
also written through `TrialLog`.

Checkpoint resume restores candidate lineage, evaluation evidence, budgets, and
random-number state. It starts a new chat and mcp-repl connection rather than
trying to serialize hidden provider or interpreter session state.

## Compose Harnesses with Omni

These are ordinary `Teleprompter` objects, so sequential `Omni()` runs can use
them as explorers or as the continuation stage. Pass their agent and runner
dependencies through the per-step argument lists:

```{r}
omni <- Omni(
  metric = metric,
  explorers = list(
    demos = BootstrapFewShotWithRandomSearch(metric = metric),
    meta = MetaHarness(metric = metric, max_iterations = 4L)
  ),
  continuation = AutoResearch(metric = metric, max_iterations = 6L)
)

compiled <- compile(
  omni,
  qa_program,
  trainset,
  valset = valset,
  .llm = task_chat,
  explorer_compile_args = list(
    meta = list(.agent_llm = research_chat, runner = runner)
  ),
  continuation_compile_args = list(
    .agent_llm = research_chat,
    runner = runner
  )
)
```

Do not use a live chat or runner with Omni's mirai `parallel = TRUE` mode.
Those stateful runtime objects are intentionally not serialized to workers.

## Inspiration and Differences

The design credits three open projects:

- Andrej Karpathy's
  [`autoresearch`](https://github.com/karpathy/autoresearch), for the explicit
  baseline, experiment, score, keep-or-revert loop;
- the
  [GEPA optimize-anything](https://github.com/gepa-ai/gepa) AutoResearch and
  Meta-Harness engines, for a common trusted evaluation boundary and
  optimizer composition;
- the
  [Meta-Harness paper](https://arxiv.org/abs/2603.28052), for the fresh-proposer
  outer-loop design; and
- Posit's [`mcp-repl`](https://github.com/posit-dev/mcp-repl), for persistent
  R execution with OS-level sandboxing.

dsprrr does not launch a particular coding-agent CLI, expose an HTTP evaluator,
or ask an agent to edit arbitrary files. It uses provider-neutral ellmer chats,
typed actions, canonical module snapshots, the package's optimizer ledger, and
safe program checkpoints. Benchmark these harnesses on your own task and
budget; results reported by the source projects do not automatically transfer
to a different model, metric, or program.
