---
title: "Flex: Optimize the Whole Program"
description: >
  Use Flex when optimization should decide which model calls, deterministic
  code, or tools a task needs—not only rewrite a prompt.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flex: Optimize the Whole Program}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Most dsprrr optimizers improve the instructions or examples inside a program
whose shape you chose. `flex()` makes the implementation itself optimizable.
GEPA can change how many predictors run, add a deterministic branch, or call a
tool you supplied.

That extra freedom is useful when you can score the outcome but do not yet know
the best decomposition. It is unnecessary when the workflow is already clear.

| What should change? | Use |
|---|---|
| Instructions or examples inside a known module | A regular module and teleprompter |
| A workflow whose steps your team should own | An explicit pipeline |
| The number or order of Predict and Chain-of-Thought steps | JSON Flex |
| The choice among R logic, predictors, and selected tools | Executable Flex |

## A router that learns when not to call a model

Suppose a support team routes tickets to four queues. Some tickets contain an
incident code with a known owner. Others need language-model judgment.

A conventional predictor sends every ticket to the model. The Flex objective
is more specific: keep routing accuracy, but avoid a predictor call when the
incident catalog already contains the answer.

### Define the task and its evidence

Use separate training and validation rows during search, then reserve a holdout
set for the final comparison:

```{r router-data}
library(dsprrr)

route_signature <- signature(
  "ticket -> queue: enum('database', 'security', 'payments', 'identity')",
  instructions = "Route each support ticket to its owning queue."
)

train <- data.frame(
  ticket = c(
    "[INC-DB-17] checkout failures in us-east",
    "Password reset links expire immediately",
    "Duplicate charge after a plan upgrade",
    "Queries time out after a schema migration"
  ),
  queue = c(
    "database", "identity", "payments", "database"
  )
)

validation <- data.frame(
  ticket = c(
    "[INC-SEC-9] unusual-login burst",
    "Card charged twice for one invoice"
  ),
  queue = c("security", "payments")
)

holdout <- data.frame(
  ticket = c(
    "[INC-PAY-4] capture queue is growing",
    "[INC-DB-17] replicas unavailable",
    "[INC-SEC-9] login spray detected",
    "Reset email arrives after its token expires",
    "Renewal created two invoices",
    "Read traffic fails after failover"
  ),
  queue = c(
    "payments", "database", "security", "identity", "payments", "database"
  )
)
```

The catalog is an ordinary R closure wrapped in an ellmer tool definition. Its
description tells GEPA what the tool returns; supplying it to Flex gives
generated code one named capability rather than the rest of the host session.

```{r router-tool}
lookup_incident_fn <- local({
  catalog <- c(
    "INC-DB-17" = "database",
    "INC-SEC-9" = "security",
    "INC-PAY-4" = "payments"
  )

  function(ticket) {
    hit <- names(catalog)[vapply(
      names(catalog),
      grepl,
      logical(1),
      x = ticket,
      fixed = TRUE
    )]
    if (!length(hit)) {
      return(list(found = FALSE))
    }
    list(found = TRUE, queue = unname(catalog[[hit[[1L]]]]))
  }
})

lookup_incident <- ellmer::tool(
  lookup_incident_fn,
  name = "lookup_incident",
  description = paste(
    "Find the owner of an incident code in a support ticket.",
    "Returns list(found = FALSE) when no known code appears; otherwise",
    "returns list(found = TRUE, queue = <owning queue>)."
  ),
  arguments = list(
    ticket = ellmer::type_string(
      description = "Complete support-ticket text."
    )
  )
)
```

### Start with one predictor

Executable Flex source defines a top-level `forward()` function. This baseline
always calls one predictor, even for a ticket whose incident code is already in
the catalog.

```{r router-baseline}
baseline_source <- paste(
  "router <- Predict(\"$outer\", instructions = \"Choose the owning queue.\")",
  "forward <- function(ticket) router(ticket = ticket)",
  sep = "\n"
)

sandbox_factory <- function() mcp_repl_runner(timeout = 45)

baseline <- flex(
  route_signature,
  module_src = baseline_source,
  tools = list(lookup_incident = lookup_incident),
  interpreter_factory = sandbox_factory,
  source_format = "r",
  max_predictor_calls = 1L,
  max_tool_calls = 1L
)
```

Install `mcptools` and [Posit
mcp-repl](https://github.com/posit-dev/mcp-repl) before running this example:

```r
install.packages("mcptools")
# In a shell:
# uv tool install posit-mcp-repl
```

mcp-repl's OS sandbox is the execution boundary for optimizer-authored R
source.

### Score the answer and the work

An exact-match metric alone cannot distinguish two correct programs. A
trace-aware metric can. Here, a correct route scores `1`; a correct route that
needed a predictor scores `0.9`; a wrong route scores `0`.

```{r router-metric}
route_metric <- metric_with_trace(
  function(prediction, expected, program_trace) {
    observed <- as.character(prediction$queue)
    wanted <- as.character(expected$queue)
    correct <- identical(observed, wanted)

    calls <- program_trace$metadata$predictor_calls
    if (is.null(calls)) {
      calls <- 0L
    }

    list(
      score = if (!correct) 0 else if (calls == 0L) 1 else 0.9,
      feedback = if (!correct) {
        sprintf("Expected '%s'; got '%s'.", wanted, observed)
      } else if (calls > 0L) {
        paste(
          "Correct, but a predictor ran.",
          "Prefer the incident catalog when it covers the ticket."
        )
      } else {
        "Correct and deterministic."
      }
    )
  },
  field = "queue"
)
```

### Optional: run a live search

The deterministic package test below is the reproducible proof. A live search
also lets GEPA propose complete `module_src` candidates, but remote model output
and runtime vary. Training rows generate feedback; validation rows decide which
candidate survives. Keep the first run deliberately small and bounded:

```{r router-compile}
llm <- ellmer::chat_openai(model = "gpt-5-mini")

optimized <- compile(
  baseline,
  GEPA(
    metric = route_metric,
    metric_threshold = 0.95,
    population_size = 2L,
    generations = 1L,
    selection = "current_best",
    seed = 20260810L,
    track_best_outputs = TRUE,
    verbose = FALSE
  ),
  trainset = train,
  valset = validation,
  .llm = llm,
  control = optimizer_control(
    max_metric_calls = 30L,
    max_provider_calls = 30L,
    max_elapsed_seconds = 300,
    num_threads = 1L,
    progress = FALSE
  )
)
```

Remote model output is stochastic, so a seed does not guarantee identical
source. Judge the result on held-out behavior, not whether it reproduces one
particular program string. The limits stop the search and return its best
partial result after 30 metric calls, 30 provider calls, or five active minutes.

### Compare quality and calls

Report ordinary accuracy separately from the efficiency objective:

```{r router-evaluate}
accuracy <- metric_exact_match(field = "queue")

before <- evaluate(
  baseline,
  holdout,
  metric = accuracy,
  .llm = llm,
  .progress = FALSE
)
after <- evaluate(
  optimized,
  holdout,
  metric = accuracy,
  .llm = llm,
  .progress = FALSE
)

count_calls <- function(result, field) {
  sum(vapply(result$metadata, function(metadata) {
    value <- metadata[[field]]
    if (is.null(value)) 0L else as.integer(value)
  }, integer(1)))
}

comparison <- data.frame(
  program = c("baseline", "optimized"),
  accuracy = c(before$mean_score, after$mean_score),
  predictor_calls = c(
    count_calls(before, "predictor_calls"),
    count_calls(after, "predictor_calls")
  ),
  tool_calls = c(
    count_calls(before, "tool_calls"),
    count_calls(after, "tool_calls")
  )
)
comparison
```

The package's deterministic regressions fix the GEPA proposal and model
responses. One compiles the baseline and selects the hybrid on disjoint
training and validation rows; the other runs both reviewed programs on six
held-out tickets:

| Reviewed program | Holdout accuracy | Predictor calls | Tool calls |
|---|---:|---:|---:|
| Baseline | 1.00 | 6 | 0 |
| Hybrid | 1.00 | 3 | 3 |

That replay verifies proposal, selection, runtime, and call accounting; it is
not a promise that every stochastic GEPA run will discover the same source. For
a live run, accept the candidate only when held-out quality is at least as good
and predictor calls fall:

```{r router-acceptance}
stopifnot(
  after$mean_score >= before$mean_score,
  count_calls(after, "predictor_calls") <
    count_calls(before, "predictor_calls")
)
```

The reveal is simple: **Flex did not find a better prompt. It found that half
the tickets did not need one.**

## Read the program GEPA selected

Always inspect executable source before promoting it. A useful candidate for
this task has a deterministic catalog path and a predictor fallback:

```{r router-winner}
optimized$module_src
```

The deterministic replay selects this shape:

```r
fallback <- Predict(
  "$outer",
  instructions = "Route tickets without a known incident code."
)
forward <- function(ticket) {
  if (grepl("INC-", ticket, fixed = TRUE)) {
    hit <- lookup_incident(ticket = ticket)
    if (isTRUE(hit$found)) {
      return(Prediction(queue = hit$queue))
    }
  }
  fallback(ticket = ticket)
}
```

`module_src` is the program Flex optimized—not an explanation of what happened.
The runtime trace supplies the evidence: predictor calls, tool calls, tokens,
and the exact source used for each row. GEPA also records the winning candidate
ID and validation score:

```{r router-audit}
optimized$config$optimizer$best_candidate_id
optimized$config$optimizer$best_scores
optimized$config$optimizer$per_val_instance_best_candidates
```

Persist callable tools and factories by registry name rather than embedding
them in the artifact:

```{r router-persist}
board <- pins::board_folder("pins")
router_registry <- list(
  lookup_incident = lookup_incident,
  sandbox_factory = sandbox_factory
)

pin_module_config(
  board,
  "issue-router-flex",
  optimized,
  registry = router_registry
)

restored <- pins::pin_read(board, "issue-router-flex") |>
  restore_module_config(registry = router_registry)
```

Use the same registry names when restoring. Saving and restoring the artifact
does not invoke the factory. The registry supplies executable authority, so
version and review its definitions as code. Record the model, package versions,
data split, metric, and seed beside the pin; pass the model again when running
the restored fallback path.

## Choose the smaller source language that fits

Flex has two source modes because not every structural question needs code.

| Mode | Use it when | Execution boundary |
|---|---|---|
| JSON (default) | GEPA may vary a bounded graph of Predict and Chain-of-Thought steps | Parsed and type-checked as data; no code interpreter |
| R (opt-in) | Candidates need branches, deterministic computation, dynamic predictors, or named tools | Fresh runner from `interpreter_factory`; enforced sandbox required by default |

JSON sources use backward references between ordered steps and map their final
values to the outer signature. See [`flex()`](../reference/flex.html) for the
schema and limits. Start there unless the optimization question truly needs R
control flow or tools.

Executable Flex exposes a small set of predictor constructors plus the named
tools you supply. Those tools execute on the host and keep their closure
environments, so treat each one as a privileged capability.

`max_predictor_calls` limits predictor invocations across the bridge, and
`max_tool_calls` limits direct host-tool requests. A single agentic predictor
such as ReAct, CodeAct, or RLM can perform additional work internally; configure
that primitive's own limits as well.

The guest source runs again from the beginning after each bridged request while
recorded host responses are replayed. Keep guest-side computation pure, avoid
external side effects, and bound loops explicitly.

Generated R must run in an enforced sandbox. `r_code_runner()` isolates a
subprocess but is for source you already trust; do not disable
`require_sandbox` for optimizer-authored code. Executable Flex is synchronous,
so use the default sequential execution; specialized token streaming is not
available.

## When Flex is the wrong tool

Do not use Flex merely because a task contains more than one step. Prefer a
regular module when prompt or demonstration optimization is enough. Prefer an
explicit pipeline when people should design and review the workflow. If a
decision is entirely deterministic and already known, write an R function.

Use Flex for the unresolved middle: the outcome is measurable, several
implementation strategies are plausible, and the choice of model calls, code,
or tools is itself what you want to optimize.

For algorithm details, see [Advanced
Optimization](../articles/advanced-optimization.html). For intentional API
differences and remaining gaps, see [dsprrr vs.
DSPy](../articles/dspy-comparison.html).
