---
title: "Structural Optimization with Flex"
description: >
  Build and validate experimental predictor graphs with dsprrr's constrained
  JSON intermediate representation.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structural Optimization with Flex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`flex()` is an experimental module whose predictor structure is part of the
program state. It gives optimizers a structural edit surface without giving
them an arbitrary code-execution surface.

That distinction is important. dsprrr Flex source is a bounded, versioned JSON
intermediate representation (IR), parsed strictly as data. It is never
evaluated as R or Python. This is intentionally narrower than DSPy's
experimental Flex module and is not a claim of full DSPy Flex parity.

## Start with the safe baseline

With no `module_src`, `flex()` creates one Predict step over the outer
signature:

```{r baseline}
library(dsprrr)

program <- flex("question -> answer")
program$module_src

result <- run(
  program,
  question = "Why does ice float?",
  .llm = ellmer::chat_openai()
)
```

The `module_src` active binding is read-only and canonical. Equivalent input
JSON is normalized to one stable representation, which supports reliable
fingerprints, comparisons, and optimizer candidates.

## Describe a two-step plan

Version 1 source has exactly three top-level fields:

- `schema_version`, currently the number `1`;
- `steps`, a non-empty ordered array; and
- `outputs`, a mapping for every output in the outer signature.

Each step declares a unique `name`, an allowlisted `primitive`, a `signature`,
and an `inputs` mapping. Optional `instructions` refine the step. Input and
output references may point to an outer input (`$input.<field>`) or an output
from an earlier step (`$step.<name>.<field>`).

```{r two-step}
source <- jsonlite::toJSON(
  list(
    schema_version = 1,
    steps = list(
      list(
        name = "draft",
        primitive = "chain_of_thought",
        signature = "question -> draft",
        instructions = "Work out the relevant causal mechanism.",
        inputs = list(question = "$input.question")
      ),
      list(
        name = "final",
        primitive = "predict",
        signature = "draft -> answer",
        instructions = "Give a concise answer for a general reader.",
        inputs = list(draft = "$step.draft.draft")
      )
    ),
    outputs = list(answer = "$step.final.answer")
  ),
  auto_unbox = TRUE
)

program <- flex(
  "question -> answer",
  module_src = source,
  max_predictor_calls = 4
)
```

Version 1 permits only `predict` and `chain_of_thought`. Step names and
references follow a restricted grammar, references can only point backwards,
and every step-signature input needs exactly one compatible binding while every
outer output needs exactly one compatible mapping. A plan may reuse or ignore
outer inputs, although runtime inputs must still match the outer signature.
`max_predictor_calls` bounds graph size before execution.
Version 1 supports ellmer string, number, integer, boolean, enum, array, and
non-empty object field types. Opaque `TypeJsonSchema` values and empty objects
are rejected explicitly; the IR does not guess whether an arbitrary JSON
Schema is assignable.

## Replace a plan transactionally

Use `bind()` to replace the active source:

```{r bind}
program$bind(source)
program$module_src # The validated, canonical source is now active
```

dsprrr parses, validates, type-checks, and compiles the complete candidate
before changing the module. If validation fails, the previous source and plan
remain active. Direct assignment is rejected:

```{r read-only}
program$module_src <- source
# Error: module_src is read-only; use bind()
```

Optimizer infrastructure can use
`program$apply_optimization_params(list(module_src = candidate))` as the same
transactional boundary. The module advertises its graph as an optimization
parameter, but an optimizer only changes `module_src` if that optimizer
explicitly supports structural candidates. Ordinary prompt optimization does
not silently rewrite the plan.

## Search complete plans with GEPA-lite

`GEPA()` explicitly supports Flex structural candidates. For a program that
contains Flex leaves, each candidate is a complete set of ordinary instruction
values and complete canonical `module_src` values. Source text is never spliced
or evaluated: dsprrr copies the program, validates and binds each source
transactionally, then evaluates the resulting program.

```{r gepa-flex}
metric <- metric_exact_match(field = "answer")
optimizer <- GEPA(
  metric = metric,
  population_size = 6L,
  generations = 2L,
  seed = 42L
)

optimized <- compile(
  optimizer,
  program,
  trainset = question_answer_data,
  .llm = llm
)

optimized$module_src
optimized$config$optimizer$component_semantics
```

The structured proposer sees row-aligned inputs, expected output, prediction,
and metric feedback. An invalid proposal receives aligned failure records for
audit and budget accounting but is never selectable. GEPA-lite's Pareto
selection ranks complete programs. It does not reproduce DSPy's independent
per-component Pareto frontiers or inference-time search; those limits are
recorded in `config$optimizer$component_semantics`.

Freezing a module removes its components from this search. If the complete
graph has no mutable components, GEPA returns a fresh unchanged program without
calling the proposer or evaluator and records
`skip_reason = "no_mutable_components"` in optimizer metadata.

## Understand the safety boundary

Flex controls what predictors run and how typed values flow between them. It
does not make the LLM calls themselves trusted, and it does not provide a
general programming environment.

In particular, Flex source cannot:

- evaluate R or Python expressions;
- call arbitrary functions, tools, files, or network services;
- construct module classes outside the allowlist;
- reference a later step or create a cycle; or
- exceed its source-size and predictor-call bounds.

If a task needs generated R code, use `program_of_thought()`, `code_act()`, or
`rlm_module()` with an explicitly selected code runner. Those modules have a
separate execution and lifecycle contract; Flex does not bypass it.

Flex currently runs through synchronous `run()`. `run_async()`,
`stream_async()`, and module `$stream()` reject a program graph containing a
Flex leaf before making a provider call. `run_stream()` without a matching
token listener preserves the ordinary synchronous `forward()` path; a matching
token-stream request is rejected before provider work so it cannot bypass
Flex's validated step workflow.

## Choose Flex deliberately

Use `flex()` when the search question is structural: for example, whether a
task benefits from an intermediate draft or a Chain-of-Thought step. Use a
regular module when one known topology plus optimized instructions is enough.
Use an explicit pipeline when humans should own and review the topology.

Because Flex and its version 1 schema are experimental, persist the canonical
`module_src`, pin package versions, and revalidate a candidate when upgrading
dsprrr.
