---
title: "How the Recursive Language Model (RLM) Works"
description: >
  The RLM execution contract: external context, iterative R code, recursive
  queries, typed submission, runner ownership, and bounded evidence.
author: "James H. Wade"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{How the Recursive Language Model (RLM) Works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

An RLM changes where context lives. A regular predictor serializes its inputs
into the model request. An RLM keeps the inputs in an R execution environment
and initially shows the model only their names, types, sizes, and previews. The
model writes R code to select evidence, observes bounded output, and repeats
until it can submit the signature's typed outputs.

That makes RLM an inference strategy for discovery, not a more elaborate way to
answer every prompt.

## Decide before you build

| Need | Prefer |
|---|---|
| Short context and a direct answer | `Predict` |
| Hard reasoning over already selected context | `ChainOfThought` |
| A known calculation expressed as generated R code | `ProgramOfThought` |
| External actions or data acquisition through tools | `ReAct` or `CodeAct` |
| Retrieval from a prepared index | RAG |
| Adaptive exploration of large or irregular in-memory objects | RLM |
| Search for a reusable implementation across labeled cases | Flex |

Use RLM when the model must discover *how* to inspect the supplied context.
Skip it when a direct R function, SQL query, retrieval rule, or ordinary module
already describes the path. Every RLM iteration adds another model call and
another code execution.

The [release-regression tutorial](tutorial-rlm-dsprrr.html) shows the intended
shape: use R to identify an anomalous cohort, inspect one relevant change
record, and validate the answer against an independent calculation.

## One invocation, step by step

An invocation has six observable stages.

1. `rlm_module()` reads the signature and fixes the iteration, recursive-call,
   and output budgets.
2. The runner loads the signature inputs into `.context`. A persistent runner
   keeps its execution state for the whole invocation.
3. The outer model receives variable metadata and the trajectory so far, then
   proposes one R code block.
4. The runner executes that code. After any stricter runner transport limit,
   bounded head-and-tail output enters the next model turn; the full source
   object does not.
5. A valid `SUBMIT(...)` ends the loop. An invalid typed submission becomes
   repairable feedback instead.
6. If the loop spends `max_iters` without a valid submission, a final extraction
   pass attempts to produce the best typed answer supported by the trajectory;
   provider or type-validation failure remains terminal.

The core interface is small:

```{r basic-module}
library(dsprrr)

explorer <- rlm_module(
  "document, question -> answer, evidence",
  interpreter_factory = function() {
    r_code_runner(timeout = 30, persistent = TRUE)
  },
  max_iters = 10,
  max_llm_calls = 6,
  max_output_chars = 10000
)
```

All signature inputs are available under `.context`:

```r
nchar(.context$document)
search(.context$document, "renewal|termination")
peek(.context$document, start = 12000, end = 15000)
```

`peek()` and `search()` are conveniences, not a separate query language. The
generated program can use ordinary R functions appropriate to the value in
`.context`.

## State belongs to one invocation

RLM code is iterative. A value calculated on one turn should remain available
on the next, so the runner used for an invocation must provide persistent state.

The recommended lifecycle is a zero-argument factory:

```{r factory-lifecycle}
explorer <- rlm_module(
  "records, question -> answer",
  interpreter_factory = function() mcp_repl_runner(),
  max_iters = 10
)
```

dsprrr creates one runner from the factory, uses it for that invocation, and
closes it exactly once on success, error, or interrupt. Factory-backed modules
can create isolated runners for concurrent invocations.

A directly supplied `runner` is caller-owned. dsprrr neither resets nor closes
it. RLM attempts to remove staged context and invocation-private variables when
a call ends. Inspect cleanup warnings and close or reset the runner if cleanup
fails. No cleanup can undo arbitrary file, option, or network side effects from
guest code. Reuse a caller-owned runner only sequentially and within one trust
boundary:

```{r caller-owned}
local({
  runner <- r_code_runner(timeout = 30, persistent = TRUE)
  on.exit(runner$close(), add = TRUE)
  explorer <- rlm_module(
    "records, question -> answer",
    runner = runner
  )
  # Run explorer sequentially inside this scope.
})
```

## Recursive calls happen through the host

The outer model is responsible for planning and code. Generated code can ask a
model to interpret a focused slice:

```r
candidate <- subset(.context$notes, component == "checkout-auth")
interpretation <- llm_query(
  "Which change could reduce successful token refreshes?",
  paste(candidate$note, collapse = "\n")
)
SUBMIT(answer = interpretation)
```

The replay bridge does not pass model credentials into the recursive call path.
Guest access to environment variables and other host resources still depends
on the selected runner: only a verified sandbox enforces that boundary. The
guest emits a nonce-bound, schema-checked request, dsprrr calls the sub-model in
the host process, and the code evaluation is replayed with the response.
Ordinary R assignments are committed only after the replay completes, and
bridged host tools execute once. Direct file, network, or other external side
effects before a query cannot be rolled back and may repeat, so keep pre-query
guest work read-only.

Bridge requests use unclassed JSON-compatible values. Missing values, `NaN`,
infinities, and classed R objects are rejected instead of being silently
coerced. A ToolDef schema helps the model form a call; the host function must
still enforce domain-specific constraints before acting.

`sub_lm = NULL` inherits the outer model passed to `run()`. Supply a separate
chat when narrow interpretation can use a cheaper model:

```{r sub-lm}
explorer <- rlm_module(
  "documents, question -> answer",
  interpreter_factory = function() mcp_repl_runner(),
  sub_lm = ellmer::chat_openai(),
  max_llm_calls = 8
)
```

`llm_query_batched()` runs independent focused questions concurrently and
returns results in input order. A per-request provider failure becomes an
ordered `[ERROR] ...` slot. An unexpected exception, malformed batch result, or
transport-level failure terminates the invocation instead of masquerading as a
model answer. Every requested prompt consumes one unit of `max_llm_calls`. Set
that budget from the economics of the task, not from the largest number the
provider permits.

## Submission is part of the contract

The signature defines both the names and types accepted by `SUBMIT()`:

```{r typed-submit}
typed <- rlm_module(
  paste(
    "logs ->",
    "error_count: integer,",
    "severity: enum('low', 'high'),",
    "evidence: array(string)"
  ),
  interpreter_factory = function() mcp_repl_runner()
)
```

Generated code can submit named values:

```r
SUBMIT(
  error_count = 7L,
  severity = "high",
  evidence = c("AUTH-401 increased", "Refresh retries fell to zero")
)
```

Missing required fields, extra fields, and incompatible values do not silently
become a final answer. Optional fields may be omitted. The validation message
joins the trajectory, giving the model a chance to repair its submission.
Positional submission is supported, but named fields are easier to audit.

RLM strictly validates explicit ellmer string, number, integer, boolean, enum,
array, and object types, including nested combinations. It rejects opaque
`TypeJsonSchema` output nodes at construction because accepting them without a
JSON Schema validator would make the repair contract misleading. Express the
output with ellmer's explicit type constructors when using RLM.

If `max_iters` is exhausted, fallback extraction reads the trajectory and
attempts to return the signature fields. Provider or output-validation failure
still terminates the invocation. Treat a successful fallback as degraded
completion: it is useful, but it is evidence that the exploration budget or
instructions may need work.

## Output is bounded evidence

`max_output_chars` limits how much runner output from each execution enters
model history. The default 10,000-character module display retains the head and
tail. A runner may impose a stricter limit first: managed MCP rejects file/pager
compaction for RLM rather than treating an incomplete preview as evidence. The
module bound is not permission to return arbitrarily large control payloads.

Ask for structured output when the trajectory matters:

```{r structured-result}
result <- run(
  explorer,
  records = records,
  question = "Which change preceded the failure spike?",
  .llm = ellmer::chat_openai(),
  .return_format = "structured"
)

result$output
result$metadata$repl_history
result$metadata$iterations
result$metadata$llm_calls
result$metadata$runner_policy
```

Each trajectory entry records the proposed reasoning, executed code, displayed
output, success state, and whether a final submission occurred. Prefer the
returned metadata over mutable module state for async or batch work.

### One investigation or many

`run()` always stages each RLM input as one REPL variable. Its R length does not
imply a batch: an atomic vector, list, matrix, data frame, or fitted model is one
context object for one investigation. This matches the way generated code reads
the inputs through `.context` and avoids silently splitting irregular objects.

Use `run_dataset()` for multiple investigations. Each data-frame row is one RLM
invocation; put rich per-row objects in list-columns. A factory-backed RLM owns
one fresh runner per row and can use isolated mirai execution. A caller-owned
runner is reused sequentially and rejects concurrent dataset execution.

The counters separate workflow steps from paid provider work.
`action_calls`, `recursive_calls`, and `extraction_calls` count logical RLM
operations. `provider_calls` sums the verified provider turns behind them, with
component fields for action, recursive, and extraction calls. A cached action
or extraction contributes zero provider calls and zero current-run usage. If a
backend cannot prove every contributing turn, provider, token, and cost totals
remain `NA` instead of reporting a partial total.

## Choose the execution boundary

RLM executes model-generated code. Runner choice is therefore part of the
program's security and data contract.

| Runner path | Strength | Limitation |
|---|---|---|
| Managed `mcp_repl_runner()` | OS sandbox, network disabled, fresh factory-owned invocation | Workspace writes remain allowed; bounded transport is intended for compact JSON-compatible context |
| Persistent `r_code_runner()` | Preserves rich R objects in a long-lived callr process | Not a security sandbox; process retains the host user's permissions |
| Custom runner | Can provide a remote sandbox or domain-specific resource loading | Must implement and truthfully report the dsprrr runner policy |

The one-call `rlm()` helper chooses a fresh managed MCP runner by default.
Install the suggested R package and external executable first:

```sh
R -q -e 'install.packages("mcptools")'
uv tool install posit-mcp-repl
```

Then run a compact investigation:

```{r safe-convenience}
answer <- rlm(
  "document, question -> answer",
  document = "Owner: team-a\nObligation: rotate signing keys quarterly",
  question = "Which obligations have no owner?",
  .llm = ellmer::chat_openai(),
  .max_iterations = 4L,
  .max_llm_calls = 0L
)
```

The managed default runs model-generated code under an OS sandbox with network
access disabled; it can still mutate allowed workspace files. The final
JSON-RPC request must fit the 7 KB wire bound. When the raw request is too
large, dsprrr first tries a gzip/base64 wrapper and rejects it before execution
only if that request still does not fit.
`SUBMIT()` values, recursive-query context slices, and host-tool arguments also
cross a 3,000-byte encoded control-frame boundary. Replayed query and tool
results instead count against the next 7 KB raw-or-compressed request. Large
data frames and fitted models can use an explicit trusted
`r_code_runner(persistent = TRUE)`. Live external-pointer resources generally
need a child-side resource loader or a custom runner rather than callr
serialization.

The runner boundary is not a privacy guarantee. RLM automatically sends up to
1,000 characters of a structural preview for each input value to the outer
model; `str()`-style previews can include sample values. Anything printed into
the trajectory, passed to `llm_query()`, or returned in the final answer can
also be sent to the configured provider. Declared host tools run in the dsprrr
host process, outside the guest sandbox, with the host's permissions.

Action and fallback predictor requests also follow dsprrr's response-cache
configuration, which can include memory and disk. For sensitive investigations,
pass `.cache = FALSE` to `run()` or configure a memory-only or disabled cache.
Recursive `llm_query()` calls and runner execution are not response-cached.

## Repairable and terminal failures

Ordinary R errors are part of exploration. A malformed regular expression or
missing column can be shown to the model so it can try a corrected expression.

Interpreter startup, transport, protocol, and shutdown failures are different.
They invalidate the execution boundary and terminate the invocation. dsprrr
does not disguise those failures as an empty observation or reuse a runner whose
state is uncertain.

Three budgets keep the ordinary loop finite:

| Argument | Bounds |
|---|---|
| `max_iters` | Outer model/code/observation turns before fallback |
| `max_llm_calls` | Host-side recursive model calls; batch items count separately |
| `max_output_chars` | Head-and-tail execution output retained per turn |

The host bridge also enforces an internal safety ceiling of 1,000 declared tool
calls in one generated R step. This is a protocol guard, not a recommended
working budget; generated steps should make a small, auditable number of calls.

Start small, inspect the returned trajectory, and increase a budget only when
the trace shows useful unfinished work.

## Relationship to DSPy

RLMs were introduced by Zhang, Kraska, and Khattab [@zhang2025rlm]. DSPy makes
the pattern a module with a per-invocation interpreter, recursive query tools,
typed final output, and trajectory metadata. dsprrr follows those execution
contracts while using R and ellmer:

- inputs appear as `.context$name` and generated programs are R;
- `sub_lm = NULL` inherits the invocation's outer ellmer chat;
- a factory creates one persistent runner per invocation; a caller-owned
  persistent runner can be reused sequentially;
- invalid typed submissions are repairable observations;
- recursive calls and host tools cross a validated replay boundary; and
- `run(..., .return_format = "structured")` returns outputs with trajectory and
  runner evidence.

The R implementation is not a line-by-line port. In particular, runner and
serialization capabilities determine which R objects can cross into an
execution environment. Those limits should be visible and testable rather than
described as unlimited context.

## The practical rule

Use RLM for high-value questions where the useful slice and computation are not
known beforehand. Inspect its trajectory as seriously as its answer. When the
same exploration pattern repeats, replace discovery with a deterministic R
function, a dsprrr pipeline, or an optimized executable program.

For a complete deterministic example, continue to
[Investigate a Release Regression with an RLM](tutorial-rlm-dsprrr.html).

## References
