---
title: "How the Recursive Language Model (RLM) Works"
description: >
  From the research origins to dsprrr's R implementation: what Recursive
  Language Models are, why they matter, and how the internals work.
author: "James H. Wade"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{How the Recursive Language Model (RLM) Works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Code blocks in this article are excerpts from the source, not runnable
  # examples. The companion tutorial (tutorial-rlm-dsprrr) uses VCR cassettes
  # for evaluated examples.
  eval = FALSE
)
```

## The Problem: Context Rot

Modern LLMs accept enormous context windows. GPT-5 handles a million tokens; Gemini stretches to two million. But bigger windows do not solve the fundamental problem.

As context grows, performance degrades: details get lost and answers go wrong. The MIT researchers who introduced Recursive Language Models call this *context rot*, the empirical observation that output quality deteriorates as prompts grow, even when the relevant information is technically within the window [see @zhang2025rlm]. The model misses what it needs with increasing frequency as input length grows.

And there is no adaptive retrieval. The model cannot decide to re-read section 14 after discovering something relevant in section 42. It processes the entire input in one pass and produces output from whatever signal survived.

## The Insight: Context as Environment

The core idea behind RLMs is simple: *don't put the context in the prompt*. Instead, store it as a variable in a programming environment and let the model write code to explore it.

A traditional call looks like this:

```r
llm$chat(paste("Summarize this document:", huge_document))
```

An RLM inverts the relationship. The document lives outside the model as a variable in an R session, and the model generates code to interact with it. When you call `run()`, dsprrr provides a REPL: the model writes R code, dsprrr executes it in a subprocess, and the printed output feeds back into the next iteration. A typical exploration might look like this:

```r
# The model generates and executes code like this:
intro <- peek(.context$document, 1, 2000)
findings <- search(.context$document, "\\b(conclusion|finding|result)\\b")
section_42 <- peek(.context$document, 85000, 90000)
SUBMIT(answer = "The document concludes that...")
```

The shift is from treating context as *input* to treating it as *environment*. The model reads what it needs, skips what it doesn't, and revisits sections as its picture of the data develops.

In dsprrr's API, the module receives input arguments (e.g., `question`), holds context variables (e.g., `document`) outside the prompt, and exposes `llm_query()` so the model can delegate sub-questions to a secondary model from generated code. In the paper's notation [@zhang2025rlm], these correspond to a query $q$, context $C$, and a recursive tool call $\text{RLM}_M(\hat{q}, \hat{C})$ that spawns an isolated sub-instance with a new query and a transformed slice of the context.

## Origin and Ecosystem

RLMs were introduced by Alex Zhang, Tim Kraska, and Omar Khattab at MIT [@zhang2025rlm]. On BrowseComp-Plus (a benchmark with 6--11 million token inputs), standard models scored 0% while an RLM powered by GPT-5 achieved 91.33%. That comparison is less "RLM beats prompting" than "RLM makes previously intractable tasks tractable"; inputs that large exceed every current model's context window. The fairer apples-to-apples result is that their post-trained RLM-Qwen3-8B outperformed the base Qwen3-8B by 28.3% on average across long-context tasks.

The idea has since spread quickly. DSPy integrated RLMs as a first-class module ([`dspy.RLM`](https://dspy.ai/api/modules/RLM/)) in version 3.1.2+, using a Pyodide WASM sandbox for code execution. Google's Agent Development Kit [re-implemented the pattern](https://medium.com/google-cloud/recursive-language-models-in-adk-d9dc736f0478) with Gemini models. The [official `rlm` Python package](https://github.com/alexzhang13/rlm), a [community implementation](https://github.com/ysz/recursive-llm), and [Prime Intellect's research program](https://www.primeintellect.ai/blog/rlm) round out the ecosystem. The [comparison table below](#how-dsprrrs-rlm-compares) summarizes the key differences.

dsprrr's `rlm_module()` brings the same approach to R, using R as the REPL
language instead of Python and structured outputs via
[ellmer](https://ellmer.tidyverse.org/). Execution is delegated to the selected
runner backend: the built-in `r_code_runner()` uses
[callr](https://callr.r-lib.org/), while the default managed
`mcp_repl_runner()` uses an OS-sandboxed persistent session.

## How dsprrr Implements RLM

The rest of this article walks through dsprrr's implementation. For a practical example, see `vignette("tutorial-rlm-dsprrr", package = "dsprrr")`, which uses `rlm_module()` to trace a theming bug across the bslib, shiny, and brand.yml codebases.

### The User-Facing API

Creating an RLM module requires a signature plus exactly one execution binding.
To reuse a caller-owned runner object, pass `runner`:

```{r}
library(dsprrr)

runner <- r_code_runner(timeout = 30)

rlm <- rlm_module(
  signature = "document, question -> answer",
  runner = runner
)
```

For tasks that benefit from recursive sub-queries, you can wire up a secondary model:

```{r}
rlm <- rlm_module(
  signature = "document, question -> answer",
  runner = runner,
  sub_lm = ellmer::chat_openai(model = "gpt-5-mini"),
  max_llm_calls = 10
)
```

You can also inject custom R functions as tools available in the REPL. The factory validates these against reserved names (`SUBMIT`, `peek`, `search`, `llm_query`, etc.) to prevent collisions.

### The REPL Loop

When you call `run(rlm, ...)`, dsprrr dispatches to the module's internal `forward()` method. This is where the REPL loop lives. A `PredictModule`'s `forward()` makes a single call; the RLM module loops, up to `max_iterations` times (default 20):

```{r}
# From R/module-rlm.R (error handling, sub-query interception, and
# fallback extraction omitted -- see those sections below)
for (iter in seq_len(self$max_iterations)) {
  # Build prompt including all previous iterations
  prompt <- private$build_iteration_prompt(system_prompt, history, iter)

  # Ask the model to generate R code
  response <- private$get_code_response(llm, prompt)

  # Execute code in isolated subprocess with RLM tools injected
  exec_result <- private$execute_with_rlm_tools(
    response$code,
    inputs,
    call_counter
  )

  # Record in history -- the model sees this on the next iteration
  history[[iter]] <- list(
    iteration = iter,
    reasoning = response$reasoning,
    code = response$code,
    output = exec_result$formatted_output,
    success = exec_result$success,
    is_final = exec_result$is_final
  )

  # SUBMIT() terminates the loop
  if (exec_result$is_final) {
    final_answer <- exec_result$final_value
    break
  }
}
```

Each iteration produces a structured response with two fields: `reasoning` (the model's explanation of its plan for that step) and `code` (R code to execute). This uses ellmer's structured output support:

```{r}
# From R/module-rlm.R -- structured code generation
output_type <- ellmer::type_object(
  reasoning = ellmer::type_string("Your thought process for this step"),
  code = ellmer::type_string("R code to execute")
)

result <- llm$chat_structured(prompt, type = output_type)
```

The accumulated history gives the model a growing record of what it has tried. Failed executions are included. The model sees its own errors and can correct course.

### REPL Tools

Before each execution, dsprrr injects a "prelude" that defines the tools available in the subprocess. These are defined in `R/rlm-tools.R`:

`peek(var, start, end)` views a slice of a variable. It dispatches on the input type: for a character vector, `start` and `end` are element indices; for a single string, they are character positions:

```{r}
# From R/rlm-tools.R
peek <- function(var, start = 1L, end = 1000L) {
  if (!is.character(var)) {
    var <- as.character(var)
  }
  if (length(var) > 1) {
    return(var[max(1L, start):min(length(var), end)])
  }
  substr(var, max(1L, start), min(nchar(var), end))
}
```

`search(var, pattern)` runs a Perl-compatible regex against a variable and returns all matches:

```{r}
# From R/rlm-tools.R
search <- function(var, pattern, ignore_case = FALSE) {
  if (!is.character(var)) {
    var <- as.character(var)
  }
  if (length(var) > 1) {
    var <- paste(var, collapse = "\n")
  }
  matches <- regmatches(
    var,
    gregexpr(pattern, var, ignore.case = ignore_case, perl = TRUE)
  )
  unlist(matches)
}
```

`SUBMIT(...)` terminates the loop and returns the final answer. It validates that the provided values match the signature's output fields, supporting both positional (`SUBMIT("my answer")`) and named (`SUBMIT(answer = "my answer")`) arguments:

```{r}
# From R/rlm-tools.R -- simplified; see source for full validation logic
SUBMIT <- function(...) {
  args <- list(...)
  arg_names <- names(args)
  has_any_names <- any(nzchar(arg_names %||% ""))

  if (!has_any_names) {
    # Positional: match by order against signature output fields
    names(args) <- .rlm_output_fields
  } else {
    # Named: validate that all required fields are present
    missing <- setdiff(.rlm_output_fields, arg_names)
    if (length(missing) > 0) {
      stop("SUBMIT() missing outputs: ", paste(missing, collapse = ", "))
    }
  }

  class(args) <- c("rlm_final", class(args))
  args
}
```

The `.rlm_output_fields` variable is not magic. It is injected into the subprocess by the same prelude that defines `peek()`, `search()`, and `SUBMIT()` itself. The prelude reads the signature's output field names and writes them as a character vector at the top of the execution script.

The `rlm_final` class is a sentinel: when the parent process sees it in the subprocess result, it exits the loop and extracts the answer.

### Runner-Selected Isolation

The example in this article uses `RCodeRunner`, so each generated code fragment
runs in an isolated R subprocess via `callr::r()`. The `RCodeRunner` class in
`R/r-code-runner.R` handles that backend:

```{r}
# From R/r-code-runner.R -- subprocess execution (simplified)
exec_result <- callr::r(
  func = private$execution_wrapper,
  args = list(code = code, context = context, ...),
  timeout = self$timeout,
  stdout = stdout_file,
  stderr = stderr_file,
  user_profile = FALSE
)
```

Inside the subprocess, a fresh environment is created with the context available as `.context`. The wrapper overrides `library()` and `require()` to enforce a package allowlist, and a pattern scanner rejects calls to `system()`, `unlink()`, `quit()`, and `download.file()`. Each iteration spawns a new subprocess, so each pays a cold-start cost (typically 200--400ms depending on platform).

### Recursive Sub-Queries

This is where the "recursive" in RLM comes from. When `sub_lm` is provided, the model can write `llm_query("What does section 3 say?", context_slice)` in its generated code. The function does not execute the sub-call inside the subprocess, which would be a security problem. Instead, it returns a marker object:

```{r}
# From R/rlm-tools.R -- returns a marker, not a result
llm_query <- function(query, context_slice = NULL) {
  structure(
    list(query = query, context = context_slice, batch = FALSE),
    class = "rlm_query_request"
  )
}
```

The parent process intercepts this marker after execution, performs the actual call, and feeds the result back on the next iteration. A batched variant, `llm_query_batched()`, allows multiple sub-questions at once, running concurrently via `ellmer::parallel_chat()` when available.

A shared call counter tracks total calls across all iterations and enforces the `max_llm_calls` budget, preventing runaway recursion.

### Fallback and Output Normalization

If the model exhausts all iterations without calling `SUBMIT()`, the module performs fallback extraction: it feeds the entire exploration trajectory back and asks for a synthesized answer from what was discovered. This uses a two-phase approach, trying structured output via `chat_structured()` first, then unstructured `chat()` if that fails.

The final answer passes through output normalization, which coerces whatever was produced into the signature's declared output fields. This handles named lists, positional lists, scalar values, and case-insensitive enum matching (e.g., `"Positive"` is mapped to `"positive"` if the signature declares `type_enum("positive", "negative")`).

### Observability

Every RLM execution records its full REPL history: reasoning, code, output, success or failure, and timing for each iteration:

```{r}
history <- rlm$get_repl_history()
last_run <- history[[length(history)]]

last_run$iterations_used
#> [1] 5
last_run$llm_calls_used
#> [1] 3
```

You can see what the model tried, where it went wrong, and how it recovered.

## How dsprrr's RLM Compares

The table below summarizes the key implementations:

| | dsprrr | DSPy | Official `rlm` | Google ADK |
|---|---|---|---|---|
| **Language** | R | Python | Python | Python |
| **REPL** | R via callr | Python via Pyodide/WASM | Python (isolated or not) | Python via ADK |
| **Sandbox** | Subprocess (callr) | Deno/WASM | Configurable | ADK orchestration |
| **Structured output** | ellmer types | DSPy signatures | Freeform | ADK tools |
| **Recursive calls** | `llm_query()` | Built-in | Built-in | Child agents |
| **Optimization** | Teleprompters, grid search | DSPy optimizers | Manual | Manual |
| **Batched sub-calls** | `llm_query_batched()` | -- | -- | -- |

The "batched sub-calls" row refers specifically to issuing multiple recursive sub-queries from a single REPL iteration and running them concurrently. All implementations support multiple sequential sub-calls; dsprrr additionally batches them via `ellmer::parallel_chat()`.

dsprrr is the only implementation that uses R as the REPL language, which matters when your context *is* R data: data frames, model objects, or package source code. It also inherits dsprrr's full optimization infrastructure (teleprompters, grid search, evaluation metrics), so you can systematically improve RLM performance, not just run it.

## When to Use RLMs (and When Not To)

The hard part of any task here is either *finding* the right context or *reasoning* about it once found. RLMs help with the first problem. If the context is already short and well-scoped, simpler approaches are faster and cheaper.

| Approach | Best for | Latency | Context limit |
|----------|----------|---------|---------------|
| `PredictModule` | Short, self-contained tasks | Low | Context window |
| `chain_of_thought()` | Complex reasoning, known context | Medium | Context window |
| `rag_module()` | Lookup in large corpora | Medium | Chunk size |
| `rlm_module()` | Exploration of large, interconnected data | High | Unlimited\* |

\*Bounded by `max_iterations` and `max_llm_calls`, not by context window size.

**Skip RLMs when the context is short.** If the document fits comfortably in the context window, a `PredictModule` or `chain_of_thought()` will be faster and cheaper. The overhead of multiple REPL round-trips is not justified when one call suffices.

**Skip RLMs when the task is well-defined.** If you know what you are looking for (extracting a specific field from a known document format, say), a prompt-optimized module will outperform an RLM. RLMs spend iterations discovering a good exploration path. If you already know the path, skip the discovery.

**Skip RLMs when cost-per-query matters.** An RLM with 15 iterations makes at least 15 calls, plus any recursive sub-queries. For a production pipeline processing thousands of inputs, that multiplier adds up. If a single-call module with good prompting gets you 80% of the accuracy at 5% of the cost, the economics favor the simpler approach.

**Skip RLMs when the context contains bad information.** RLMs gather more evidence than simpler approaches, which is usually beneficial. But more evidence also means more surface area for misleading content. If the context contains contradictions, outdated facts, or adversarial content, a `chain_of_thought()` module with curated context gives explicit control over what the model sees.

## Improvement Opportunities

dsprrr's RLM implementation is functional but young. The items below are split into design constraints that affect deployment and API improvements that would make the module more ergonomic.

### Design Constraints

**Choose the runner as a security boundary.** `r_code_runner()` provides a fresh
callr subprocess and package allowlist, but it retains the host user's
permissions and is only for trusted code. For model-generated or otherwise
untrusted code, `mcp_repl_runner()` launches Posit's mcp-repl with an
OS-enforced workspace-write sandbox and network access disabled. A user-supplied
`repl` function is reported as unverified and is not accepted where dsprrr
requires sandboxed execution.

**Runner ownership is explicit.** A directly supplied `runner` is caller-owned
and reused across separate `forward()` calls; dsprrr never closes it. The
backend determines whether execution state persists and whether `reset()` is
available. mcp-repl retains variables and loaded packages until you call
`runner$reset()`. Reset that persistent backend between logically isolated
jobs and never share it across concurrent invocations.

For isolated invocations, pass `interpreter_factory` instead. It must be a
zero-argument function returning a valid runner. dsprrr calls it once per
invocation, owns that fresh runner, and closes it exactly once when the
invocation ends, including after a failure:

```{r rlm-interpreter-factory, eval=FALSE}
rlm <- rlm_module(
  "document, question -> answer",
  interpreter_factory = function() r_code_runner(timeout = 30)
)
```

`runner` and `interpreter_factory` are mutually exclusive, and one is required.

**mcp-repl control results are intentionally bounded.** RLM submit and recursive
query frames are capped at 3,000 encoded bytes to stay inline. If incidental
output triggers an mcp-repl file preview or pager, the iteration fails closed
rather than trusting a partial control frame. Keep `SUBMIT()` values compact;
use a transport with structured out-of-band results when larger payloads are a
requirement.

**Fresh subprocesses have cold-start overhead.** `r_code_runner()` pays process
startup cost on each iteration in exchange for clean process state. A
persistent mcp-repl runner avoids that repeated startup, but requires the reset
and concurrency discipline above.

### API Improvements

**`peek()` dual-dispatch API.** `peek()` silently changes meaning depending on whether the input is a single string (character positions) or a character vector (element indices). The model has to infer which form `.context$document` is in, and if it guesses wrong, the slice is nonsensical. Splitting into `peek_chars()` and `peek_lines()` (or adding a `unit` argument) would make the contract explicit.

**`search()` returns raw matches, not locations.** The model gets the matched text but not the byte offset or surrounding context, so it often has to follow up with a `peek()` call to figure out *where* the match occurred. Returning match positions or a concordance-style snippet would save an iteration.

## What's Next

RLMs trade latency for reach. They can explore contexts that no model handles well in a single pass, with subprocess isolation, recursive sub-queries, and full optimization support through dsprrr's teleprompters and grid search.

For a hands-on walkthrough, see `vignette("tutorial-rlm-dsprrr", package = "dsprrr")`, which uses `rlm_module()` to trace a real theming bug across bslib, shiny, and brand.yml: nearly 4 million characters of source, explored in under 15 iterations.

## References
