---
title: "Composing Optimizers with Omni"
description: >
  Explore several dsprrr teleprompters from one seed, compare them fairly,
  and continue from the strongest candidate.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composing Optimizers with Omni}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

No optimizer wins on every task. Demo selection may be the decisive lever for
one module, while instruction search or reflective mutation wins for another.
`Omni()` treats that uncertainty as part of the optimization problem: it runs
several teleprompters from the same seed program, compares every result with
one validation metric, and gives a fresh continuation optimizer the strongest
candidate.

This design is inspired by the
[`omni` meta-optimizer](https://gepa-ai.github.io/gepa/blog/2026/07/22/optimize-anything-omni/)
from the open-source [GEPA project](https://github.com/gepa-ai/gepa). The GEPA
team showed that different optimization engines plateau on different
Frontier-CS tasks, and that exploring several engines before continuing from
the best result can outperform any one engine under a matched budget.

dsprrr adapts that composition pattern to `Module` objects and
`Teleprompter`s. It is not a port of GEPA's arbitrary-text engine API, and the
published Frontier-CS result should not be assumed to transfer to an R module
or a different evaluation budget. Benchmark the composition on your own task.

## The Optimization Shape

For explorers `A`, `B`, and `C`, Omni performs this sequence:

```text
                         +-> explorer A --+
                         |                |
seed program + baseline -+-> explorer B --+-> shared validation -> winner
                         |                |                         |
                         +-> explorer C --+                         v
                                                        fresh continuation
                                                                  |
                                                                  v
                                                 shared validation + best
```

There are three important contracts:

1. Every explorer receives an independent copy of the same seed program.
2. The seed and all successful optimizer outputs are re-scored with the same
   metric and validation rows.
3. The seed remains eligible through final selection, so a regressing explorer
   or continuation step cannot replace a better candidate on that comparison
   set.

The continuation optimizer is a fresh teleprompter object run from the
exploration winner. It does not resume an explorer's internal search state.

## A Complete Configuration

Start with optimizers that search in meaningfully different ways. The example
below combines demo search, coordinate instruction search, and reflective
evolution before continuing with a new GEPA run.

```{r}
library(dsprrr)

metric <- metric_exact_match(field = "answer")

omni <- Omni(
  metric = metric,
  explorers = list(
    demos = BootstrapFewShotWithRandomSearch(
      metric = metric,
      num_candidate_programs = 8L
    ),
    instructions = COPRO(
      metric = metric,
      breadth = 8L,
      depth = 2L
    ),
    reflection = GEPA(
      metric = metric,
      population_size = 8L,
      generations = 3L
    )
  ),
  continuation = GEPA(
    metric = metric,
    population_size = 8L,
    generations = 3L
  ),
  seed = 42L
)

compiled <- compile(
  qa_module,
  omni,
  trainset,
  valset = valset,
  .llm = llm
)
```

Explorer names are part of the result provenance, so use short names that
describe the strategy rather than its position in the list.

If an optimizer needs arguments at compile time, pass them by explorer name.
Arguments cannot replace Omni's core `program`, `trainset`, `valset`, or
`.llm` inputs.

```{r}
compiled <- compile(
  qa_module,
  omni,
  trainset,
  valset = valset,
  .llm = llm,
  explorer_compile_args = list(
    demos = list(control = optimizer_control(max_metric_calls = 40L)),
    reflection = list(control = optimizer_control(max_metric_calls = 40L))
  ),
  continuation_compile_args = list(
    control = optimizer_control(max_metric_calls = 40L)
  )
)
```

## Use an Explicit Validation Set

Omni can split `trainset` automatically with `valset_ratio`, but an explicit
validation set makes comparisons easier to reproduce and audit:

```{r}
split <- split_dataset(examples, prop = 0.8, seed = 42L)
trainset <- split$train
valset <- split$val
```

All explorers train on `trainset`; Omni uses `valset` to choose the exploration
winner and the final program. Keep a third, untouched test set for the final
estimate if you will try several Omni configurations. Repeatedly choosing
explorers against the same validation rows can overfit the choice itself.

## Budget the Comparison Pass

`Omni()` does not impose one universal budget because dsprrr teleprompters
expose different native controls. Give explorers comparable budgets before
constructing Omni, then account for its shared comparison work separately.

With `k` successful explorers, the common scoring pass performs up to `k + 2`
full validation evaluations:

- one for the original seed;
- one for each successful explorer;
- one for the continuation result.

Those evaluations are additional to work performed inside each optimizer.
For a fair benchmark, record at least:

| Quantity | Why it matters |
|---|---|
| Metric calls | Makes search effort comparable |
| Provider calls and tokens | Captures optimizer-side reflection and generation |
| Cost | Allows matched-budget comparisons across models |
| Elapsed time | Shows the value of parallel exploration |
| Seed, standalone, and Omni scores | Separates composition gains from optimizer strength |

Compare each explorer standalone against the same explorer inside Omni. A
strong result is not merely "Omni won"; it shows whether diversity plus
continuation improved the best standalone candidate at a similar total budget.

## Inspect Candidate Provenance

The selected module stores a row for the baseline, every explorer, and the
continuation:

```{r}
candidates <- compiled$config$optimizer$candidate_programs
candidates[, c("phase", "optimizer", "score", "selected", "error")]
```

The `program` list-column contains a copy of each successful candidate for
deeper inspection:

```{r}
winner <- candidates$program[[which(candidates$selected)]]
winner$config

compiled$config$optimizer$exploration_winner
compiled$config$optimizer$best_phase
compiled$config$optimizer$best_optimizer
```

An explorer failure is recorded in `error` and does not stop the other
branches. If continuation fails, Omni returns the best exploration candidate.
The `flag_compilation_error_occurred` field makes either case visible without
discarding a usable result.

## Parallel Exploration

Set `parallel = TRUE` when branch runtime dominates and the provider can handle
the added concurrency:

```{r}
omni_parallel <- Omni(
  metric = metric,
  explorers = omni@explorers,
  continuation = omni@continuation,
  parallel = TRUE,
  num_workers = 3L,
  seed = 42L
)

compiled <- compile(
  qa_module,
  omni_parallel,
  trainset,
  valset = valset,
  .llm = NULL
)
```

Live ellmer chat objects are not serialized to mirai workers. Parallel
exploration therefore requires `.llm = NULL` and a worker-visible
`OPENAI_API_KEY`, `ANTHROPIC_API_KEY`, or `GOOGLE_API_KEY`. A chat configured
only with `set_default_chat()` in the main R session is not available inside
workers.

Parallel execution reduces wall-clock time, not evaluation cost. Provider rate
limits and simultaneous request caps may make fewer workers faster in
practice.

## Choosing Explorers

Favor complementary search behavior over several minor variations of the same
teleprompter:

| Search need | Candidate explorer |
|---|---|
| Select or bootstrap demonstrations | `BootstrapFewShotWithRandomSearch` |
| Rewrite instructions directly | `COPRO` |
| Focus on hard examples | `SIMBA` |
| Reflect on failures and evolve candidates | `GEPA` |
| Tune a known, finite parameter grid | `GridSearchTeleprompter` |

Two or three deliberately different explorers are usually a better first
experiment than a large portfolio. Add another branch only when it represents
a plausible search behavior that the current set does not cover.

## Where dsprrr Differs from GEPA Omni

| GEPA `optimize_anything` | dsprrr `Omni()` |
|---|---|
| Composes engines over arbitrary text artifacts | Composes teleprompters over dsprrr modules |
| Ships GEPA, AutoResearch, and Meta-Harness engines | Ships `GEPA()`, `AutoResearch()`, and `MetaHarness()` as composable dsprrr teleprompters |
| Terrarium enforces a shared task, server, and budget contract | Users align heterogeneous teleprompter budgets |
| Published Frontier-CS results use a matched $20 budget | No cross-task performance claim; benchmark locally |
| Continues the best artifact with a selected engine | Compiles a fresh continuation teleprompter from the winning module |

The shared idea is optimizer diversity: when search strategies fail in
different ways, the strongest partial result can give a fresh optimizer a
better starting point than the original seed.
