---
title: "Recursive Language Models: Exploring Codebases with dsprrr"
output: rmarkdown::html_vignette
---

```{r}
#| include: false
# Replay recorded HTTP responses (VCR cassettes) during site builds;
# skip evaluation entirely during R CMD check to avoid API dependencies.
force_eval <- identical(Sys.getenv("VITALS_SHOULD_EVAL"), "true")
in_check <- !force_eval &&
  (nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(Sys.getenv("NOT_CRAN"), "false"))

should_eval <- !in_check && dsprrr:::eval_vignette()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = should_eval
)

if (should_eval) {
  if (!is.null(vcr::current_cassette())) {
    try(vcr::eject_cassette(), silent = TRUE)
  }
  vcr::vcr_configure(
    dir = "_vcr",
    filter_request_headers = c("Authorization", "x-api-key"),
    match_requests_on = c("method", "uri")
  )
  vcr::setup_knitr()
}
```

LLM context windows have a hard ceiling. A large codebase either doesn't fit,
or, if it does, accuracy degrades silently: the model keeps producing confident
output while dropping details buried in the middle.

Recursive Language Models (RLMs) take a different approach. Instead of pasting
context into the prompt, they give the model programmatic tools to explore it.
The model writes R code to read files, search for patterns, and accumulate
relevant excerpts iteratively, turning long-context problems into coding
problems. For the conceptual background (what context rot is, how the ecosystem
evolved, and how dsprrr's internals work), see
the [How the RLM Works](how-rlm-works.html) article.

```
Traditional:  [entire document] -> LLM -> answer
RLM:          [query only]      -> LLM -> code -> [selective reads] -> LLM -> ...
```

This tutorial uses `rlm_module()` to explore three interconnected R package
codebases (bslib, shiny, and brand.yml) and investigate a real open issue
that spans all three. By Step 8, we show that repeated RLM runs converge on
the same five-phase exploration pattern, a finding you can use to design
deterministic pipelines.

**Time**: 30--45 minutes

**Topics covered:**

- Setting up an RLM module for codebase exploration
- Comparing RLM results against a curated-context baseline
- Tracing a real bug across three interconnected R packages
- Delegating interpretive work with recursive sub-queries
- Extracting stable exploration patterns from RLM traces

## The Problem: Contributing to bslib

[bslib issue #1123](https://github.com/rstudio/bslib/issues/1123): setting
the `primary` color in `bs_theme()` changes the navbar in `page_navbar()` but
is ignored by `page_sidebar()`. A user reported it with this example:

```r
# This changes the navbar color:
page_navbar(theme = bs_theme(preset = "flatly", primary = "#95a5a6"))

# This does NOT:
page_sidebar(theme = bs_theme(preset = "flatly", primary = "#95a5a6"))
```

Fixing this requires tracing how a brand color flows through three packages:

- **brand.yml**: reads `_brand.yml` files and generates Sass variables
- **bslib**: compiles those variables into Bootstrap CSS
- **shiny**: renders the themed UI components

The bslib package alone has over 500 R and SCSS files. Adding shiny and
brand.yml pushes the total context to nearly 4 million characters, well
beyond what fits in a single prompt.

## Step 1: Setup and Load the Codebases

```{r setup}
library(dsprrr)
library(ellmer)
```

Create an `RCodeRunner` for executing the code the RLM generates:

```{r runner-setup}
runner <- r_code_runner(timeout = 30)
```

Pull the source code for all three packages. The `read_package_source()`
helper below clones a repo, concatenates its R and SCSS files into a single
string (preserving file paths), and returns the result. Expand the fold to
see the implementation, or skip ahead; the important thing is what the
function returns, not how it works.

<details>
<summary>Definition of <code>read_package_source()</code></summary>

```{r read-package-source, cache = TRUE}
read_package_source <- function(repo, ref = "main", subdirs = c("R", "inst")) {
  dir <- tempfile()
  status <- system2(
    "git",
    c(
      "clone",
      "--depth=1",
      "--branch",
      ref,
      paste0("https://github.com/", repo, ".git"),
      dir
    ),
    stdout = FALSE,
    stderr = FALSE
  )
  if (status != 0) {
    stop("git clone failed for ", repo, " (exit code ", status, ")")
  }

  files <- unlist(lapply(subdirs, function(subdir) {
    list.files(
      file.path(dir, subdir),
      pattern = "\\.(R|r|scss)$",
      recursive = TRUE,
      full.names = TRUE
    )
  }))

  contents <- vapply(
    files,
    function(f) {
      path <- sub(paste0(dir, "/"), "", f)
      paste0("--- FILE: ", path, " ---\n", paste(readLines(f), collapse = "\n"))
    },
    character(1),
    USE.NAMES = FALSE
  )

  paste(contents, collapse = "\n\n")
}
```

</details>

```{r load-code, cache = TRUE}
bslib_source <- read_package_source("rstudio/bslib")
shiny_source <- read_package_source("rstudio/shiny")
brandyml_source <- read_package_source(
  "posit-dev/brand-yml",
  subdirs = "pkg-r/R"
)
```

These three strings sit in programmatic space; none enter the context window
until the model requests a specific slice:

```{r context-sizes}
format_size <- function(source, label) {
  n_files <- length(gregexpr("--- FILE:", source)[[1]])
  cli::cli_li("{.strong {label}}: {format(nchar(source), big.mark = ',')} characters ({n_files} files)")
}
cli::cli_ul()
format_size(bslib_source, "bslib")
format_size(shiny_source, "shiny")
format_size(brandyml_source, "brand.yml")
cli::cli_end()
```

Nearly 4 million characters total, well beyond what any model handles
accurately in a single pass.

## Step 2: Baseline, What a Developer Would Try First

Before reaching for an RLM, a competent developer would grep for the relevant
functions and feed the results to a model. Let's try that:

```{r baseline}
#| cassette: true
# Extract the context a developer would actually assemble:
# definitions and nearby code for both page functions, plus Sass variable usage
relevant_lines <- function(source, patterns, context_chars = 3000) {
  slices <- lapply(patterns, function(pat) {
    positions <- gregexpr(pat, source, perl = TRUE)[[1]]
    positions <- positions[positions > 0]
    if (length(positions) == 0) {
      return(character(0))
    }
    # Take first 3 matches, with surrounding context
    positions <- head(positions, 3)
    vapply(
      positions,
      function(pos) {
        start <- max(1, pos - context_chars %/% 2)
        substr(source, start, start + context_chars)
      },
      character(1)
    )
  })
  paste(unlist(slices), collapse = "\n\n---\n\n")
}

curated_context <- paste(
  relevant_lines(bslib_source, c("page_navbar", "page_sidebar", "\\$primary")),
  relevant_lines(shiny_source, c("navbarPage", "navbar")),
  sep = "\n\n=== shiny ===\n\n"
)

baseline <- module(
  signature(
    "codebase, question -> analysis",
    instructions = "You are an expert R developer analyzing package source code."
  )
)

result <- run(
  baseline,
  codebase = curated_context,
  question = paste(
    "In bslib, why does setting primary in bs_theme() change the navbar",
    "in page_navbar() but not in page_sidebar()? Trace the Sass variable",
    "chain from primary through to the navbar background."
  ),
  .llm = chat_openai(model = "gpt-5-mini")
)

result$analysis
```

The model gets targeted context: function definitions for both `page_navbar()`
and `page_sidebar()`, plus Sass variable references and relevant shiny code.
Better than stuffing in a random 50K prefix, but still incomplete. The grep
captures mentions of `$primary` but not the chain of SCSS imports and mixins
that connect it (or fail to connect it) to `$navbar-bg`. The model can see the
endpoints but not the plumbing between them.

A coding agent could search files iteratively, but it needs files on disk. An
RLM works on arbitrary in-memory data: combined source strings from multiple
repos, API responses, scraped content, anything you can load into a variable.
And because `rlm_module()` is a dsprrr module, its traces feed into
`compile()`, `evaluate()`, and the rest of the optimization framework.

## Step 3: Set Up the RLM

```{r rlm-setup}
investigator <- rlm_module(
  signature(
    "bslib_source, shiny_source, brandyml_source, question -> analysis",
    instructions = paste(
      "You are an expert R/Sass developer investigating a theming bug across",
      "three interconnected R packages. Explore the source code systematically",
      "to trace how Sass variables flow between packages."
    )
  ),
  runner = runner,
  max_iterations = 15,
  verbose = TRUE
)
```

The module takes three context variables (one per package) plus a question.
Inside the REPL, these mechanisms are available:

| Mechanism | Purpose |
|-----------|---------|
| `.context$<var>` | Access a context variable (e.g., `.context$bslib_source`) |
| `peek(var, start, end)` | View a slice of a variable; dispatches on type (character positions for strings, element indices for vectors). Default: first 1000 chars |
| `search(var, pattern)` | Perl-compatible regex search; returns all matching substrings |
| `llm_query(query, context_slice)` | Delegate a sub-question to a secondary model (requires `sub_lm`) |
| `llm_query_batched(queries, slices)` | Batch multiple sub-questions in parallel (requires `sub_lm`) |
| `SUBMIT(...)` | Return the final answer and terminate the REPL loop; validates against signature output fields |

The model writes R code using these mechanisms. Each iteration, the code
executes and the output feeds back as context for the next step.

## Step 4: Run the Investigation

```{r run-rlm}
#| cassette: true
result <- run(
  investigator,
  bslib_source = bslib_source,
  shiny_source = shiny_source,
  brandyml_source = brandyml_source,
  question = paste(
    "In bslib, setting `primary` in bs_theme() changes the navbar color in",
    "page_navbar() but NOT in page_sidebar(). This is GitHub issue #1123.",
    "Trace the complete Sass variable chain from `$primary` to the navbar",
    "background in both page functions. Identify exactly where and why the",
    "chain breaks for page_sidebar(). Include specific file names and line",
    "references."
  ),
  .llm = chat_openai(model = "gpt-5-mini")
)

cli::cli_h3("Analysis")
cli::cli_verbatim(result$analysis)
```

With `verbose = TRUE`, each iteration prints as it runs.

## Step 5: Inside the REPL

The RLM runs a loop: generate code, execute it, observe results, repeat. It
does not read everything at once:

```{r repl-history}
history <- investigator$get_repl_history()
latest <- history[[length(history)]]

cli::cli_alert_info("Iterations used: {latest$iterations_used} / {investigator$max_iterations}")

# Helper for displaying iteration history
show_iteration <- function(entry, n, label = NULL) {
  header <- if (!is.null(label)) {
    paste0("Iteration ", n, " (", label, ")")
  } else {
    paste0("Iteration ", n)
  }
  cli::cli_h3(header)
  cli::cli_text("{.strong Reasoning}:")
  cli::cli_verbatim(entry$reasoning)
  cli::cli_text("{.strong Code}:")
  cli::cli_code(entry$code)
  if (!isTRUE(entry$success)) {
    cli::cli_alert_danger("Failed")
    if (!is.null(entry$output) && nzchar(entry$output)) {
      cli::cli_text("{.strong Output}:")
      cli::cli_verbatim(entry$output)
    }
  }
}
```

Each iteration records the model's reasoning and the code it wrote. The
walkthrough below is drawn from the recorded run above. Not every iteration
succeeds: the model makes wrong turns, hits R string-escaping errors, and
occasionally wastes a step. That is normal. The REPL loop is designed around
the assumption that individual steps will fail.

### Early iterations: Broad search

Nearly 4 million characters sit in programmatic space; zero are in token
space. The model typically starts by mapping the terrain:

```{r iter-early, echo = FALSE}
show_iteration(latest$history[[2]], 2)
```

`search()` returns only matching substrings, not entire files. Each result is
a targeted transfer from programmatic to token space.

### Mid-iterations: Locate definitions, trace variables

As the model accumulates results, it narrows in on specific definitions and
the surrounding code:

```{r iter-mid, echo = FALSE}
show_iteration(latest$history[[4]], 4)
```

The model has access to all of R, not just the provided REPL tools. It
frequently uses base R functions like `gregexpr()`, `grepl()`, or
`regmatches()` to refine its searches, and sometimes writes helper functions
or splits files by header.

### Failures and recovery

Not every iteration succeeds. Let's find one that failed and see how the
model recovered:

```{r iter-fail, echo = FALSE}
failed <- which(!vapply(latest$history, function(h) h$success, logical(1)))
if (length(failed) > 0) {
  i <- failed[1]
  show_iteration(latest$history[[i]], i, label = "failed")
  if (i < length(latest$history)) {
    show_iteration(latest$history[[i + 1]], i + 1, label = "recovery")
  }
} else {
  cli::cli_alert_success("This run had no failed iterations (unusual but possible).")
}
```

Across multiple runs, two failure modes recur:

**String escaping errors.** The model writes `"\("` instead of `"\\("`, or
`"\$"` instead of `"\\$"`. R rejects the code, the error feeds back, and the
model self-corrects on the next iteration.

**Lost state.** Each iteration runs in a fresh environment. A helper function
or parsed data structure defined in iteration 7 does not exist in iteration 8.
The model encounters this empirically: after a "not found" error, it re-creates
the object. This costs iterations but is part of the REPL's design. Stateless
execution prevents accumulated errors from compounding.

A typical run includes 2--4 failed iterations out of 10--15 total.

### Final iteration: SUBMIT

Once the model has gathered enough evidence, it calls `SUBMIT()` with the
answer:

```{r iter-final, echo = FALSE}
show_iteration(latest$history[[length(latest$history)]], length(latest$history), label = "final")
```

`SUBMIT()` returns the final analysis and terminates the REPL loop. If
`max_iterations` is reached without a `SUBMIT()` call, dsprrr extracts a
best-effort answer from the full exploration history.

## Step 6: Add Recursive Sub-queries

The analysis above identifies the broken Sass variable chain. To go further
(proposing a specific code fix, say), the root model may need help
interpreting complex SCSS mixins or Bootstrap conventions from raw character
slices. A secondary model handles these focused sub-questions:

```{r rlm-recursive}
#| cassette: true
deep_investigator <- rlm_module(
  signature(
    "bslib_source, shiny_source, brandyml_source, question -> analysis, fix_proposal",
    instructions = paste(
      "You are an expert R/Sass developer. Investigate the bug and propose a",
      "specific code fix. Use llm_query() to get help interpreting complex",
      "Sass logic or understanding Bootstrap conventions."
    )
  ),
  runner = runner,
  max_iterations = 20,
  sub_lm = chat_openai(model = "gpt-5-mini"), # Secondary model for sub-queries
  max_llm_calls = 10,
  verbose = TRUE
)

result <- run(
  deep_investigator,
  bslib_source = bslib_source,
  shiny_source = shiny_source,
  brandyml_source = brandyml_source,
  question = paste(
    "Investigate bslib issue #1123 and propose a fix.",
    "The page_sidebar() title bar should respect the primary color the same",
    "way page_navbar() does. What's the minimal change to fix this?"
  ),
  .llm = chat_openai(model = "gpt-5-mini")
)

cli::cli_h3("Analysis")
cli::cli_verbatim(result$analysis)
cli::cli_h3("Proposed Fix")
cli::cli_verbatim(result$fix_proposal)
```

With `sub_lm` set, the root model can delegate interpretive tasks to a
secondary model. For example, when it encounters a complex SCSS mixin:

```r
llm_query(
  "In Bootstrap 5 Sass, what is the difference between $navbar-bg and
   $navbar-light-bg? When would each be used?",
  context_slice = scss_snippet
)
```

The root model orchestrates exploration; the sub-model handles focused
interpretation. A smaller, cheaper model is usually sufficient for these
queries, since the sub-questions are narrow and well-scoped.

`llm_query_batched()` sends multiple sub-questions in parallel.

## Step 7: Inspect Costs and Trajectory

RLMs trade latency for accuracy:

```{r costs}
history <- deep_investigator$get_repl_history()
latest <- history[[length(history)]]

cli::cli_ul()
cli::cli_li("Iterations used: {latest$iterations_used} / {deep_investigator$max_iterations}")
cli::cli_li("LLM sub-calls: {latest$llm_calls_used} / {deep_investigator$max_llm_calls}")
cli::cli_end()
```

A typical run uses 10--15 iterations. Each involves one call to generate
code plus the execution itself; 2--4 of those iterations will fail (string
escaping errors, lost state, timeouts) and self-correct. Recursive sub-queries
add additional calls. Total token usage is a fraction of what stuffing all
three codebases into one prompt would require.

The tradeoff is wall-clock time. Each iteration is a sequential round-trip
(generate code, execute, observe result): expect 2--5 minutes for a full run,
depending on model latency and how many iterations the model needs.

## Step 8: From Traces to Agent Designs

There is a secondary use for `get_repl_history()` beyond debugging. As
[Breunig (2026)](https://www.dbreunig.com/2026/02/09/the-potential-of-rlms.html)
observes, running an RLM on the same task multiple times reveals repeatable
exploration patterns.

We ran the bslib investigation four times with `gpt-5-mini`. The code varied
across runs, but the exploration structure converged:

| Phase | Run 1 | Run 2 | Run 3 | Run 4 |
|-------|-------|-------|-------|-------|
| 1. Orient | `search("page_navbar")` | `search("page_navbar")` | `peek(bslib, 1, 5000)` | `search("page_sidebar")` |
| 2. Locate definitions | `gregexpr("page_navbar")` + `peek` | `search("page_sidebar")` + `peek` | `search("page_navbar\\b")` | `gregexpr("page_navbar")` + `peek` |
| 3. Find Sass chain | `search("\\$navbar-bg")` | `search("navbar-bg")` | `search("\\$primary")` | `search("\\$navbar-bg")` |
| 4. Cross-reference | `search(shiny, "navbar")` | `search(shiny, "navbarPage")` | `search(shiny, "navbar")` | `search(brandyml, "primary")` |
| 5. Identify gap | Compare page_navbar vs page_sidebar SCSS | Compare $navbar-bg vs sidebar vars | Compare preset mappings | Compare $navbar-bg chain |

All four runs searched for `page_navbar` and `$navbar-bg` within the first
four iterations. All four cross-referenced at least one other package. All four
converged on the same diagnosis. The specific code and ordering differed, but
the five-phase structure (orient, locate, trace Sass, cross-reference, identify
gap) was stable.

That stable structure is a specification you can extract and formalize into a
deterministic pipeline, trading the RLM's flexibility for speed and
reliability.

This connects to dsprrr's optimization story. A `compile()` call with a
teleprompter tunes a module's parameters against a dataset. RLM traces offer a
complementary path: instead of optimizing *within* a module, you observe the
module's behavior to design a *new* module, or a chain of modules, that
encodes the discovered strategy directly.

## Summary

The investigation traced how `bs_theme(primary = ...)` flows through bslib's
Sass pipeline and found the gap: `page_navbar()` picks up `$primary` via the
flatly preset's mapping to `$navbar-bg`, but `page_sidebar()`'s title bar
defaults to `$secondary` with no equivalent link. Three packages, nearly
4 million characters of source, and the model identified the disconnect in
a handful of iterations.

More importantly, the traces revealed a stable five-phase exploration pattern
that converged across multiple runs, the kind of structure you can extract
and formalize into a deterministic pipeline.

For guidance on when RLMs are the right tool (and when simpler approaches
win), see the decision framework in
the [How the RLM Works](how-rlm-works.html) article.

## Try It Yourself

The snippet below uses `read_package_source()` from Step 1. You already have
the bslib, shiny, and brand.yml source loaded; try a second investigation
with the same data. There are several
[open theming issues](https://github.com/rstudio/bslib/issues?q=is%3Aopen+label%3A%22theming%22)
in bslib that require the same kind of cross-package tracing. For example:

```{r try-it, eval = FALSE}
# Investigate another theming issue with the same data
run(
  investigator,
  bslib_source = bslib_source,
  shiny_source = shiny_source,
  brandyml_source = brandyml_source,
  question = paste(
    "How does bs_theme()'s `font_scale` argument propagate through bslib's",
    "Sass pipeline? Which components respect it and which ignore it?"
  ),
  .llm = chat_openai(model = "gpt-5-mini")
)
```

Or load your own package source:

```{r try-it-own, eval = FALSE}
my_source <- read_package_source("your-org/your-package")

explorer <- rlm_module(
  "codebase, question -> answer",
  runner = r_code_runner(timeout = 30),
  max_iterations = 10
)

run(
  explorer,
  codebase = my_source,
  question = "How does the authentication middleware work?",
  .llm = chat_openai(model = "gpt-5-mini")
)
```

## Further Reading

- [Zhang, Kraska & Khattab (2025). "Recursive Language Models."](https://arxiv.org/abs/2512.24601)
  The paper introducing the RLM approach.
- [Breunig (2026). "The Potential of RLMs."](https://www.dbreunig.com/2026/02/09/the-potential-of-rlms.html)
  Practical experience with RLMs at scale (400MB+ contexts), plus the
  trace-to-pipeline idea. Breunig draws a useful analogy: RLMs are to
  long-context problems what chain-of-thought was to reasoning, a test-time
  strategy that works today and will improve as models are trained to exploit it.
- `vignette("advanced-modules", package = "dsprrr")`: ChainOfThought,
  BestOfN, and other reasoning patterns in dsprrr
- `vignette("reasoning-models", package = "dsprrr")`: Using reasoning
  models (o1, o3, GPT-5) with dsprrr
- `vignette("rag-workflows", package = "dsprrr")`: When retrieval-based
  approaches are a better fit
